### 2020-10-02 SingleSite_SingleDisturb.Rmd
Initial commit of simple toy case of disturbance model

---
title: "Single Site, Single Disturbance"
author: "Michael Dietze"
date: "7/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Initial code for this started from the "Assimilation of disturbance" example in <https://github.com/EcoForecast/EF_BookCode/blob/master/Ch02_Logitic.Rmd> which itself derived from simple simulations and KF developed for a 2013 talk at the Ecological Society of America annual meeting.

### Assimilation of disturbance

"True" run

```{r,echo=1:2, fig.width=7, fig.height=6}
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NT = 100
time = 1:NT
xlim = c(5,25)
n = rep(nIC,NT)
d = rep(0,NT)
d[10] = 1  ## prescribe a disturbance
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K) + rnorm(1,0,n.sd)
#  d[t] = rbinom(1,1,p.dist)
  if(d[t]) n[t] = abs(rnorm(1,n0,0.5))
}

plot(time,n,type='l',ylim=range(n),xlim=xlim,xlim=xlim,ylab="LAI",lwd=3)
obs.sd = 0.75
n.true=n
n.obs = rnorm(NT,n.true,obs.sd)
points(time,n.obs,pch=18,col=2,cex=1.5)
legend("topleft",legend=c("truth","obs"),lwd=c(2,NA),pch=c(NA,18),col=1:2)
## Simulate less frequent observations
#na.seq = 1:NT; na.seq=na.seq[-seq(1,NT,by=5)]
#n.fia = n.obs;n.fia[na.seq] <- NA
#points(time,n.fia,pch=18,col="blue",cex=2)
```

## "blind" forecast w/o disturbance

```{r}
n0.sd = n.sd
NE = 1000
n0s = rlnorm(NE,log(nIC),n0.sd)
ne0 = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    ne0[i,t] = abs(ne0[i,t-1] + r*ne0[i,t-1]*(1-ne0[i,t-1]/K) + rnorm(1,0,n.sd))
#    if(rbinom(1,1,p.dist)) ne[i,t] = abs(rnorm(1,n0,0.5))
  }
}
yrng = c(0,8) #range(ne0)
n0.stats = apply(ne0,2,quantile,c(0.025,0.5,0.975))
plot(time,n.obs,type='n',ylim=yrng,xlim=xlim)
polygon(cbind(c(time, rev(time), 1), c(n0.stats[1, ], rev(n0.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n0.stats[2,],lwd=2,col="blue4")
points(time,n.obs,pch=18,col=2,cex=1.5)
```

## standard EnKF w/o disturbance

```{r,echo=FALSE, fig.width=7, fig.height=6}
## initial ensemble
ens = matrix(NA,NT*2+1,NE)
ens[1,] = rlnorm(NE,log(nIC),n0.sd)
Xae = Pae = 0
Xfe = mean(ens[1,])
Pfe = var(ens[1,])
for(i in 1:NT){

  ##Approximate forecast with Normal
  k = 2*(i-1)+1
  Xfe[i] = mean(ens[k,])
  Pfe[i] = var(ens[k,])

  ## analysis step: combine previous forecast with observed data
  H <- matrix(1,ncol=1)
  Kg <- Pfe[i] * t(H) %*% solve(H%*%Pfe[i]%*%t(H) + obs.sd)
  if(!is.na(n.obs[i])){
    Xae[i] <- Xfe[i] + Kg*(n.obs[i] - Xfe[i]) 
  }else{
    Xae[i] <- Xfe[i]
  }
  Pae[i] <- (1-Kg %*% H)*Pfe[i]

  ## generate new ensemble
  ens[k+1,] = rnorm(NE,Xae[i],sqrt(Pae[i]))

  ## Forecast step: predict to next step from current
  ens[k+2,] = ens[k+1,] + r*ens[k+1,]*(1-ens[k+1,]/K) + rnorm(NE,0,n.sd) 
#  disturbed = as.logical(rbinom(NE,1,p.dist))
#  ens[k+2,disturbed] = abs(rnorm(sum(disturbed),n0,0.5)) ## disturbance

}

## forecast
fx0 = ens[2*(1:NT)-1,]
fx0.stats = apply(fx0,1,quantile,c(0.025,0.25,0.5,0.75,0.975))
plot(time,fx0.stats[5,],col="blue",ylim=yrng,xlim=xlim,type='n',ylab="LAI",cex.lab=1.5)  
polygon(cbind(c(time, rev(time), 1), c(n0.stats[1, ], rev(n0.stats[3, ]),      ## Free run, no disturbance
  n0.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(fx0.stats[1, ], rev(fx0.stats[5, ]),    ## Forecast 95%
  fx0.stats[1, 1])), border = NA, col = 'lightgreen')
polygon(cbind(c(time, rev(time), 1), c(fx0.stats[2, ], rev(fx0.stats[4, ]),    ## Forecast IQR
  fx0.stats[2, 1])), border = NA, col = 'green4')
#lines(time[-NT],fx.stats[2,-1],lwd=4,col="green")
lines(time,n0.stats[2,],lwd=2,col="blue4")                                     ## Free run mean
lines(time,n.true,type='l',lwd=2)                                              ## truth
points(time,n.obs,pch=18,col=2)                                                ##obs
legend("topleft",legend=c("True","Obs","95%","IQR","Model"),lty=c(1,0,1,1,1),col=c(1,2,"lightgreen","green4","lightblue"),lwd=c(1,NA,15,15,15),cex=1.2,pch=c(NA,18,NA,NA,NA))

## analysis
an0 = ens[2*(1:NT),]
an0.stats = apply(an0,1,quantile,c(0.025,0.25,0.5,0.75,0.975))
plot(time,an0.stats[5,],col="blue",ylim=yrng,xlim=xlim,type='n',ylab="LAI",cex.lab=1.5)
polygon(cbind(c(time, rev(time), 1), c(n0.stats[1, ], rev(n0.stats[3, ]),
  n0.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(an0.stats[1, ], rev(an0.stats[5, ]),
  an0.stats[1, 1])), border = NA, col = 'lightgreen')
polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
  an0.stats[2, 1])), border = NA, col = 'green4')
lines(time,n0.stats[2,],lwd=2,col="blue4")
lines(time,n.true,type='l',lwd=2)
points(time,n.obs,pch=18,col=2)
legend("topleft",legend=c("True","Obs","95%","IQR","Model"),lty=c(1,0,1,1,1),col=c(1,2,"lightgreen","green4","lightblue"),lwd=c(1,NA,15,15,15),cex=1.2,pch=c(NA,18,NA,NA,NA))

```

## "blind" forecast w/ disturb

```{r}
n0.sd = n.sd
NE = 1000
n0s = rlnorm(NE,log(nIC),n0.sd)
ne = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    ne[i,t] = abs(ne[i,t-1] + r*ne[i,t-1]*(1-ne[i,t-1]/K) + rnorm(1,0,n.sd))
    if(rbinom(1,1,p.dist)) ne[i,t] = abs(rnorm(1,n0,0.5))
  }
}
yrng = range(ne)
n.stats = apply(ne,2,quantile,c(0.025,0.5,0.975))
plot(time,n.obs,type='n',xlim=xlim)
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2,col="blue4")
points(time,n.obs,pch=18,col=2,cex=1.5)
```

## standard EnKF + Disturb

```{r,echo=FALSE, fig.width=7, fig.height=6}
## initial ensemble
ens = matrix(NA,NT*2+1,NE)
ens[1,] = rlnorm(NE,log(nIC),n0.sd)
Xae = Pae = 0
Xfe = mean(ens[1,])
Pfe = var(ens[1,])
for(i in 1:NT){

  ##Approximate forecast with Normal
  k = 2*(i-1)+1
  Xfe[i] = mean(ens[k,])
  Pfe[i] = var(ens[k,])

  ## analysis step: combine previous forecast with observed data
  H <- matrix(1,ncol=1)
  Kg <- Pfe[i] * t(H) %*% solve(H%*%Pfe[i]%*%t(H) + obs.sd)
  if(!is.na(n.obs[i])){
    Xae[i] <- Xfe[i] + Kg*(n.obs[i] - Xfe[i]) 
  }else{
    Xae[i] <- Xfe[i]
  }
  Pae[i] <- (1-Kg %*% H)*Pfe[i]

  ## generate new ensemble
  ens[k+1,] = rnorm(NE,Xae[i],sqrt(Pae[i]))

  ## Forecast step: predict to next step from current
  ens[k+2,] = ens[k+1,] + r*ens[k+1,]*(1-ens[k+1,]/K) + rnorm(NE,0,n.sd) 
  disturbed = as.logical(rbinom(NE,1,p.dist))
  ens[k+2,disturbed] = abs(rnorm(sum(disturbed),n0,0.5)) ## disturbance
#  Xfe[i+1] = m*Xa[i] + beta[2]
#  Pfe[i+1] = q + m*Pa[i]*m

}

## forecast
fx = ens[2*(1:NT)-1,]
fx.stats = apply(fx,1,quantile,c(0.025,0.25,0.5,0.75,0.975))
plot(time,fx.stats[5,],col="blue",ylim=yrng,xlim=xlim,type='n',ylab="LAI",cex.lab=1.5)
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(fx.stats[1, ], rev(fx.stats[5, ]),
  fx.stats[1, 1])), border = NA, col = 'lightgreen')
polygon(cbind(c(time, rev(time), 1), c(fx.stats[2, ], rev(fx.stats[4, ]),
  fx.stats[2, 1])), border = NA, col = 'green4')
#lines(time[-NT],fx.stats[2,-1],lwd=4,col="green")
lines(time,n.stats[2,],lwd=2,col="blue4")
lines(time,n.true,type='l',lwd=2)
points(time,n.obs,pch=18,col=2)
legend("topleft",legend=c("True","Obs","95%","IQR","Model"),lty=c(1,0,1,1,1),col=c(1,2,"lightgreen","green4","lightblue"),lwd=c(1,NA,15,15,15),cex=1.2,pch=c(NA,18,NA,NA,NA))

## analysis
an = ens[2*(1:NT),]
an.stats = apply(an,1,quantile,c(0.025,0.25,0.5,0.75,0.975))
plot(time,an.stats[5,],col="blue",ylim=yrng,xlim=xlim,type='n',ylab="LAI",cex.lab=1.5)
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(an.stats[1, ], rev(an.stats[5, ]),
  an.stats[1, 1])), border = NA, col = 'lightgreen')
polygon(cbind(c(time, rev(time), 1), c(an.stats[2, ], rev(an.stats[4, ]),
  an.stats[2, 1])), border = NA, col = 'green4')
polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),  ## Analysis IQR- no disturb
  an0.stats[2, 1])), border = NA, col = 'orange')
lines(time,n.stats[2,],lwd=2,col="blue4")
lines(time,n.true,type='l',lwd=2)
points(time,n.obs,pch=18,col=2)
legend("topleft",legend=c("True","Obs","95%","IQR","Model"),lty=c(1,0,1,1,1),col=c(1,2,"lightgreen","green4","lightblue"),lwd=c(1,NA,15,15,15),cex=1.2,pch=c(NA,18,NA,NA,NA))

```

## Offline State-Space Model

Start by fitting the full state-space model to the data, then move on to figuring out how to approximate that iteratively

```{r}
library(rjags)

SSSD <- "
model{
  q  ~ dgamma(aq,bq)
  r  ~ dgamma(ar,br)
  R  ~ dgamma(aR,bR)
  K  ~ dgamma(aK,bK)
  p  ~ dbeta(ap,bp)
  for(t in 1:nt){
    Y[t] ~ dnorm(n[t],r)
  }
  n[1] <- N0            ## should add some uncertainty
  for(t in 2:nt){
    muN[t] <- n[t-1] + R*n[t-1]*(1-n[t-1]/K)
    muD[t] ~ dnorm(mu0,pa0)
    D[t] ~ dbern(p)
    mu[t] <- D[t]*muD[t] + (1-D[t])*muN[t]
    n[t] ~ dnorm(mu[t],q)
  }
}"
## Run individual models
data = list(Y=n.obs,nt=NT,
            mu0=1,pa0=1,  ## put a bit more confidence in the low value
            aq=0.1,bq=0.1,
            ar=0.1,br=0.1,
            aR=r,bR=1,
            aK=K,bK=1,
            ap=p.dist*50,bp=(1-p.dist)*50, ## try with an informative prior
            N0 = nIC)  
 mod <- jags.model(file=textConnection(SSSD),
                   data=data,
                  n.adapt=1000,n.chains=3,
                  init=list(R = r, K = K, q = n.sd,p=p.dist))
 jdat <- coda.samples(mod,variable.names=c("R","K","p","r","q"),n.iter=3000) ## burnin
 #plot(jdat)
 jdat <- coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=30000,thin=10) ## samples
```

```{r}
## summarize output
CI <- apply(as.matrix(jdat),2,quantile,c(0.025,0.5,0.975))
CIN <- CI[,grep(pattern = "n",colnames(CI))]
col=col2rgb("blue")
plot(data$Y,ylim = range(rbind(CIN,data$Y)),xlim=xlim)
ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
lines(CIN[2,],col="blue",lwd=2)

## Disturbance probability
Dsamp <- as.matrix(jdat)
Dsamp <- Dsamp[,grep(pattern = "^D",colnames(Dsamp))]
muD <- apply(Dsamp,2,mean)

CID <- CI[,grep(pattern = "^D",colnames(CI))]
col=col2rgb("blue")
plot(2:NT,muD,ylim = range(CID),xlim=xlim,type='l')
ecoforecastR::ciEnvelope(2:NT,CID[1,],CID[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))

if(FALSE){  ## didn't work as expected because D is 0/1, want to plot p[t]
Dden <- data.frame(val=as.vector(Dsamp),year=rep(1:ncol(Dsamp),each=nrow(Dsamp)))
ggplot2::ggplot(Dden, ggplot2::aes(x = val, y = year,  group = year)) +
  ggridges::geom_density_ridges() +
  ggridges::theme_ridges() + 
  ggplot2::theme(legend.position = "none")
}

## Disturb vs Non
CImuN <- CI[,grep(pattern = "muN",colnames(CI))]
CImuD <- CI[,grep(pattern = "muD",colnames(CI))]
colN=col2rgb("green")
colD=col2rgb("orange")
plot(data$Y,ylim = range(rbind(CImuN,CImuD,data$Y[-1])),xlim=xlim,
     ylab="Disturbance Probability")
ecoforecastR::ciEnvelope(2:NT,CImuD[1,],CImuD[3,],col=rgb(colD[1],colD[2],colD[3],0.3*256,maxColorValue=256))
ecoforecastR::ciEnvelope(2:NT,CImuN[1,],CImuN[3,],col=rgb(colN[1],colN[2],colN[3],0.3*256,maxColorValue=256))

## comparison to EnKF
plot(data$Y,ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
     ylab="LAI",xlab="time")
polygon(cbind(c(time, rev(time), 1), c(an0.stats[1, ], rev(an0.stats[5, ]),
  an0.stats[1, 1])), border = NA, col = '#FFBF6C')
polygon(cbind(c(time, rev(time), 1), c(an.stats[1, ], rev(an.stats[5, ]),
  an.stats[1, 1])), border = NA, col = ecoforecastR::col.alpha('#b2df8a',0.7))
ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
lines(an0.stats[3,],col="orange",lwd=2)
lines(an.stats[3,],col="#b2df8a",lwd=2)
lines(CIN[2,],col="blue",lwd=2)
points(data$Y,pch=18,col=2)
legend("topright",legend=c("EnKF, no disturb","EnKF, disturb","Multinomial","data"),
       lwd=5,col=c("orange","#b2df8a","blue","red"))

```

## Forward Filter version of SSSD, Known Params

```{r}
FilterSSSD <- "
model{
  N0 ~ dnorm(mu0,pa0)  ## initial conditions
  ## Forecast
  muN <- N0 + R*N0*(1-N0/K)
  muD <- 1
  D ~ dbern(p)
  mu <- D[t]*muD + (1-D)*muN
  n ~ dnorm(mu,q)
  ## Analysis
  Y ~ dnorm(n,r)
}"
## Forecast Cycle
CI.LFP = matrix(NA,3,nt)
Xl = data$mu0
pl = data$pa0
aqp = bqp = arp = brp = 0.1
mu.m = 0
p.m  = 1e-6
for(t in 1:nt){
  ## Forecast + Analysis
  update = list(Y=Y[t],mu0=Xl,pa0=pl,aq=aqp[t],bq=bqp[t],ar=arp[t],br=brp[t],mu.m=mu.m[t],tau.m=p.m[t])
  mod <- jags.model(file=textConnection(LinFilterParam),
                   data=update,
                  n.adapt=1000,n.chains=3,
                  init=list(X.ic=Xlf))
 jdat <- coda.samples(mod,variable.names=c("m","r","q","X"),n.iter=10000) 
# plot(jdat)

  ## update parameters  
  dat = as.matrix(jdat)
  Xl  = mean(dat[,"X"])
  pl  = 1/var(dat[,"X"])
  CI.LFP[,t] = quantile(dat[,"X"],c(0.025,0.5,0.975))
  mq = mean(dat[,"q"])
  vq = var(dat[,"q"])
  aqp[t+1] = mq*mq/vq
  bqp[t+1] = mq/vq
  mr = mean(dat[,"r"])
  vr = var(dat[,"r"])
  arp[t+1] = mr*mr/vr
  brp[t+1] = mr/vr
  mu.m[t+1] = mean(dat[,"m"])
  p.m[t+1] = 1/var(dat[,"m"])
}

```

## Todo

-   Figure on what happens when a SDA model that doesn't include disturbance encounters disturbance
-   More detail on what a model w/ disturbance but EnKF does during disturbance
-   Get the iterative filter working to see how it compares to the other two
-   Tweak model to have a simple C budget -- where does the C go when the model doesn't assimilate disturbance?
-   Apply to SIPNET SDA

### 2020-10-03 SingleSite_SingleDisturb.Rmd
Iterative JAGS filter (single site, single disturb) working!

@@
 model{
-  N0 ~ dnorm(mu0,pa0)  ## initial conditions
+  N ~ dnorm(muIC,paIC)  ## initial conditions
   ## Forecast
-  muN <- N0 + R*N0*(1-N0/K)
-  muD <- 1
+  muN <- N + R*N*(1-N/K)
+  muD ~ dnorm(mu0,pa0)
   D ~ dbern(p)
-  mu <- D[t]*muD + (1-D)*muN
+  mu <- D*muD + (1-D)*muN
   n ~ dnorm(mu,q)
@@
 ## Forecast Cycle
-CI.LFP = matrix(NA,3,nt)
-Xl = data$mu0
-pl = data$pa0
-aqp = bqp = arp = brp = 0.1
-mu.m = 0
-p.m  = 1e-6
-for(t in 1:nt){
+CI.fN = matrix(NA,5,NT)
+CI.fD = matrix(NA,5,NT)
+Xl = nIC         ## set IC
+pl = 1/n.sd^2
+Dbar = 0
+#mu0 = data$mu0   ## set disturbed state
+#pa0 = data$pa0
+#aqp = data$aq
+#bqp = data$bq
+#arp = data$ar
+#brp = data$br
+for(t in 1:NT){
   ## Forecast + Analysis
-  update = list(Y=Y[t],mu0=Xl,pa0=pl,aq=aqp[t],bq=bqp[t],ar=arp[t],br=brp[t],mu.m=mu.m[t],tau.m=p.m[t])
-  mod <- jags.model(file=textConnection(LinFilterParam),
+  update = list(Y=data$Y[t],
+                muIC=Xl,paIC=pl,
+                mu0=data$mu0,pa0=data$pa0,
+                #aq=aqp[t],bq=bqp[t],ar=arp[t],br=brp[t],
+                R=r, K=K,p = p.dist, q = 1/n.sd^2, r=1/obs.sd^2     ## hard coding these to start
+                )
+  mod <- jags.model(file=textConnection(FilterSSSD),
                    data=update,
-                  n.adapt=1000,n.chains=3,
-                  init=list(X.ic=Xlf))
- jdat <- coda.samples(mod,variable.names=c("m","r","q","X"),n.iter=10000) 
+                  n.adapt=1000,n.chains=3)#,
+#                  init=list(X.ic=Xlf))
+ jdat <- coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000) 
 # plot(jdat)
@@
   dat = as.matrix(jdat)
-  Xl  = mean(dat[,"X"])
-  pl  = 1/var(dat[,"X"])
-  CI.LFP[,t] = quantile(dat[,"X"],c(0.025,0.5,0.975))
-  mq = mean(dat[,"q"])
-  vq = var(dat[,"q"])
-  aqp[t+1] = mq*mq/vq
-  bqp[t+1] = mq/vq
-  mr = mean(dat[,"r"])
-  vr = var(dat[,"r"])
-  arp[t+1] = mr*mr/vr
-  brp[t+1] = mr/vr
-  mu.m[t+1] = mean(dat[,"m"])
-  p.m[t+1] = 1/var(dat[,"m"])
+  Xl  = mean(dat[,"n"])
+  pl  = 1/var(dat[,"n"])
+  CI.fN[,t] = quantile(dat[,"n"],c(0.025,0.25,0.5,0.75,0.975))
+  CI.fD[,t] = quantile(dat[,"D"],c(0.025,0.25,0.5,0.75,0.975))
+  Dbar[t] = mean(dat[,"D"])
+#  mu0[t+1] = mean(dat[,"muD"])
+#  pa0[t+1] = 1/var(dat[,"muD"])
+
+#  mq = mean(dat[,"q"])
+#  vq = var(dat[,"q"])
+#  aqp[t+1] = mq*mq/vq
+#  bqp[t+1] = mq/vq
+#  mr = mean(dat[,"r"])
+#  vr = var(dat[,"r"])
+#  arp[t+1] = mr*mr/vr
+#  brp[t+1] = mr/vr
 }
@@
 
+```{r}
+plot(0,0,xlim=xlim,ylim=c(0,1))
+ecoforecastR::ciEnvelope(time,CI.fD[1,],CI.fD[5,],col=col.alpha(4,0.5))
+ecoforecastR::ciEnvelope(time,CI.fD[2,],CI.fD[4,],col=4)
+lines(time,CI.fD[2,])
+```
+
+```{r}
+## comparison to EnKF
+plot(data$Y,ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
+     ylab="LAI",xlab="time")
+polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
+  an0.stats[1, 1])), border = NA, col = '#FFBF6C')
+polygon(cbind(c(time, rev(time), 1), c(an.stats[2, ], rev(an.stats[4, ]),
+  an.stats[1, 1])), border = NA, col = ecoforecastR::col.alpha('#b2df8a',0.7))
+#ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
+ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
+
+lines(an0.stats[3,],col="orange",lwd=2)
+lines(an.stats[3,],col="#b2df8a",lwd=2)
+lines(CI.fSSSD[2,],col="blue",lwd=2)
+lines(Dbar,lwd=2)
+points(data$Y,pch=18,col=2)
+legend("topright",legend=c("EnKF, no disturb","EnKF, disturb","Multinomial","data","P(dist)"),
+       lwd=5,col=c("orange","#b2df8a","blue","red","black"))
+```
+
 ## Todo
-
--   Figure on what happens when a SDA model that doesn't include disturbance encounters disturbance
--   More detail on what a model w/ disturbance but EnKF does during disturbance
--   Get the iterative filter working to see how it compares to the other two
+-   Visualize bimodal forecast and analysis distributions through time for simple example
 -   Tweak model to have a simple C budget -- where does the C go when the model doesn't assimilate disturbance?

### 2020-10-03 SingleSite_SingleDisturb.Rmd
Added ridgeline and violin visualizations of multimodal Analysis

@@
 knitr::opts_chunk$set(echo = TRUE)
+library(rjags)
+library(ecoforecastR)
+library(tidyverse)
 ```
@@
 #brp = data$br
+Analysis <- list()
 for(t in 1:NT){
@@
 #  brp[t+1] = mr/vr
+
+  ## save Analysis posterior
+  Analysis[[t]] <- dat
+
 }
@@
 
+### Comparision to EnKF
 ```{r}
-## comparison to EnKF
 plot(data$Y,ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
@@
 
+### Ridgeline plot of Analysis
+```{r}
+## organize Analysis state var (n) into a wide format
+n.wide <- sapply(Analysis,function(x){x[,"n"]})
+colnames(n.wide) <- paste0("t",1:NT)
+n.wide <- as.data.frame(n.wide)
+
+## transform to a long format
+n.long <- tidyr::pivot_longer(n.wide[,xlim[1]:xlim[2]],
+                              cols=1:(diff(xlim)+1),
+                              names_to = "time",names_prefix = "t",
+                              values_to = "n")
+n.long$time <- as.numeric(n.long$time)
+
+## ridge plot
+ggplot(n.long) + aes(x=n,y=time,group=time) +
+  ggridges::geom_density_ridges(bandwidth = 0.02) +
+  labs(x = "LAI", y = "Time")
+
+## violin plot
+n.long %>%
+ filter(n >= 0 & n <= 8) %>%
+ ggplot() +
+ aes(x = "", y = n, group = time) +
+ geom_violin(adjust = 0.5, scale = "area", fill = "#0c4c8a") +
+ labs(x = "Time", y = "LAI") +
+ theme_minimal()
+
+```
+
 ## Todo
--   Visualize bimodal forecast and analysis distributions through time for simple example
 -   Tweak model to have a simple C budget -- where does the C go when the model doesn't assimilate disturbance?

### 2020-10-04 SingleSite_SingleDisturb.Rmd
flushing out to-do more

@@
 ## Todo
+-   Move forecast out of JAGS
 -   Tweak model to have a simple C budget -- where does the C go when the model doesn't assimilate disturbance?
 -   Apply to SIPNET SDA
+-   Calibrate state and transition probabilities for a single region
+-   Upgrade code to multinomial
+-   Add PFT switching
+-   Add ensemble weighting so that we can manage ensemble sizes
+-   Add "fusion" if a case becomes too rare to worry about
+-   Think about whether we need to add LULC state to state variables & how LULC state would be assimilated
+-   Test against subseasonal dynamics (e.g. distinguishing phenology from disturbance)
+-   investigate whether we need a "jumpback" capacity once disturbance event type and timing is resolved
+-

### 2020-10-17 MultiSite_MultiPFT_MultiAge.Rmd
Some thoughts on how to move to the landscape model

---
title: "MultiSite_MultiPFT_MultiAge"
author: "Mike Dietze"
date: "10/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plan

* Consider a PFT by age class matrix
* An initial ensemble run of a model would generate IC predictions, and thus a (multivariate state) ensemble PDF of each class
* Each pixel on a landscape could be assigned a probability of being in each of class based on `n` data constraints. For example a pixel with high LAI may be more likely for a range of forest age classes and unlikely for a grassland or recent disturbance. A high AGB associated with the same pixel would push the probability to older age forests, but it may be hard to distinguish 80-90, 90-100, 100-150, etc. so they may have similar weight.
* Also store the posterior quantile associated with the most likely class as a check on fit -- if you're a bad fit to your most likely fit this means the model's missing something.
* We then forward simulate the landscape to the next step. Each element in the PFTxAGE matrix has a probability of transitioning to the other, though many will be 0. Task is to update the per-pixel probabilities of being in each class, while simultaneously updating the system state.
 + for most cases, additional information will increase the probability associated with a given age class
 + Hopefully we capture aging naturally -- e.g. ~10% of a 10yr age bin should shift older
 + Us what we learned at the single site scale to capture the probability of transition to a disturbed state or shift to a different PFT
 + Might need to have multiple AGE 0 options if the transitions FROM are really different and predict different transitions TO. i.e. may need to start with all possible age 0 classes and will need to revisit patch fusion
 + At the same time, need to do the Assimilation step for all ensemble members, using their weights (probabilities) appropriately in updating state. Need to make sure this doesn't do stupid things like locking in a class and preventing it from maturing
 + Should try adding age class to the single site model to get ready for this
 + Thankfully, not all transitions are possible, and not all are equally likely, so hoping this will put real, strong constraints on the problem
 + Have to shift to treating p(disturbance) as a hierarchical model, where each year's posterior is a mixed effect (deterministic predictors + random variability) and then we update the model each year as we learn about disturbance drivers and as the probabilities shift (like Beckage fire paper, the underlying risk needs to evolve through time)

### 2020-10-19 SingleSite_SingleDisturb.Rmd
SSSD: separate R model from JAGS forecast

@@
 
+## Foreward filter, separate process model
+
+In the previous code, both the forecast and analysis steps were done within JAGS.
+In this version we're pulling out the Forecast step and running that in R since (i) that's how most models work and (ii) it will allow us to more easily distinguish the Forecast prior and Analysis posteriors
+
+```{r}
+AnalysisSSSD <- "
+model{
+  ## prior
+  muN ~ dnorm(mufN,pfN)
+  muD ~ dnorm(mufD,pfD)
+  D ~ dbern(p)
+  ## Analysis
+  n <- D*muD + (1-D)*muN
+  Y ~ dnorm(n,r)
+}"
+## Storage
+CI.fN = matrix(NA,5,NT)
+CI.fD = matrix(NA,5,NT)
+Dbar = 0
+Analysis <- Forecast <- list()
+## set IC
+nMC = 5000
+Xl = nIC         
+pl = 1/n.sd^2
+Na = rnorm(nMC,Xl,n.sd)
+Nf = N*NA
+
+for(t in 1:NT){
+  ## Forecast 
+  d.prior <- rbinom(nMC,1,p.dist)   ## prior on disturbance status
+  sel <- !as.logical(d.prior)
+  Nf[!sel] <- rnorm(sum(d.prior),data$mu0,1/sqrt(data$pa0))  ## draw disturbed predictions
+  Nf[sel]  <- rnorm(sum(sel),Na[sel] + r*Na[sel]*(1-Na[sel]/K),n.sd) ## draw undisturbed
+  Nf[Nf < 0] <- 0 ## Tobit
+
+  ## priors
+  priors <- list(p    = mean(d.prior),
+                 mufN = mean(Nf[sel]),
+                 mufD = mean(Nf[!sel]),
+                 pfN  = 1/var(Nf[sel]),
+                 pfD  = 1/var(Nf[!sel])
+                 )
+  Forecast[[t]] <- c(list(d.prior=d.prior,Nf=Nf),priors)
+
+  ## Analysis
+  update = c(list(Y=data$Y[t],r=1/obs.sd^2),priors)
+  mod <- jags.model(file=textConnection(AnalysisSSSD),
+                   data=update,
+                  n.adapt=1000,n.chains=3)
+ jdat <- coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000) 
+# plot(jdat)
+
+  ## update parameters  
+  dat = as.matrix(jdat)
+  Xl  = mean(dat[,"n"])
+  pl  = 1/var(dat[,"n"])
+  CI.fN[,t] = quantile(dat[,"n"],c(0.025,0.25,0.5,0.75,0.975))
+  CI.fD[,t] = quantile(dat[,"D"],c(0.025,0.25,0.5,0.75,0.975))
+  Dbar[t] = mean(dat[,"D"])
+
+  ## save Analysis posterior
+  Analysis[[t]] <- dat
+
+  ## draw new IC
+  Na <- sample(dat[,"n"],nMC,replace=TRUE)
+
+}
+
+```
+
+```{r}
+## organize Analysis state var (n) into a wide format
+n.wide <- sapply(Analysis,function(x){x[,"n"]})
+colnames(n.wide) <- paste0("t",1:NT)
+n.wide <- as.data.frame(n.wide)
+
+## transform to a long format
+n.long <- tidyr::pivot_longer(n.wide[,xlim[1]:xlim[2]],
+                              cols=1:(diff(xlim)+1),
+                              names_to = "time",names_prefix = "t",
+                              values_to = "n")
+n.long$time <- as.numeric(n.long$time)
+
+## ridge plot
+ggplot(n.long) + aes(x=n,y=time,group=time) +
+  ggridges::geom_density_ridges(bandwidth = 0.02) +
+  labs(x = "LAI", y = "Time")
+
+## violin plot
+n.long %>%
+ filter(n >= 0 & n <= 8) %>%
+ ggplot() +
+ aes(x = "", y = n, group = time) +
+ geom_violin(adjust = 0.5, scale = "area", fill = "#0c4c8a") +
+ labs(x = "Time", y = "LAI") +
+ theme_minimal()
+```
+
+### Comparision to EnKF
+```{r}
+plot(data$Y,ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
+     ylab="LAI",xlab="time")
+polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
+  an0.stats[1, 1])), border = NA, col = '#FFBF6C')
+polygon(cbind(c(time, rev(time), 1), c(an.stats[2, ], rev(an.stats[4, ]),
+  an.stats[1, 1])), border = NA, col = ecoforecastR::col.alpha('#b2df8a',0.7))
+#ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
+ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
+
+lines(an0.stats[3,],col="orange",lwd=2)
+lines(an.stats[3,],col="#b2df8a",lwd=2)
+lines(CI.fSSSD[2,],col="blue",lwd=2)
+lines(Dbar,lwd=2)
+points(data$Y,pch=18,col=2)
+legend("topright",legend=c("EnKF, no disturb","EnKF, disturb","Multinomial","data","P(dist)"),
+       lwd=5,col=c("orange","#b2df8a","blue","red","black"))
+```
+
 ## Todo
--   Move forecast out of JAGS
 -   Tweak model to have a simple C budget -- where does the C go when the model doesn't assimilate disturbance?

### 2020-11-12 SingleSite_SingleDisturb.Rmd
SSSD: update todo with more detail about SIPNET test & need for real world SSSD-MultiPool: migrate code to work with VSEM

@@
 -   Tweak model to have a simple C budget -- where does the C go when the model doesn't assimilate disturbance?
--   Apply to SIPNET SDA
 -   Calibrate state and transition probabilities for a single region
@@
 -   Add PFT switching
+-   Apply to SIPNET SDA
+    - version of code into PEcAn
+    - test coupling to SIPNET
+-   Real-world data examples
 -   Add ensemble weighting so that we can manage ensemble sizes

### 2020-11-12 SingleSite_SingleDisturb_MultiPool.Rmd
SSSD: update todo with more detail about SIPNET test & need for real world SSSD-MultiPool: migrate code to work with VSEM

---
title: "Single Site, Single Disturbance, MultiPool"
author: "Mike Dietze"
date: "10/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BayesianTools)
set.seed(42)
```

## Todo

-   Tweak model to have a simple C budget -- where does the C go when the model doesn't assimilate disturbance?

## helper functions

```{r}
tobit <- function(x){
  x[x < 0] <- 0
  x
}

disturbance <- function(x){ ## x = Bleaf, Bsoil, Bstem
  old.biomass <- x[c(1,3)]
  new.biomass <- tobit(mvtnorm::rmvnorm(1,mu0,V0)) ## draw disturbed leaf and stem
  check.max <- which(new.biomass > old.biomass)
  if(length(check.max)>0) new.biomass[check.max] = old.biomass[check.max]
  residual = sum(old.biomass-new.biomass)
  x[c(1,3)] <- new.biomass
  x[2] <- x[2] + residual*alloc.soil
  removal <- residual*(1-alloc.soil)
  return(x)
}

```

## drivers

```{r}
NT = 25
nt = 365*NT
time = 1:nt
nd   = 31 ## number of driver ensembles
PAR <- matrix(NA,nd,nt)
for(i in 1:nd){
  PAR[i,] <- VSEMcreatePAR(time)  
}
plot(PAR[1,], main = "PAR (driving the model)", xlab = "Day",cex=0.35)
```

## Parameters

```{r}
refPars <- VSEMgetDefaults()
refPars$lower[9:11] = 1
np = 50
sigmaP = diag((refPars$upper - refPars$lower)/10)^2 ## assume range is +/- 5 SD
Theta <- as.data.frame(mvtnorm::rmvnorm(np,refPars$best,sigmaP))
colnames(Theta) <- rownames(refPars)
par(mfrow=c(3,4))
for(i in 1:11){
  Theta[Theta[,i] < refPars$lower[i] , i] <- refPars$lower[i]
  Theta[Theta[,i] > refPars$upper[i] , i] <- refPars$upper[i]
  hist(Theta[,i],main=colnames(Theta)[i])
}
```

## initial conditions

```{r}
## spin up
ne = 100
met = sample.int(nd,ne,replace=TRUE)
par = sample.int(np,ne,replace=TRUE)
Na <- matrix(NA,ne,3)
for(i in 1:ne){
  Na[i,] = VSEM(as.numeric(Theta[par[i],]),PAR = PAR[met[i],])[nt,2:4]
}
Na.orig = Na
Nmu = Nf = Na*NA
par(mfrow=c(1,3))
hist(Na[,1],main="Cleaf")
hist(Na[,2],main="Csoil")
hist(Na[,3],main="Cstem")
```

## set up storage and parameters

```{r}
## hard coded 
Q = diag(c(0.1,0.1,0.1)) ## process error
R = diag(c(.25,10,1))      ## observation error
THETA <- as.matrix(Theta[par,]) ## ensemble parameters
mu0 <- c(1,1)            ## mean leaf & stem biomass after disturbance
V0  <- diag(c(0.15,0.25))^2## var in post disturbance leaf and stem biomass
alloc.soil = 0.25        ## fraction of removed C that goes in the soil
p.dist = 0.05

## storage
Dbar = 0
Analysis <- Forecast <- list()
removal = matrix(0,ne,NT)

## pseudodata
No = matrix(NA,NT,3)                   ## storage for observed state
Ntrue = matrix(NA,NT,3)                ## storage for latent "true" state
true = 42                              ## reference ensemble member
D.true = rep(0,NT)                     ## true disturbance timeseries
D.true[10] = 1                         ## prescribe a disturbance
refpar = as.numeric(Theta[par[true],]) ## grab the "true" parameters
refpar[9:11] = Na[true,]               ## set initial biomass pools
for(t in 1:NT){
  tsel <- (t-1)*365 + 1:365            ## select 1 yr of met  
  ref.mu = VSEM(refpar,PAR[met[true],tsel])[365,2:4]    ## forward simulation
  ref.f  = mvtnorm::rmvnorm(1,ref.mu,Q)## process error
  if(D.true[t]){
    ref.f = disturbance(ref.f)
  }
  ref.f = tobit(ref.f)
  refpar[9:11] = ref.f                 ## update IC for next iteration
  Ntrue[t,] = ref.f
  No[t,] = tobit(mvtnorm::rmvnorm(1,ref.f,R)) ## observation error
}
par(mfrow=c(3,1))
plot(No[,1])
plot(No[,2])
plot(No[,3])
```

```{r,echo=FALSE,results='hide',message=FALSE}
AnalysisSSSDMP <- "  ## SingleSite SingleDisturbance MultiPool
model{
  ## prior
  muN ~ dmnorm(mufN,pfN)                ## forecast prior on Not disturbed
  muD ~ dmnorm(mufD,pfD)                ## forecast prior on Disturbed
  D ~ dbern(p)                          ## disturbance probability
  ## Analysis 
  n <- D*muD + (1-D)*muN                ## select between disturbed and undisturbed
  Y ~ dmnorm(n,R)                       ## Observation error
}"

for(t in 1:NT){

  tsel <- (t-1)*365 + 1:365               ## Select days from driver files

  ## update IC
  THETA[,9:11] <- Na

  ## Forecast 
  for(i in 1:ne){
    Nmu[i,] = VSEM(THETA[i,],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
    Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
  }

  ## Disturbance
  d.prior <- rbinom(ne,1,p.dist)             ## prior on disturbance status
  sel <- !as.logical(d.prior)                ## select undisturbed ensemble members
  if(sum(!sel)>2){
    sel.dist = which(!sel)
    for(j in sel.dist){                      ## run disturbance model
      Nf[j,] = disturbance(Nf[j,])
    }
    mufD = colMeans(Nf[!sel,])
    pfD = solve(diag(c(0.15,0.5,0.25)^2)) #try(solve(var(Nf[!sel,])))
  } 
  if(sum(!sel) <= 2 || class(pfD) == "try-error") {   ## backup values if the ensemble stats fail
      mufD = c(mu0[1],mean(Nf[,2]),mu0[2])             
      pfD  = solve(diag(c(0.15,0.5,0.25)^2))
  }
  Nf <- tobit(Nf)

  ## priors
  priors <- list(p    = mean(d.prior),         ## disturbance rate
                 mufN = colMeans(Nf[sel,]),    ## Not disturbed forecast means
                 mufD = mufD,   ## Disturbed forecast means
                 pfN  = solve(var(Nf[sel,])),  ## Not disturbed forecast precision
                 pfD  = pfD  ## Disturbed forecast precision
  )
  Forecast[[t]] <- c(list(d.prior=d.prior,Nf=Nf),priors)

  ## Analysis
  update = c(list(Y=No[t,],R=solve(R)),priors)
  mod <- rjags::jags.model(file=textConnection(AnalysisSSSDMP),
                           data=update,
                           n.adapt=1000,n.chains=3)
  jdat <- rjags::coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000)

  ## update parameters & save Analysis posterior
  dat = as.matrix(jdat)
  nsel <- grep(pattern="^n",x=colnames(dat))
  Analysis[[t]] <- dat

  ## draw new IC     *** REPLACE WITH ENSEMBLE ADJUSTMENT ***
  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
  Na <- dat[sel.ic,nsel]
}

```

```{r}
pool.names=c("Cleaf","Csoil","Croot")
for(i in 1:3){

  CI.Na = sapply(X = Analysis,function(x){
    sel.n = which(colnames(x) == paste0("n[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  #CI.Na = sapply(X = Analysis,function(x){
  #  apply(x,2,quantile,c(0.025,0.975))
  #})

  plot(No[,i],
       #ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[2,],col="#b2df8a",lwd=2)
  lines(Ntrue[,i],lty=2,lwd=2)
}
```

## EnKF, no disturbance

```{r,echo=FALSE}
Na = Na.orig
Forecast.EnKF = Analysis.EnKF = list()
for(t in 1:NT){

  tsel <- (t-1)*365 + 1:365               ## Select days from driver files

  ## update IC
  THETA[,9:11] <- Na

  ## Forecast 
  for(i in 1:ne){
    Nmu[i,] = VSEM(THETA[i,],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
    Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
  }

  muf = colMeans(Nf[sel,])    ## Not disturbed forecast means
  Pf  = var(Nf[sel,])
  pf  = solve(Pf)             ## Not disturbed forecast precision
  Forecast.EnKF[[t]] <- list(d.prior=d.prior,Nf=Nf,muf=muf,pf=pf)

  ## Analysis
  I = H <- diag(rep(1,3))
  K <- Pf %*% t(H) %*% solve(H%*%Pf%*%t(H) + R)
  mua <- muf + K%*%(No[t,] - muf) 
  Pa <- (I-K %*% H)%*%Pf    
  pa = solve(Pa)

  ## draw new IC     *** REPLACE WITH ENSEMBLE ADJUSTMENT ***
  Na <- mvtnorm::rmvnorm(ne,mua,Pa)

  ## save Analysis posterior
  Analysis.EnKF[[t]] <- list(mua=mua,pa=pa,Pa=Pa,Na=Na)

}

```

### 2020-11-12 Paper.Rmd
paper outline

---
title: "Disturbance Assimilation"
author: "Michael Dietze"
date: '`r Sys.Date()`'
output:
  html_document: 
    fig_caption: yes
    number_sections: yes
    toc: yes
    keep_tex: true
bibliography: bibliography.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Outline

# Abstract

# Intro

**Borrow text from CMS proposals**

1. The context: MRV and predicting Terrestrial C
2. Status quo: LULC * look up table
  + very coarse, doesn't capture spatiotemporal varibility
  + rightly captures that disturbance and land use change both critical to C accounting and, more importantly, a place where we can exert some control
3. Alternative: SDA
  + models capture process
  + data constrains (data driven, model as cov)
  + data fusion
  + seamless transition from reanalysis to nowcast to forecast [@Dietze2017a]
  + Disturbance is exactly why you would want to nudge a model back to reality
4. The problem: Conventional SDA doesn't capture jumps
  + series of nudges
  + potentially misattributes pools and fluxes
  + "drags down" neighbors **no sims of this yet**
5. The solution:
  + assimilate P(dist)
  + update C | dist
6. This paper
  + Illustration of problem
  + Derivation Multinomial
  + Simulated data experiments (known disturbance)
  + Real-world example

# Methods

1. Illustration of problem
2. Derivation Multinomial

Conceptual figure

3. Simulated data experiments (known disturbance)
  + Single Site, Single Disturbance
  + SSSD Multi Pool
  + SS Multi-Disturbance w/ PFT switching
  + Simple Multi-site: spatial cov
4. Real-world examples
  + Describe Model & PEcAn
  + Calibrate state-and-transition prior
    + Landtrendr
    ftp://islay.ceoas.oregonstate.edu/change_agent_attribution
  + Applications: PNW
    + Logging
    + Fire
    + Land use transition (forest -> non?)

Email to Robert:

I need to extract some example time series for single pixels (e.g. AGB product) where the disturbance it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance). From Landtrendr itself I’m looking to know when you attributed the year, severity, and cause so we can see if our assimilation is getting it “right”. Bonus would be if you could suggest places where you know there’s ground data we could use to validate the model (disturbances at flux towers, LTER sites, FIA plots, etc).

<!---
```{r child = 'methods.Rmd'}
```
--->

# Results

1. Simulated data experiments (known disturbance)
  + Single Site, Single Disturbance

Figures showing the current problem, solution, and that we've got a genuine multimodal distribution and P(dist)

  + SSSD Multi Pool

Figures showing pool cov, C accounting / attribution

  + SS Multi-Disturbance w/ PFT switching

Figure showing PFT switching vs not, ability to detect disturbance type

  + Simple Multi-site: spatial cov

Figure showing "dragging down" of neighbors in conventional and breaking of COV in Multinomial

2. Real-world examples
  + Calibrate state-and-transition prior
    * data: (samples from) maps of disturbance type by year
    * model: Multinomial Dirlichet

Figure showing map and estimated priors

  + Applications
    + Logging
    + Fire
    + Land use transition (forest -> non?)

# Discussion

* Interpretation
  * TBD based on results

* Implications  (borrow text from CMS proposals)
  * Improving MRV
  * Recovery forecasting
  * Disturbance nowcasting
    * requires reducing temporal resolution

* Future directions
  * Disturbance priors need to be spatially heterogenous and evolve temporally
    * Connect to state-and-transition literature
  * Scaling up -> spatially implicit
    * ED2
    * Contagion

# Acknowledgements

NASA CMS, PEcAn, NEFI

# Bibliography

### 2020-11-17 SingleSite_SingleDisturb_MultiPool.Rmd
SSSD-MV: added disturbed EnKF and ensemble adjustment to both EnKFs

@@
 Na.orig = Na
-Nmu = Nf = Na*NA
+Nmu = NmuK = NfK = Nf = Na*NA
 par(mfrow=c(1,3))
@@
 Analysis <- Forecast <- list()
+Forecast.KD = Analysis.KD = list() ## EnKF with Disturbance
 removal = matrix(0,ne,NT)
@@
 
+Na <- NaK <- Na.orig
+THETAK <- THETA
 for(t in 1:NT){
@@
   ## update IC
-  THETA[,9:11] <- Na
+  THETA[,9:11]  <- Na    ## Multinomial
+  THETAK[,9:11] <- NaK  ## EnKF w/ disturbance
 
@@
   for(i in 1:ne){
-    Nmu[i,] = VSEM(THETA[i,],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
-    Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
+    Nmu[i,]  = VSEM(THETA[i,], PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
+    NmuK[i,] = VSEM(THETAK[i,],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
+    epsilon  = mvtnorm::rmvnorm(1,rep(0,3),Q)                   ## process error  ** shared between versions **
+    Nf[i,]   = Nmu[i,]  + epsilon
+    NfK[i,]  = NmuK[i,] + epsilon
   }
@@
       Nf[j,]  = disturbance(Nf[j,])
+      NfK[j,] = disturbance(NfK[j,])
     }
@@
   if(sum(!sel) <= 2 || class(pfD) == "try-error") {   ## backup values if the ensemble stats fail
-      mufD = c(mu0[1],mean(Nf[,2]),mu0[2])             
+      mufD = c(mu0[1], mean(Nf[,2]),  mu0[2])            
       pfD  = solve(diag(c(0.15,0.5,0.25)^2))
@@
   Nf <- tobit(Nf)
+
+  ## EnKF forecast stats
+  NfK <- tobit(NfK)
+  mufK = colMeans(NfK)
+  PfK  = var(NfK)
+  pfK  = solve(PfK)             ## Not disturbed forecast precision
+  Forecast.KD[[t]] <- list(d.prior=d.prior,Nf=NfK,muf=mufK,pf=pfK)
 
@@
   Na <- dat[sel.ic,nsel]
+
+  ## EnKF Analysis
+  I = H <- diag(rep(1,3))
+  K <- PfK %*% t(H) %*% solve(H%*%PfK%*%t(H) + R)
+  muaK <- mufK + K%*%(No[t,] - mufK) 
+  PaK <- (I-K %*% H)%*%PfK    
+  paK = solve(PaK)
+  NaK <- PEcAn.assim.sequential::adj.ens(Pf = PfK,X = NfK,mu.f = mufK,mu.a = muaK,Pa = PaK)
+  Analysis.KD[[t]] <- list(mua=muaK,pa=paK,Pa=PaK,Na=NaK)
+
 }
@@
 
-  muf = colMeans(Nf[sel,])    ## Not disturbed forecast means
-  Pf  = var(Nf[sel,])
+  muf = colMeans(Nf)    ## Not disturbed forecast means
+  Pf  = var(Nf)
   pf  = solve(Pf)             ## Not disturbed forecast precision
@@
 
-  ## draw new IC     *** REPLACE WITH ENSEMBLE ADJUSTMENT ***
-  Na <- mvtnorm::rmvnorm(ne,mua,Pa)
+  ## draw new IC     *** REPLACED WITH ENSEMBLE ADJUSTMENT ***
+#  Na <- mvtnorm::rmvnorm(ne,mua,Pa)
+  Na <- PEcAn.assim.sequential::adj.ens(Pf = Pf,X = Nf,mu.f = muf,mu.a = mua,Pa = Pa)
 

### 2020-11-17 Paper.Rmd
Methods outline to methods.Rmd

@@
 
-# Methods
-
-1. Illustration of problem
-2. Derivation Multinomial
-
-Conceptual figure
-
-3. Simulated data experiments (known disturbance)
-  + Single Site, Single Disturbance
-  + SSSD Multi Pool
-  + SS Multi-Disturbance w/ PFT switching
-  + Simple Multi-site: spatial cov
-4. Real-world examples
-  + Describe Model & PEcAn
-  + Calibrate state-and-transition prior
-    + Landtrendr
-    ftp://islay.ceoas.oregonstate.edu/change_agent_attribution
-  + Applications: PNW
-    + Logging
-    + Fire
-    + Land use transition (forest -> non?)
-
-Email to Robert:
-
-I need to extract some example time series for single pixels (e.g.
-AGB product) where the disturbance it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance).
-From Landtrendr itself I’m looking to know when you attributed the year, severity, and cause so we can see if our assimilation is getting it “right”.
-Bonus would be if you could suggest places where you know there’s ground data we could use to validate the model (disturbances at flux towers, LTER sites, FIA plots, etc).
-
-<!---
+<!---- METHODS --->
 ```{r child = 'methods.Rmd'}
 ```
---->
 

### 2020-11-17 methods.Rmd
Methods outline to methods.Rmd

# Methods

1. Illustration of problem
2. Derivation Multinomial

Conceptual figure

3. Simulated data experiments (known disturbance)
  + Single Site, Single Disturbance
  + SSSD Multi Pool
  + SS Multi-Disturbance w/ PFT switching
  + Simple Multi-site: spatial cov
4. Real-world examples
  + Describe Model & PEcAn
  + Calibrate state-and-transition prior
    + Landtrendr
    ftp://islay.ceoas.oregonstate.edu/change_agent_attribution
  + Applications: PNW
    + Logging
    + Fire
    + Land use transition (forest -> non?)

Email to Robert:

I need to extract some example time series for single pixels (e.g. AGB product) where the disturbance it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance). From Landtrendr itself I’m looking to know when you attributed the year, severity, and cause so we can see if our assimilation is getting it “right”. Bonus would be if you could suggest places where you know there’s ground data we could use to validate the model (disturbances at flux towers, LTER sites, FIA plots, etc).

### 2020-11-17 SingleSite_SingleDisturb_MultiPool.Rmd
Added multi.ens.adj function to update disturbance class assigments and state variables

@@
 library(BayesianTools)
+library(magrittr)
 set.seed(42)
@@
 ```{r}
+source("multi.ens.adj.R")
+
 tobit <- function(x){
@@
 
-  ## draw new IC     *** REPLACE WITH ENSEMBLE ADJUSTMENT ***
-  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
-  Na <- dat[sel.ic,nsel]
+  ## draw new IC
+  Na <- multi.ens.adj(
+          Xf   = Nf,
+          cf   = d.prior,
+          mu.f = rbind(priors$mufN,priors$mufD),
+          Pf   = abind::abind(solve(priors$pfN),
+                 solve(priors$pfD),
+                 along=3),
+          Xa = dat[,nsel],
+          ca = dat[,"D"])
+  Ca <- Na[,1] ## updated class assignments
+  Na <- tobit(Na[,-1])
+#  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
+#  Na <- dat[sel.ic,nsel]
 
@@
   paK = solve(PaK)
-  NaK <- PEcAn.assim.sequential::adj.ens(Pf = PfK,X = NfK,mu.f = mufK,mu.a = muaK,Pa = PaK)
+  NaK <- PEcAn.assim.sequential::adj.ens(Pf = PfK,X = NfK,mu.f = mufK,mu.a = muaK,Pa = PaK) %>% as.matrix()
   Analysis.KD[[t]] <- list(mua=muaK,pa=paK,Pa=PaK,Na=NaK)
@@
 
-  ## draw new IC     *** REPLACED WITH ENSEMBLE ADJUSTMENT ***
+  ## draw new IC
   #  Na <- mvtnorm::rmvnorm(ne,mua,Pa)
-  Na <- PEcAn.assim.sequential::adj.ens(Pf = Pf,X = Nf,mu.f = muf,mu.a = mua,Pa = Pa)
+  Na <- PEcAn.assim.sequential::adj.ens(
+    Pf = Pf,X = Nf,mu.f = muf,mu.a = mua,Pa = Pa)
 

### 2020-11-17 multi.ens.adj.R

@@ -1,0 +1,125 @@
+##' @title multi.ens.adj
+##' @name  multi.ens.adj
+##' @author Michael Dietze \email{dietze@@bu.edu}
+##' 
+##' 
+##' @param Xf Dataframe or matrix of forecast state variables for different ensembles.
+##' @param cf vector assigning forecast disturbance class to each ensemble member
+##' @param mu.f A vector with forecast mean estimates of state variables.
+##' @param Pf  A cov matrix of forecast state variables.  
+##' @param Xa A matrix of posterior samples of state variables.
+##' @param ca A vector of posterior samples of disturbance class
+##' 
+##' @return Returns a matrix of adjusted analysis mean estimates of state variables and class assignments
+##' @export
+
+multi.ens.adj<-function(Xf, cf, mu.f, Pf, Xa, ca){
+  
+  if(FALSE){  ## set params
+    Xf = Nf
+    mu.f = rbind(priors$mufN,priors$mufD)
+    Pf   = abind::abind(solve(priors$pfN),
+                 solve(priors$pfD),
+                 along=3)
+    cf = d.prior
+    Xa = dat[,nsel]
+    ca = dat[,"D"]
+  }
+  if(FALSE){ ## figures
+    plot(density(Xf[,1]),main="")
+    lines(density(Xa[,1]),col=2)
+    points(Xf[,1],rep(0,nrow(Xf)),pch="|")
+    plot(ecdf(Xf[,1]))
+    lines(ecdf(Xa[,1]),col=2)
+  }
+  
+  ## reassign classes
+  ff = table(cf)/length(cf) ## class frequency in the forecast
+  uc = sort(unique(cf))
+  fa = table_by(ca,uc)/length(ca) ## class frequency in the posterior
+  df = pmax(fa-ff,0) ## difference in frequency
+  df = df/sum(df)    ## posterior reassignment frequency
+  cA = cf            ## class assigned in the Analysis
+  for(i in seq_along(uc)){
+    if(fa[i] < ff[i]){ ## if a class decreases in frequency in the analysis
+      sel.c = which(cf == uc[i])
+      cA[sel.c] = sample(uc,length(sel.c),prob=df,replace = TRUE)
+    }
+  }
[…]

### 2020-11-18 SingleSite_SingleDisturb.Rmd
Single Site Multi-Disturbance Multi-Pool

@@
 
-plot(time,n,type='l',ylim=range(n),xlim=xlim,xlim=xlim,ylab="LAI",lwd=3)
+plot(time,n,type='l',ylim=range(n),xlim=xlim,ylab="LAI",lwd=3)
 obs.sd = 0.75
@@
 polygon(cbind(c(time, rev(time), 1), c(n0.stats[1, ], rev(n0.stats[3, ]),
-  n.stats[1, 1])), border = NA, col = 'lightblue')
+  n0.stats[1, 1])), border = NA, col = 'lightblue')
 lines(time,n0.stats[2,],lwd=2,col="blue4")
@@
 ```{r}
-plot(data$Y,ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
+plot(data$Y,ylim = range(rbind(CI.fN,data$Y)),xlim=xlim,type='n',
      ylab="LAI",xlab="time")
@@
 lines(an.stats[3,],col="#b2df8a",lwd=2)
-lines(CI.fSSSD[2,],col="blue",lwd=2)
+#lines(CI.fSSSD[2,],col="blue",lwd=2)
+lines(CI.fN[3,],col="blue",lwd=2)
 lines(Dbar,lwd=2)

### 2020-11-18 SingleSite_SingleDisturb_MultiPool.Rmd
Single Site Multi-Disturbance Multi-Pool

@@
 ## spin up
-ne = 100
+ne = 200
 met = sample.int(nd,ne,replace=TRUE)
@@
   for(i in 1:ne){
-    Nmu[i,]  = VSEM(THETA[i,], PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
-    NmuK[i,] = VSEM(THETAK[i,],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
+    fx       = VSEM(THETA[i,], PAR = PAR[met[i],tsel]) ## deterministic simulation
+    Nmu[i,]  =  fx[365,2:4]                         ## state variables at time of assim
+    fxK      = VSEM(THETAK[i,],PAR = PAR[met[i],tsel]) ## deterministic simulation
+    NmuK[i,] = fxK[365,2:4] ## deterministic simulation
     epsilon  = mvtnorm::rmvnorm(1,rep(0,3),Q)                   ## process error  ** shared between versions **
@@
   nsel <- grep(pattern="^n",x=colnames(dat))
-  Analysis[[t]] <- dat
 
@@
 #  Na <- dat[sel.ic,nsel]
+  Analysis[[t]] <- list(MCMC=dat,
+                        Na = Na,
+                        Ca = Ca)
 
@@
 
+### time series
+
 ```{r}
-pool.names=c("Cleaf","Csoil","Croot")
+pool.names=c("Cleaf","Csoil","Cstem")
 for(i in 1:3){
 
-  CI.Na = sapply(X = Analysis,function(x){
+  CI.Na = sapply(X = Analysis,function(y){
+    x = y$MCMC
     sel.n = which(colnames(x) == paste0("n[",i,"]"))
@@
 ```
+### Ensemble adj at time of disturbance
 
+```{r,fig.asp=1,fig.width=3}
+#par(mfrow=c(1,3))
+
+plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("red",alpha=0.5),
+     ylim=range(Forecast[[10]]$Nf[,3]),
+     xlim=range(Forecast[[10]]$Nf[,1]),
+     xlab=pool.names[1],ylab=pool.names[3],
+     cex.lab=1.25,asp = 1)
+points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=0.5)
+points(No[10,1],No[10,3],cex=3,col=ecoforecastR::col.alpha("purple",alpha = 0.75),pch=18)
+
+## ensemble adjustment
+sel.a=sample.int(ne,10)
+arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
+       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
+legend("topleft",
+       legend=c("Forecast","Analysis","Data"),
+       pch=c(1,3,18),col=c(1,2,"purple"),
+       cex = 1.5)
+
+```
 ## EnKF, no disturbance

### 2020-11-18 methods.Rmd
Single Site Multi-Disturbance Multi-Pool

@@
 
+*Priors*
+
+  $$X^L_k \sim MVN(\vec{\mu}_{f,k},P_{f,k})$$
+  $$\vec{\rho} \sim Multinomial(\vec{p})$$ 
+*Process Model*
+
+  $$X = X^L \vec{\rho}$$
+
+*Likelihood*
+  $$Y \sim MVN(X,R)$$
+### Ensemble Adjustment
+
+[@Anderson2001]
+
 Conceptual figure
@@
 3. Simulated data experiments (known disturbance)
-  + Single Site, Single Disturbance
-  + SSSD Multi Pool
-  + SS Multi-Disturbance w/ PFT switching
+  + x Single Site, Single Disturbance  (add a case where EnKF has disturb but it's more rare)
+  + x SSSD Multi Pool
+  + x SS Multi-Disturbance - distinct
+  + SS Multi-Disturbance w/ distinct, PFT switching
+  + SSMD - initially overlapping but diverge
+      * how to handle updating of assignment based on new information?
+  + SSMD - transition probabilities change with state
   + Simple Multi-site: spatial cov
+  + rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
 4. Real-world examples

### 2020-11-18 disturbance.R

@@ -1,0 +1,28 @@
+tobit <- function(x){
+  x[x < 0] <- 0
+  x
+}
+
+disturbance <- function(x){ ## x = Bleaf, Bsoil, Bstem
+  old.biomass <- x[c(1,3)]
+  new.biomass <- tobit(mvtnorm::rmvnorm(1,mu0,V0)) ## draw disturbed leaf and stem
+  check.max <- which(new.biomass > old.biomass)
+  if(length(check.max)>0) new.biomass[check.max] = old.biomass[check.max]
+  residual = sum(old.biomass-new.biomass)
+  x[c(1,3)] <- new.biomass
+  x[2] <- x[2] + residual*alloc.soil
+  removal <- residual*(1-alloc.soil)
+  return(x)
+}
+
+disturbance2 <- function(x,type){ ## x = Bleaf, Bsoil, Bstem
+  old.biomass <- x[c(1,3)]
+  new.biomass <- tobit(mvtnorm::rmvnorm(1,mu0[,type]*old.biomass,V0[,,type])) ## draw disturbed leaf and stem
+  check.max <- which(new.biomass > old.biomass)
+  if(length(check.max)>0) new.biomass[check.max] = old.biomass[check.max]
+  residual = sum(old.biomass-new.biomass)
+  x[c(1,3)] <- new.biomass
+  x[2] <- x[2] + residual*alloc.soil
+  removal <- residual*(1-alloc.soil)
+  return(x)
+}

### 2020-11-18 multi.ens.adj.R

@@ -16,6 +16,7 @@
 multi.ens.adj<-function(Xf, cf, mu.f, Pf, Xa, ca){
   
   if(FALSE){  ## set params
+    ## single
     Xf = Nf
     mu.f = rbind(priors$mufN,priors$mufD)
     Pf   = abind::abind(solve(priors$pfN),
@@ -24,6 +25,17 @@ multi.ens.adj<-function(Xf, cf, mu.f, Pf, Xa, ca){
     cf = d.prior
     Xa = dat[,nsel]
     ca = dat[,"D"]
+    ## multi class
+    Xf   = Nf
+    cf   = d.class
+    mu.f = t(priors$muf)
+    Pf   = abind::abind(    ## find a way to do this with apply(priors$pf,3,solve)
+      solve(priors$pf[,,1]),
+      solve(priors$pf[,,2]),
+      solve(priors$pf[,,3]),
+      along=3)
+    Xa = dat[,Xsel]
+    ca = d.class.post
   }
   if(FALSE){ ## figures
     plot(density(Xf[,1]),main="")

### 2020-11-18 SingleSite_MultiDisturb_MultiPool.Rmd
SSMDMP

---
title: "Single Site, Multiple Disturbance, MultiPool"
author: "Mike Dietze"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BayesianTools)
library(magrittr)
set.seed(42)
## helper functions
source("multi.ens.adj.R")
source("disturbance.R")
```

## Todo

* Multinomial with distinct disturbances
* Multinomial with distinct disturbances that are different PFTs
* Multinomial with initially-overlapping disturbances that then diverge

## SET-UP

```{r}
ne = 200     ## number of ensemble members
NT = 25      ## number of years
nt = 365*NT
time = 1:nt

## DRIVERS
nd   = 31 ## number of driver ensembles
PAR <- matrix(NA,nd,nt)
for(i in 1:nd){
  PAR[i,] <- VSEMcreatePAR(time)  
}

## Parameters
refPars <- VSEMgetDefaults()
refPars$lower[9:11] = 1
np = 50
sigmaP = diag((refPars$upper - refPars$lower)/10)^2 ## assume range is +/- 5 SD
Theta <- as.data.frame(mvtnorm::rmvnorm(np,refPars$best,sigmaP))
colnames(Theta) <- rownames(refPars)
par(mfrow=c(3,4))
for(i in 1:11){
  Theta[Theta[,i] < refPars$lower[i] , i] <- refPars$lower[i]
  Theta[Theta[,i] > refPars$upper[i] , i] <- refPars$upper[i]
  hist(Theta[,i],main=colnames(Theta)[i])
}

## initial conditions
met = sample.int(nd,ne,replace=TRUE)
par = sample.int(np,ne,replace=TRUE)
Na <- matrix(NA,ne,3)
for(i in 1:ne){
  Na[i,] = VSEM(as.numeric(Theta[par[i],]),PAR = PAR[met[i],])[nt,2:4]
}
Na.orig = Na
par(mfrow=c(1,3))
hist(Na[,1],main="Cleaf")
hist(Na[,2],main="Csoil")
hist(Na[,3],main="Cstem")
```

# Multinomial with distinct disturbances

## set up storage, parameters, pseudodata
```{r}
## hard coded 
Q = diag(c(0.1,0.1,0.1)) ## process error
R = diag(c(.25,10,1))      ## observation error
THETA <- as.matrix(Theta[par,]) ## ensemble parameters

## disturbance
mu0 <- matrix(c(0.1,0.1,
                0.1,0.95),2,2)    ## mean leaf & stem biomass after disturbance as % of original
V1  <- diag(c(0.15,0.25))^2       ## var in post disturbance leaf and stem biomass (one for each type)
V2  <- diag(c(0.15,0.01))^2
V0  <- abind::abind(V1,V2,along=3)       ## combine covs
alloc.soil = 0.25        ## fraction of removed C that goes in the soil
p.dist = c(0.9,0.05,0.05)  ## Multinomial transition probability; first is no change

## storage
Dbar = 0
Analysis <- Forecast <- list()
Nmu = NmuK = NfK = Nf = Na*NA

## pseudodata
No = matrix(NA,NT,3)                   ## storage for observed state
Ntrue = matrix(NA,NT,3)                ## storage for latent "true" state
true = 42                              ## reference ensemble member
D.true = rep(0,NT)                     ## true disturbance timeseries
D.true[10] = 1                         ## prescribe a disturbance
refpar = as.numeric(Theta[par[true],]) ## grab the "true" parameters
refpar[9:11] = Na[true,]               ## set initial biomass pools
for(t in 1:NT){
  tsel <- (t-1)*365 + 1:365            ## select 1 yr of met  
  ref.mu = VSEM(refpar,PAR[met[true],tsel])[365,2:4]    ## forward simulation
  ref.f  = mvtnorm::rmvnorm(1,ref.mu,Q)## process error
  if(D.true[t]){
      ref.f = disturbance2(ref.f,type=D.true[t])
  }
  ref.f = tobit(ref.f)
  refpar[9:11] = ref.f                 ## update IC for next iteration
  Ntrue[t,] = ref.f
  No[t,] = tobit(mvtnorm::rmvnorm(1,ref.f,R)) ## observation error
}
par(mfrow=c(3,1))
plot(No[,1])
plot(No[,2])
plot(No[,3])
```

```{r,echo=FALSE,results='hide',message=FALSE}
AnalysisSSMDMP <- "  ## SingleSite MultiDisturbance MultiPool
model{
  ## prior
  for(i in 1:N){
    x[1:3,i]  ~ dmnorm(muf[,i],pf[,,i]) ## forecast priors for different cases
  }
#  rho[] ~ dmulti(p[], 1)          ## disturbance probability
## dmulti throws error, so decomposing by hand
 rho[2] ~ dbern(p[2])
 rho[3] ~ dbin(p[3]/(1-p[2]),1-rho[2])
 rho[1] <- 1 - rho[2] - rho[3]
  ## Analysis 
  X <- rho[1]*x[1:3,1] + rho[2]*x[1:3,2] + rho[3]*x[1:3,3] ## select between cases
#  X <- x%*%rho
  Y ~ dmnorm(X,R)                       ## Observation error
}"

Na <- Na.orig
muf <- matrix(NA,3,3)
pf  <- array(0,c(3,3,3)) ## 3rd variable is class
for(t in 1:NT){

  tsel <- (t-1)*365 + 1:365               ## Select days from driver files

  ## update IC
  THETA[,9:11]  <- Na    ## Multinomial

  ## Forecast 
  for(i in 1:ne){
    fx       = VSEM(THETA[i,], PAR = PAR[met[i],tsel]) ## deterministic simulation
    Nmu[i,]  =  fx[365,2:4]                         ## state variables at time of assim
    Nf[i,]   =  mvtnorm::rmvnorm(1,Nmu[i,],Q)
  }

  ## Disturbance
  d.prior <- rmultinom(ne,1,p.dist)             ## prior on disturbance status
  d.class <- apply(d.prior,2,which.max)
  for(d in 2:3){  ## loop over disturbance classes
    sel.dist <- which(d.class == d)              ## select disturbed ensemble members
    if(length(sel.dist)>3){
      for(j in sel.dist){                      ## run disturbance model
        Nf[j,]  = disturbance2(Nf[j,],type = d-1)
      }
      muf[,d]  = colMeans(Nf[sel.dist,])
      pf[,,d] =  try(solve(var(Nf[sel.dist,])))
    } 
    if(sum(sel.dist) <= 3 || class(pfD) == "try-error") {   ## backup values if the ensemble stats fail
      muf[,d] = colMeans(Nf[d.class==1,])*c(mu0[1,d-1],1,mu0[2,d-1])
      pf[,,d]  = solve(diag(c(0.15,0,0.15)^2))
    }
  }
  ## set undisturbed
  sel.dist = which(d.class == 1)
  muf[,1] = colMeans(Nf[sel.dist,])
  pf[,,1] = solve(var(Nf[sel.dist,]))
  Nf <- tobit(Nf)

  ## priors
  priors <- list(p   = rowMeans(d.prior),         ## disturbance rate
                 muf = muf,    ## forecast means
                 pf  = pf      ## forecast precision
  )
  Forecast[[t]] <- c(list(d.prior=d.prior,Nf=Nf),priors)

  ## Analysis
  update = c(list(Y=No[t,],R=solve(R),N=3),priors)
  mod <- rjags::jags.model(file=textConnection(AnalysisSSMDMP),
                           data=update,
                           n.adapt=1000,n.chains=3)
  jdat <- rjags::coda.samples(mod,n.iter=2000) ## burn-in 
  jdat <- rjags::coda.samples(mod,variable.names=c("X","rho","x"),n.iter=10000)

  ## update parameters & save Analysis posterior
  dat = as.matrix(jdat)
  Xsel <- grep(pattern="^X",x=colnames(dat))

  ## draw new IC
  d.post <- dat[,grep(pattern="rho",x=colnames(dat))]
  d.class.post <- apply(d.post,1,which.max)
  Na <- multi.ens.adj(
          Xf   = Nf,
          cf   = d.class,
          mu.f = t(priors$muf),
          Pf   = abind::abind(    ## find a way to do this with apply(priors$pf,3,solve)
                    solve(priors$pf[,,1]),
                    solve(priors$pf[,,2]),
                    solve(priors$pf[,,3]),
                 along=3),
          Xa = dat[,Xsel],
          ca = d.class.post)
  Ca <- Na[,1] ## updated class assignments
  Na <- tobit(Na[,-1])
#  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
#  Na <- dat[sel.ic,nsel]
  Analysis[[t]] <- list(MCMC=dat,
                        Na = Na,
                        Ca = Ca)
}

```

### time series

```{r}
post.dist = sapply(X=Analysis,function(y){
  colMeans(y$MCMC[,grep("rho",colnames(y$MCMC))])
})
plot(post.dist[1,],ylim=c(0,1),type='l')
lines(post.dist[2,],col=2)
lines(post.dist[3,],col=3)
legend("topright",legend=c("undisturbed","harvest","defoliation"),
       lwd=2,col=1:3)

pool.names=c("Cleaf","Csoil","Cstem")
for(i in 1:3){

  CI.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("X[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  #CI.Na = sapply(X = Analysis,function(x){
  #  apply(x,2,quantile,c(0.025,0.975))
  #})

  plot(No[,i],
       #ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[2,],col="#b2df8a",lwd=2)
  lines(Ntrue[,i],lty=2,lwd=2)
}
```
### Ensemble adj at time of disturbance

```{r,fig.asp=1,fig.width=3}
#par(mfrow=c(1,3))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("red",alpha=0.5),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25,asp = 1)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=0.5)
points(No[10,1],No[10,3],cex=3,col=ecoforecastR::col.alpha("purple",alpha = 0.75),pch=18)

## ensemble adjustment
sel.a=sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("topleft",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c(1,2,"purple"),
       cex = 1.5)

```

### 2020-11-19 SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
Single Site, Multi-[disturbance, pool, pft]. Works but would benifit from convergence checks.

---
title: "Single Site, Multiple Disturbance, MultiPool, Multi PFT"
author: "Mike Dietze"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BayesianTools)
library(magrittr)
set.seed(42)
## helper functions
source("multi.ens.adj.R")
source("disturbance.R")
```

## Todo

* Multinomial with distinct disturbances that are different PFTs
* Multinomial with initially-overlapping disturbances that then diverge

## SET-UP

```{r}
ne = 200     ## number of ensemble members
NT = 25      ## number of years
nt = 365*NT
time = 1:nt

## DRIVERS
nd   = 31 ## number of driver ensembles
PAR <- matrix(NA,nd,nt)
for(i in 1:nd){
  PAR[i,] <- VSEMcreatePAR(time)  
}

## Parameters
refPars <- VSEMgetDefaults()
refPars$lower[9:11] = 1
pft1 <- refPars$best; pft1[7] = pft1[7]*10
refPars$upper[7] =  refPars$upper[7]*10
np = 50
sigmaP = diag((refPars$upper - refPars$lower)/10)^2 ## assume range is +/- 5 SD
epsilon <- mvtnorm::rmvnorm(np,rep(0,nrow(sigmaP)),sigmaP)
Theta <- list()
Theta[[1]] <- as.data.frame(t(pft1         + t(epsilon))) ## PFT 1
Theta[[2]] <- as.data.frame(t(refPars$best + t(epsilon))) ## PFT 2
colnames(Theta[[1]]) <- rownames(refPars)
colnames(Theta[[2]]) <- rownames(refPars)
par(mfrow=c(3,4))
for(i in 1:11){
  for(p in 1:2){
    Theta[[p]][Theta[[p]][,i] < refPars$lower[i] , i] <- refPars$lower[i]
    Theta[[p]][Theta[[p]][,i] > refPars$upper[i] , i] <- refPars$upper[i]
  }
  hist(Theta[[1]][,i],main=colnames(Theta[[1]])[i],probability = TRUE)
  lines(density(Theta[[2]][,i]))
}
pft.f = pft.a = rep(1,ne)

## initial conditions
met = sample.int(nd,ne,replace=TRUE)
par = sample.int(np,ne,replace=TRUE)
Na <- matrix(NA,ne,3)
for(i in 1:ne){
  Na[i,] = VSEM(as.numeric(Theta[[1]][par[i],]),PAR = PAR[met[i],])[nt,2:4]
}
Na.orig = Na
par(mfrow=c(1,3))
hist(Na[,1],main="Cleaf")
hist(Na[,2],main="Csoil")
hist(Na[,3],main="Cstem")
```

# Multinomial with distinct disturbances w/ different PFTs

## set up storage, parameters, pseudodata
```{r}
## hard coded 
Q = diag(c(0.1,0.1,0.1)) ## process error
R = diag(c(.25,10,.5))      ## observation error
THETA1 <- as.matrix(Theta[[1]][par,]) ## ensemble parameters
THETA2 <- as.matrix(Theta[[1]][par,]) 
THETA  <- THETA1
THETA[pft.f == 2,] <- THETA2[pft.f==2,]

## disturbance
       ## mean leaf & stem biomass after disturbance as % of original
mu0 <- matrix(c(0.1,0.05,          ## removal, switch to PFT2
                0.1,0.95),2,2)     ## defoliation, PFT1
V1  <- diag(c(0.15,0.20))^2       ## var in post disturbance leaf and stem biomass (one for each type)
V2  <- diag(c(0.15,0.01))^2
V0  <- abind::abind(V1,V2,along=3)       ## combine covs
alloc.soil = 0.25          ## fraction of removed C that goes in the soil
p.dist = c(0.9,0.05,0.05)  ## Multinomial transition probability; first is no change
dist.pft = c(0,2,0)        ## PFT to assign post-disturbance; 0 = no change

## storage
Dbar = 0
Analysis <- Forecast <- list()
Nmu = Nf = Na*NA

## pseudodata
No = matrix(NA,NT,3)                   ## storage for observed state
Ntrue = matrix(NA,NT,3)                ## storage for latent "true" state
true = 42                              ## reference ensemble member
D.true = rep(0,NT)                     ## true disturbance timeseries
D.true[10] = 1                         ## prescribe a disturbance
refpar = as.numeric(Theta[[1]][par[true],]) ## grab the "true" parameters
refpar[9:11] = Na[true,]               ## set initial biomass pools
for(t in 1:NT){
  tsel <- (t-1)*365 + 1:365            ## select 1 yr of met  
  ref.mu = VSEM(refpar,PAR[met[true],tsel])[365,2:4]    ## forward simulation
  ref.f  = mvtnorm::rmvnorm(1,ref.mu,Q)## process error
  if(D.true[t]){
      ref.f = disturbance2(ref.f,type=D.true[t])
      refpar = as.numeric(Theta[[2]][par[true],]) ## switch PFT parameters
  }
  ref.f = tobit(ref.f)
  refpar[9:11] = ref.f                 ## update IC for next iteration
  Ntrue[t,] = ref.f
  No[t,] = tobit(mvtnorm::rmvnorm(1,ref.f,R)) ## observation error
}
par(mfrow=c(3,1))
plot(No[,1])
plot(No[,2])
plot(No[,3])
```

```{r,echo=FALSE,results='hide',message=FALSE}
AnalysisSSMDMP <- "  ## SingleSite MultiDisturbance MultiPool
model{
  ## prior
  for(i in 1:N){
    x[1:3,i]  ~ dmnorm(muf[,i],pf[,,i]) ## forecast priors for different cases
  }
#  rho[] ~ dmulti(p[], 1)          ## disturbance probability
## dmulti throws error, so decomposing by hand
 rho[2] ~ dbern(p[2])
 rho[3] ~ dbin(p[3]/(1-p[2]),1-rho[2])
 rho[1] <- 1 - rho[2] - rho[3]
  ## Analysis 
  X <- rho[1]*x[1:3,1] + rho[2]*x[1:3,2] + rho[3]*x[1:3,3] ## select between cases
#  X <- x%*%rho
  Y ~ dmnorm(X,R)                       ## Observation error
}"

Na <- Na.orig
muf <- matrix(NA,3,3)
pf  <- array(0,c(3,3,3)) ## 3rd variable is class
for(t in 1:NT){

  tsel <- (t-1)*365 + 1:365               ## Select days from driver files

  ## update IC & PFT
  pft.f = pft.a
  THETA  <- THETA1
  THETA[pft.f == 2,] <- THETA2[pft.f==2,]
  THETA[,9:11]  <- Na    ## Multinomial

  ## Forecast 
  for(i in 1:ne){
    fx       = VSEM(THETA[i,], PAR = PAR[met[i],tsel]) ## deterministic simulation
    Nmu[i,]  =  fx[365,2:4]                         ## state variables at time of assim
    Nf[i,]   =  mvtnorm::rmvnorm(1,Nmu[i,],Q)
  }

  ## Disturbance
  d.prior <- rmultinom(ne,1,p.dist)             ## prior on disturbance status
  d.class <- apply(d.prior,2,which.max)
  for(d in 2:3){  ## loop over disturbance classes
    sel.dist <- which(d.class == d)              ## select disturbed ensemble members
    if(length(sel.dist)>3){
      for(j in sel.dist){                      ## run disturbance model
        Nf[j,]  = disturbance2(Nf[j,],type = d-1)
      }
      muf[,d]  = colMeans(Nf[sel.dist,])
      pf[,,d] =  try(solve(var(Nf[sel.dist,])))
    } 
    if(sum(sel.dist) <= 3 || class(pf) == "try-error") {   ## backup values if the ensemble stats fail
      muf[,d] = colMeans(Nf[d.class==1,])*c(mu0[1,d-1],1,mu0[2,d-1])
      pf[,,d]  = solve(diag(c(0.15,0,0.15)^2))
    }
    if(dist.pft[d] > 0){  ## some disturbances switch PFT
      pft.f[sel.dist] <- dist.pft[d]
    }
  }
  ## set undisturbed
  sel.dist = which(d.class == 1)
  muf[,1] = colMeans(Nf[sel.dist,])
  pf[,,1] = solve(var(Nf[sel.dist,]))
  Nf <- tobit(Nf)

  ## priors
  priors <- list(p   = rowMeans(d.prior),         ## disturbance rate
                 muf = muf,    ## forecast means
                 pf  = pf      ## forecast precision
  )
  Forecast[[t]] <- c(list(d.prior=d.prior,Nf=Nf,pft=pft.f),priors)

  ## Analysis
  update = c(list(Y=No[t,],R=solve(R),N=3),priors)
  mod <- rjags::jags.model(file=textConnection(AnalysisSSMDMP),
                           data=update,
                           n.adapt=1000,n.chains=3)
  jdat <- rjags::coda.samples(mod,variable.names="X",n.iter=2000) ## burn-in 
  jdat <- rjags::coda.samples(mod,variable.names=c("X","rho","x"),n.iter=10000)

  ## update parameters & save Analysis posterior
  dat = as.matrix(jdat)
  Xsel <- grep(pattern="^X",x=colnames(dat))

  ## draw new IC
  d.post <- dat[,grep(pattern="rho",x=colnames(dat))]
  d.class.post <- apply(d.post,1,which.max)
  Na <- multi.ens.adj(
          Xf   = Nf,
          cf   = d.class,
          mu.f = t(priors$muf),
          Pf   = abind::abind(    ## find a way to do this with apply(priors$pf,3,solve)
                    solve(priors$pf[,,1]),
                    solve(priors$pf[,,2]),
                    solve(priors$pf[,,3]),
                 along=3),
          Xa = dat[,Xsel],
          ca = d.class.post)
  Ca <- Na[,1] ## updated class assignments
  Na <- tobit(Na[,-1])
  sel.pft = which(dist.pft[Ca]>0)
  if(length(sel.pft)>0){
    pft.a[sel.pft] = dist.pft[Ca[sel.pft]]  ## update PFT assignment
  }
  Analysis[[t]] <- list(MCMC=dat,
                        Na  = Na,
                        Ca  = Ca,
                        pft = pft.a)
}

```

### time series

```{r}
post.dist = sapply(X=Analysis,function(y){
  colMeans(y$MCMC[,grep("rho",colnames(y$MCMC))])
})
plot(post.dist[1,],ylim=c(0,1),type='l')
lines(post.dist[2,],col=2)
lines(post.dist[3,],col=3)
legend("topright",legend=c("undisturbed","harvest","defoliation"),
       lwd=2,col=1:3)

post.pft = sapply(X=Analysis,function(y){mean(y$pft)})
plot(post.pft,type='l')

pool.names=c("Cleaf","Csoil","Cstem")
for(i in 1:3){

  CI.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("X[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  #CI.Na = sapply(X = Analysis,function(x){
  #  apply(x,2,quantile,c(0.025,0.975))
  #})

  plot(No[,i],
       #ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[2,],col="#b2df8a",lwd=2)
  lines(Ntrue[,i],lty=2,lwd=2)
}
```

### Ensemble adj at time of disturbance

```{r,fig.asp=1,fig.width=6}
#par(mfrow=c(1,3))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("red",alpha=0.5),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25,asp = 1)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=0.5)
points(No[10,1],No[10,3],cex=3,col=ecoforecastR::col.alpha("purple",alpha = 0.75),pch=18)

## ensemble adjustment
sel.a=sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("topleft",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c(1,2,"purple"),
       cex = 1.5)

```

### 2020-11-19 TwoSite_SingleDisturb_MultiPool.Rmd
Two Site, single disturbance, multi-pool

---
title: "Two Site, Single Disturbance, MultiPool"
author: "Mike Dietze"
date: "11/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BayesianTools)
library(magrittr)
set.seed(42)
## helper functions
source("multi.ens.adj.R")
source("disturbance.R")
```

## SET-UP

```{r}
NT = 25
nt = 365*NT
time = 1:nt

## drivers
nd   = 31 ## number of driver ensembles
PAR <- matrix(NA,nd,nt)
for(i in 1:nd){
  PAR[i,] <- VSEMcreatePAR(time)  
}
plot(PAR[1,], main = "PAR (driving the model)", xlab = "Day",cex=0.35)

## Parameters
refPars <- VSEMgetDefaults()
refPars$lower[9:11] = 1
np = 50
sigmaP = diag((refPars$upper - refPars$lower)/10)^2 ## assume range is +/- 5 SD
Theta <- as.data.frame(mvtnorm::rmvnorm(np,refPars$best,sigmaP))
colnames(Theta) <- rownames(refPars)
par(mfrow=c(3,4))
for(i in 1:11){
  Theta[Theta[,i] < refPars$lower[i] , i] <- refPars$lower[i]
  Theta[Theta[,i] > refPars$upper[i] , i] <- refPars$upper[i]
  hist(Theta[,i],main=colnames(Theta)[i])
}

## initial conditions
## spin up
ne = 500
met = sample.int(nd,ne,replace=TRUE)
par = sample.int(np,ne,replace=TRUE)
Na <- matrix(NA,ne,3)
for(i in 1:ne){
  Na[i,] = VSEM(as.numeric(Theta[par[i],]),PAR = PAR[met[i],])[nt,2:4]
}
Na.orig = Na
par(mfrow=c(1,3))
hist(Na[,1],main="Cleaf")
hist(Na[,2],main="Csoil")
hist(Na[,3],main="Cstem")
```

## set up storage and parameters

```{r}
## hard coded 
Q = diag(c(0.1,0.1,0.1)) ## process error
R = diag(c(.5,10,1))      ## observation error
R = as.matrix(Matrix::bdiag(R,R))  ## nsite = 2
THETA <- as.matrix(Theta[par,]) ## ensemble parameters
THETA = abind::abind(THETA,THETA,along = 3)  ## n=2 sites
mu0 <- c(1,1)            ## mean leaf & stem biomass after disturbance
V0  <- diag(c(0.15,0.25))^2## var in post disturbance leaf and stem biomass
alloc.soil = 0.25        ## fraction of removed C that goes in the soil
p.dist = 0.05

## storage
Dbar = 0
Analysis <- Forecast <- list()
Forecast.KD = Analysis.KD = list() ## EnKF with Disturbance

## pseudodata
nsite = 2
No = matrix(NA,NT,3*nsite)                   ## storage for observed state
Ntrue = matrix(NA,NT,3*nsite)                ## storage for latent "true" state
true = 42                                    ## reference ensemble member
D.true = matrix(0,nsite,NT)                  ## true disturbance timeseries
D.true[1,10] = 1                             ## prescribe a disturbance
refpar = matrix(as.numeric(Theta[par[true],]),nsite,ncol(Theta),byrow = TRUE) ## grab the "true" parameters
refpar[,9:11] = Na[true+(1:nsite)-1,]        ## set initial biomass pools
ref.f = list()
for(t in 1:NT){
  tsel <- (t-1)*365 + 1:365                  ## select 1 yr of met  
  for(s in 1:nsite){
    ref.mu = VSEM(refpar[s,],PAR[met[true],tsel])[365,2:4]    ## forward simulation
    ref.f[[s]]  = mvtnorm::rmvnorm(1,ref.mu,Q)## process error
    if(D.true[s,t]){
     ref.f[[s]] = disturbance(ref.f[[s]])
    }
    ref.f[[s]] = tobit(ref.f[[s]])
    refpar[s,9:11] = ref.f[[s]]                 ## update IC for next iteration
  }
  Ntrue[t,] = unlist(ref.f)
  No[t,] = tobit(mvtnorm::rmvnorm(1,Ntrue[t,],R)) ## observation error
}
par(mfrow=c(3,1))
plot(No[,1])
lines(No[,4],col=2)
plot(No[,2])
lines(No[,5],col=2)
plot(No[,3])
lines(No[,6],col=2)
```

```{r,echo=FALSE,results='hide',message=FALSE}
Analysis2SMDMP <- "  ## SingleSite MultiDisturbance MultiPool
model{
  ## prior
  for(i in 1:N){
    x[1:6,i]  ~ dmnorm(muf[,i],pf[,,i]) ## forecast priors for different cases
  }
 rho[2] ~ dbern(p[2])
 rho[3] ~ dbin(p[3]/(1-p[2]),1-rho[2])
 rho[4] ~ dbin(p[4]/(1-p[2]-p[3]),1-rho[2]-rho[3])
 rho[1] <- 1 - rho[2] - rho[3] - rho[4]
  X <- rho[1]*x[1:6,1] + rho[2]*x[1:6,2] + rho[3]*x[1:6,3] + rho[4]*x[1:6,4] ## select between cases
  Y ~ dmnorm(X,R)                       ## Observation error
}"

Na <- NaK <- abind::abind(Na.orig,
                          Na.orig[sample.int(ne,ne),],  ## resample IC for site 2
                          along = 3)
NfK = Nf = Na*NA
THETAK <- THETA
d.prior = matrix(nrow=ne,ncol=2)
muf <- matrix(NA,3*nsite,4)
pf  <- array(0,c(3*nsite,3*nsite,4)) ## 3rd variable is class
for(t in 1:NT){

  tsel <- (t-1)*365 + 1:365               ## Select days from driver files

  ## update IC
  THETA[,9:11,]  <- Na    ## Multinomial
  THETAK[,9:11,] <- NaK  ## EnKF w/ disturbance

  ## Forecast 
  for(i in 1:ne){
    for(s in 1:nsite){
      Nmu  =  VSEM(THETA[i,,s], PAR = PAR[met[i],tsel])[365,2:4]    
      NmuK      = VSEM(THETAK[i,,s],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
      epsilon  = mvtnorm::rmvnorm(1,rep(0,3),Q)                   ## process error: shared between versions, currently independent by site
      Nf[i,,s]   = Nmu  + epsilon
      NfK[i,,s]  = NmuK + epsilon
    }
  }

  ## Disturbance
  for(i in 1:ne) d.prior[i,] <- rcbern(rep(p.dist,2),rho=0.5)   ## prior on disturbance status
  d.class <- d.prior[,1] + 2*d.prior[,2]
  for(d in 2:4){  ## loop over disturbance classes
    sel.dist <- which(d.class == d-1)              ## select disturbed ensemble members
    if(length(sel.dist)> (3*nsite)){
      for(j in sel.dist){                      ## run disturbance model
        switch (as.character(d),
            "2" = {Nf[j,,1] = disturbance(Nf[j,,1])
                   NfK[j,,1]= disturbance(NfK[j,,1])
                  },
            "3" = {Nf[j,,2] = disturbance(Nf[j,,2])
                   NfK[j,,2]= disturbance(NfK[j,,2])
                  },
            "4" = {Nf[j,,1] = disturbance(Nf[j,,1])
                   Nf[j,,2] = disturbance(Nf[j,,2])
                   NfK[j,,1]= disturbance(NfK[j,,1])
                   NfK[j,,2]= disturbance(NfK[j,,2])
                  }
        )
      }
      muf[,d]  = apply(Nf[sel.dist,,],3,colMeans)
      pfTMP    = try(solve(var(cbind(Nf[sel.dist,,1],Nf[sel.dist,,2]))))
    } 
    if(sum(sel.dist) <= (3*nsite) || class(pfTMP) == "try-error") { 
      ## set to highly unlikely values that will be selected out
      muf[,d]  = rep(0,nsite*3)
      pfTMP    = solve(diag(rep(0.01,nsite*3)^2))
    }
    pf[,,d]    = pfTMP
  }
  ## set undisturbed
  sel.dist = which(d.class == 0)
  muf[,1] = apply(Nf[sel.dist,,],3,colMeans)
  pf[,,1] = solve(var(cbind(Nf[sel.dist,,1],Nf[sel.dist,,2])))
  Nf <- tobit(Nf)

  ## EnKF forecast stats
  NfK <- tobit(NfK)
  NK = cbind(NfK[,,1],NfK[,,2])
  mufK = colMeans(NK)
  PfK  = var(NK)
  pfK  = solve(PfK)             ## Not disturbed forecast precision
  Forecast.KD[[t]] <- list(d.prior=d.prior,Nf=NfK,muf=mufK,pf=pfK)

  ## priors
  priors <- list(N = 4,  ## 2 sites x 2 disturbance classes
                 p   = table(d.class)/ne,         ## disturbance rate
                 muf = muf,    ## forecast means
                 pf  = pf      ## forecast precision
  )
  Forecast[[t]] <- c(list(d.prior=d.prior,Nf=Nf),priors)

  ## Analysis
  update = c(list(Y=No[t,],R=solve(R)),priors)
  mod <- rjags::jags.model(file=textConnection(Analysis2SMDMP),
                           data=update,
                           n.adapt=1000,n.chains=3)
  jdat <- rjags::coda.samples(mod,variable.names="X",n.iter=2000) ## burn-in 
  jdat <- rjags::coda.samples(mod,variable.names=c("X","rho","x"),n.iter=10000)

  ## update parameters & save Analysis posterior
  dat = as.matrix(jdat)
  Xsel <- grep(pattern="^X",x=colnames(dat))

  ## draw new IC
  d.post <- dat[,grep(pattern="rho",x=colnames(dat))]
  d.class.post <- apply(d.post,1,which.max)
  Na <- multi.ens.adj(
          Xf   = cbind(Nf[,,1],Nf[,,2]),
          cf   = d.class+1,
          mu.f = t(priors$muf),
          Pf   = abind::abind(    ## find a way to do this with apply(priors$pf,3,solve)
                    solve(priors$pf[,,1]),
                    solve(priors$pf[,,2]),
                    solve(priors$pf[,,3]),
                    solve(priors$pf[,,4]),
                 along=3),
          Xa = dat[,Xsel],
          ca = d.class.post)
  Ca <- Na[,1] ## updated class assignments
  Na <- tobit(Na[,-1])
#  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
#  Na <- dat[sel.ic,nsel]
  Analysis[[t]] <- list(MCMC=dat,
                        Na = Na,
                        Ca = Ca)
  Na <- abind::abind(Na[,1:3],Na[,4:6],along = 3)

  ## EnKF Analysis
  I = H <- diag(rep(1,3*nsite))
  K <- PfK %*% t(H) %*% solve(H%*%PfK%*%t(H) + R)
  muaK <- mufK + K%*%(No[t,] - mufK) 
  PaK <- (I-K %*% H)%*%PfK    
  paK = solve(PaK)
  NaK <- PEcAn.assim.sequential::adj.ens(Pf = PfK,X = NK,mu.f = mufK,mu.a = muaK,Pa = PaK) %>% as.matrix()
  Analysis.KD[[t]] <- list(mua=muaK,pa=paK,Pa=PaK,Na=NaK)
  NaK <- abind::abind(NaK[,1:3],NaK[,4:6],along = 3)

}

```

### time series

```{r}
pool.names=c("Cleaf","Csoil","Cstem")
for(i in 1:6){

  CI.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("X[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  #CI.Na = sapply(X = Analysis,function(x){
  #  apply(x,2,quantile,c(0.025,0.975))
  #})
  CI.EnKF = sapply(Analysis.KD,function(x){
    mu = x$mua[i]
    sd = sqrt(x$Pa[i,i])
    return(c(mu-1.96*sd,mu,mu+1.96*sd))
  })

  plot(No[,i],
       #ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.EnKF),CI.EnKF[1,],CI.EnKF[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.EnKF[2,],col="#b2df8a",lwd=2)
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('blue',0.5))
  lines(CI.EnKF[2,],col="blue",lwd=2)
  abline(v=10)  
  lines(Ntrue[,i],lty=2,lwd=2)
}
```
### Ensemble adj at time of disturbance

```{r,fig.asp=1}
#par(mfrow=c(1,3))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("red",alpha=0.25),
     ylim=range(Forecast[[10]]$Nf[,3,1]),
     xlim=range(Forecast[[10]]$Nf[,1,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25,asp = 1)
points(Forecast[[10]]$Nf[,1,1],Forecast[[10]]$Nf[,3,1],cex=0.5,col=ecoforecastR::col.alpha("black",alpha=0.25))
points(No[10,1],No[10,3],cex=3,col=ecoforecastR::col.alpha("purple",alpha = 0.75),pch=18)

## ensemble adjustment
sel.a=sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1,1],Forecast[[10]]$Nf[sel.a,3,1],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("topleft",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c(1,2,"purple"),
       cex = 1.5)

cov2cor(solve(Forecast.KD[[10]]$pf))

```
## EnKF, no disturbance

```{r,echo=FALSE}
Na = Na.orig
Forecast.EnKF = Analysis.EnKF = list()
for(t in 1:NT){

  tsel <- (t-1)*365 + 1:365               ## Select days from driver files

  ## update IC
  THETA[,9:11] <- Na

  ## Forecast 
  for(i in 1:ne){
    Nmu[i,] = VSEM(THETA[i,],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
    Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
  }

  muf = colMeans(Nf)    ## Not disturbed forecast means
  Pf  = var(Nf)
  pf  = solve(Pf)             ## Not disturbed forecast precision
  Forecast.EnKF[[t]] <- list(d.prior=d.prior,Nf=Nf,muf=muf,pf=pf)

  ## Analysis
  I = H <- diag(rep(1,3))
  K <- Pf %*% t(H) %*% solve(H%*%Pf%*%t(H) + R)
  mua <- muf + K%*%(No[t,] - muf) 
  Pa <- (I-K %*% H)%*%Pf    
  pa = solve(Pa)

  ## draw new IC
  #  Na <- mvtnorm::rmvnorm(ne,mua,Pa)
  Na <- PEcAn.assim.sequential::adj.ens(
    Pf = Pf,X = Nf,mu.f = muf,mu.a = mua,Pa = Pa)

  ## save Analysis posterior
  Analysis.EnKF[[t]] <- list(mua=mua,pa=pa,Pa=Pa,Na=Na)

}

```

### 2020-11-19 disturbance.R

@@ -26,3 +26,19 @@ disturbance2 <- function(x,type){ ## x = Bleaf, Bsoil, Bstem
   removal <- residual*(1-alloc.soil)
   return(x)
 }
+
+# random correlated bernouli
+rcbern <- function(p,rho){
+  
+  ## calculate cov
+  D = as.matrix(dist(seq_along(mu),diag = TRUE,upper = TRUE))
+  SIGMA <- 1/(1-rho^2)*rho^D
+  
+  ## draw rMVN
+  x = mvtnorm::rmvnorm(1,rep(0,length(p)),SIGMA)
+  
+  ## inverse distribution transform
+  y = pnorm(x)
+  qbinom(y,1,p)
+  
+}

### 2020-11-19 multi.ens.adj.R

@@ -36,6 +36,18 @@ multi.ens.adj<-function(Xf, cf, mu.f, Pf, Xa, ca){
       along=3)
     Xa = dat[,Xsel]
     ca = d.class.post
+    ## two site
+    Xf   = cbind(Nf[,,1],Nf[,,2])
+    cf   = d.class+1
+    mu.f = t(priors$muf)
+    Pf   = abind::abind(    ## find a way to do this with apply(priors$pf,3,solve)
+      solve(priors$pf[,,1]),
+      solve(priors$pf[,,2]),
+      solve(priors$pf[,,3]),
+      solve(priors$pf[,,4]),
+      along=3)
+    Xa = dat[,Xsel]
+    ca =d.class.post
   }
   if(FALSE){ ## figures
     plot(density(Xf[,1]),main="")

### 2020-11-19 Paper.Rmd
small text update, pulled out Discussion to own file

@@
 
-# Discussion
-
-* Interpretation
-  * TBD based on results
-
-* Implications  (borrow text from CMS proposals)
-  * Improving MRV
-  * Recovery forecasting
-  * Disturbance nowcasting
-    * requires reducing temporal resolution
-
-* Future directions
-  * Disturbance priors need to be spatially heterogenous and evolve temporally
-    * Connect to state-and-transition literature
-  * Scaling up -> spatially implicit
-    * ED2
-    * Contagion
+<!---- DISCUSSION --->
+```{r child = 'discussion.Rmd'}
+```
 

### 2020-11-19 discussion.Rmd
small text update, pulled out Discussion to own file

# Discussion

* Interpretation
  * TBD based on results

* Issues
  * Need a more detail exploration of rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
  * Would like to add look-back capabilities (ensemble backward smoothing) to update assignments based on more recent info
    * related: with multi-PFT need to check for, and protect against, long term PFT leakage
  * Computational demand it high for problems with many grid cells
    * Efficient (semi-)analytic approximations or LUT
    * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly [@McCabe]
  * Multisite needs a way of estimating $\rho$ by site with cross site correlations, rather than unique $\rho$ for all permutations
    * for $\rho$ handle the covariances of nearby sites -- disturbance in one usually increases the probability of disturbance in others
    * similarly, need to be able to compose Multisite $\mu_f$ and $P_f$, from individual site values with reasonable approximations of cross-[site x pool] correlations

* Implications  (borrow text from CMS proposals)
  * Improving MRV
  * Recovery forecasting
  * Disturbance nowcasting
    * requires reducing temporal resolution

* Future directions
  * Disturbance priors need to be spatially heterogenous and evolve temporally
    * Connect to state-and-transition literature
  * Scaling up -> spatially implicit
    * ED2
    * Contagion

### 2020-11-19 methods.Rmd
small text update, pulled out Discussion to own file

--- 2020-11-17 methods.Rmd
@@
 
+*Priors*
+
+  $$X^L_k \sim MVN(\vec{\mu}_{f,k},P_{f,k})$$
+  $$\vec{\rho} \sim Multinomial(1,\vec{p})$$ 
+*Process Model*
+
+  $$X = X^L \vec{\rho}$$
+
+*Likelihood*
+  $$Y \sim MVN(X,R)$$
+### Ensemble Adjustment
+
+[@Anderson2001]
+
+*reassignment of classes*
+
+To reassign forecasted ensemble members to new disturbance classes we first identify those that decreased between the prior probability, $p_k$, and the posterior frequency, $\bar{\rho}_k$.
+
+$$\delta p_k = max\{0,\bar{\rho}_k-p_k \}$$
+
+Classes that decreased in relative abundance are reassigned to new classes according to a Multinomial draw
+
+$$c_{A|k} = Multinomial\left( n_k, {{\delta p}\over{\sum \delta p}}  \right)$$
+
+*Ensemble adjustment by class*
+
 Conceptual figure
@@
 3. Simulated data experiments (known disturbance)
-  + Single Site, Single Disturbance
-  + SSSD Multi Pool
-  + SS Multi-Disturbance w/ PFT switching
+  + x Single Site, Single Disturbance  (add a case where EnKF has disturb but it's more rare)
+  + x SSSD Multi Pool
+  + x SS Multi-Disturbance - distinct
+  + x SS Multi-Disturbance w/ distinct, PFT switching
+      * mention parallel PFT ensembles
+      * need to protect against long-term PFT leakage; related to more general need to "look back"
   + Simple Multi-site: spatial cov
+  + SSMD - initially overlapping but diverge
+      * how to handle updating of assignment based on new information?
+  + SSMD - transition probabilities change with state
+  + rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
 4. Real-world examples

### 2020-11-20 TwoSite_SingleDisturb_MultiPool.Rmd
two site model. Some text

@@
 ## spin up
-ne = 500
+ne = 1000
 met = sample.int(nd,ne,replace=TRUE)
@@
 alloc.soil = 0.25        ## fraction of removed C that goes in the soil
-p.dist = 0.05
+p.dist = 0.025
 
@@
 pf  <- array(0,c(3*nsite,3*nsite,4)) ## 3rd variable is class
-for(t in 1:NT){
+for(t in 1:10){
 
@@
   ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('blue',0.5))
-  lines(CI.EnKF[2,],col="blue",lwd=2)
+  lines(CI.Na[2,],col="blue",lwd=2)
   abline(v=10)  

### 2020-11-20 methods.Rmd
two site model. Some text

@@
 3. Simulated data experiments (known disturbance)
+  + describe model
   + x Single Site, Single Disturbance  (add a case where EnKF has disturb but it's more rare)
@@
       * need to protect against long-term PFT leakage; related to more general need to "look back"
-  + Simple Multi-site: spatial cov
+  + x Simple Multi-site: spatial cov
   + SSMD - initially overlapping but diverge
@@
 4. Real-world examples
-  + Describe Model & PEcAn
   + Calibrate state-and-transition prior
     + Landtrendr
-    ftp://islay.ceoas.oregonstate.edu/change_agent_attribution
+    http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html
+    + Location:
+    tile h03v04
+xmin       : -2115585 
+xmax       : -1965585 
+ymin       : 2564805 
+ymax       : 2714805 
+Upper Left  (-2115585.000, 2714805.000) (123d 6'32.29"W, 44d41'17.09"N)
+Lower Left  (-2115585.000, 2564805.000) (122d35' 7.51"W, 43d23'46.18"N)
+Upper Right (-1965585.000, 2714805.000) (121d16'45.91"W, 45d 3'17.45"N)
+Lower Right (-1965585.000, 2564805.000) (120d47'15.95"W, 43d45'18.18"N)
+Center      (-2040585.000, 2639805.000) (121d56'25.42"W, 44d13'35.07"N)
+> res(cover)
+[1] 30 30
+> crs(cover)
+CRS arguments:
+ +proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83
++units=m +no_defs 
+    + Biomass [@Kennedy2018]
+      1990-2017
+    + Cover
+      1990-2017
+    + Disturbance
+      1984-2011
+      - nearest neighbor resampled to same as cover
+    + LAI??
+  + Calibrate model
   + Applications: PNW

### 2020-11-20 StateTransition.Rmd
initial code for state-and-transition calibration. Not much yet, currently just loading the data

---
title: "State & Transition"
author: "Michael Dietze"
date: "11/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
```

## Objectives

* Use data on land cover, disturbance, and biomass to calibrate a simple state and transition model to drive disturbance probabilities for real-world examples
* set up points and time series for assimilation examples

Metadata: downloaded tile_h03v04 from http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html on 11/19/2020

## load data

```{r}
# cover
cover = raster("conus_land_cover_ARD_tile_h03v04/conus_land_cover_ARD_tile_h03v04.tif")
extent(cover)
nlayers(cover)
res(cover)
crs(cover)
image(cover)

# disturbance
dis.processed = "dis.h03v04.RData"
if(file.exists(dis.processed)){
  load(dis.processed)
} else {
  ## process disturbance data to match cover's extent and resolution
  dis = raster("Cmonster_disturbance_class_ARD_tile_h03v04/Cmonster_disturbance_class_ARD_tile_h03v04.tif")
  extent(dis)
  nlayers(dis)
  res(dis)
  crs(dis)

  ## match disturbance to cover
  dis <- raster::resample(dis,cover,method="ngb")
  save(dis,file=dis.processed)

  compareRaster(dis, cover, res = TRUE)
  compareRaster(dis,cover,extent = TRUE)
}

```

```{r}
### biomass
#biomass = raster("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
#extent(biomass)
#res(biomass)
#nlayers(biomass)
#summary(biomass)
```

## estimate cover frequencies, select those of interest
```{r}
f = freq(cover)
## most frequent class is 42 = evergreen forest

nbands(cover)
b = bandnr(cover)

cvr = getValues(cover)
cvr = as.matrix(cover[[1]])

```

## Estimate disturbance probability for each cover class
```{r}
fd = freq(dis)

#### disturbance aggregations ####

## 
dis[dis==60] <-

```

## Determine transitions (cover / pft)

## Reduce dimensionality, save

## select sites of interest, save

### 2020-12-01 StateTransition.Rmd
State and Transition prior

@@
 library(raster)
+options(scipen=999)
 ```
@@
 cover = raster("conus_land_cover_ARD_tile_h03v04/conus_land_cover_ARD_tile_h03v04.tif")
+covS  = stack("conus_land_cover_ARD_tile_h03v04/conus_land_cover_ARD_tile_h03v04.tif")
+ctime = 1990:2017
 extent(cover)
@@
   ## process disturbance data to match cover's extent and resolution
-  dis = raster("Cmonster_disturbance_class_ARD_tile_h03v04/Cmonster_disturbance_class_ARD_tile_h03v04.tif")
-  extent(dis)
-  nlayers(dis)
-  res(dis)
-  crs(dis)
+  #dis = raster("Cmonster_disturbance_class_ARD_tile_h03v04/Cmonster_disturbance_class_ARD_tile_h03v04.tif")
+  #extent(dis)
+  #nlayers(dis)
+  #res(dis)
+  #crs(dis)
+
+  disS = stack(x = "Cmonster_disturbance_class_ARD_tile_h03v04/Cmonster_disturbance_class_ARD_tile_h03v04.tif")
 
@@
 ### biomass
-#biomass = raster("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
+biomass = raster("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
+bioS = stack("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
 #extent(biomass)
 #res(biomass)
-#nlayers(biomass)
+nlayers(bioS)
 #summary(biomass)
@@
 
+Disturbance class info from Robert Kennedy (email 2020-11-24)
+```
+AGENTS
+ = {
+    1:   'Undisturbed',
+    10:  'Unkwn Agent',
+    11:  'Other ',
+    20:  'Clearcut',
+    21:  'Partial Harvest',
+    22:  'Salvage',
+    30:  'Development',
+    31:  'Road',
+    40:  'Fire',
+    50:  'Insect/Disease',
+    51:  'MPB-29',
+    52:  'MPB-239',
+    53:  'WSB-29',
+    54:  'WSB-239',
+    61:  'Water',
+    62:  'Wind',
+    70:  'Avalanche-Chute',
+    71:  'Avalanche-Runout',
+    72:  'Debris Flow',
+    73:  'Landslide',
+    100: 'Ukn Slow Disturbance',
+    110: 'Ukn Abrupt Disturbance',
+    160: 'Recovery',
+    200: 'No Vis Change',
+    201: 'False Change',
+    254: 'Err No Code',
+    255: 'Err Undefined'
+}
+
+Classes 1 through 50 and 61-73 are based on patches labeled using our methodology (like that 2015 paper you found) to link interpreted patches to agents using random forests.
+Unknown agent is a confirmed change that the interpret saw but couldn't attribute.
+51-54 are special cases -- mountain pine beetle and western spruce budworm -- that we took from my former PhD student's work to link ADS data to our time-series stuff.  
+Classes 200 and 201 are also based on the interpreter rules for patches -- if they could not determine that there was a change, or if they could determine that the likely change was false -- like a seasonality or snow issue.
+Basically both are "no change"
+
+Classes 100-160 are based on pixel-level labels based on the temporal signature alone.  They are smaller than our minimum mapping unit of 11 pixels. 
+```
+
 ## Estimate disturbance probability for each cover class
 ```{r}
+dtime = 1985:2012
 fd = freq(dis)
+fdS = freq(disS)
+dclass <- unique(unlist(sapply(fdS,function(x){x[,1]}))) ## disturbance classes
+dcount = matrix(0,length(dclass),length(fdS))
+for(t in 1:length(fdS)){
+  sel = match(fdS[[t]][,1],dclass)
+  dcount[sel,t] = fdS[[t]][,2]
+}
+rownames(dcount)=dclass
+colnames(dcount)=dtime
+d.prior = rowSums(dcount)
+d.prior = d.prior/sum(d.prior)
+d.prior[d.prior>0.001]   ## classes to worry about
 
 #### disturbance aggregations ####
+rcl = matrix(
+    ## Untriggered  
+       c(1,   0, # Undisturbed
+         100, 0, # Ukn Slow Disturbance
+         160, 0, # Recovery
+         200, 0, # No Vis Change
+         201, 0, # False Change
+         254, 0, # Err No Code
+         255, 0, # Err Undefined
+    ## Clearcut
+         20, 20, # 'Clearcut',
+         21, 20, # Partial Harvest
+         22, 20, # Salvage
+    ## Fire
+         40, 40, # Fire'
+    ## Pest
+         50, 50, #  Insect/Disease',
+         51, 50, # 'MPB-29',
+         52, 50, # 'MPB-239',
+         53, 50, # 'WSB-29'
+         54, 50, # 'WSB-239'
+    ## OTHER
+         10, 10,  # 'Unkwn Agent',
+         11, 10,  # 'Other '
+         30, 10, # Development'      ## Development
+         31, 10, # Road
+         61, 10,  # 'Water',
+         62, 10,  # 'Wind',
+         70, 10,  # 'Avalanche-Chute',
+         71, 10,  # 'Avalanche-Runout',
+         72, 10,  # 'Debris Flow',
+         73, 10  # 'Landslide',
+         ),ncol=2,byrow = TRUE)
 
-## 
-dis[dis==60] <-
+disR = raster::subs(disS,as.data.frame(rcl))
+save(disR,file=dis.processed)
+
+if(FALSE){
+disS[disS==1]   = 0   # Undisturbed
+disS[disS==100] = 0   # Ukn Slow Disturbance
+disS[disS==160] = 0   # Recovery
+disS[disS==200] = 0   # No Vis Change
+disS[disS==201] = 0   # False Change
+disS[disS==254] = 0   # Err No Code
+disS[disS==255] = 0   # Err Undefined
+
+## Cut
+##    20:  'Clearcut',
+disS[disS==21]  = 20  # Partial Harvest
+disS[disS==22]  = 20  # Salvage
+
+## Development
+##    30:  'Development',
+disS[disS==31]  = 30  # Road
+
+## Fire
+##    40:  'Fire',
+
+## Pest
+## 50:  'Insect/Disease',
+disS[disS==51]  = 50  # 'MPB-29',
+disS[disS==52]  = 50  # 'MPB-239',
+disS[disS==53]  = 50  # 'WSB-29'
+disS[disS==54]  = 50  # 'WSB-239'
+
+## OTHER
+##  10:  'Unkwn Agent',
+disS[disS==11]  = 10  # 'Other '
+disS[disS==61]  = 10  # 'Water',
+disS[disS==62]  = 10  # 'Wind',
+disS[disS==70]  = 10  # 'Avalanche-Chute',
+disS[disS==71]  = 10  # 'Avalanche-Runout',
+disS[disS==72]  = 10  # 'Debris Flow',
+disS[disS==73]  = 10  # 'Landslide',
+}
+
+### RECALCULATE FREQUENCIES
+
+fdS = freq(disR)
+save(disR,fdS,file=dis.processed)
+dclass <- unique(as.vector(unlist(sapply(fdS,function(x){x[,1]})))) ## disturbance classes
+dcount = matrix(0,length(dclass),length(fdS))
+for(t in 1:length(fdS)){
+  sel = match(fdS[[t]][,1],dclass)
+  dcount[sel,t] = fdS[[t]][,2]
+}
+rownames(dcount)=dclass
+colnames(dcount)=dtime
+d.prior = rowSums(dcount)
+d.prior = d.prior/sum(d.prior)
+d.prior[d.prior>0.001]   ## classes to worry about
+
+for(i in seq_along(dclass)){
+  plot(dtime,dcount[i,]*(30^2/100^2),main=dclass[i],ylab="hectares",type='l')
+}
+
+  ## match disturbance to cover
+  disR <- raster::resample(disR,cover,method="ngb")
+  dis.resamp = "disR.resamp.RData"
+  save(disR,fdS,file=dis.resamp)
 
@@
 ## Determine transitions (cover / pft)
+## looking at a sample, most disturbances correspond to cover changes THAT YEAR
+## exception is fire, where there seems to be a nontrivial number that are forest that year but nonforest e.g barren the following year
+## do see a good number of cover transitions
+```{r}
+train = list()
+nd = length(dclass)-1
+tmat = matrix(0,10,nd) ## transition matrix
+bioBar=bioSq=pbioBar=pbioSq = tmat ## sums for biomass and proportional biomass change
+nsamp = 100
 
-## Reduce dimensionality, save
+## loop over years
+ttime = ctime[ctime<2004][-1] ## training years
+for(t in seq_along(ttime)){
+  print(t)
+
+  if(t>length(train)){ 
+## sample
+    dtmp <- disR[[which(dtime==ttime[t])]]  ## disturbances for that year
+    stmp = raster::sampleStratified(dtmp,size = nsamp,exp=1000) ## sample of cell #'s
+
+## eliminate disturbances that did not apply to forests
+    ctmp <- covS[[which(ctime==ttime[t])]]  ## cover for previous year
+    ftmp <- stmp[ctmp[stmp[,1]]==42,]       ## only keep conifer
+    train[[t]] = list(sample=ftmp)
+  } else {
+    ftmp = train[[t]]$sample   # restart if already created sample
+  }
+  save(train,file="stateTrain.RData")
+
+## extract cover and biomass (with restart capability)
+  if(is.null(train[[t]][["cover"]])){
+    cetmp = raster::extract(covS,ftmp[,1])
+    train[[t]]$cover = cetmp
+  } else {cetmp = train[[t]][["cover"]]}
+  if(is.null(train[[t]][["biomass"]])){
+    betmp = raster::extract(bioS,ftmp[,1])
+    train[[t]]$biomass = betmp
+  } else {betmp = train[[t]][["biomass"]]}
+  save(train,file="stateTrain.RData")
+
+  bm0    = apply(betmp[,ctime %in% (ttime[t]-(1:3)),drop=FALSE],1,max)
+#  bm1    = betmp[,ctime == ttime[t]]
+#  bm2    = betmp[,ctime == (ttime[t]+1)]
+#  bm = pmin(bm1,bm2)
+  bm  = apply(betmp[,ctime %in% (ttime[t]+(0:2)),drop=FALSE],1,min)
+  br    = bm/bm0
+  bm[is.na(bm)] = 0
+  br[is.na(br)] = 0
+
+  for(c in seq_along(dclass)){
+    selc = which(ftmp[,2] == dclass[c])
+    plot(ctime,ctime,ylim=range(betmp,na.rm = TRUE),type='n',main=paste(ttime[t],dclass[c]))
+    for(i in seq_along(selc)) lines(ctime,betmp[selc[i],],col=floor(cetmp[selc[i],t+1]/10))
+    abline(v=ttime[t])
+    print(c(t,dclass[c],mean(br[selc])))
+  }
+
+## loop over disturbance types
+  for(c in seq_along(dclass)){
+    print(c(t,c))
+    selc = which(ftmp[,2] == dclass[c])
+
+    ## tally within-year transitions
+    trans = cetmp[selc,ctime %in% (ttime[t]+0:1),drop=FALSE]
+    if(nrow(trans) == 0) next
+    trans = floor(trans/10)  ## round down
+    trans[trans[,2]==4,1] = 4 ## eliminate 1-yr only transitions
+    t1 = unique(trans[,1])
+    for(lc in t1){
+      if(lc == 4){
+        ## for forest, tally next-year transitions
+        t2sel = which(trans[,1]==4)
+        t2 = unique(trans[t2sel,2])
+        tmat[t2,c]    = tmat[t2,c]    + tapply(bm[selc[t2sel]],  trans[,2],length)
+        ## for each transition, tally biomass
+        bioBar[t2,c]  = bioBar[t2,c]  + tapply(bm[selc[t2sel]],  trans[,2],sum)
+        bioSq[t2,c]   = bioSq[t2,c]   + tapply(bm[selc[t2sel]]^2,trans[,2],sum)
+        pbioBar[t2,c] = pbioBar[t2,c] + tapply(br[selc[t2sel]],  trans[,2],sum)
+        pbioSq[t2,c]  = pbioSq[t2,c]  + tapply(br[selc[t2sel]]^2,trans[,2],sum)
+      } else {
+        ## tally this-year transitions
+        t1sel = which(trans[,1]!=4)
+        t0 = unique(trans[t1sel,1])
+        tmat[t0,c]    = tmat[t0,c]    + tapply(bm[selc[t1sel]],  trans[,1],length)
+        ## for each transition, tally biomass, biomass^2, propBiomass, and propBiomass^2
+        bioBar[t0,c]  = bioBar[t0,c]  + tapply(bm[selc[t1sel]],  trans[,1],sum)
+        bioSq[t0,c]   = bioSq[t0,c]   + tapply(bm[selc[t1sel]]^2,trans[,1],sum)
+        pbioBar[t0,c] = pbioBar[t0,c] + tapply(br[selc[t1sel]],  trans[,1],sum)
+        pbioSq[t0,c]  = pbioSq[t0,c]  + tapply(br[selc[t1sel]]^2,trans[,1],sum)
+      }
+
+    } ## end loop over lc
+
+  } ## end loop over disturbance class c
+} ## end loop over time
+TRANS = t(t(tmat)/colSums(tmat))
+BBAR  = bioBar/tmat
+PBBAR = pbioBar/tmat
+BVAR  = sqrt((bioSq -  (bioBar  *  bioBar)/tmat) / (tmat - 1))
+PBVAR = sqrt((pbioSq - (pbioBar * pbioBar)/tmat) / (tmat - 1))
+BVAR/BBAR - PBVAR
+## in all cases, disturbance CV was higher on an absolute than a relative basis
+save(PBBAR,PBVAR,file="STprior.RData")
+```
 
 ## select sites of interest, save
+```{r}
+## sample disturbances for target year
+nsamp = 100
+d04 <- disR[[which(dtime==2004)]]
+samp = raster::sampleStratified(d04,size = nsamp,exp=1000)
+
+## eliminate disturbances that did not apply to forests
+c03 <- covS[[which(ctime==2003)]]
+csamp <- c03[samp[,1]]
+fsamp <- samp[csamp==42,]
+
+dextract = raster::extract(disR,fsamp[,1])
+
+## plot timeseries for each samp to get a feel for variability and timing
+for(c in seq_along(dclass)){
+  selc = which(fsamp[,2] == dclass[c])
+  plot(dtime,dtime,ylim=range(dextract,na.rm = TRUE),type='n',main=dclass[c])
+  for(i in seq_along(selc)) lines(dtime,dextract[selc[i],])
+}
+
+## similarly, plot how biomass and cover change
+cextract = raster::extract(covS,fsamp[,1])
+for(c in seq_along(dclass)){
+  selc = which(fsamp[,2] == dclass[c])
+  plot(ctime,ctime,ylim=range(cextract,na.rm = TRUE),type='n',main=dclass[c])
+  for(i in seq_along(selc)) lines(ctime,jitter(cextract[selc[i],],amount=0.5))
+  abline(v=2004)
+}
+
+bextract = raster::extract(bioS,fsamp[,1])
+for(c in seq_along(dclass)){
+  selc = which(fsamp[,2] == dclass[c])
+  plot(ctime,ctime,ylim=range(bextract,na.rm = TRUE),type='n',main=dclass[c])
+  for(i in seq_along(selc)) lines(ctime,bextract[selc[i],],col=floor(cextract[selc[i],16]/10))
+  abline(v=2004)
+  print(c(dclass[c],mean(bextract[selc,15]/bextract[selc,14])))
+}
+
+```

### 2020-12-23 OR_original_model.Rmd
Code and data extractions to start doing test of the model against real data (landtrendr)

---
title: "OR Multi Disturbance"
author: "Michael Dietze"
date: "12/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("disturbance.R")
source("multi.ens.adj.R")
```

## set up
```{r}
## Load disturbance rate, state&transition, disturbed biomass stats
load("STprior.RData")
colnames(dextract) <- paste0("t",dtime)
dextract[is.na(dextract)] <- 0
d.prior["UN"] = d.prior["UN"] + d.prior[6]                         ## change at runtime (removes <NA>)
d.prior = d.prior[-6]

## Load Calibration
load("fit.RData")

## Load extracted site timeseries
load("extract.h03v04.RData")
```

## drivers
```{r}
latlon = points@coords
metFile = "daymet.RData"
if(file.exists(metFile)){
  load(metFile)
} else {
  dm = list()
  for(i in 1:nrow(latlon)){
    dm[[i]] = daymetr::download_daymet(site="h03v04",latlon[i,2],latlon[i,1],start=1985,end=2017,internal=TRUE,simplify=TRUE)
  }
  save(dm,file=metFile)
}

## extract Shortwave and convert to daily PAR
SW2PAR = 0.45
srad = subset(dm[[1]],measurement == "srad..W.m.2.")   
metTime <- as.POSIXct(paste(srad$year,srad$yday,sep="-"),format="%Y-%j")
rad <- matrix(NA,length(dm),length(metTime))
for(i in seq_along(dm)){
  srad = subset(dm[[i]],measurement == "srad..W.m.2.")   
  rad[i,] = srad$value*1e-6*86400*SW2PAR  # -> MJ /m2 /day PAR
}
plot(metTime,rad[1,],type='l')
## plot(date,PAR,col=2) ## calibration data
```

## functions
```{r,echo=FALSE,results='hide',message=FALS}
##' Single Site, Single Disturbance
##' @param Y observed data
##' @param R observation error
##' @param Q process error
##' @param Xic state initial condition matrix
##' @param THETA parameter matrix
##' @param dist  disturbance list: p.dist, mu0[3], V0[3,3], alloc soil 
SSSD.model <- function(Y,R,Q,PAR,Xic,THETA,dist){

  ## init
  NT = length(PAR)/365
  Nf = Nmu = Na = Xic
  ne = nrow(Xic)
  if(!is.matrix(THETA)) THETA = matrix(THETA,nrow=ne,ncol=length(THETA),byrow = TRUE)
  Forecast = Analysis = list()

  AnalysisSSSDMP <- "  ## SingleSite SingleDisturbance MultiPool
    model{
      ## prior
      muN ~ dmnorm(mufN,pfN)                ## forecast prior on Not disturbed
      muD ~ dmnorm(mufD,pfD)                ## forecast prior on Disturbed
      D ~ dbern(p)                          ## disturbance probability
      ## Analysis 
      n <- D*muD + (1-D)*muN                ## select between disturbed and undisturbed
      Y[H] ~ dmnorm(n[H],R[H,H])                       ## Observation error
  }"

  for(t in 1:2){

    ## update IC
    THETA[,9:11]  <- Na    ## Multinomial

    ## Forecast 
    for(i in 1:ne){
      tsel    = (t-1)*365 + 1:365             ## Select days from driver files
      Nmu[i,] = VSEM(THETA[i,], PAR = PAR[tsel])[365,2:4] ## deterministic simulation
      Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
    }

    ## Disturbance
    d.forecast <- rep(0,ne); d.forecast[1:ceiling(ne*0.1)] <- 1 #rbinom(ne,1,dist$p.dist)             ## prior on disturbance status
    sel <- !as.logical(d.forecast)                ## select undisturbed ensemble members
    if(sum(!sel)>2){
      sel.dist = which(!sel)
      for(j in sel.dist){                      ## run disturbance model
        Nf[j,]  = disturbance.p(Nf[j,],mu0=dist$mu0,V0 = dist$V0,alloc.soil = dist$alloc.soil)
      }
      mufD  = colMeans(Nf[!sel,])
      pfD = solve(var(Nf[!sel,]))#solve(diag(c(0.15,0.5,0.25)^2))
    } 
    if(sum(!sel) <= 2 || class(pfD) == "try-error") {   ## backup values if the ensemble stats fail
      mufD = c(dist$mu0[1], mean(Nf[,2]),  dist$mu0[2])            
      pfD  = solve(diag(c(0.15,0.5,0.25)^2))
    }
    Nf <- tobit(Nf)

    ## priors
    priors <- list(p    = dist$p.dist,         ## disturbance rate
                   mufN = colMeans(Nf[sel,]),    ## Not disturbed forecast means
                   mufD = mufD,   ## Disturbed forecast means
                   pfN  = solve(var(Nf[sel,])),  ## Not disturbed forecast precision
                   pfD  = pfD  ## Disturbed forecast precision
    )
    Forecast[[t]] <- c(list(d.forecast=d.forecast,Nf=Nf),priors)

    ## Analysis
    update = c(list(Y=Y[t,],R=solve(R)),priors,H = which(!is.na(Y[t,])))
    if(length(update$H) > 0){ ## data observed, run filter
      mod <- rjags::jags.model(file=textConnection(AnalysisSSSDMP),
                               data=update,
                               n.adapt=1000,n.chains=3,quiet=TRUE)
      jdat <- rjags::coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000)

      ## update parameters & save Analysis posterior
      dat = as.matrix(jdat)
      nsel <- grep(pattern="^n",x=colnames(dat))
      Xa = dat[,nsel]
      ca = dat[,"D"]
      run.adjustment = TRUE
      if(FALSE){
        d3 <-density(Xa[,3]) 
        plot(d3)
        lines(d3$x,dnorm(d3$x,update$Y[3],1/sqrt(R[3,3])),col=2)
        lines(d3$x,dnorm(d3$x,priors$mufN[3],1/sqrt(priors$pfN[3,3])),col="blue")
        p.kf = R[3,3]+priors$pfN[3,3]
        mu.kf = priors$mufN[3]*priors$pfN[3,3]/p.kf+
          update$Y[3]*R[3,3]/p.kf
        lines(d3$x,dnorm(d3$x,mu.kf,1/sqrt(p.kf)),col=3,lty=2)

        jtest <- "model{
         Y ~ dmnorm(n[H],R[H,H])
        }"
        jtest <- "  ## SingleSite SingleDisturbance MultiPool
    model{
      ## prior
      n[3] ~ dnorm(mufN[3],pfN[3,3])  
      Y[H] ~ dnorm(n[H],R[H,H])
      dummy <- mufD[1] + pfD[1,1] + p
  }"
      jtest <- "  ## SingleSite SingleDisturbance MultiPool
    model{
      ## prior
      n[3] ~ dnorm(mufN[3],pfN[3,3])  
      Y[H] ~ dnorm(n[3],R[H,H])
  }"
        update2 = list(mufN = update$mufN,
                       pfN = update$pfN,
                       Y=update$Y,R=R,H=3)
        mtest <- rjags::jags.model(file=textConnection(jtest),
                                   data = update,
#                               data=list(n = c(NA,NA,mu.kf),
#                                         R=matrix(p.kf,3,3),H=3),
                               n.adapt=1000,n.chains=3,quiet=TRUE)
        mdat <- rjags::coda.samples(mtest,variable.names=c("n"),n.iter=10000) 
      summary(mdat)  
      }
    } else {
      Xa = Nf
      ca = 1
      run.adjustment = FALSE
    }

      ## draw new IC
    if(run.adjustment){
    Na <- multi.ens.adj(
      Xf   = Nf,
      cf   = d.forecast,
      mu.f = rbind(priors$mufN,priors$mufD),
      Pf   = abind::abind(solve(priors$pfN),
                            solve(priors$pfD),
                            along=3),
      Xa = Xa,
      ca = ca)
    } else {
      Na = cbind(rep(0,nrow(Nf)),Nf)
    }
    Ca <- Na[,1] ## updated classf assignments
    Na <- tobit(Na[,-1])
    #  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
    #  Na <- dat[sel.ic,nsel]
    Analysis[[t]] <- list(Na = Na,
                          Ca = Ca,
                          CI = apply(dat,2,quantile,c(0.025,0.25,0.5,0.75,0.975))
    )
  }
  return(list(Forecast=Forecast,Analysis=Analysis))
}
```

## initial conditions & observations
```{r}
## BIOMASS
B2C = 0.5 ##biomass to carbon ratio
Yw <- bextract*B2C*1000*1e-4 ## convert Mg/ha  to  kg C 
Rw <- (84*B2C*1000*1e-4)^2 ## Observation Error: Kennedy et al 2018 Environ. Res. Lett. 13 025004; Fig 2 RMSE
## units: Mg/ha -> kg/m2

## grab the distribution of AGB for the first time point from all pixels
AGBprior <- Yw[,1]
ne = length(AGBprior) ## number of ensemble members

## spin up
PARlong <- as.vector(t(rad))
spin = VSEM(theta,PARlong)
thin = seq(nrow(spin)/2,nrow(spin),length=ne)
spinIC = spin[thin,2:4]
#for(i in 2:4){
#  hist(spin[thin,i],type='l')
#}

## disturbance recovery trajectory
nNBG = 200 ## years
NBG <- array(NA,c(ne,nNBG,3))
for(i in 1:ne){
  param = theta
  param[9:11] <- spinIC[i,]*c(0.01,1,0.01) ## set leaf and wood to 1%
  raw = VSEM(param,PARlong[1:(nNBG*365.25)])
  NBG[i,,] = raw[round(seq(365,nNBG*365.25,length=nNBG)),2:4]  ## grab pools at the end of each year
}

## for each AGBprior, grab the matching leaf and soil
ICprior <- matrix(NA,ne,3)
age = rep(NA,ne)
for(i in 1:ne){
  age[i] <- which.min((AGBprior[i]-NBG[i,,3])^2)  
  ICprior[i,] <- c(NBG[i,age[i],1:2],AGBprior[i])
}
hist(age)
colnames(ICprior) <- c("Cv","Cs","CR") 
pairs(ICprior)

```

## analyze sites - just one disturbance
```{r}
i = 181  # 177:263 = #fire
NT = floor(ncol(rad)/365)
Yi = matrix(NA,NT,3)  ## observed data
Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
R = diag(rep(Rw/1000,3))    ## observation error   ### div 1000 is test
Q = diag(c(Ql,Qs,Qw)^2)*10  ## process error
dist <- list(              ## disturbance parameters
  p.dist = d.prior[4],     ## disturbance probability
  mu0 = c(0.2,PBBAR[4,4]),   ## % retention [leaf,agb], first parameter made up, should include soil
  V0 =  diag(c(0.15,PBVAR[4,4])^2), #variance in retention
  alloc.soil = 0.05        ## made up  
)
test = SSSD.model(Y = Yi,R=R,Q=Q,
                  PAR=rad[i,],Xic=ICprior,THETA=theta[1:11],
                  dist = dist)
```

## visualization
```{r}
i = 181
pool.names=c("Cleaf","Csoil","Cstem")
for(i in 1:3){

    CI.Na = sapply(X = test$Analysis,function(x){
      y = x$CI
      sel.n = which(colnames(y) == paste0("n[",i,"]"))
      y[,sel.n]
    })

    CI.D = sapply(X = test$Analysis,function(x){
      y = x$CI
      sel.n = which(colnames(y) == paste0("D"))
      y[,sel.n]
    })

    CI.Nf = sapply(X = test$Forecast, function(x){
      mu = x$mufN[i]
      sd = 1/sqrt(x$pfN[i,i])
      return(mu + c(-1.96,0,1.96)*sd)
    })

  plot(Yi[,i],
       ylim = range(rbind(CI.Na,Yi[,i]),na.rm=TRUE),
       type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[3,],col="#b2df8a",lwd=2)
  points(Yi[,i])
  lines(CI.D[3,])

}

```

 TODO:
 * why is Analysis not contracting around the data?
    * reducing obs error by /10 didn't solve
 * why are the disturbance forecast pools crazy?
 * assess disturbance posterior

## Computer experiments
* Run under defaults
* Run under a grid of inc process error and decrease obs error and increase disturbance probability

## postprocessing

false positives

false negatives

classification success

## Next steps:
* Run with multiple disturbance options and PFT switching

### 2020-12-23 Paper.Rmd
Code and data extractions to start doing test of the model against real data (landtrendr)

@@
   + rightly captures that disturbance and land use change both critical to C accounting and, more importantly, a place where we can exert some control
+  + BULC
+  + state-and-transition modeling
+  + process-model projections
 3. Alternative: SDA

### 2020-12-23 StateTransition.Rmd
Code and data extractions to start doing test of the model against real data (landtrendr)

@@
 ## in all cases, disturbance CV was higher on an absolute than a relative basis
-save(PBBAR,PBVAR,file="STprior.RData")
+colnames(TRANS)<-c("UN","OTHER","CUT","FIRE","PEST")
+names(d.prior)[1:5] <- colnames(TRANS)
+rownames(TRANS)<-c("water","devel","barren","forest","shrub","---","herb","crop","wetland","NA")
+TRANS
+save(d.prior,TRANS,PBBAR,PBVAR,file="STprior.RData")
 ```
@@
 
+coord = xyFromCell(bioS,fsamp[,1])
+points = spTransform(SpatialPoints(coord,crs(bioS)),crs("+proj=longlat +datum=WGS84"))
+
+save(points,bextract,cextract,dextract,file="extract.h03v04.RData")
 ```

### 2020-12-23 calibrate.Rmd
Code and data extractions to start doing test of the model against real data (landtrendr)

---
title: "Calibration"
author: "Michael Dietze"
date: "11/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BayesianTools)
citation("BayesianTools")
set.seed(42)
```

## Data stream 1: Eddy covariance flux & met
```{r}
AF = read.csv("Metolius.csv",header=TRUE)

AF.biomass = raster::extract(bioS,AF[,c("Lat.","Long.")])

LAI.BADM = c(1.87, 1.96, 2.35, 2.37, 2.7, 2.8, 2.87, 3.01)

SoilCcon = c(0.37192982456,0.18974358974,0.10518518519)  ##gC/kgSoil
SoilBD   = c(1.14,1.17,1.35)/1000 ## kgSoil/cm3  
SoilDepth= c(20,30,50)  ## cm
SoilC = sum(SoilCcon*SoilBD*SoilDepth*1e4) ## gC/kgSoil * kgSoil/cm3 * cm3/m2  ->kgC/m2

flux = read.csv("Ameriflux/AMF_US-Me2_BASE-BADM_14-5/AMF_US-Me2_BASE_HH_14-5.csv",header = TRUE,skip=2,na.strings = c(-9999,-6999))
time = as.POSIXlt(as.character(flux$TIMESTAMP_START),format="%Y%m%d%H%M")

## PAR data
ein = 2.17e-5 #J mol-1 from ED2 wiki
ein = 0.327 * 1e6 #J/umol * 1e6 umol/mol #units from Apogee 
date.i = as.Date(time)
date = as.Date(names(tapply(date.i,date.i,mean)))
PAR = tapply(flux$PPFD_IN,date.i,mean)*ein*86400*1e-12 ## µmolPhoton m-2 s-1 -> MJ /m2 /day

## Gapfill
doy <- tapply(lubridate::yday(time),date.i,mean)
PARclim <- tapply(PAR,doy,mean,na.rm=TRUE)
PAR[is.na(PAR)] <- PARclim[doy[is.na(PAR)]]

plot(date,PAR)

## NEE
NEE = tapply(flux$FC,date.i,mean,na.rm=TRUE)
NEEp = tapply(flux$NEE_PI,date.i,mean,na.rm=TRUE)
cnt = tapply(is.na(flux$FC),date.i,sum)
cntp = tapply(is.na(flux$NEE_PI),date.i,sum)
NEE2 = NEE
NEE2[cnt<1] = NA
#plot(time,flux$FC)
plot(date,NEE)
points(date,NEE2,col=2)

plot(doy,NEE)
points(doy,NEE2,col=2)

## aggregate flux to annual and change units to kgC/m2/year
years <- lubridate::year(date)
aNEE = tapply(NEE,years,mean,na.rm=TRUE)*1.2e-8 * 86400 *365 ## umol/m2/sec * kg/umol * sec/day
year = as.numeric(names(aNEE))
plot(year,aNEE)

## leaf and wood biomass
Cw = Cl = rep(NA,19)
Cl[c(1,2,3,4,9,10,5,6,7,8)] = c(452.80554869,460.10972745,469.24831463,477.98289883,525.91518937,535.12650483, 485.76130987,496.26140507,506.1776831,514.03435729)/1000
Cw[c(1,2,3,4,10,5,6,7,8,9)] = c(4435.1119994,4793.8053523,4968.8831826,5140.6894352,6293.0243086,5305.4365914,5504.2730671,5699.6935828,5871.0926993,6100.944853)/1000

```

## Data stream 2: LAI

## Data stream 3: AGB

## Data stream 4: spatial meteorology

## Define Likelihood, priors
```{r}
parSel = c(3, 4, 6, 7, 8, 12:14)

## Priors
refPars <- VSEMgetDefaults()
refPars[12,] <- c(.1, 0.01, 4)  ## sigma NEE
refPars[13,] <- c(.2, 0.01, 4)  ## sigma Cl
refPars[14,] <- c(.5, 0.01, 4)  ## sigma Cw
rownames(refPars)[12:14] <- c("sd_NEE","sd_leaf","sd_wood")
refPars["LUE","best"] <- 0.00022 ## tuning
refPars["GAMMA",] <- c(0.55,0.2,0.7) ## tuning
refPars["tauS",] <- c(5e+04,4000,1e5) 
refPars["LAR","best"] <- 1000/127/0.5   ## Ameriflux BADM says LMA is 127 g m-2  -> m2/kgC
refPars["tauV","best"] <- 4.7*365.25 ## Wikipedia says var. ponderosa has a longevity of 4.7+/-0.14 years
refPars["tauR",] <- c(100,5,400)*356 ## sylvics says 300-year lifespan, but there will be branchfall and fine litter
refPars["Av","best"] <- 113.81026786/(113.8 + 206.2)*0.95 ## Ameriflux litter prod/(litter+wood prod) * tuning fudge factor
refPars["Cv","best"] <- Cl[1]*1.5-0.5*Cl[2] #0.4528 ## Ameriflux gC/m2 -> kgC/m2 ## LAI.BADM[1]/refPars["LAR","best"]
refPars["Cr","best"] <- Cw[1]*1.5-0.5*Cw[2]#4.4351119994 ## Ameriflux
refPars["Cs","best"] <- SoilC/10  ## BADM value seems too high
prior <- createUniformPrior(lower = refPars$lower[parSel], 
                            upper = refPars$upper[parSel], best = refPars$best[parSel])

## Likelihood

# here is the likelihood 
likelihood <- function(par, sum = TRUE){
  # set parameters that are not calibrated on default values 
  x = refPars$best
  x[parSel] = par
  pred <- VSEM(x[1:11], PAR) # replace here VSEM with your model
  NEEm = tapply(pred[,"NEE"],years,sum)
  Cl.m  = tapply(pred[,"Cv"],years,mean)
  Cw.m  = tapply(pred[,"CR"],years,mean)
  llValues <- c(0.25*dnorm(x = aNEE, mean = NEEm, sd = x[12], log = TRUE),  ## downweight because less important
              dnorm(Cl,Cl.m,x[13],log=TRUE),
              4*dnorm(Cw,Cw.m,x[14],log=TRUE)) ## upweight (destroys prob interpretation, but var most important)
  llValues[is.nan(llValues)]<- -Inf
  return(sum(llValues,na.rm = TRUE))
}

# optional, you can also directly provide lower, upper in the createBayesianSetup, see help
bayesianSetup <- createBayesianSetup(likelihood, prior, 
                                     names = rownames(refPars)[parSel])#,parallel = TRUE)#,parallelOptions = list(variables=list(refPars=,"parSel","PAR","years","aNEE","Cl","Cw")))
```

## Precalibration runs
```{r}
pred = VSEM(refPars[1:11,"best"],PAR = PAR)
NEEm = tapply(pred[,"NEE"],years,sum)

plot(NEEm,aNEE)
abline(0,1,col=2,lty=2,lwd=3)

#plot(doy,NEE)
#points(doy,pred[,"NEE"],pch="+",col=2)

plot(year,tapply(pred[,"Cv"],years,mean))
points(year,Cl,col=2,pch="+")

plot(year,tapply(pred[,"CR"],years,mean))
points(year,Cw,col=2,pch="+")

plot(pred[,"Cs"])

```

## Fit via MLE
```{r}
negLL <- function(par){-1*likelihood(par)}
mle = optim(refPars[parSel,"best"],negLL)
#mle = optim(refPars[parSel,"best"],negLL,method="L-BFGS-B",lower=refPars[parSel,"lower"],upper=refPars[parSel,"upper"])
cbind(refPars[parSel,"best"],mle$par)

likelihood(refPars[parSel,"best"])
likelihood(mle$par)

```

## Fit via Bayesian Tools
```{r}
settings <- list(iterations = 25000, nrChains = 2)#,parallel=TRUE)
out <- runMCMC(bayesianSetup = bayesianSetup, sampler = "DEzs", settings = settings)
foo = summary(out)
```

```{r}
plot(out)
```

##postcalibrartion
```{r}
ref = refPars[,"best"]
#fit = BayesianTools::MAP(out)$parametersMAP
fit = mle$par
ref[12:14] = fit[6:8]
likelihood(ref[parSel])
likelihood(fit)

theta = refPars[,"best"]
cbind(theta[parSel],fit)
theta[parSel] = fit
names(theta) <- rownames(refPars)
post = VSEM(theta[1:11],PAR = PAR)
NEEm = tapply(post[,"NEE"],years,sum)

plot(NEEm,aNEE)
abline(0,1,col=2,lty=2,lwd=3)

#plot(doy,NEE)
#points(doy,pred[,"NEE"],pch="+",col=2)

Cl.pred = tapply(pred[,"Cv"],years,mean)
Cl.post = tapply(post[,"Cv"],years,mean)
plot(year,Cl.pred,ylim=range(c(Cl,Cl.post,Cl.pred),na.rm=TRUE))
lines(year,Cl.post,col=3)
points(year,Cl,col=2,pch="+")

Cw.post = tapply(post[,"CR"],years,mean)
plot(year,tapply(pred[,"CR"],years,mean))
lines(year,Cw.post,col=3)
points(year,Cw,col=2,pch="+")

Cs.post = tapply(post[,"Cs"],years,mean)
plot(pred[,"Cs"])
lines(Cs.post)
abline(h=SoilC/10)

## Rough approximation of process error: RMSE on delta
Ql <- sqrt(mean((diff(Cl)-diff(Cl.post))^2,na.rm=TRUE))
Qw <- sqrt(mean((diff(Cw[-1])-diff(Cw.post[-1]))^2,na.rm=TRUE)) ## first value bit of an outlier
Qs <- sqrt(mean(diff(Cs.post))) ## assume 100% error (i.e. true delta Soil C = 1)

## Cv of Q
Ql/mean(diff(Cl),na.rm=TRUE)
Qw/mean(diff(Cw),na.rm=TRUE)

save(theta,Ql,Qw,Qs,file="fit.RData")
```

## Ameriflux Download Manifest
```
#downloadDate:20201206113353
#numberOfSitesDownloaded:6
#emailForSitePIs:bev.law@oregonstate.edu
#dataFiles
SITE_ID,filename,start_year,end_year,version,filetype,timestamp
US-Me1,AMF_US-Me1_BASE_HH_3-5.csv,2004,2005,3-5,FLUX-MET,201902011245
US-Me2,AMF_US-Me2_BASE_HH_14-5.csv,2002,2020,14-5,FLUX-MET,202011171249
US-Me3,AMF_US-Me3_BASE_HH_4-5.csv,2004,2009,4-5,FLUX-MET,201811191702
US-Me4,AMF_US-Me4_BASE_HH_4-1.csv,1996,2000,4-1,FLUX-MET,201602060503
US-Me5,AMF_US-Me5_BASE_HH_2-1.csv,1999,2002,2-1,FLUX-MET,201605150147
US-Me6,AMF_US-Me6_BASE_HH_12-5.csv,2010,2020,12-5,FLUX-MET,202011171249
```

### 2020-12-23 methods.Rmd
Code and data extractions to start doing test of the model against real data (landtrendr)

@@
 1. Illustration of problem
-2. Derivation Multinomial
+
+## Derivation of Multinomial Filter
 
@@
 
-*Ensemble adjustment by class*
+Conceptual figure
 
-Conceptual figure
+## Class-based Ensemble adjustment
+
+## VSEM Description
+
+## Simulated data experiments
 
@@
 Bonus would be if you could suggest places where you know there’s ground data we could use to validate the model (disturbances at flux towers, LTER sites, FIA plots, etc).
+
+Things to try:
+* Run just the Assimilation step for simulated disturbance that vary in magnitude, process error, observation error, etc to detect sensitivity
+* Approximation of not simulating disturbed ensemble members and relying SOLEY on ensemble adjustment
+* Analytical approximation of sequential beta-binomial then KF vs full joint model
+
+## Calibration of VSEM
+*
+
+[@BayesianTools]
+
+## Comparison to real-world data
+
+* LandTrendr tile
+* Construction of frequency and transition matrix priors

### 2020-12-23 disturbance.R

@@ -3,9 +3,34 @@ tobit <- function(x){
   x
 }
 
-disturbance <- function(x){ ## x = Bleaf, Bsoil, Bstem
+tobit2 <- function(x,upper){
+  x[x < 0] <- 0
+  x[x>upper] <- upper
+  x
+}
+
+
+disturbance <- function(x,                         ## x = Bleaf, Bsoil, Bstem
+                        mu0 = c(1,1),              ## mean leaf & stem biomass after disturbance
+                        V0 = diag(c(0.15,0.25))^2, ## var in post disturbance leaf and stem biomass, 
+                        alloc.soil = 0.25){        ## fraction of removed C that goes in the soil
   old.biomass <- x[c(1,3)]
   new.biomass <- tobit(mvtnorm::rmvnorm(1,mu0,V0)) ## draw disturbed leaf and stem
+  check.max <- which(new.biomass > old.biomass)
+  if(length(check.max)>0) new.biomass[check.max] = old.biomass[check.max]
+  residual = sum(old.biomass-new.biomass)
+  x[c(1,3)] <- new.biomass
+  x[2] <- x[2] + residual*alloc.soil
+  removal <- residual*(1-alloc.soil)
+  return(x)
+}
+
+disturbance.p <- function(x,                         ## x = Bleaf, Bsoil, Bstem
+                        mu0 = c(1,1),              ## mean leaf & stem % retention after disturbance
+                        V0 = diag(c(0.15,0.25))^2, ## var in post disturbance leaf and stem biomass, 
+                        alloc.soil = 0.25){        ## fraction of removed C that goes in the soil
+  old.biomass <- x[c(1,3)]
+  new.biomass <- tobit(mvtnorm::rmvnorm(1,mu0*old.biomass,tcrossprod(old.biomass,old.biomass)*V0)) ## draw disturbed leaf and stem
   check.max <- which(new.biomass > old.biomass)
   if(length(check.max)>0) new.biomass[check.max] = old.biomass[check.max]
   residual = sum(old.biomass-new.biomass)

### 2020-12-30 OR_original_model.Rmd
Flushed out analysis of basic real-world application. Methods: model description

@@
 load("fit.RData")
+Qsoil = Qs
 
@@
 
-  for(t in 1:2){
+  for(t in 1:NT){
 
@@
                           Ca = Ca,
-                          CI = apply(dat,2,quantile,c(0.025,0.25,0.5,0.75,0.975))
+                          CI = apply(dat,2,quantile,c(0.025,0.25,0.5,0.75,0.975)),
+                          mu = apply(dat,2,mean,na.rm=TRUE)
     )
@@
 ```{r}
-i = 181  # 177:263 = #fire
+Rmult = c(1,0.1,0.01,0.001)
+Qmult = c(1,10)
+Pmult = c(1)
+Ri = 4 #4
+Qi = 1 #2
+Pi = 1
+#i = 181  # 177:263 = #fire
+od = c(83:176,264:356)
+for(i in 83:356){
+  ofile = paste0("OR_SD_",
+                 formatC(i, width = 3, format = "d", flag = "0"),
+                 "_P",Pi,"Q",Qi,"R",Ri,".RData")
+  print(ofile)
   NT = floor(ncol(rad)/365)
@@
   Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
-R = diag(rep(Rw/1000,3))    ## observation error   ### div 1000 is test
-Q = diag(c(Ql,Qs,Qw)^2)*10  ## process error
+  R = diag(rep(Rw,3))*Rmult[Ri]    ## observation error
+  Q = diag(c(Ql,Qsoil,Qw)^2)*Qmult[Qi]  ## process error
+  dtype = which(names(d.prior) == dextract[i,"t2004"])
+  if(length(dtype)==0) break
   dist <- list(              ## disturbance parameters
-  p.dist = d.prior[4],     ## disturbance probability
-  mu0 = c(0.2,PBBAR[4,4]),   ## % retention [leaf,agb], first parameter made up, should include soil
-  V0 =  diag(c(0.15,PBVAR[4,4])^2), #variance in retention
+    p.dist = d.prior[dtype]*Pmult[Pi],     ## disturbance probability
+    mu0 = c(0.2,PBBAR[4,dtype]),   ## % retention [leaf,agb], first parameter made up, should include soil
+    V0 =  diag(c(0.15,PBVAR[4,dtype])^2), #variance in retention
     alloc.soil = 0.05        ## made up  
@@
                   dist = dist)
+  save(test,file=ofile)
+}
 ```
@@
 
+```{r}
+## Detect set of experiments that have been run
+runs <- dir(pattern="*.RData")
+runs <- runs[grepl("^OR",runs)]
+exp = sapply(strsplit(runs,"_",fixed=TRUE),function(x){
+  sub(".RData","",x[4],fixed = TRUE)
+})
+exps <- unique(exp)
+
+## load experiments that haven't yet been summarized
+# elist <- list()   ## run once
+t04 = 15            ## time of disturbance assessement (2004)
+tlo = 28            ## time of last observation
+for(i in seq_along(exps)){
+  if(is.null(elist[[exps[i]]])){
+    esel <- which(exp==exps[i])
+    distrate = rep(NA,nrow(dextract))
+    divergence = predist = distsize = matrix(NA,nrow(dextract),6) ## pre and post disturbance biomass pools (w/uncertainty)
+    for(j in seq_along(esel)){
+      ## load file
+      load(runs[esel[j]])
+      k = as.numeric(unlist(strsplit(runs[esel[j]],"_",fixed=TRUE))[3])
+
+      ## extract detection rate
+      distrate[k]=test$Analysis[[t04]]$mu["D"]
+
+      ## capture of disturbance change
+      predist[k,1]    = test$Analysis[[t04-1]]$mu["n[3]"]
+      predist[k,2:6]  = test$Analysis[[t04-1]]$CI[,"n[3]"]
+      distsize[k,1]   = test$Analysis[[t04]]$mu["n[3]"]
+      distsize[k,2:6] = test$Analysis[[t04]]$CI[,"n[3]"]
+
+      ## Filter divergence
+      divergence[k,1]   = test$Analysis[[tlo]]$mu["n[3]"]
+      divergence[k,2:6] = test$Analysis[[tlo]]$CI[,"n[3]"]
+    } ## end loop over individual runs
+
+    elist[[exps[i]]] = list(distrate=distrate,predist=predist,
+                            distsize=distsize,divergence=divergence)
+
+  } ## end if not processed
+
+} ## end loop over experiments
+
+## grab matching data
+obs.pre  = Yw[,t04-1]
+obs.dist = Yw[,t04]
+obs.tlo  = Yw[,tlo]
+obs.class= dextract[,"t2004"]/10
+```
+
+## analyses
+
+```{r}
+pdf("OR_exps_diagnostics.pdf")
+trim <- function(x,lo=0.0001,hi=0.9999){
+  x[x<lo] <- lo
+  x[x>hi] <- hi
+  return(x)
+}
+betatrim <- function(y){
+    n.obs <- sum(!is.na(y))
+    (y * (n.obs - 1) + 0.5) / n.obs
+}
+stats <- list()
+for(i in seq_along(exps)){
+
+  ## divergence
+  div = (elist[[i]]$divergence[,1] - obs.tlo)/
+    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
+  div.reg    = lm(div~obs.tlo)
+  div.median = median(div,na.rm = TRUE)
+  div.cov    = sum(div < 1 & div > -1,na.rm=TRUE)/sum(!is.na(div))
+
+  hist(div,main=exps[i])
+  abline(v=c(-1,1),lty=2)
+  plot(obs.tlo,div,col=obs.class,main=exps[i])
+  abline(h=c(-1,1),lty=2)
+  abline(div.reg,col=2)
+
+  ## Disturbance detection
+  distrate = elist[[i]]$distrate
+  hist(distrate,main=exps[i])
+
+  ## Disturbance skill
+  ord <- order(distrate)
+  plot(distrate[ord],col=obs.class[ord],main=exps[i])
+  dist.bar   = mean(distrate,na.rm = TRUE)
+  dist.class = tapply(distrate,obs.class,mean,na.rm=TRUE)
+
+  ## vs predisturbance biomass
+  distrate2 = betatrim(distrate)
+  pred.seq = data.frame(obs.pre=seq(0,max(obs.pre,na.rm = TRUE),length=500),
+                        distprop = seq(-1,1,length=500))
+  dist.reg.pre  = lm(distrate ~ obs.pre)
+  dist.beta.pre = betareg::betareg(distrate2 ~ obs.pre,
+                                   na.action=na.omit,link="logit")
+  plot(obs.pre,distrate,col=obs.class,main=exps[i])
+  abline(dist.reg.pre,col=2)
+  lines(pred.seq$obs.pre,predict(dist.beta.pre,newdata=pred.seq),col="green")
+
+  ## vs proportional disturbance size
+  distprop = 1 - elist[[i]]$distsize[,1]/elist[[i]]$predist[,1] ## predicted
+  distprop = 1 - obs.dist/obs.pre ## observed
+  dist.beta.prop = betareg::betareg(distrate2 ~ distprop,
+                                   na.action=na.omit,link="logit")
+  plot(distprop,distrate,col=obs.class,main=exps[i])
+  lines(pred.seq$distprop,predict(dist.beta.prop,newdata=pred.seq),col="green")
+
+  ## vs ABSOLUTE disturbance size
+  distabs = obs.pre-obs.dist
+  pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
+  dist.beta.abs = betareg::betareg(distrate2 ~ distabs | distabs,
+                                   na.action=na.omit,
+                                   link="logit",
+                                   link.phi = "identity",
+                                   start=c(0,5,0.5,0),
+                                   type="BR")
+  dist.qb.abs = glm(distrate ~ distabs,
+                    na.action=na.omit,
+                    family=quasibinomial("logit")
+  )
+  dist.bin.abs = glm(round(distrate) ~ distabs,
+                    na.action=na.omit,
+                    family=binomial("logit")
+  )
+  plot(distabs,distrate,col=obs.class,main=exps[i])
+  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile"),col="green")
+  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.75),col="green",lty=2)
+  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.25),col="green",lty=2)
+  lines(pred.seq$distabs,predict(dist.qb.abs,newdata=pred.seq,type="response"),col="purple")
+  lines(pred.seq$distabs,predict(dist.bin.abs,newdata=pred.seq,type="response"),col="blue")
+
+  ## vs divergence
+  absdiv = abs(div)
+  dist.beta.absdiv = betareg::betareg(distrate2 ~ absdiv,
+                                   na.action=na.omit,link="logit")
+  dist.bin.absdiv = glm(round(distrate) ~ absdiv,
+                                   na.action=na.omit,
+                        family = binomial("logit"))
+  pred.seq$absdiv = seq(0,max(absdiv,na.rm=TRUE),length=500)
+  plot(absdiv,distrate,col=obs.class,main=exps[i])
+  abline(v=c(-1,1),lty=2,col=3)
+  abline(lm(distrate ~ absdiv),col=2)
+  lines(pred.seq$absdiv,predict(dist.beta.absdiv,newdata=pred.seq),col="green")
+  lines(pred.seq$absdiv,predict(dist.bin.absdiv,newdata=pred.seq,type="response"),col="blue")
+
+ dist.qb.combo = glm(distrate2 ~ obs.pre + distprop +
+                        distabs + absdiv + obs.class,
+                      na.action=na.omit,
+                      family=quasibinomial("logit"))
+  dist.bin.combo = glm(round(distrate) ~ obs.pre + distprop +
+                        distabs + absdiv + as.factor(obs.class),
+                      na.action=na.omit,
+                      family=binomial("logit"))
+
+# dist.beta.combo = betareg::betareg(distrate2 ~ obs.pre + obs.class + distabs + absdiv,
+#                                   na.action=na.omit,link="logit")
+
+  stats[[i]] <- list(
+    div.reg = div.reg,
+    div.median = div.median,
+    div.cov = div.cov,
+    dist.bar = dist.bar,
+    dist.class = dist.class,
+    dist.pre =  dist.beta.pre,
+    dist.prop = dist.beta.prop,
+    dist.beta.abs =  dist.beta.abs,
+    dist.bin.abs  = dist.bin.abs,
+    dist.absdiv   = dist.beta.absdiv,
+    dist.combo = dist.bin.combo
+  )
+
+}
+dev.off()
+```
+
+factors affecting detection of disturbance
+```{r}
+Ps = as.numeric(substr(names(elist),2,2))
+Qs = as.numeric(substr(names(elist),4,4))
+Rs = as.numeric(substr(names(elist),6,6))
+
+## P
+
+## Q
+
+## R
+
+## Divergence
+div.cov = sapply(stats,"[[","div.cov")
+plot(Rmult[Rs],div.cov,col=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage")
+
+div.median = sapply(stats,"[[","div.median")
+plot(Rmult[Rs],div.median,col=Qs,log="x",xlab="Variance Multiplier",ylab="Median Divergance")
+
+div.slope = sapply(stats,function(x){coef(x$div.reg)[2]})
+plot(Rmult[Rs],div.slope,col=Qs,log="x",xlab="Variance Multiplier",ylab="Divergance Slope")
+## Disturbance @ default
+summary(stats[[4]]$dist.combo)
+## nothing matters
+
+## Disturbance @ P1Q2R2 (intermediate)
+summary(stats[[4]]$dist.combo)
+## detection varies with divergence & disturbance type
+
+## Disturbance magnitude: P1Q1R4
+i = 3
+summary(stats[[i]]$dist.combo)  ## absolute magnitude of divergence is best predictor
+summary(stats[[i]]$dist.bin.abs)
+beta = coef(stats[[i]]$dist.bin.abs)
+## threshold mx+b = 0
+thresh = -beta[1]/beta[2]
+
+## Disturbance type: P1Q1R4 (reduced divergence)
+stats[[i]]$dist.class
+tapply(distabs,obs.class,summary) ## obs dist size by class
+thresh.prop = tapply(distabs,obs.class,function(x){sum(x>thresh,na.rm = TRUE)/sum(!is.na(x))}) ## prop > detection threshold
+thresh.prop
+stats[[i]]$dist.class
+dclass = stats[[i]]$dist.class
+#dclass[1] = 0
+dcol = as.numeric(names(stats[[i]]$dist.class)) + 1
+plot(thresh.prop,dclass,col=dcol,pch=19)
+abline(0,1,lty=2,col=2)
+legend("topleft",legend = dcol-1,col=dcol,pch=19)
+
+## additional experiments to run (P1Q1R4)
+```
+
 false positives

### 2020-12-30 methods.Rmd
Flushed out analysis of basic real-world application. Methods: model description

@@
 # Methods
+
+## VSEM (Very Simple Ecosystem Model)
+
+To illustrate the problem of assimilating disturbance, and present potential solutions, we make use of the Very Simple Ecosystem Model (VSEM), a daily-timestep point-scale model developed by David Cameron that is available in the BayesianTools R package [@BayesianTools].
+VSEM contains three carbon pools, leaf (Cl), wood (Cw), and soil (Cs), and calculates leaf area index based on specific leaf area (SLA), $LAI = SLA*Cl$.
+Gross Primary Productivity (GPP) is estimated based on incoming photosynthetically active radiation (PAR), light use efficiency (LUE), and Beer's Law light interception ($1-exp(-K_{ext}\cdot LAI)$) with a extinction coefficient $K_{ext}$.
+Autotrophic respiration, Ra, is a fixed fraction, $\Gamma$ of GPP.
+Of the remaining carbon (net primary productivity), a fixed fraction, $Av$, is allocated to leaves and the remainder is allocated to wood.
+Leaves and wood turnover into soil C based on fixed longevities, $\tau_l$ and $\tau_w$, while heterotrophic respiration, Rh, occurs based on a fixed residence time $\tau_s$.
+Net Ecosystem Exchange follows the atmospheric sign convention $Ra + Rh - GPP$.
+The VSEM was selected for its speed and ease of understanding, but its fundamental structure is analogous to much more complex, large-scale terrestrial carbon models.
+
+## Ensemble Adjustment Kalman Filter
+
+To
 

### 2021-02-02 SingleSite_SingleDisturb.Rmd
update SSSD to lower disturbance rate, larger ensemble, and saving of key results to a RData file that's read by the paper code (knits into paper itself)

@@
 
-```{r,echo=1:2, fig.width=7, fig.height=6}
+```{r,echo=1:2}
+#fig.width=7, fig.height=6
 set.seed(1066)
-p.dist = 0.05
+p.dist = 0.005
 n0 = 1
@@
 points(time,n.obs,pch=18,col=2,cex=1.5)
-legend("topleft",legend=c("truth","obs"),lwd=c(2,NA),pch=c(NA,18),col=1:2)
+legend("topright",legend=c("truth","obs"),lwd=c(2,NA),pch=c(NA,18),col=1:2)
 ## Simulate less frequent observations
@@
 n0.sd = n.sd
-NE = 1000
+NE = 10000
 n0s = rlnorm(NE,log(nIC),n0.sd)
@@
 Na = rnorm(nMC,Xl,n.sd)
-Nf = N*NA
+Nf = Na*NA
 
@@
        lwd=5,col=c("orange","#b2df8a","blue","red","black"))
+
+### Save outputs required to include figure in Results
+save(data,CI.fN,time,an0.stats,an.stats,NT,Dbar,file="SSSD.RData")
+
 ```

### 2021-02-02 methods.Rmd
Methods: initial KF text, SSSD figure, cleaning up of outline for real-world analysis

@@
 
-## Ensemble Adjustment Kalman Filter
+## Ensemble Kalman Filter
 
-To
+The Ensemble Kalman Filter (EnKF) [CITE] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approached have with capturing discrete disturbances and as our jumping off point for developing more flexible methods.
+Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabalistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g.
+carbon pools) and sometimes model parameters.
+These updated variables are then used as the initial conditions for the next Forecast round.
+
+Within the EnKF, the probabalistic Forecast step is performed using an ensemble of model runs that vary in at least initial conditions, but potentially also model parameters, drivers, and process error.
+The ensemble forecast approach is easy to apply to any model, is particularly good at capturing nonlinear predictions, and unlike many other approaches does not require solving for the adjoint of the model.
+In EnKF the forecast, $f$, at any point in time, $t$, is then summarized in terms of a vector of means across ensemble members, $\mu_{f,t}$ and the sample covariance, $P_{f,t}$.
+In our example we are interested in the three state variables in the VSEM, ${C_l,C_w,C_s}$, at a single site and assume an ensemble size of 200.
+Therefore $\mu_{f,t}$ is a vector of length 3 and $P_{f,t}$ is a $3 \times 3$ covariance matrix.
+If we were making a spatial forecast at $n_s$ locations, $\mu_{f,t}$ would be a vector of length $3 n_s$ and $P_{f,t}$ would be a $3 n_s \times 3 n_s$ covariance matrix that would capture the full set of correlations across both variables and space.
+An important difference between data assimilation and spatial statistics is that in data assimilation the spatial covariance matrix is coming from the mechanistic understanding embedded in our process model, and the spatial covariance in the forecast inputs (initial condition, parameters, drivers, process error), rather than an empirical semivariogram.
+
+The Analysis step in the EnKF is the same as that in the standard Kalman Filter
 
 1. Illustration of problem
+
+```{r, echo=FALSE}
+SSSD_figure <- function(){
+  load("SSSD.RData")
+  plot(data$Y,ylim = range(rbind(CI.fN,data$Y)),xlim=xlim,type='n',
+     ylab="LAI",xlab="time")
+  polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
+    an0.stats[1, 1])), border = NA, col = '#FFBF6C')
+  polygon(cbind(c(time, rev(time), 1), c(an.stats[2, ], rev(an.stats[4, ]),
+    an.stats[1, 1])), border = NA, col = ecoforecastR::col.alpha('#b2df8a',0.7))
+  #ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
+  ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
+
+  lines(an0.stats[3,],col="orange",lwd=2)
+  lines(an.stats[3,],col="#b2df8a",lwd=2)
+  #lines(CI.fSSSD[2,],col="blue",lwd=2)
+  lines(CI.fN[3,],col="blue",lwd=2)
+  lines(Dbar,lwd=2)
+  points(data$Y,pch=18,col=2)
+  legend("topright",legend=c("EnKF, no disturb","EnKF, disturb","Multinomial","data","P(dist)"),
+       lwd=5,col=c("orange","#b2df8a","blue","red","black"))
+}
+SSSD_figure()
+```
 
@@
 *Priors*
-
   $$X^L_k \sim MVN(\vec{\mu}_{f,k},P_{f,k})$$
   $$\vec{\rho} \sim Multinomial(1,\vec{p})$$
+
 *Process Model*
-
   $$X = X^L \vec{\rho}$$
@@
   $$Y \sim MVN(X,R)$$
+
+**ADD Conceptual figure**
+
 ### Ensemble Adjustment
@@
 
-Conceptual figure
+## Simulated data experiments (known disturbance)
 
-## Class-based Ensemble adjustment
-
-## VSEM Description
-
-## Simulated data experiments
-
-3. Simulated data experiments (known disturbance)
-  + describe model
-  + x Single Site, Single Disturbance  (add a case where EnKF has disturb but it's more rare)
-  + x SSSD Multi Pool
+  + x Single Site, Single Disturbance (SSSD)
+      vs. EnKF with and without Disturbance
+  + x SSSD Multi Pool (VSEM)
   + x SS Multi-Disturbance - distinct
@@
   + x Simple Multi-site: spatial cov
-  + SSMD - initially overlapping but diverge
+  + Future
+      * SSMD - initially overlapping but diverge
         * how to handle updating of assignment based on new information?
-  + SSMD - transition probabilities change with state
-  + rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
-4. Real-world examples
-  + Calibrate state-and-transition prior
+      * SSMD - transition probabilities change with state
+        * rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
+
+## Real-world examples
+
+Extract some example time series for single pixels (e.g.
+AGB product) where the disturbance (2004) it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance).
+
+### Study region and data sources
+
     + Landtrendr
@@
       - nearest neighbor resampled to same as cover
-    + LAI??
-  + Calibrate model
+
  + Applications: PNW
@@
 
-Email to Robert:
+### Construction of disturbance frequency and transition matrix priors
 
-I need to extract some example time series for single pixels (e.g.
-AGB product) where the disturbance it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance).
-From Landtrendr itself I’m looking to know when you attributed the year, severity, and cause so we can see if our assimilation is getting it “right”.
-Bonus would be if you could suggest places where you know there’s ground data we could use to validate the model (disturbances at flux towers, LTER sites, FIA plots, etc).
+* recoded to a smaller set of disturbances: none, fire, pest, clearcut, other
+* Calculate frequencies of different disturbance types
+* Calculate disturbance-by-landCover matrix of transition probabilities
+* Use these as disturbance priors
 
-Things to try:
+### Calibration of VSEM
+
+[@BayesianTools]
+
+* Calibrate at US-Me2, 2002-2019 
+* Initialize with BADM soil, leaf, and stem C
+* Priors (and some fixed parameters) using BADM; default uniform for rest
+* Likelihood against tower _annual_ NEE and BADM leaf and stem biomass
+* MCMC using DEzs algorithm for 25000 iterations of 2 chains
+* GBR, effective sample
+
+### Comparison to real-world data
+
+* Selected conifer forest land cover type
+* Sampled ~100 points from each disturbance type for 2004
+* Daymet PAR as driver
+* IC
+  * ensemble spin up
+  * construct disturbance recovery trajectory
+  * Extract rs map AGB
+  * Match leaf and soil C from spin-up trajectory
+
+* Experiments
+  * run SSSDMP model for each site independently
+  * run other versions varying R, Q, and the disturbance probability
+  * Score: filter divergence, disturbance detection & skill
+  * Regress vs: pre-disturbance biomass, size of disturbance (abs & rel), type, etc
+
+* RESULTS
+  * needed to reduce obs and process error to work
+  * abs magnitude of disturbance is best predictor of assim prob.
+    * given magnitude, dist type n.s.
+
+* FUTURE DIR
+  * disturbance products are 'seeing' more than is in AGB product alone
+    * also would benifit from more products (e.g. LAI)
+  * would be good to add 'memory' to algorithm so further evidence of disturbance reinforces initial detection (similar to BLC); would allow reduction in time step and greater sensor fusion
+
+Future Notes:
 * Run just the Assimilation step for simulated disturbance that vary in magnitude, process error, observation error, etc to detect sensitivity
@@
 * Analytical approximation of sequential beta-binomial then KF vs full joint model
-
-## Calibration of VSEM
-*
-
-[@BayesianTools]
-
-## Comparison to real-world data
-
-* LandTrendr tile
-* Construction of frequency and transition matrix priors

### 2021-02-02 Paper.Rmd
Pushed results outline to own file, small updates to discussion and methods

@@
 
-# Results
-
-1. Simulated data experiments (known disturbance)
-  + Single Site, Single Disturbance
-
-Figures showing the current problem, solution, and that we've got a genuine multimodal distribution and P(dist)
-
-  + SSSD Multi Pool
-
-Figures showing pool cov, C accounting / attribution
-
-  + SS Multi-Disturbance w/ PFT switching
-
-Figure showing PFT switching vs not, ability to detect disturbance type
-
-  + Simple Multi-site: spatial cov
-
-Figure showing "dragging down" of neighbors in conventional and breaking of COV in Multinomial
-
-2. Real-world examples
-  + Calibrate state-and-transition prior
-    * data: (samples from) maps of disturbance type by year
-    * model: Multinomial Dirlichet
-
-Figure showing map and estimated priors
-
-  + Applications
-    + Logging
-    + Fire
-    + Land use transition (forest -> non?)
+<!---- RESULTS --->
+```{r child = 'results.Rmd'}
+```
 

### 2021-02-02 discussion.Rmd
two site model. Some text
Pushed results outline to own file, small updates to discussion and methods

@@
 
+* Discussion of similar approaches
+[@Ye2020]
+
 * Implications  (borrow text from CMS proposals)
@@
     * Connect to state-and-transition literature
+    * shared spatial information (e.g. dispersal kernels)
   * Scaling up -> spatially implicit
@@
     * Contagion
+
+* OREGON: FUTURE DIR
+  * disturbance products are 'seeing' more than is in AGB product alone
+    * also would benifit from more products (e.g. LAI)
+  * would be good to add 'memory' to algorithm so further evidence of disturbance reinforces initial detection (similar to BLC); would allow reduction in time step and greater sensor fusion
+
+* Run just the Assimilation step for simulated disturbance that vary in magnitude, process error, observation error, etc to detect sensitivity
+* Approximation of not simulating disturbed ensemble members and relying SOLEY on ensemble adjustment
+* Analytical approximation of sequential beta-binomial then KF vs full joint model

### 2021-02-02 methods.Rmd
Pushed results outline to own file, small updates to discussion and methods

@@
   * Regress vs: pre-disturbance biomass, size of disturbance (abs & rel), type, etc
-
-* RESULTS
-  * needed to reduce obs and process error to work
-  * abs magnitude of disturbance is best predictor of assim prob.
-    * given magnitude, dist type n.s.
-
-* FUTURE DIR
-  * disturbance products are 'seeing' more than is in AGB product alone
-    * also would benifit from more products (e.g. LAI)
-  * would be good to add 'memory' to algorithm so further evidence of disturbance reinforces initial detection (similar to BLC); would allow reduction in time step and greater sensor fusion
-
-Future Notes:
-* Run just the Assimilation step for simulated disturbance that vary in magnitude, process error, observation error, etc to detect sensitivity
-* Approximation of not simulating disturbed ensemble members and relying SOLEY on ensemble adjustment
-* Analytical approximation of sequential beta-binomial then KF vs full joint model

### 2021-02-02 results.Rmd
Pushed results outline to own file, small updates to discussion and methods

# Results

1. Simulated data experiments (known disturbance)
  + Single Site, Single Disturbance

Figures showing the current problem, solution, and that we've got a genuine multimodal distribution and P(dist)

  + SSSD Multi Pool

Figures showing pool cov, C accounting / attribution

  + SS Multi-Disturbance w/ PFT switching

Figure showing PFT switching vs not, ability to detect disturbance type

  + Simple Multi-site: spatial cov

Figure showing "dragging down" of neighbors in conventional and breaking of COV in Multinomial

2. Real-world examples
  + Calibrate state-and-transition prior
    * data: (samples from) maps of disturbance type by year
    * model: Multinomial Dirlichet

Figure showing map and estimated priors

  + Applications
    + Logging
    + Fire
    + Land use transition (forest -> non?)

* RESULTS
  * needed to reduce obs and process error to work
  * abs magnitude of disturbance is best predictor of assim prob.
    * given magnitude, dist type n.s.

### 2021-06-04 methods.Rmd
spent the morning working on the Methods section

@@
 
-The Ensemble Kalman Filter (EnKF) [CITE] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approached have with capturing discrete disturbances and as our jumping off point for developing more flexible methods.
+The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approached have with capturing discrete disturbances and as our jumping off point for developing more flexible methods.
 Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabalistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g.
@@
 If we were making a spatial forecast at $n_s$ locations, $\mu_{f,t}$ would be a vector of length $3 n_s$ and $P_{f,t}$ would be a $3 n_s \times 3 n_s$ covariance matrix that would capture the full set of correlations across both variables and space.
-An important difference between data assimilation and spatial statistics is that in data assimilation the spatial covariance matrix is coming from the mechanistic understanding embedded in our process model, and the spatial covariance in the forecast inputs (initial condition, parameters, drivers, process error), rather than an empirical semivariogram.
+An important difference between data assimilation and spatial statistics is that in data assimilation the spatial covariance matrix comes from (1) the mechanistic understanding embedded in our process model and (2) the spatial covariance in the forecast inputs (initial conditions, parameters, drivers, process error), rather than an empirical semivariogram.
 
-The Analysis step in the EnKF is the same as that in the standard Kalman Filter
+The Analysis step in the EnKF is the same as that in the standard Kalman Filter.
+Because the forecast step from t-1 to t occurs prior to the observed data at time t, $Y_t$, the forecast of our three state variables is used as an informative _prior_ distribution and is assumed to follow a multivariate Normal (MVN) distribution with mean $\mu_{f,t}$ and covariance $P_{f,t}$.
+The EnKF also acknowledges that the observed data at time t, $Y_t$, is measured imperfectly.
+Observation error is assumed to follow a MVN distribution with an observation error covariance matrix $R$.
+Furthermore, it is assumed that $R$ is known, which is not unreasonable as many derived data products and sensors report uncertainties, and uncertainties from field data can often be inferred from sampling theory or Monte Carlo resampling / bootstrapping.
+In many cases new data do not exist for every state variable (e.g.
+we may observe leaves and wood but not soil C) and so an observation operator matrix, $H$ is also introduced, which in most cases is just a $n \times m$ indicator matrix matching the observed $Y$'s (n=2 in this example) with the vector of state variables (m=3 in this example).
+Indeed, one of the powerful features of data assimilation is its ability to update unobserved state variables based on the covariances between observed and unobserved variables.
+Putting this all together, Bayes' Theorem can then be used to update our MVN forecast prior, given a MVN Likelihood of $Y_t$ with covariance $R$, giving an analytically-tractable conjugate MVN posterior with mean $\mu_{a,t}$ and covariance $P_{a,t}$:
 
-1. Illustration of problem
+$$P_{a,t}^{-1} = H_t^T R_t^{-1} H_t + P_{f,t}^{-1}$$
+$$\mu_{a,t} =  P_{a,t}^{-1}\left( H_t^T R^{-1} Y_{t} + P_{f,t}^{-1} \mu_{f,t} \right) $$
 
-```{r, echo=FALSE}
+The subscript $a$ is used to refer to the Analysis posterior and distinguish this from the prior forecast $f$ mean and covariance for the same timestep.
+While the matrix notation in these equations can make these solutions look complicated, the underlying meaning is accessible.
+First, the precision of the analysis (i.e.
+1/variance) is the sum of the observation precision and the forecast precision, meaning that the analysis is always more precise than the data or forecast alone.
+Second, the mean of the analysis is a weighted average between the data, $Y_t$ and the forecast $mu_{f,t}$, each weighted by their precision.
+This averaging can also be viewed as a 'nudge', pushing the forecast toward the new data, with the magnitude of that nudge determined by the uncertainties in both the forecast and data (i.e.
+highly precise data = big nudge, highly uncertain data = very small nudge).
+
+While this fusion of forecast and new data is an incredibly powerful concept, as alluded to in the _Introduction_ there are times that the Kalman Normal-Normal "nudge" produces results that don't always make sense.
+Consider as a simple example a simulated data experiment where LAI growth follows a simple stochastic logistic model with growth rate $r$=0.1, equilibrium $K$=6, process error standard deviation 0.25, and observation error standard deviation 0.75.
+Under normal conditions, EnKF would have no trouble nudging a LAI forecast back on track every time a new observation is made, but what if a discrete disturbance event occurs knocking LAI from steady-state down to $1.0 \pm 0.5$?
+In this case the compromise between the forecast ($LAI=6$) and observations ($LAI=1$) produces a compromise ($LAI \approx 5$ that is a poor representation of the state of the system (Figure \@ref(fig:enkf)).
+Furthermore, as the true LAI slowly recovers upward, the EnKF keeps gradually nudging LAI downward until the two converge.
+Not only is the assimilated state wrong, but so is our estimate of the process -- a discrete disturbance and rapid recovery is behaving like gradual degradation, which will lead to incorrect inferences about related processes (litter, soil C, NPP, NEE, ET, albedo, etc.).
+Not also that including a disturbance probability in the process model does not solve this problem.
+Assuming the disturbance probability is low (e.g.
+a background rates of <1%), at every point in time the forecast would include a small number of disturbed ensemble members, but the overwhelming majority would be undisturbed.
+Therefore $\mu_{f,t}$ will be relatively unchanged and $P_{f,t}$ would see a small increase in variance.
+The fundamental problem is that even though the true forecast is now bimodal (a large peak at the undisturbed state and a much smaller peak at the disturbed state), the EnKF is still representing this as a unimodal Normal.
+Indeed, if the you increased the disturbance probability substantially (e.g.
+50%), the EnKF would treat this as a unimodal forecast half way between the disturbed and undisturbed states, a place where no ensemble members are present.
+
+```{r enkf, echo=FALSE, fig.cap="Example assimilation of a disturbance event shows that conventional assimilation approaches, such as EnKF, take time to converge to new states after disturbance (orange).
+By modeling the probability of disturbance (black) our Multinomial assimilation (blue) can correctly jump between states during disturbance events (e.g.
+P(disturbance) = 0.977 at time 10) while remaining resilient to potential false-positive events (e.g.
+time 12 and 15)."}
 SSSD_figure <- function(){
-  load("SSSD.RData")
+  load("SSSD.RData")      ## simulation from SingleSite_SingleDisturbance.Rmd
   plot(data$Y,ylim = range(rbind(CI.fN,data$Y)),xlim=xlim,type='n',
@@
   polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
-    an0.stats[1, 1])), border = NA, col = '#FFBF6C')
-  polygon(cbind(c(time, rev(time), 1), c(an.stats[2, ], rev(an.stats[4, ]),
-    an.stats[1, 1])), border = NA, col = ecoforecastR::col.alpha('#b2df8a',0.7))
-  #ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
-  ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
-
-  lines(an0.stats[3,],col="orange",lwd=2)
-  lines(an.stats[3,],col="#b2df8a",lwd=2)
-  #lines(CI.fSSSD[2,],col="blue",lwd=2)
-  lines(CI.fN[3,],col="blue",lwd=2)
-  lines(Dbar,lwd=2)
-  points(data$Y,pch=18,col=2)
-  legend("topright",legend=c("EnKF, no disturb","EnKF, disturb","Multinomial","data","P(dist)"),
-       lwd=5,col=c("orange","#b2df8a","blue","red","black"))
+    an0.stats[1, 1])), border = NA, col = '#FFBF6C')   ## EnKF interval
+  ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256)) ## Multinomial interval
+  lines(an0.stats[3,],col="orange",lwd=2)   ## EnKF median
+  lines(CI.fN[3,],col="blue",lwd=2)         ## Multinomial median
+  lines(Dbar,lwd=2)                         ## Multinomial disturbance rate
+  points(data$Y,pch=18,col=2)               ## data
+  legend("topright",legend=c("EnKF","Multinomial","data","P(dist)"),
+       lwd=5,col=c("orange","blue","red","black")) ##"#b2df8a"
 }
@@
 
-## Derivation of Multinomial Filter
+## Derivation of the Multinomial Filter
 
-*Priors*
-  $$X^L_k \sim MVN(\vec{\mu}_{f,k},P_{f,k})$$
-  $$\vec{\rho} \sim Multinomial(1,\vec{p})$$
+To address this fundamental flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodel, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges.
+To begin, let's use $X$ to represent the latent (unobserved) "true" state of the system.
+Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent continuous state variables of interest, which we'll refer to as $X^L_{k,t}$.
+Similar to the EnKF we'll assume that the ensemble prediction for any one of these categories can be represented as a multivariate Normal based on the ensemble sample mean and covariance:
 
-*Process Model*
-  $$X = X^L \vec{\rho}$$
+$$X^L_{k,t} \sim MVN(\vec{\mu}_{f,k,t},P_{f,k,t})$$
 
-*Likelihood*
-  $$Y \sim MVN(X,R)$$
+However, because we have $k$ categories we end up with a multimodel prediction.
+Furthermore, these modes need not be distinct; it is perfectly fine in concept for these MVNs to be overlapping.
+In practice they more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabalistically.
+Indeed, to be able to represent our multimodel prediction as a single probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.
 
-**ADD Conceptual figure**
+In practice, an individual location can only be in one of these $k$ states, which we'll model using a Multinomial distribution with a sample size of 1.
+
+$$\vec{\rho_t} \sim Multinomial(1,\vec{p_t})$$
+
+This is a generalization of the Bernoulli distribution (Binomial with n=1) and for any individual draw from the Multinomial $\vec{\rho_t}$ will be a vector of length $k$ with a 1 in the category being assigned and a 0 for all other categories.
+
+Within the analysis step we will retain the assumption that the observed data $Y_t$ follow a multivariate Normal distribution around the true latent $X_t$ with observation error $R_t$
+
+$$Y_t \sim MVN(X_t,R_t)$$
+
+We then model $X_t$ as the product of the matrix $X^L_t$ and the vector $\vec{\rho_t}$
+
+$$X_t = X^L_t \vec{\rho_t}$$
+which for any individual draw of $\vec{\rho_t}$ assigns $X_t$ that specific $X^L_{k,t}$.
+In practice, we will use Bayes' Theorem to integrate over the uncertainties in both our categorical assignment and continuous state variables, producing posterior Analysis probability distributions for both $X_t$ and $\vec{\rho_t}$.
+Unlike in the EnKF, where conjugacy allows us to calculate an analytical solution for the posterior, there is not simple analytical solution for the joint posterior distribution of $X_t$ and $\vec{\rho_t}$.
+In this case we will use well-developed Bayesian Markov Chain Monte Carlo (MCMC) numerical algorithms to sample from the posterior distribution.
+Specifically, numerical analyses in this paper were conducted using JAGS v4.2.0 [@Plummer03] accessed from R v4.0.3 [@Rcore] using the rjags package v4.10 [@rjags].
 
@@
 
-[@Anderson2001]
+When making ensemble forecasts using the Analysis from the traditional EnKF, one needs to sample new initial conditions from the Analysis joint posterior distribution.
+This work fine if the initial conditions are complete (100% of the model variables represented) and when there is no covariance with any of the other uncertainties being propagated (e.g.
+sampling ensemble drivers or posterior parameter distributions).
+In practice, this assumption is frequently violated, for example an above average ensemble member might be above average not just because of its initial condition, but also because it had favorable parameters or drivers, creating a complex covariance between states, parameters, and drivers that would be destroyed by randomly resampling initial conditions.
+Because of this Anderson et al.
+[-@Anderson2001] developed an alternative know as _ensemble adjustment_ that nudges individual ensemble members toward the posterior analysis mean and covariance.
+Mathematically, the approach is essentially the multivariate generalization of using the prior forecast mean and variance to calculate a Z-score for every ensemble member (i.e.
+express as an standard Normal anomaly with mean 0 and standard deviation 1), and then apply that Z-score to the new posterior analysis mean and variance.
+In our Multinomial filter it is important to generalize this concept of ensemble adjustment to preserve covariances when updating the discrete categorical assignments of individual ensemble members.
+
+when individual ensemble members are moved fro
 

### 2021-06-05 methods.Rmd
more methods: ensemble adjustment, table of simulated experiments, text for first two.

@@
 express as an standard Normal anomaly with mean 0 and standard deviation 1), and then apply that Z-score to the new posterior analysis mean and variance.
-In our Multinomial filter it is important to generalize this concept of ensemble adjustment to preserve covariances when updating the discrete categorical assignments of individual ensemble members.
 
-when individual ensemble members are moved fro
+In our Multinomial filter it is important to generalize this concept of ensemble adjustment to both discretely move ensemble members from one category to another, and then to preserve the covariances among state variables, and with drivers and parameters, when updating the discrete categorical assignments of individual ensemble members.
+To reassign forecasted ensemble members to new disturbance classes we first identify the "sink" categories that increased in probability between the prior, $p_{k,t}$, and the posterior, $\bar{\rho}_{k,t}$.
 
-*reassignment of classes*
+$$\delta p_{k,t} = max\{0,\bar{\rho}_{k,t}-p_{k,t} \}$$
 
-To reassign forecasted ensemble members to new disturbance classes we first identify those that decreased between the prior probability, $p_k$, and the posterior frequency, $\bar{\rho}_k$.
+For "source" ensemble members that belong to categories that decreased in relative abundance, the excess is reassigned to new classes according to a Multinomial draw.
 
-$$\delta p_k = max\{0,\bar{\rho}_k-p_k \}$$
+$$c_{A|k} = Multinomial\left( n_k, {\vec{\delta p_t}}\over{\sum \vec{\delta p_t}}  \right)$$
 
-Classes that decreased in relative abundance are reassigned to new classes according to a Multinomial draw
+Conditional on this reassignment, the continuous states are updated according to [@Anderson2001] using the posterior statistics for that $X^L_{k,t}$.
+It is also worth noting that some cases of class reassignment may also require keeping track of categorical switches in Plant Functional Type (PFT), such as a transition from forest to agriculture, which implies a large jump in PFT parameters as well.
+One could apply the same [@Anderson2001] style adjustment to PFT parameters, but for simplicity here we just maintained alternative parameter ensembles for each PFT.
 
-$$c_{A|k} = Multinomial\left( n_k, {{\delta p}\over{\sum \delta p}}  \right)$$
+## Simulated data experiments
 
-## Simulated data experiments (known disturbance)
+To test our algorithm we applied it to a series of increasingly complex simulated data experiments where we know the "true" state of the system, including the timing and category of disturbance events.
+The first of these is the logistic LAI model described previously in the EnKF section (single pool, single site, single disturbance.
 
-  + x Single Site, Single Disturbance (SSSD)
-      vs. EnKF with and without Disturbance
-  + x SSSD Multi Pool (VSEM)
+```{r pseudoTable,echo=FALSE}
+## table summarizing simulated data experiments
+exps = data.frame(exp=1:5,
+                  pools= c(1,3,3,3,3),
+                  disturbances = c(1,1,2,2,1),
+                  PFTs = c(1,1,1,2,1),
+                  sites = c(1,1,1,1,2)
+                  )
+knitr::kable(exps,caption="Simulated data experiments")
+```
+
+For the second experiment, we used the VSEM to extend the Multinomial assimilation model to a single site, single disturbance example with multiple pools.
+In this and the following simulated data experiments we used VSEM's default parameters, drawing ensemble members from Normal distributions centered on the prior mean and assuming a standard deviation that was 1/10th the default uniform prior range.
+The goal here was not to replicate any specific system, but to provide an amount of parameter uncertainty that was realistic but not so large as to generate trivial ensemble forecasts.
+Model runs were for 25 years using VSEM's default simulated PAR as the driver and with data assimilation occurring annually.
+Model initial conditions were generated by an initial 25 year ensemble spin-up prior to the 25 year experiment.
+Process error was added to the simulation annually assuming a zero-bound tobit distribution with a process error covariance matrix that was diagonal with variances of 0.1 for all pools.
+Ensemble members experienced disturbance stochastically with a 5% annual probability.
+For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).
+
+Pseudodata for assimilation was generated by selecting the parameters and initial conditions from one ensemble member at random and running a forward simulation with an imposed disturbance event in year 10.
+Observation error was then randomly added to the model's predictions using a tobit distribution with the same known observation error matrix, R, as used in the assimilation, which was a diagonal covariance matrix with variances of 0.25, 10, and 1 for leaf, soil, and wood respectively.
+Within the assimilation the prior on the disturbance rate was 5%.
+
+Our third experiment was similar to the second but introduced a second disturbance type (single-site, multi-disturbance, multi-pool) ....
+
+Our forth experiment extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT....
+
+Our fifth and final experiment extended the analysis to a two-site scenario with spatial covariance across sites and pools, but with a disturbance occurring in one site but not at the other.
+To keep things simpler, this experiment reverted to a single disturbance type.
+This scenario was designed to test a concern we had in previous research that a disturbance occurring at one site could cause the biomass to decline at other undisturbed sites because of the cross-site covariances [CITE Viskari, Hamze]....
+
   + x SS Multi-Disturbance - distinct

### 2021-06-05 methods.Rmd
methods: finished simulated data text

@@
 It is also worth noting that some cases of class reassignment may also require keeping track of categorical switches in Plant Functional Type (PFT), such as a transition from forest to agriculture, which implies a large jump in PFT parameters as well.
-One could apply the same [@Anderson2001] style adjustment to PFT parameters, but for simplicity here we just maintained alternative parameter ensembles for each PFT.
+One could apply the same [@Anderson2001] style adjustment to PFT parameters, but for simplicity here we just maintained alternative parameter ensembles for each PFT, with each ensemble member constructed with the identical anomalies around each PFT's mean..
 
@@
 Ensemble members experienced disturbance stochastically with a 5% annual probability.
-For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).
+For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 [UNITS?] for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).
 
@@
 
-Our third experiment was similar to the second but introduced a second disturbance type (single-site, multi-disturbance, multi-pool) ....
+Our third experiment was similar to the second but introduced a second disturbance type (single-site, multi-disturbance, multi-pool), both with a 5% probability.
+For both disturbances leaf biomass was reduced to 10% of their pre-disturbance values with a standard deviation of 0.15; in the first disturbance type wood biomass was also reduced to 10% with a standard deviation of 0.25, similar to a stand-clearing disturbance, while in the second 95% of the original wood biomass was retained, with a standard deviation of 0.01, similar to a defoliation event.
+The prescribed disturbance in the pseudodata was the first type.
+In this scenario, the Multinomial assimilation needs to leverage the wood data to distinguish between disturbance types because the leaf biomass reduction would be identical.
 
-Our forth experiment extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT....
+Our forth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT.
+Specifically PFT 2 was identical to the VSEM default, but PFT 1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT.
+In addition, we established a transition matrix where disturbance 1 (stand clearing) transitions from PFT1 to PFT2, while disturbance 2 (defoliation) keeps the current PFT is.
+As with the previous scenario, the prescribed disturbance in the pseudodata was type 1 (stand clearing, transition from forest to shrub).
 
 Our fifth and final experiment extended the analysis to a two-site scenario with spatial covariance across sites and pools, but with a disturbance occurring in one site but not at the other.
-To keep things simpler, this experiment reverted to a single disturbance type.
-This scenario was designed to test a concern we had in previous research that a disturbance occurring at one site could cause the biomass to decline at other undisturbed sites because of the cross-site covariances [CITE Viskari, Hamze]....
+To keep things simpler, this experiment reverted to a single disturbance type, same as experiment 2, except we dropped the disturbance probability to 2.5%.
+This scenario was designed to test a concern we had in previous research that a disturbance occurring at one site could cause the biomass to decline at other undisturbed sites because of the cross-site covariances [CITE Viskari, Hamze].
 

### 2021-06-06 discussion.Rmd
Methods: starting text on RS data & calibration

@@
 * Analytical approximation of sequential beta-binomial then KF vs full joint model
+
+Leftover text from Methods:
+  + x SS Multi-Disturbance - distinct
+  + x SS Multi-Disturbance w/ distinct, PFT switching
+      * mention parallel PFT ensembles
+      * need to protect against long-term PFT leakage; related to more general need to "look back"
+  + x Simple Multi-site: spatial cov
+  + Future
+      * SSMD - initially overlapping but diverge
+        * how to handle updating of assignment based on new information?
+      * SSMD - transition probabilities change with state
+        * rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?

### 2021-06-06 methods.Rmd
Methods: starting text on RS data & calibration

@@
 
-  + x SS Multi-Disturbance - distinct
-  + x SS Multi-Disturbance w/ distinct, PFT switching
-      * mention parallel PFT ensembles
-      * need to protect against long-term PFT leakage; related to more general need to "look back"
-  + x Simple Multi-site: spatial cov
-  + Future
-      * SSMD - initially overlapping but diverge
-        * how to handle updating of assignment based on new information?
-      * SSMD - transition probabilities change with state
-        * rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
+## Remotely-sensed disturbance in Oregon
 
-## Real-world examples
-
-Extract some example time series for single pixels (e.g.
-AGB product) where the disturbance (2004) it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance).
+Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades.
+In summary, we: (1) calibrated VSEM against field data, then (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transition, then (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types.
+We then assessed the ability of the algorithm to detect different disturbances types and magnitudes.
 
@@
 
-    + Landtrendr
-    http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html
-    + Location:
-    tile h03v04
-xmin       : -2115585 
-xmax       : -1965585 
-ymin       : 2564805 
-ymax       : 2714805 
-Upper Left  (-2115585.000, 2714805.000) (123d 6'32.29"W, 44d41'17.09"N)
-Lower Left  (-2115585.000, 2564805.000) (122d35' 7.51"W, 43d23'46.18"N)
-Upper Right (-1965585.000, 2714805.000) (121d16'45.91"W, 45d 3'17.45"N)
-Lower Right (-1965585.000, 2564805.000) (120d47'15.95"W, 43d45'18.18"N)
-Center      (-2040585.000, 2639805.000) (121d56'25.42"W, 44d13'35.07"N)
-> res(cover)
-[1] 30 30
-> crs(cover)
-CRS arguments:
- +proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83
-+units=m +no_defs 
-    + Biomass [@Kennedy2018]
-      1990-2017
-    + Cover
-      1990-2017
-    + Disturbance
-      1984-2011
-      - nearest neighbor resampled to same as cover
+Our study region was a Landsat tile located between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude.
+For this tile we extracted data from three derived Landtrendr datasets (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html, tile h03v04):  biomass, cover, and disturbance.
+All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover.
+The biomass product ...
+[@Kennedy2018].
+The cover product ...
+[CITE].
+To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of ponderosa pine.
+The disturbance product ...
+[CITE].
+To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.
+
+```{r map, echo=FALSE}
+## maps of pre-disturbance biomass, cover, and 2004 disturbance
+
+```
 
@@
 
+VSEM was calibrated primarily using data from 2002-2019 from the Ameriflux site US-ME2 (lat.lon), a mature Ponderosa pine stand.
+VSEM pools (leaf, stem, and soil C) were initialized using the earliest available site-specific data from the Ameriflux BADM (Biomass, Ancillary, Disturbance, Management) dataset.
+
 [@BayesianTools]
@@
 * Calibrate at US-Me2, 2002-2019 
-* Initialize with BADM soil, leaf, and stem C
 * Priors (and some fixed parameters) using BADM; default uniform for rest
@@
 ### Comparison to real-world data
+
+Extract some example time series for single pixels (e.g.
+AGB product) where the disturbance (2004) it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance).
 

### 2021-06-10 Paper.Rmd
spent the morning working on the Methods section
model output for EnKF figure; config to render paper to PDF

@@
 output:
+  pdf_document: 
+    fig_caption: yes
+    number_sections: yes
+    toc: yes
+    keep_tex: true
   html_document: 
@@
 
-# Outline
-
-# Abstract
+# Abstract {-}
 

### 2021-06-14 SingleSite_MultiDisturb_MultiPool.Rmd
start of Results: maps and initial figures

@@
 
-* Multinomial with distinct disturbances
 * Multinomial with distinct disturbances that are different PFTs
@@
 ```{r}
-ne = 200     ## number of ensemble members
+ne = 500     ## number of ensemble members
 NT = 25      ## number of years
@@
                 0.1,0.95),2,2)    ## mean leaf & stem biomass after disturbance as % of original
-V1  <- diag(c(0.15,0.25))^2       ## var in post disturbance leaf and stem biomass (one for each type)
+V1  <- diag(c(0.15,0.15))^2       ## var in post disturbance leaf and stem biomass (one for each type)
 V2  <- diag(c(0.15,0.01))^2
@@
 Ntrue = matrix(NA,NT,3)                ## storage for latent "true" state
-true = 42                              ## reference ensemble member
+true = 43                              ## reference ensemble member
 D.true = rep(0,NT)                     ## true disturbance timeseries
@@
   ## priors
-  priors <- list(p   = rowMeans(d.prior),         ## disturbance rate
+  priors <- list(p   = p.dist, #rowMeans(d.prior),         ## disturbance rate
                  muf = muf,    ## forecast means
@@
 
+  ## initial conditions: disturbance likelihood
+  Lp = NA
+  for(i in 1:3){ Lp[i] = mvtnorm::dmvnorm(No[t,],muf[,i],pf[,,i])}
+  rho.IC = rep(0,3); rho.IC[which.max(Lp)] = 1; rho.IC[1] = NA
+
   ## Analysis
@@
                            data=update,
-                           n.adapt=1000,n.chains=3)
-  jdat <- rjags::coda.samples(mod,n.iter=2000) ## burn-in 
+                           inits = list(rho=rho.IC),
+                           n.adapt=3000,n.chains=3)
+  jdat <- rjags::coda.samples(mod,variable.names = "rho",n.iter=2000) ## burn-in 
+  plot(jdat)
   jdat <- rjags::coda.samples(mod,variable.names=c("X","rho","x"),n.iter=10000)
@@
 
+```
+
+```{r}
+save(Forecast,Analysis,No,Ntrue,R,file="exp3.RData")
 ```

### 2021-06-14 StateTransition.Rmd
start of Results: maps and initial figures

@@
 
+## extract year for paper figure and recode to NLCD
+cov04 = covS[[which(ctime==2004)]]
+cov04 = ratify(cov04)
+rat <- levels(cov04)[[1]]
+nlcd <- FedData::pal_nlcd()
+colnames(nlcd)[2] <- "ID"
+nlcd$ID = as.numeric(nlcd$ID)
+rat = dplyr::left_join(rat,nlcd,by="ID")
+levels(cov04) <- rat 
+cov04 = deratify(cov04,"description")
+cov04 <- raster::projectRaster(cov04,crs=crs("+proj=longlat +datum=WGS84"))
+raster::writeRaster(cov04,"cov04.tif",format="GTiff",overwrite=TRUE)
+save(rat,file="rat.RData")
+
 # disturbance
@@
 #summary(biomass)
+
+bio04 = bioS[[which(ctime==2004)]]
+raster::writeRaster(bio04,"bio04.tif",format="GTiff",overwrite=TRUE)
+
+raster::plot(bio04,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main="Biomass")
+
 ```
@@
 save(disR,file=dis.processed)
+raster::writeRaster(disR,"dis.processed.tif",format="GTiff",overwrite=TRUE)
 
@@
   save(disR,fdS,file=dis.resamp)
+
+## save year for figure
+dis04 = disR[[which(dtime==2004)]]
+#dis04 = ratify(dis04)
+#drat <- levels(dis04)[[1]]
+#drat$class <- c("Intact","Other","Cut","Fire","Pest")
+#levels(dis04) <- drat 
+#dis04 = deratify(dis04,"class")
+#dis04 <- raster::projectRaster(dis04,crs=crs("+proj=longlat +datum=WGS84"))
+raster::writeRaster(dis04,"dis04.tif",format="GTiff",overwrite=TRUE)
+
+dname = c("Intact","Other","Cut","Fire","Pest")
+raster::plot(dis04,col=c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a'),axis.args=list(at=seq(3,48,length=5),labels=dname,cex.axis=0.4),asp=1,main="Disturbance") #"#a6cee3" '#33a02c'
 

### 2021-06-14 methods.Rmd
start of Results: maps and initial figures

@@
 
-Our study region was a Landsat tile located between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude.
+Our study region was a Landsat tile over central Oregon, USA located between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude.
 For this tile we extracted data from three derived Landtrendr datasets (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html, tile h03v04):  biomass, cover, and disturbance.
 All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover.
-The biomass product ...
-[@Kennedy2018].
-The cover product ...
-[CITE].
-To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of ponderosa pine.
-The disturbance product ...
-[CITE].
+The biomass product calibrated Landtrendr-processed Landsat data against field plot data using the gradient nearest neighbor (GNN) methodology to produce an annual biomass estimates [@Kennedy2018].
+The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S.
+National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018].
+To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_).
+The disturbance product **need text and citation from Robert** [CITE].
 To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.
@@
 
+## Cover
+if(!exists("cov04")) cov04 = raster::raster("cov04.tif")
+load("rat.RData")
+raster::plot(cov04,col=rat$color,axis.args=list(at=seq_along(rat$description),labels=rat$description,cex.axis=0.4),asp=1,main="Land Cover")
+
+## Disturbance
+if(!exists("dis04")) dis04 = raster::raster("dis04.tif")
+dname = c("Intact","Other","Cut","Fire","Pest")
+raster::plot(dis04,col=c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a'),axis.args=list(at=seq(3,48,length=5),labels=dname,cex.axis=0.4),asp=1,main="Disturbance") #"#a6cee3" '#33a02c'
+
+## Biomass
+if(!exists("bio04")) bio04 = raster::raster("bio04.tif")
+raster::plot(bio04,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main="Biomass")
+
 ```
-
- + Applications: PNW
-    + Logging
-    + Fire
-    + Land use transition (forest -> non?)
 
@@
 
-* recoded to a smaller set of disturbances: none, fire, pest, clearcut, other
-* Calculate frequencies of different disturbance types
-* Calculate disturbance-by-landCover matrix of transition probabilities
-* Use these as disturbance priors
+Our Multinomial disturbance assimilation model takes as priors the disturbance probability and the post-disturbance pool sizes (mean and precision).
+The algorithm also allows the user to specify a transition matrix for pre- and post-disturbance changes in vegetation type.
+For some ecosystem models, these quantities may be generated dynamically by the model itself as part of the ensemble forecast.
+Because VSEM does not internally predict disturbances we instead generated these priors using the Landtrendr disturbance product.
+The disturbance probability prior was calculated simply as the frequency of each disturbance type over the whole time period and the subset of spatial domain that was classified as evergreen forest.
+The post-disturbance pool size and transition matrix priors were similarly estimated using the remotely sensed biomass and cover estimates (respectively), this time focusing on a sample of 100 disturbance transitions per year for the years 1990-2003 because of the greater computational cost of this estimation process.
+Because of the potential lags involved in when a disturbance occurred within a year, we considered an evergreen forest disturbance to be a disturbed pixel that was evergreen forest in the year of disturbance or the prior year.
+Similarly, we considered the pixel to transition out of forest if the cover type in the year of disturbance or the year following disturbance to be non-forest, but filtered out disturbances where that transition only lasted one year before returning to forest (i.e.
+these were considered forest-to-forest transitions in the matrix, as were pixels that stayed forest the whole time).
+Similarly, in order to assess the change in biomass associated with disturbance we calculated the maximum biomass in the three years prior to disturbance and the minimum biomass in either the year of the disturbance or the two years after.
+For all disturbance types we calculated the mean and standard deviation of the change in biomass both in absolute terms and as a proportional change.
 
@@
 VSEM was calibrated primarily using data from 2002-2019 from the Ameriflux site US-ME2 (lat.lon), a mature Ponderosa pine stand.
-VSEM pools (leaf, stem, and soil C) were initialized using the earliest available site-specific data from the Ameriflux BADM (Biomass, Ancillary, Disturbance, Management) dataset.
+VSEM pools (leaf, wood, and soil C) were initialized using the earliest available site-specific data from the Ameriflux BADM (Biomass, Ancillary, Disturbance, Management) dataset.
+An additional ten years of leaf and wood biomass data were also available from the BADM, which we included as part of the calibration dataset.
+Also included was the cumulative annual tower NEE (2002-2019).
+Because we were working with annual data we assumed Normal likelihoods for all three response variables, and because our priority was the correct prediction of aboveground biomass we upweighted the wood likelihood by 4 and downweighted the NEE likelihood by 1/4.
 
-[@BayesianTools]
+Prior to calibration we set the leaf area ratio (LAR) to 15.7 m2/kgC and leaf residence time (tauL) to 4.7 years based on data from the BADM and left the light extinction coefficient (KEXT) at its default (0.5).
+The remaining fit parameters were the light use efficiency (LUE), the fraction of GPP allocated to autotrophic respiration (GAMMA), the fraction of NPP allocated to leaves (Av), the wood and soil C residence times (tauW, tauS), and the residual standard deviations for each of our data constraints (sd_NEE, sd_leaf, sd_wood).
+The fit was done taking a Maximum Likelihood approach using base R's `optim` function.
+Finally, because the process error within a dynamical data assimilation system is not identical to the calibration residual error, we estimated the leaf and wood process errors (Ql, Qw) based on the difference between the predicted and observed net annual changes in each of these pools.
+Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state.
 
-* Calibrate at US-Me2, 2002-2019 
-* Priors (and some fixed parameters) using BADM; default uniform for rest
-* Likelihood against tower _annual_ NEE and BADM leaf and stem biomass
-* MCMC using DEzs algorithm for 25000 iterations of 2 chains
-* GBR, effective sample
+```{r paramTable,echo=FALSE}
+load("fit.RData")
+varName = c(names(theta),"Ql","Qw","Qs")
+varName[c(5,7)] = c("tauL","tauW")
+paramTable = data.frame(varName,fit=format(c(theta,Ql,Qw,Qs),digits=3,scientific=TRUE))
+knitr::kable(paramTable, caption = "VSEM parameter estimates")
+```
 
-### Comparison to real-world data
+### Assimilation of remotely-sensed data
 
-Extract some example time series for single pixels (e.g.
-AGB product) where the disturbance (2004) it is far enough in the past that there’s an interesting recovery trajectory but also where it’s not so early (so that the data assimilation has time to come into steady-state before the disturbance).
+To assess the performance of the Multinomial assimilation algorithm against real-world data, we sampled ~100 evergreen forest points for each disturbance class.
+We specifically sampled disturbance locations from the year 2004, which is sufficiently deep in the past that there is an observable disturbance recovery trajectory, but with a sufficiently long time-series prior to disturbance for the data assimilation to come into steady-state with the biomass data product.
+For each selected location we extracted a time-series from the remotely-sensed biomass product for use as the data constraint within the assimilation algorithm.
+The cover and disturbance products were not assimilated, but were just used to construct priors (see above) and assess performance.
 
-* Selected conifer forest land cover type
-* Sampled ~100 points from each disturbance type for 2004
-* Daymet PAR as driver
-* IC
-  * ensemble spin up
-  * construct disturbance recovery trajectory
-  * Extract rs map AGB
-  * Match leaf and soil C from spin-up trajectory
+For each location we next extracted PAR driver data from the Daymet product [CITE] using the daymetr R package [@daymetr].
+For each location the carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale.
+To start we used the first time point in each biomass time-series as the initial condition for the wood pool.
+Next, using the US-ME2 initial conditions we performed a >10k year spin-up, and then sampled carbon pool ensemble members from the second half of this time-series.
+Next, for each ensemble member we imposed at disturbance that reduced wood and leaf biomass to 1% of their equilibrum values, then simulated a 200-year recovery trajectory.
+Finally, for each ensemble member we selected the stand age that minimized the squared error between the modeled biomass and the initial biomass, and then selected the leaf and soil C pools at that stand age.
+This results is an ensemble of leaf, wood, and soil carbon pools with a covariance structure that is consistent with the calibrated model's dynamics.
 
-* Experiments
-  * run SSSDMP model for each site independently
-  * run other versions varying R, Q, and the disturbance probability
+For the assimilation experiment, each site was assimilated separately.
+In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^2$ based on [@Kennedy2018] Figure 2.
+Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017).
+The difference between the predicted and observed biomass at the end of the assimilation was taken as a metric of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations.
+**ONCE RESULTS ARE FINALIZED, COME BACK AND FILL IN WHICH SPECIFIC ANALYSES WERE USED**.
+
+Finally, based on preliminary results we performed a series of additional assimilation experiments where we manually altered the observation error, R, or process error, Q, to assess the sensitivity of our results.
+Table (\@ref(table:QRexp)) lists the multipliers that were used to adjust R and Q.
+The skill of these assimilation experiments were assessed the same as the original.
+In addition, changes in skill across experiments were assessed as a function of R and Q.
+
+```{r QRexp,echo=FALSE}
+## summary table of experiments
+load("QRexp.RData")
+QRexp = data.frame(experiment=1:6,Qmult=Qmult[Qs],Rmult=Rmult[Rs])
+knitr::kable(QRexp,caption="Data Assimilation experiments")
+```
+
   * Score: filter divergence, disturbance detection & skill

### 2021-06-14 multi.ens.adj.R

@@ -74,7 +74,7 @@ multi.ens.adj<-function(Xf, cf, mu.f, Pf, Xa, ca){
   
   Z <- Xf*0
   
-  for(k in seq_along(uc)){
+  for(k in seq_along(uc)){   ## loop over disturbance classes
     sel.c = which(cf == uc[k])
     
     ## SVD of forecast covariances
@@ -105,15 +105,15 @@ multi.ens.adj<-function(Xf, cf, mu.f, Pf, Xa, ca){
     if(length(sel.c) == 0) next
     
     ## analysis (by class)
-    mu.a <- colMeans(Xa[ca == uc[k],])
-    Pa   <- cov(Xa[ca == uc[k],]) ## calculate analysis cov for each class
+    mu.a <- colMeans(Xa[sel.c,])
+    Pa   <- cov(Xa[sel.c,]) ## calculate analysis cov for each class
     S_a  <- svd(Pa)
     L_a  <- S_a$d
     V_a  <- S_a$v
   
     ## analysis ensemble 
     for(i in sel.c){
-      # she decomposed Pa - then it's putting it back together but with a different Z which comes from the likelihood of that ens    
+      # we decomposed Pa - then it's putting it back together but with a different Z which comes from the likelihood of that ens    
       X_a[i,] <- V_a %*%diag(sqrt(L_a))%*%Z[i,] + mu.a
     }
     

### 2021-06-14 SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
Two Site, single disturbance, multi-pool
start of Results: maps and initial figures

@@
 
-* Multinomial with distinct disturbances that are different PFTs
 * Multinomial with initially-overlapping disturbances that then diverge
@@
                            data=update,
+                           inits = list(rho=rho.IC),
                            n.adapt=1000,n.chains=3)
@@
 
+```
+
+```{r}
+save(Forecast,Analysis,No,Ntrue,R,file="exp4.RData")
 ```

### 2021-06-14 discussion.Rmd
start of Results: maps and initial figures

@@
 [@Ye2020]
+
+* Calibration was "quick and dirty" because the focus was on disturbance detection not default dynamics.
+For actual carbon monitoring and forecasting we would recommend a more extensive multi-site Bayesian calibration so that parameter uncertainty can be propagated [CITE istem 2018, dietze 2017], preferably employing hierarchical approaches to also capture parameter heterogeneity [CITE istem 2021].
+Similarly, would recommend a more robust estimation and propagation of process error as part of the data assimilation approach itself [CITE raiho].
 

### 2021-06-14 results.Rmd
start of Results: maps and initial figures

@@
 # Results
+
+## Simulated data experiments
 
@@
 
-Figures showing pool cov, C accounting / attribution
+```{r, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
+load("exp2.Rdata") # Generated by SingleSite_SingleDisturb_MultiPool.Rmd
+#Figures showing all 3 pools cov, C accounting / attribution 
+pool.names=c("Cleaf","Csoil","Cwood")
+for(i in 1:3){
+
+  CI.Na = sapply(X = Analysis,function(y){
+    x = y$MCMC
+    sel.n = which(colnames(x) == paste0("n[",i,"]"))
+    quantile(x[,sel.n],c(0.025,0.5,0.975))
+  })
+  D.Na = sapply(X = Analysis,function(y){
+    x = y$MCMC
+    sel.n = which(colnames(x) == paste0("D"))
+    mean(x[,sel.n])
+  })
+
+  rng = range(CI.Na)
+  if(rng[1]<1) rng[1] = 0
+  plot(No[,i],
+       ylim = rng,type='n',
+       ylab=pool.names[i],xlab="time")
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
+  lines(CI.Na[2,],col="#b2df8a",lwd=2)
+  points(No[,i])
+  #lines(Ntrue[,i],lty=2,lwd=2)
+  #lines(D.Na,lwd=2)
+  if(i==1){
+    legend("bottomright",legend=c("Assimilation","pseudodata"),
+       lwd=5,col=c("#b2df8a","black"))
+  }
+}
+```
+
+```{r,echo=FALSE, fig.asp=1,fig.width=3, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
+Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
+load("exp2.Rdata")
+
+plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+     ylim=range(Forecast[[10]]$Nf[,3]),
+     xlim=range(Forecast[[10]]$Nf[,1]),
+     xlab=pool.names[1],ylab=pool.names[3],
+     cex.lab=1.25,asp = 1)
+points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
+points(No[10,1],No[10,3],cex=3,col=1)
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
+
+## ensemble adjustment
+sel.a= seq(1,ne,length=10)#sample.int(ne,10)
+arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
+       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
+legend("topleft",
+       legend=c("Forecast","Analysis","Data"),
+       pch=c(1,3,18),col=c("purple","#b2df8a",1),
+       cex = 1.5)
+```
 
@@
 Figure showing "dragging down" of neighbors in conventional and breaking of COV in Multinomial
+
+```{r, echo=TRUE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance"}
+## figure with 3 rows, 2 col showing timeseries in rt and bivariate in left
+
+########experiment 3 - two disturb
+load("exp3.Rdata") ## Generated by SingleSite_MultiDisturb_MultiPool.Rmd
+pool.names=c("Cleaf","Csoil","Cwood")
+i = 3
+CI.Na = sapply(X = Analysis,function(y){
+  x = y$MCMC
+  sel.n = which(colnames(x) == paste0("X[",i,"]"))
+  quantile(x[,sel.n],c(0.025,0.5,0.975))
+})
+rng = range(CI.Na)
+if(rng[1]<1) rng[1] = 0
+plot(No[,i],
+     ylim = rng,type='n',
+     ylab=pool.names[i],xlab="time",
+     main="Multi-Disturbance")
+ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
+lines(CI.Na[2,],col="#b2df8a",lwd=2)
+points(No[,i])
+legend("bottomleft",legend=c("Assimilation","pseudodata"),
+       lwd=5,col=c("#b2df8a","black"))
+
+plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+     ylim=range(Forecast[[10]]$Nf[,3]),
+     xlim=range(Forecast[[10]]$Nf[,1]),
+     xlab=pool.names[1],ylab=pool.names[3],
+     cex.lab=1.25,asp = 1)
+points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
+points(No[10,1],No[10,3],cex=3,col=1)
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
+
+## ensemble adjustment
+sel.a= seq(1,ne,length=10)#sample.int(ne,10)
+arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
+       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
+legend("topleft",
+       legend=c("Forecast","Analysis","Data"),
+       pch=c(1,3,18),col=c("purple","#b2df8a",1),
+       cex = 1.5)
+
+##### experiment 4 - two disturb, pft switch
+load("exp4.Rdata") ## SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
+pool.names=c("Cleaf","Csoil","Cwood")
+i = 3
+CI.Na = sapply(X = Analysis,function(y){
+  x = y$MCMC
+  sel.n = which(colnames(x) == paste0("X[",i,"]"))
+  quantile(x[,sel.n],c(0.025,0.5,0.975))
+})
+rng = range(CI.Na)
+if(rng[1]<1) rng[1] = 0
+plot(No[,i],
+     ylim = rng,type='n',
+     ylab=pool.names[i],xlab="time",
+     main="Multi-Disturbance, PFT switching")
+ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
+lines(CI.Na[2,],col="#b2df8a",lwd=2)
+points(No[,i])
+legend("bottomleft",legend=c("Assimilation","pseudodata"),
+       lwd=5,col=c("#b2df8a","black"))
+
+plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+     ylim=range(Forecast[[10]]$Nf[,3]),
+     xlim=range(Forecast[[10]]$Nf[,1]),
+     xlab=pool.names[1],ylab=pool.names[3],
+     cex.lab=1.25,asp = 1)
+points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
+points(No[10,1],No[10,3],cex=3,col=1)
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
+
+## ensemble adjustment
+sel.a= seq(1,ne,length=10)#sample.int(ne,10)
+arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
+       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
+legend("topleft",
+       legend=c("Forecast","Analysis","Data"),
+       pch=c(1,3,18),col=c("purple","#b2df8a",1),
+       cex = 1.5)
+
+#experiment 5 - one disturb, 2 sites
+## Generated by
+
+```
 

### 2021-06-14 TwoSite_SingleDisturb_MultiPool.Rmd
Results: exp 5 two site plot

@@
 pf  <- array(0,c(3*nsite,3*nsite,4)) ## 3rd variable is class
-for(t in 1:10){
+for(t in 1:NT){
 
@@
   ## Disturbance
-  for(i in 1:ne) d.prior[i,] <- rcbern(rep(p.dist,2),rho=0.5)   ## prior on disturbance status
+  for(i in 1:ne) d.prior[i,] <- rcbern(rep(p.dist,2),rho=0.5,mu=1:2)   ## prior on disturbance status
   d.class <- d.prior[,1] + 2*d.prior[,2]
@@
   ## EnKF forecast stats
+  if(FALSE){
     NfK <- tobit(NfK)
@@
     Forecast.KD[[t]] <- list(d.prior=d.prior,Nf=NfK,muf=mufK,pf=pfK)
-
+  }
   ## priors
@@
 
+  ## initial conditions: disturbance likelihood
+  Lp = NA
+  for(i in 1:4){ Lp[i] = mvtnorm::dmvnorm(No[t,],muf[,i],pf[,,i])}
+  rho.IC = rep(0,4); rho.IC[which.max(Lp)] = 1; rho.IC[1] = NA
+
   ## Analysis
@@
                            data=update,
+                           inits = list(rho=rho.IC),
                            n.adapt=1000,n.chains=3)
-  jdat <- rjags::coda.samples(mod,variable.names="X",n.iter=2000) ## burn-in 
+  jdat <- rjags::coda.samples(mod,variable.names="rho",n.iter=2000) ## burn-in 
   jdat <- rjags::coda.samples(mod,variable.names=c("X","rho","x"),n.iter=10000)
@@
   ## EnKF Analysis
+  if(FALSE){
     I = H <- diag(rep(1,3*nsite))
@@
     NaK <- abind::abind(NaK[,1:3],NaK[,4:6],along = 3)
+  }
 
 }
+
+```
+
+```{r}
+save(Forecast,Analysis,No,Ntrue,R,file="exp5.RData")
+Forecast10 = Forecast[[10]]
+Analysis10 = Analysis[[10]]
+CI.Na = list()
+for(i in c(3,6)){
+  CI.Na[[i]] = sapply(X = Analysis,function(y){
+    x = y$MCMC
+    sel.n = which(colnames(x) == paste0("X[",i,"]"))
+    quantile(x[,sel.n],c(0.025,0.5,0.975))
+  })
+}
+save(Forecast10,Analysis10,CI.Na,No,Ntrue,R,file="exp5trim.RData")
 

### 2021-06-14 results.Rmd
Results: exp 5 two site plot

@@
 ## ensemble adjustment
+ne = nrow(Forecast[[10]]$Nf)
 sel.a= seq(1,ne,length=10)#sample.int(ne,10)
@@
 
-```{r, echo=TRUE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance"}
+```{r, echo=FALSE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance"}
 ## figure with 3 rows, 2 col showing timeseries in rt and bivariate in left
@@
 
-#experiment 5 - one disturb, 2 sites
-## Generated by
+######### experiment 5 - one disturb, 2 sites
+load("exp5trim.Rdata") ##Generated by TwoSite_SingleDisturb_MultiPool.Rmd
+pool.names=c("Cleaf","Csoil","Cwood")
+rng = range(CI.Na)
+if(rng[1]<1) rng[1] = 0
+plot(No[,i],
+     ylim = rng,type='n',
+     ylab=pool.names[i],xlab="time",
+     main="Multi-Site, Single-Disturbance")
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[3]]),CI.Na[[3]][1,],CI.Na[[3]][3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[6]]),CI.Na[[6]][1,],CI.Na[[6]][3,],col=ecoforecastR::col.alpha("green",0.7))
+lines(CI.Na[[3]][2,],col="#b2df8a",lwd=2)
+lines(CI.Na[[6]][2,],col="green",lwd=2)
+points(No[,3])
+points(No[,6],pch=2)
+legend("bottomleft",legend=c("Site 1","Site 2","pseudodata"),
+       lwd=5,col=c("#b2df8a","green","black"))
 
+### cov plot
+ss = floor(seq(1,nrow(Analysis10$Na),length=200)) ## subsample
+plot(Analysis10$Na[,1],Analysis10$Na[,3],pch=Analysis10$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+     ylim=range(Forecast10$Nf[,3,1]),
+     xlim=range(Forecast10$Nf[,1,1]),
+     xlab=pool.names[1],ylab=pool.names[3],
+     cex.lab=1.25,asp = 1,type='n')
+points(Forecast10$Nf[ss,1,1],Forecast10$Nf[ss,3,1],cex=0.5,col="purple",pch=18)
+points(Forecast10$Nf[ss,1,2],Forecast10$Nf[ss,3,2],cex=0.5,col="violet",pch=18)
+points(Analysis10$Na[ss,1],Analysis10$Na[ss,3],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("#b2df8a",alpha=0.8),cex=0.5)
+points(Analysis10$Na[ss,4],Analysis10$Na[ss,6],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("green",alpha=0.8),cex=0.5)
+points(No[10,1],No[10,3],cex=3,col=1)
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
+points(No[10,4],No[10,6],cex=3,col=1)
+lines(No[10,4]+c(-1,1)*sqrt(R[1,1]),rep(No[10,6],2))
+lines(rep(No[10,4],2),No[10,6]+c(-1,1)*sqrt(R[3,3]))
+
+## ensemble adjustment
+sel.a= ss[seq(1,length(ss),length=10)]#sample.int(ne,10)
+arrows(Forecast10$Nf[sel.a,1,1],Forecast10$Nf[sel.a,3,1],
+       Analysis10$Na[sel.a,1],Analysis10$Na[sel.a,3],length = 0.1,lwd = 0.5)
+arrows(Forecast10$Nf[sel.a,1,2],Forecast10$Nf[sel.a,3,2],
+       Analysis10$Na[sel.a,4],Analysis10$Na[sel.a,6],length = 0.1,lwd = 0.5)
+legend("topleft",
+       legend=c("Forecast","Analysis","Data"),
+       pch=c(1,3,18),col=c("purple","#b2df8a",1),
+       cex = 1)
 ```
+
+## Remotely-sensed disturbance in Oregon
 

### 2021-06-14 disturbance.R

@@ -53,7 +53,7 @@ disturbance2 <- function(x,type){ ## x = Bleaf, Bsoil, Bstem
 }
 
 # random correlated bernouli
-rcbern <- function(p,rho){
+rcbern <- function(p,rho,mu){
   
   ## calculate cov
   D = as.matrix(dist(seq_along(mu),diag = TRUE,upper = TRUE))

### 2021-06-14 results.Rmd

@@
 ```{r, echo=FALSE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance"}
+
 ## figure with 3 rows, 2 col showing timeseries in rt and bivariate in left

### 2023-01-01 SingleSite_SingleDisturb.Rmd
Pushed results outline to own file, small updates to discussion and methods
Results: simulated data experiments

@@
 ### Save outputs required to include figure in Results
-save(data,CI.fN,time,an0.stats,an.stats,NT,Dbar,file="SSSD.RData")
+save(data,CI.fN,time,an0.stats,an.stats,NT,Dbar,xlim,col,file="SSSD.RData")
+
+```
+
+## Conceptual Figure
+```{r, echo=FALSE}
+## params
+set.seed(1066)
+p.dist = 0.05
+n0 = 1
+nIC = 6
+r = .1
+K = 6
+n.sd = 0.25
+NE = 1000
+obs=1
+obs.sd = 0.75
+
+## generate IC ensemble
+ens = rnorm(NE,nIC,n.sd)
+
+## Forecast step
+LAI = ens + r*ens*(1-ens/K) + rnorm(NE,0,n.sd) 
+disturbed = as.logical(rbinom(NE,1,p.dist))
+LAI[disturbed] = abs(rnorm(sum(disturbed),n0,0.5)) ## disturbance
+
+##Approximate forecast with Normal
+xseq = seq(0,8,length=1000)
+muf = mean(LAI)
+Pf = var(LAI)
+
+## analysis step: combine previous forecast with observed data
+H <- matrix(1,ncol=1)
+K <- Pf * t(H) %*% solve(H%*%Pf%*%t(H) + obs.sd^2)
+mua <- muf + K*(obs - muf) 
+Pa <- (1-K %*% H)*Pf
+
+## EnKF
+hist(LAI,probability = TRUE)#,ylim=c(0,dnorm(mua,mua,sqrt(Pa))))
+lines(xseq,dnorm(xseq,muf,sqrt(Pf)),col="#b2df8a",lwd=3)
+lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
+lines(xseq,dnorm(xseq,mua,sqrt(Pa)),col='orange',lwd=3)
+legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","orange"))
+
+## Forecast: Mixture of normals
+mu1 = mean(LAI[!disturbed])
+P1 = var(LAI[!disturbed])
+mu2 = mean(LAI[disturbed])
+P2 = var(LAI[disturbed])
+
+## assimilation
+priors <- list(p = p.dist,
+                 mufN = mu1,
+                 mufD = mu2,
+                 pfN  = 1/P1,
+                 pfD  = 1/P2,
+                 Y = obs,
+                 r = 1/obs.sd^2
+                 )
+mod <- jags.model(file=textConnection(AnalysisSSSD),
+                   data=priors,n.adapt=1000,n.chains=3)
+jdat <- coda.samples(mod,variable.names=c("n"),n.iter=10000)
+
+hist(LAI,probability = TRUE,ylim=c(0,dnorm(mu1,mu1,sqrt(P1))))
+lines(xseq,(1-p.dist)*dnorm(xseq,mu1,sqrt(P1)),col="#b2df8a",lwd=3)
+lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="brown",lwd=3)
+lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
+lines(density(as.matrix(jdat)),col="blue",lwd=3)
+legend("top",legend=c("undisturbed","disturbed","data","assimilation"),lwd=3,col=c("#b2df8a","brown","red","blue"))
+
+#legend("top",legend=c("undisturbed","disturbed"),lwd=3,col=c("#b2df8a","brown"))
 

### 2023-01-01 methods.Rmd
Results: simulated data experiments

@@
 Indeed, if the you increased the disturbance probability substantially (e.g.
-50%), the EnKF would treat this as a unimodal forecast half way between the disturbed and undisturbed states, a place where no ensemble members are present.
+50%), the EnKF would treat this as a unimodal forecast half way between the disturbed and undisturbed states, a place where no ensemble members are present (Figure \@ref(fig:concept), top).
 
@@
 SSSD_figure()
+```
+
+```{r concept, echo=FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
+Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities.
+The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
+## params
+set.seed(1066)
+p.dist = 0.05
+n0 = 1
+nIC = 6
+r = .1
+K = 6
+n.sd = 0.25
+NE = 1000
+obs=1
+obs.sd = 0.75
+
+## generate IC ensemble
+ens = rnorm(NE,nIC,n.sd)
+
+## Forecast step
+LAI = ens + r*ens*(1-ens/K) + rnorm(NE,0,n.sd) 
+disturbed = as.logical(rbinom(NE,1,p.dist))
+LAI[disturbed] = abs(rnorm(sum(disturbed),n0,0.5)) ## disturbance
+
+##Approximate forecast with Normal
+xseq = seq(0,8,length=1000)
+muf = mean(LAI)
+Pf = var(LAI)
+
+## analysis step: combine previous forecast with observed data
+H <- matrix(1,ncol=1)
+K <- Pf * t(H) %*% solve(H%*%Pf%*%t(H) + obs.sd^2)
+mua <- muf + K*(obs - muf) 
+Pa <- (1-K %*% H)*Pf
+
+## EnKF
+hist(LAI, probability = TRUE, main = "Conventional EnKF")#,ylim=c(0,dnorm(mua,mua,sqrt(Pa))))
+lines(xseq,dnorm(xseq,muf,sqrt(Pf)),col="#b2df8a",lwd=3)
+lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
+lines(xseq,dnorm(xseq,mua,sqrt(Pa)),col='orange',lwd=3)
+legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","orange"))
+
+## Forecast: Mixture of normals
+mu1 = mean(LAI[!disturbed])
+P1 = var(LAI[!disturbed])
+mu2 = mean(LAI[disturbed])
+P2 = var(LAI[disturbed])
+
+## assimilation
+priors <- list(p = p.dist,
+                 mufN = mu1,
+                 mufD = mu2,
+                 pfN  = 1/P1,
+                 pfD  = 1/P2,
+                 Y = obs,
+                 r = 1/obs.sd^2
+                 )
+mod <- jags.model(file=textConnection(AnalysisSSSD),
+                   data=priors,n.adapt=1000,n.chains=3)
+jdat <- coda.samples(mod,variable.names=c("n"),n.iter=10000)
+
+hist(LAI,probability = TRUE,ylim=c(0,dnorm(mu1,mu1,sqrt(P1))), main="Multinomial Filter")
+lines(xseq,(1-p.dist)*dnorm(xseq,mu1,sqrt(P1)),col="#b2df8a",lwd=3)
+lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="brown",lwd=3)
+lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
+lines(density(as.matrix(jdat)),col="blue",lwd=3)
+legend("top",legend=c("undisturbed","disturbed","data","assimilation"),lwd=3,col=c("#b2df8a","brown","red","blue"))
 ```
@@
   * Regress vs: pre-disturbance biomass, size of disturbance (abs & rel), type, etc
+
+Divergence
+
+As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval.
+If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.

### 2023-01-01 results.Rmd
Results: simulated data experiments

@@
 
-1. Simulated data experiments (known disturbance)
-  + Single Site, Single Disturbance
+Overall, the results from the simulated data experiments (Table \@ref(tab:pseudoTable)) demonstrate that the proposed multinomial ensemble filter is able to successfully detect disturbances under a range of different conditions.
 
-Figures showing the current problem, solution, and that we've got a genuine multimodal distribution and P(dist)
+Experiment 1, which aimed to capture a single disturbance at a single site for a single pool, was already depicted in Figure \@ref(fig:enkf) and Figure \@ref(fig:concept) when discussing the challenges facing traditional Kalman Filter variants.
+Specifically, Figure \@ref(fig:enkf) shows that the multinomial filter (blue) is able to capture the discrete nature of the disturbance at time 10, with the posterior probability of disturbance jumping from near zero before this event to 0.977 for the disturbance, and then dropping back down to near zero in subsequent years.
+Figure \@ref(fig:concept) bottom illustrates what is occurring in this disturbance year.
+As in the prior 9 years the forecast itself is multimodal, with most of the ensemble members predicting no disturbance and a small number predicting a disturbance event.
+The multinomial filter assigns a prior mean and variance to each mode based on the ensemble summary statistics, which is the same as a traditional EnKF except for the stratification by disturbance type, and then assigns each state a prior probability based on the ensemble frequency of being in each state (disturbed = dark red, undisturbed = green).
+When the observed data (red) suggests a large drop in LAI, the posterior (blue) not only updates the ensemble mean and variance of each mode, but also the probability of being in each mode.
+This posterior is still multimodal, but in this specific instance the weight assigned the undisturbed mode (0.023) is too low to be visually notable.
+Noteworthy is that the LAI in years 12 and 15 is actually lower than that in the disturbance year but the multinomial, yet the model still assigns this as being undisturbed.
+This occurs because in these years the observed low LAI is compatible with the observation error around the undisturbed forecast, which has a higher prior probability than the odds of subsequent disturbance.
 
-  + SSSD Multi Pool
+Experiment 2 expands the assimilation to a multivariate forecast, in this case including joint predictions for leaf, wood, and soil carbon, and with a disturbance to both the leaf and wood pools (Figure \@ref(fig:sssd)).
+Figure \@ref(fig:sssdBi) shows the multivariate nature of the detection of the disturbance event at time 10, with the bimodal forecast ensemble shown in purple (undisturbed in the top right, disturbed in the bottom left), an observation in black that is consistent with the disturbed state, and the posterior ensemble in green.
+Arrows depict the adjustment of a random subset of individual ensemble members from the forecast prior to the analysis posterior.
 
-```{r, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
+```{r sssd, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
 load("exp2.Rdata") # Generated by SingleSite_SingleDisturb_MultiPool.Rmd
@@
 
-```{r,echo=FALSE, fig.asp=1,fig.width=3, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
+```{r sssdBi,echo=FALSE, fig.asp=1,fig.width=3, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
 Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
@@
 
-  + SS Multi-Disturbance w/ PFT switching
+Experiment 3 (Figure \@ref(fig:multiPseudo) top row) illustrates the ability of the multinomial model to select between more than one disturbance.
+In the bivariate plot (top right panel) we can see that the model can successfully distinguishing between two types of disturbance, one that involves the loss of both wood and leaf biomass (e.g.
+fire, logging) and the other that just removes leaf biomass (e.g.
+defoliating forest pest).
+Experiment 4 (Figure \@ref(fig:multiPseudo) middle row) is similar to 3 but extends the woody biomass removing disturbance to one that causes a switch in PFT from a forest to a low-stature vegetation type (e.g.
+grassland, shrubland, agriculture), demonstrating that the model can handle land use and land cover changes.
+Finally, experiment 5 (Figure \@ref(fig:multiPseudo) bottom row) illustrates the extension of the assimilation to multiple sites, one that experiences and disturbance and one that does not.
+In the years prior to the disturbance the multivariate nature of the assimilation allows us to borrow strength across sites, with observations at one site reducting uncertainty not just at that site but at the other site as well based on the cross-site forecast covariance.
+That said,this example illustrates the ability of the multinomial model to avoid the impact of spurious correlations across sites during disturbance events.
+The presence of the disturbance at site 1 did not cause a reduction in biomass in site 2 during the disturbance, nor did it increase the productivity in site 2 in the years post-disturbance despite the large increase observed at site 1.
 
-Figure showing PFT switching vs not, ability to detect disturbance type
-
-  + Simple Multi-site: spatial cov
-
-Figure showing "dragging down" of neighbors in conventional and breaking of COV in Multinomial
-
-```{r, echo=FALSE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance"}
+```{r multiPseudo, echo=FALSE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.height=4,fig.width=3}
 
 ## figure with 3 rows, 2 col showing timeseries in rt and bivariate in left
+#par(mfrow=c(3,2))
+m = matrix(c(1,1,2,3,3,4,5,5,6),nrow=3,byrow = TRUE)
+layout(m)
 
@@
        Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
-legend("topleft",
+legend("bottomright",
        legend=c("Forecast","Analysis","Data"),
        pch=c(1,3,18),col=c("purple","#b2df8a",1),
-       cex = 1.5)
+       cex = 1.25)
 
@@
      xlab=pool.names[1],ylab=pool.names[3],
-     cex.lab=1.25,asp = 1)
+     cex.lab=1.25)
 points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
@@
 
+## VSEM Calibration
+
 ## Remotely-sensed disturbance in Oregon
 
-2. Real-world examples
-  + Calibrate state-and-transition prior
-    * data: (samples from) maps of disturbance type by year
-    * model: Multinomial Dirlichet
+```{r ORorig, echo=FALSE}
+## summary figure showing didn't work great at default error rate
 
-Figure showing map and estimated priors
+### Row 1: Example time-series ###
+site = 188
+exp.name = c("P1Q1R1","P1Q1R4")
+exp.pretty.name = c("Original","Low Obs Error")
+pool.names=c("Cleaf","Csoil","AGB")
+i = 3 ## Stem
+for(ex in seq_along(exp.name)){
+  load(paste0("OR_TS_",site,"_",exp.name[ex],".RData"))
+  plot(Yi[,i],
+       ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
+       type='n',
+       ylab=pool.names[i],xlab="time",
+       main=exp.pretty.name[ex])
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
+  lines(CI.Na[3,],col="#b2df8a",lwd=2)
+  points(Yi[,i])
+  if(ex == 1) {
+    legend("bottomleft",legend=c("Assimilation","Data"),
+       lwd=5,col=c("#b2df8a","black"),cex=0.7)
+  }
+}
 
-  + Applications
-    + Logging
-    + Fire
-    + Land use transition (forest -> non?)
+### Row 2: Divergence ###
+load("stats.RData")
+load("elist.RData")
+disturb.names = c("Intact","Other","Cut","Fire","Pest")
+disturb.col = c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a')
+disturb.col.al <- tapply(disturb.col,1:5,FUN=ecoforecastR::col.alpha,0.7); disturb.col.al[1] = disturb.col[1]
+for(i in 1:2){
+  ## divergence
+  div = (elist[[i]]$divergence[,1] - obs.tlo)/
+    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
+  div.reg    = lm(div~obs.tlo)
+
+  dh = density(na.omit(div))
+#  abline(v=c(-1,1),lty=2)
+  plot(obs.tlo,div,col=disturb.col.al[obs.class],
+       xlab="final AGB",ylab="Divergence",
+       pch=16)
+  abline(h=c(-1,1),lty=2)
+  abline(div.reg,col="purple",lwd=3)
+  abline(v=0,lty=2)
+  lines(dh$y*10,dh$x,lwd=3)
+  if(i == 1) legend("topright",legend=disturb.names,
+                  col=disturb.col,pch=16,cex=0.7,
+                  text.col=c(1,disturb.col[2:5]))
+}
+
+### Row 3: Disturbance magnitude vs P(detection) ###
+pred.seq=list()
+distabs = obs.pre-obs.dist
+pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
+for(i in 1:2){
+  distrate = distrate = elist[[i]]$distrate
+  dist.qb.abs = glm(distrate ~ distabs,
+                    na.action=na.omit,
+                    family=quasibinomial("logit")
+  )
+  plot(distabs,distrate,col=disturb.col.al[obs.class],
+       xlab="AGB disturbance magnitude",ylab="P(detect)",
+       pch=16)
+  lines(pred.seq$distabs,predict(dist.qb.abs,newdata=pred.seq,type="response"),col="purple")
+  if(i == 1) legend("topleft",legend=disturb.names,
+                  col=disturb.col,pch=16,cex=0.7,
+                  text.col=c(1,disturb.col[2:5]))
+}
+```
 
 * RESULTS
+  * limited success under default parameters
+    * high frequency of filter divergence [summary fig, example divergence fig]
+    * divergence coverage original `r stats[[1]]$div.cov`
+    * divergence coverage low error `r stats[[2]]$div.cov`
+    * the smaller the observed biomass at the end of the timeseries, the larger the divergence [include scatterplot?]
+    * low rate of detected disturbance [calc stat?, hist? cdf?]
+    * what disturbances were detected tended to be associated with high pre-disturbance biomass and a large reduction in biomass (both absolute and proportional).
+Most commonly seen in dist class "red"
   * needed to reduce obs and process error to work
-  * abs magnitude of disturbance is best predictor of assim prob.
+  * example "successful" scenario: P1Q1R4
+    * reduced observation error
+    * abs magnitude of disturbance is best predictor of assim prob. [fig]
       * given magnitude, dist type n.s.  
+  * comparative analysis

### 2023-01-02 Paper.Rmd
Results: start of calibration

--- 2021-02-02 Paper.Rmd
@@
 ---
-title: "Disturbance Assimilation"
-author: "Michael Dietze"
+title: "Assimilating disturbance: Toward real-time monitoring and forecasting"
+author:
+  - Michael C. Dietze, Earth and Environment, Boston University, Boston, MA
+  - Shawn P. Serbin, Environmental and Climate Sciences Department, Brookhaven National Laboratory, Upton, NY
+  - Robert Kennedy, Geography, Environmental Sciences, and Marine Resource Management, Oregon State University, Corvallis, OR
+  - Beverly E. Law, Forest Ecosystems & Society, Oregon State University, Corvallis, OR
 date: '`r Sys.Date()`'
 output:
+  pdf_document: 
+    fig_caption: yes
+    number_sections: yes
+    toc: yes
+    keep_tex: true
   html_document: 
@@
 
-# Outline
+# Abstract {-}
 
-# Abstract
+Current approaches to bottom-up disturbance monitoring rely heavily on the detection of land-use, land-use change, and forestry (LULUCF) through remote sensing, but often account for ecosystem impacts using simple look-up tables.
+By contrast, process models are frequently used to analyze and predict disturbance dynamics in greater detail.
+Once observations are available, however, we need to update predictions, especially for stochastic processes such as disturbance.
+State data assimilation (SDA) is designed specifically to update predictions, nudging modeled states back toward reality in proportion to the uncertainties in the model and the data, but current SDA algorithms are designed to update continuous states, not discrete disturbances.
+Here we develop a new Bayesian SDA algorithm that combines a discrete Multinomial state-and-transition framework with conventional ensemble filtering SDA approaches.
+To demonstrate the potential for assimilating disturbance, we applied the Multinomial SDA to the Very Simple Ecosystem Model (VSEM), performing both simulated data experiments with known disturbances and testing the algorithm against real-world disturbances detected in the LandTrender data product for central Oregon.
+
+With simulated disturbance we demonstrate the ability to not only detect discrete disturbance events but also avoid false positives.
+We also demonstrated the ability to fuse multiple data types to successfully distinguish different disturbance types, and to probabilistically capture vegetation type ‘switching’ events within the assimilation and ensemble forecast.
+To apply this to real-world data we calibrated VSEM against eddy-covariance and ancillary data from the Ameriflux US-Me2 tower.
+We then selected 356 conifer forest sites for testing, using the Landtrendr disturbance product to stratify by four disturbance types (cut, burn, pest, and other).
+We then assimilated the 30m Landtrendr annual aboveground biomass product from 1990-2017 and assessed the rate of disturbance detection.
+Assimilating just AGB, our assimilation was sensitive to disturbances that reduced biomass by 1.5 kg/m2 but underpredicted defoliation disturbances, which we expect would be improved by also assimilating LAI.
+Moving forward, the SDA framework provides an exciting opportunity to fuse multiple data sources to holistically improve real-time disturbance detection, impact assessment (e.g.
+carbon sequestration), and forecasts of both disturbance events and post-disturbance recovery within a single integrated system.
 
@@
 
+<!---- SUPPLEMENTAL MATERIAL --->
+```{r child = 'supplemental.Rmd'}
+```
+
 # Bibliography

### 2023-01-02 calibrate.Rmd
Results: simulated data experiments
Results: start of calibration

@@
 
+bioS = stack("conus_biomass_ARD_tile_h03v04/conus_biomass_ARD_tile_h03v04.tif")
 AF.biomass = raster::extract(bioS,AF[,c("Lat.","Long.")])
@@
 
+## calibration only
+Cw.post = tapply(post[,"CR"],years,mean)
+plot(year,Cw.post,ylab="AGB (kgC/m2)",type='l',lwd=3,
+     xlim=c(2002,2012),ylim=range(Cw,na.rm=TRUE))
+points(year,Cw,col=2,pch="+",cex=2)
+legend("topleft",legend=c("model","data"),pch=c(NA,"+"),lty=c(1,NA),col=1:2)
+
 ## Rough approximation of process error: RMSE on delta
@@
 
-save(theta,Ql,Qw,Qs,file="fit.RData")
+## R2
+RMSE = sqrt(c(mean((Cl.post-Cl)^2,na.rm=TRUE),
+            mean((Cw.post-Cw)^2,na.rm=TRUE),
+            mean((NEEm-aNEE)^2,na.rm=TRUE)))
+1-RMSE^2/c(var(Cl,na.rm = TRUE),var(Cw,na.rm = TRUE),var(aNEE))
+
+save(theta,Ql,Qw,Qs,NEEm,aNEE,year,Cl,Cl.post,Cw,Cw.post,mle,file="fit.RData")
 ```

### 2023-01-02 methods.Rmd
Results: start of calibration

@@
 The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
+library(rjags)
 ## params
@@
 ## assimilation
+AnalysisSSSD <- "
+model{
+  ## prior
+  muN ~ dnorm(mufN,pfN)
+  muD ~ dnorm(mufD,pfD)
+  D ~ dbern(p)
+  ## Analysis
+  n <- D*muD + (1-D)*muN
+  Y ~ dnorm(n,r)
+}"
 priors <- list(p = p.dist,

### 2023-01-02 results.Rmd
Results: start of calibration

@@
 
+Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from Ameriflux site US-ME2.
+Best fit parameters are given in Supplemental Table \@ref(tab:parameters)
+
+RMSE
+
+The closest fit to the observed data was for woody biomass, which had a RMSE of 0.08947263 $kgC/m^2$ and a $R^2$ of 0.9772768 (Supplemental figure \@ref(fig:calibrate))
+
+RMSE
+> RMSE
+[1] 0.06703855 0.08947263 0.61244709
+
+Process error was approximated as...
+> Ql
+[1] 0.01040558
+> Qw
+[1] 0.03424142
+> Qs
+[1] 0.1499524
+
+## Cv of Q
+> Ql/mean(diff(Cl),na.rm=TRUE)
+[1] 1.137623
+> Qw/mean(diff(Cw),na.rm=TRUE)
+[1] 0.1658704
+
+Systematic bias in leaf C (model is ~2x too productive)
+
+Systematic bias in NEE, most likely associated with ...
+
+Overall, while the calibration does suggest room for improvement in the underlying model, it performed well for our target variable (AGB) and appears adequate for the job of testing the multinomial model against real AGB data for a range of different disturbance events.
+
 ## Remotely-sensed disturbance in Oregon

### 2023-01-02 supplemental.Rmd
Results: start of calibration

# Supplemental material

```{r calibration,fig.cap="Predicted and observed woody biomass, leaf biomass, and NEE at US-Me2",echo=FALSE}
load("fit.RData")

plot(year,Cw.post,ylab="AGB (kgC/m2)",type='l',lwd=3,
     xlim=c(2002,2012),ylim=range(Cw,na.rm=TRUE))
points(year,Cw,col=2,pch="+",cex=2)
legend("topleft",legend=c("model","data"),pch=c(NA,"+"),lty=c(1,NA),col=1:2)

plot(year,Cl.post,ylab="Leaf Biomass (kgC/m2)",
     ylim=range(c(Cl,Cl.post),na.rm=TRUE))
points(year,Cl,col=2,pch="+")
legend("topleft",legend=c("model","data"),pch=c(NA,"+"),lty=c(1,NA),col=1:2)

plot(year,NEEm,ylab="annual NEE ()",
     ylim=range(c(NEEm,aNEE),na.rm=TRUE))
points(year,aNEE,col=2,pch="+")
legend("bottomright",legend=c("model","data"),pch=c(NA,"+"),lty=c(1,NA),col=1:2)
```

```{r parameters,echo=FALSE}
param.names = c("LUE","GAMMA","tauS","tauW","Av","sd_NEE","sd_leaf","sd_wood")
param.desc = c("Light Use Efficiency","Fraction of GPP allocated to autotrophic respiration","Soil C residence time","Wood C residence time","Fraction of NPP allocated to leaves","NEE standard deviation","leaf C standard deviation","wood C standard deviation")
param.table = cbind(param.names,param.desc,mle$par)
colnames(param.table) <- c("Parameter","Description","Value")
knitr::kable(param.table,caption="Calibrated model parameters")
```

### 2023-01-06 OR_original_model.Rmd
old code: OR postprocessing

@@
 obs.class= dextract[,"t2004"]/10
+
+save(elist,obs.pre,obs.dist,obs.tlo,obs.class,file="elist.RData")
 ```
@@
 dev.off()
+save(stats,file="stats.RData")
 ```
@@
 
+## preprocess timeseries example for visualization
+```{r}
+## visualization to help select a site to look at
+plotly::plot_ly(data.frame(obs.pre=obs.pre,obs.post=obs.dist),x=~obs.pre,y=~obs.post)
+which.min((obs.pre-13.3)^2+(obs.dist-5.1)^2) ## select a point
+
+## settings
+site = 188
+exp.name = "P1Q1R4" #"P1Q1R1"  ## manually run for each experiment **
+
+## data
+NT = 33
+Yi = matrix(NA,NT,3)  ## observed data
+Yi[1:ncol(Yw),3] = Yw[site,]  ## only Cw observed
+
+## model
+load(paste0("OR_SD_",site,"_",exp.name,".RData"),envir = viz)
+pool.names=c("Cleaf","Csoil","Cstem")
+test = viz$test
+
+for(i in 1:3){
+
+    CI.Na = sapply(X = test$Analysis,function(x){
+      y = x$CI
+      sel.n = which(colnames(y) == paste0("n[",i,"]"))
+      y[,sel.n]
+    })
+
+    CI.D = sapply(X = test$Analysis,function(x){
+      y = x$CI
+      sel.n = which(colnames(y) == paste0("D"))
+      y[,sel.n]
+    })
+
+    CI.Nf = sapply(X = test$Forecast, function(x){
+      mu = x$mufN[i]
+      sd = 1/sqrt(x$pfN[i,i])
+      return(mu + c(-1.96,0,1.96)*sd)
+    })
+
+  plot(Yi[,i],
+       ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
+       type='n',
+       ylab=pool.names[i],xlab="time")
+  ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
+  lines(CI.Na[3,],col="#b2df8a",lwd=2)
+  points(Yi[,i])
+  lines(CI.D[3,])
+
+}
+save(Yi,CI.Na,CI.D,file=paste0("OR_TS_",site,"_",exp.name,".RData"))
+```
+
+## summary info for filter divergence figures
+```{r}
+exp.name = "P1Q1R4" #"P1Q1R1"
+
+for(i in seq_along(exps)){
+
+  ## divergence
+  div = (elist[[i]]$divergence[,1] - obs.tlo)/
+    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
+  div.reg    = lm(div~obs.tlo)
+  div.median = median(div,na.rm = TRUE)
+  div.cov    = sum(div < 1 & div > -1,na.rm=TRUE)/sum(!is.na(div))
+
+}
+```
+
 ## Next steps:

### 2023-01-12 OR_original_model.Rmd
Results: flushed out first draft text, added a couple citations. Methods: tried to tweak OR maps

@@
 plot(Rmult[Rs],div.cov,col=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage")
+legend("topright",legend=paste0(Qmult[1:2],"x"))
 
@@
 plot(Rmult[Rs],div.median,col=Qs,log="x",xlab="Variance Multiplier",ylab="Median Divergance")
+legend("topleft",legend=paste0(Qmult[1:2],"x"))
 

### 2023-01-12 methods.Rmd
Results: flushed out first draft text, added a couple citations. Methods: tried to tweak OR maps

@@
 library(rjags)
+par(mfrow=c(2,1))
 ## params
@@
 
-```{r map, echo=FALSE}
+```{r map, echo=FALSE, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances"}
 ## maps of pre-disturbance biomass, cover, and 2004 disturbance
@@
 load("rat.RData")
-raster::plot(cov04,col=rat$color,axis.args=list(at=seq_along(rat$description),labels=rat$description,cex.axis=0.4),asp=1,main="Land Cover")
+raster::plot(cov04,col=rat$color,asp=1,main="NLCD Land Cover",
+            axis.args=list(at=seq_along(rat$description),labels=rat$description,cex.axis=0.4))
+#cov04 = terra::rast("cov04.tif")
+#terra::plot(cov04,col=rat$color,asp=1,main="NLCD Land Cover")
 
 ## Disturbance
-if(!exists("dis04")) dis04 = raster::raster("dis04.tif")
+if(FALSE){ 
+  ## both these options seem to interpolate/regrid during the projection, turning integer classes into continuous values that don't map the same
+  dis04 = terra::rast("dis04.tif")
+  dis04ll <- terra::project(dis04,crs(cov04))
+
+  dis04 = raster::raster("dis04.tif")
+  dis04ll = raster::projectRaster(dis04,crs=crs(cov04))
+  #terra::plot(dis04ll,col=c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a'),asp=1,main="Landtrendr Disturbance",plg=list(legend=dname)) #"#a6cee3" '#33a02c'
+}
+if(!exists("dis04")) 
 dname = c("Intact","Other","Cut","Fire","Pest")
-raster::plot(dis04,col=c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a'),axis.args=list(at=seq(3,48,length=5),labels=dname,cex.axis=0.4),asp=1,main="Disturbance") #"#a6cee3" '#33a02c'
+raster::plot(dis04,col=c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a'),asp=1,
+            axis.args=list(at=seq(-5,50,length=5),labels=dname,cex.axis=0.4), 
+            main="Landtrendr Disturbance") #"#a6cee3" '#33a02c'
 
@@
 if(!exists("bio04")) bio04 = raster::raster("bio04.tif")
-raster::plot(bio04,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main="Biomass")
+raster::plot(bio04,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main="Landtrendr Biomass")
 
@@
 Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state.
-
-```{r paramTable,echo=FALSE}
-load("fit.RData")
-varName = c(names(theta),"Ql","Qw","Qs")
-varName[c(5,7)] = c("tauL","tauW")
-paramTable = data.frame(varName,fit=format(c(theta,Ql,Qw,Qs),digits=3,scientific=TRUE))
-knitr::kable(paramTable, caption = "VSEM parameter estimates")
-```
 

### 2023-01-12 results.Rmd
Results: flushed out first draft text, added a couple citations. Methods: tried to tweak OR maps

--- 2023-01-01 results.Rmd
@@
 
-```{r sssdBi,echo=FALSE, fig.asp=1,fig.width=3, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
+```{r sssdBi,echo=FALSE, fig.asp=1,fig.width=6, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
 Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
+## for html set fig.width=3
 load("exp2.Rdata")
@@
 
+Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from Ameriflux site US-ME2.
+Best fit parameters are given in Supplemental Table \@ref(tab:parameters).
+The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibrate)).
+Achieving this fit came at the cost of nontrivial systematic biases in both leaf carbon (0.064 $kgC/m^2$, 13%) and NEE (0.528 $kgC/m^2/year$, 69%).
+The RMSEs for leaf carbon (0.067 $kgC/m^2$, 13.6%) and NEE (0.612 $kgC/m^2/year$, 80%) were similar to the biases, suggesting that most of the model error was systematic, particularly for leaf carbon.
+
+The dynamic process errors for the leaf, wood, and soil pools, which were approximated as the RMSE of the _change_ in each pool each year, were 0.01, 0.03, and 0.15 $kgC/m^2/yr$.
+This represents a dynamic error of 2.1%, 0.6%, and 0.7% respectively relative to the size of the pool and a CV of 114% and 17% for leaf and wood relative to the observed annual biomass increment.
+
+Overall, while the inability of the calibration to capture multiple data constraints simultaneously does suggest room for improvement in the underlying model (@Cameron2022), the model performed well for our target variable (AGB) and appears adequate for the job of testing the multinomial model against real AGB data for a range of different disturbance events.
+
+```{r paramTable,echo=FALSE}
+# load("fit.RData")
+# varName = c(names(theta),"Ql","Qw","Qs")
+# varName[c(5,7)] = c("tauL","tauW")
+# paramTable = data.frame(varName,fit=format(c(theta,Ql,Qw,Qs),digits=3,scientific=TRUE))
+# knitr::kable(paramTable, caption = "VSEM parameter estimates")
+```
+
 ## Remotely-sensed disturbance in Oregon
 
-```{r ORorig, echo=FALSE}
+```{r ORorig, echo=FALSE,fig.asp=0.75,fig.cap="Filter divergence and disturbance detection for central Oregon test case.
+Left column is the default configuration while the right column is a reduced observation error experiment (Exp 2).
+Top row is the assimilation for an example site.
+Middle row shows the across-site distribution of divergence metrics and how these vary by disturbance magnitude (purple) and disturbance type (symbols).
+Bottom row shows how the probability of detecting the disturbance varies with disturbance magnitude"}
 ## summary figure showing didn't work great at default error rate
+m = matrix(1:6,nrow=3,byrow = TRUE)
+layout(m)
 
@@
 exp.name = c("P1Q1R1","P1Q1R4")
-exp.pretty.name = c("Original","Low Obs Error")
+exp.pretty.name = c("(a) Example site, Original","(c) Example site, Low Obs Error")
 pool.names=c("Cleaf","Csoil","AGB")
 i = 3 ## Stem
+time = 1985:2017
 for(ex in seq_along(exp.name)){
   load(paste0("OR_TS_",site,"_",exp.name[ex],".RData"))
-  plot(Yi[,i],
+  plot(time,Yi[,i],
        ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
@@
        main=exp.pretty.name[ex])
-  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
-  lines(CI.Na[3,],col="#b2df8a",lwd=2)
-  points(Yi[,i])
+  ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
+  lines(time,CI.Na[3,],col="#b2df8a",lwd=2)
+  points(time,Yi[,i])
   if(ex == 1) {
@@
 disturb.col.al <- tapply(disturb.col,1:5,FUN=ecoforecastR::col.alpha,0.7); disturb.col.al[1] = disturb.col[1]
+div.pretty.name = c("(b) Divergence, Original","(e) Divergence, Low Obs Error")
+div.reg <- list()
+div.class <- list()
+obs.class.re = obs.class; obs.class.re[obs.class.re ==4] = 3; obs.class.re[obs.class.re ==5] = 4; obs.class.re = obs.class.re + 1
 for(i in 1:2){
@@
     (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
-  div.reg    = lm(div~obs.tlo)
+  div.reg[[i]]    = lm(div~obs.tlo)
+  div.class[[i]]    = lm(div~obs.tlo + as.factor(obs.class))
 
@@
 #  abline(v=c(-1,1),lty=2)
-  plot(obs.tlo,div,col=disturb.col.al[obs.class],
+  plot(obs.tlo,div,col=disturb.col.al[obs.class.re],
        xlab="final AGB",ylab="Divergence",
-       pch=16)
+       pch=16, main=div.pretty.name[i])
   abline(h=c(-1,1),lty=2)
-  abline(div.reg,col="purple",lwd=3)
+  abline(div.reg[[i]],col="purple",lwd=3)
   abline(v=0,lty=2)
@@
 pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
+detect.pretty.name = c("(c) Probability of detection, Original","(f) Probability of Detection, Low Obs Error")
+dist.qb.abs <- list()
 for(i in 1:2){
   distrate = distrate = elist[[i]]$distrate
-  dist.qb.abs = glm(distrate ~ distabs,
+  dist.qb.abs[[i]] = glm(distrate ~ distabs,
                     na.action=na.omit,
@@
   )
-  plot(distabs,distrate,col=disturb.col.al[obs.class],
+  plot(distabs,distrate,col=disturb.col.al[obs.class.re],
        xlab="AGB disturbance magnitude",ylab="P(detect)",
-       pch=16)
-  lines(pred.seq$distabs,predict(dist.qb.abs,newdata=pred.seq,type="response"),col="purple")
+       pch=16, main=detect.pretty.name[i],xlim=c(0,17))
+  lines(pred.seq$distabs,predict(dist.qb.abs[[i]],newdata=pred.seq,type="response"),col="purple")
   if(i == 1) legend("topleft",legend=disturb.names,
@@
 ```
+In total we assimilated 365 example AGB timeseries from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively.
+Initial applications of the multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries.
+Filter divergence is a known issue for Kalman Filters, not something unique to our multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data.
+Figure \@ref(fig:ORorig)a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed.
+Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:ORorig)b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`).
+In other words, sites with low final biomass (i.e., large previous disturbance) showed a larger divergence and the divergence declined as the final biomass increased.
+Divergence also varied by disturbance type (p < `r format(anova(div.class[[1]])$"Pr(>F)"[2],digits=3,scientific=TRUE)`), with clear cuts having the lowest divergence for a given disturbance size and forest pests being lower than the fires and "other".
+Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:ORorig)c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`).
+However, detection was rarely successful for AGB disturbances that removed less than 7.5 $kgC/m^2$ and did not exceed 50% until disturbances were larger than `r bm=coef(dist.qb.abs[[1]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
 
-* RESULTS
-  * limited success under default parameters
-    * high frequency of filter divergence [summary fig, example divergence fig]
-    * divergence coverage original `r stats[[1]]$div.cov`
-    * divergence coverage low error `r stats[[2]]$div.cov`
-    * the smaller the observed biomass at the end of the timeseries, the larger the divergence [include scatterplot?]
-    * low rate of detected disturbance [calc stat?, hist? cdf?]
-    * what disturbances were detected tended to be associated with high pre-disturbance biomass and a large reduction in biomass (both absolute and proportional).
-Most commonly seen in dist class "red"
-  * needed to reduce obs and process error to work
-  * example "successful" scenario: P1Q1R4
-    * reduced observation error
-    * abs magnitude of disturbance is best predictor of assim prob. [fig]
-      * given magnitude, dist type n.s.  
-  * comparative analysis
+The causes of filter divergence are known to be some combination of model overconfidence (e.g., insufficient process error, Q) and high observation error (R) in the data.
+As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(table:QRexp)) to verify that without filter divergence the multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so.
+Figure \@ref(fig:ORorig)d-f show the results for Experiment 2, where we experimented with reducing the observation error, R.
+In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:ORorig)d).
+We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:ORorig)e).
+Divergence is no longer significantly affected by disturbance magnitude (p < `r format(summary(div.class[[2]])$coefficients[2,4],digits=3)`) and the only significant effect of disturbance type was a small reduction in divergence for clearcuts (p < `r format(summary(div.class[[2]])$coefficients[3,4],digits=3)`).
+Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:ORorig)f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
+
+Looking across all of our experiments, the coverage of our filter divergence statistic improved as the observation error decreased (Figure \@ref(fig:divExp)a).
+Increasing process error had little impact at the original observation error, but improved coverage once the observation error was decreased.
+The threshold disturbance magnitude required to detect disturbances also improved (i.e., became smaller) as the observation error decreased (Figure \@ref(fig:divExp)b) but was insensitive to process error, at least for the values we explored in our computational experiments.
+
+```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=0.75 }
+m = matrix(1:3,nrow=3,byrow = TRUE)
+layout(m)
+## Divergence
+div.cov = sapply(stats,"[[","div.cov")
+plot(Rmult[Rs],div.cov,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage",main="(a) Filter Coverage")
+legend("topright",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2,title="Q")
+
+#div.median = sapply(stats,"[[","div.median")
+#plot(Rmult[Rs],div.median,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab="Median Divergance",main="(b) Filter Divergence")
+#legend("topleft",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2)
+
+## Detection probability
+for(i in 3:6){
+  distrate = distrate = elist[[i]]$distrate
+  dist.qb.abs[[i]] = glm(distrate ~ distabs,
+                    na.action=na.omit,
+                    family=quasibinomial("logit")
+  )
+}
+thresh = sapply(dist.qb.abs,function(x){bm=coef(x); -bm[1]/bm[2]})
+plot(Rmult[Rs],thresh,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab=expression("Detection Threshold" ~ (kgC/m^2)),main="(c) Disturbance Detection Threshold")
+legend("topleft",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2)
+```

### 2023-01-12 supplemental.Rmd
Results: flushed out first draft text, added a couple citations. Methods: tried to tweak OR maps

@@
 param.desc = c("Light Use Efficiency","Fraction of GPP allocated to autotrophic respiration","Soil C residence time","Wood C residence time","Fraction of NPP allocated to leaves","NEE standard deviation","leaf C standard deviation","wood C standard deviation")
-param.table = cbind(param.names,param.desc,mle$par)
+param.table = cbind(param.names,param.desc,format(mle$par,digits=3,scientific=TRUE))
 colnames(param.table) <- c("Parameter","Description","Value")

### 2023-01-12 Paper.Rmd
Draft introduction

@@
 
-**Borrow text from CMS proposals**
+Researchers have devoted considerable time and resources to developing "bottom-up" terrestrial carbon (C) cycle data products at landscape- to global-scales (**CITE**, e.g.
+SOCCR2).
+Such efforts have drawn upon various combinations of remote sensing, inventory data, field-based sensors (e.g., eddy covariance), and process-based models with the aims of improving our understanding of C cycle processes, improving predictions of future changes in the C cycle, and improving monitoring, reporting, and verification (MRV) or C stocks and fluxes (**CITE**).
+MRV in particular is driven by a combination of local, national, and international reporting requirements (e.g.
+UN Framework Convention on Climate Change) and increasingly by the demand for natural climate solutions and carbon market certification (**CITE**).
 
-1. The context: MRV and predicting Terrestrial C
-2. Status quo: LULC * look up table
-  + very coarse, doesn't capture spatiotemporal varibility
-  + rightly captures that disturbance and land use change both critical to C accounting and, more importantly, a place where we can exert some control
-  + BULC
-  + state-and-transition modeling
-  + process-model projections
-3. Alternative: SDA
-  + models capture process
-  + data constrains (data driven, model as cov)
-  + data fusion
-  + seamless transition from reanalysis to nowcast to forecast [@Dietze2017a]
-  + Disturbance is exactly why you would want to nudge a model back to reality
-4. The problem: Conventional SDA doesn't capture jumps
-  + series of nudges
-  + potentially misattributes pools and fluxes
-  + "drags down" neighbors **no sims of this yet**
-5. The solution:
-  + assimilate P(dist)
-  + update C | dist
-6. This paper
-  + Illustration of problem
-  + Derivation Multinomial
-  + Simulated data experiments (known disturbance)
-  + Real-world example
+A particularly critical part of carbon MRV is the impact of land use, land use change, forestry (LULUCF) and natural disturbance (e.g.
+wildfire, pests, wind), which can have major impacts on the carbon cycle, quickly turning C sinks into large sources, and are responsible for a large fraction of interannual C cycle variability (CITE).
+They are also places where management and policy can have the greatest impact on carbon pools and fluxes, in terms of avoiding transitions, increasing resilience, and managing restoration and recovery after transitions occur.
+Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense, and there is mounting evidence that C stocks are likely to be more sensitive to climate effects on disturbance regimes than they are to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell, @Andrews2018].
+While there are a variety of approaches employed to monitor and understand the C cycle processes for intact ecosystems, when it comes to disturbance the status quo is based on detecting disturbances (e.g., via remote sensing) and then using a simple look up table to quantify C cycle impacts (CITE).
+While the closely related fields of disturbance detection and land use, land-use change, and forestry (LULUCF) monitoring have made considerable advances in recent years (CITE), the reliance on simple static look-up tables to quantify emissions is clearly missing considerable variability across space, time, and disturbance intensity.
+
+The alternative to look-up tables is to rely on process-based models to capture the spatiotemporal in the C cycle.
+However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), with major classes of disturbance being absent from most models (e.g., forest pests, [@Hicke2022, @DietzeMatthes2014]).
+Ultimately, MRV needs to be done with the disturbances that actually happened, not those that were predicted.
+On the other hand, prescribing disturbance in models ignores the uncertainty in disturbance detection and brings us back to using look-up tables to prescribe how the disturbance is represented in the model.
+
+State data assimilation offers a promising path forward for addressing these challenges and is already being to see increasing applications in other aspects of carbon MRV [@Baatz2021].
+The basic idea is that a process model is run numerous times to make a probabilistic prediction of the full carbon cycle (pools and fluxes) from one point in time to the next, sampling over the uncertainties about the current state of the system and possibly also parameter, driver, and process (residual) uncertainties.
+This multivariate prediction represents our (Bayesian) _prior_ understanding of the carbon cycle that is then updated, using Bayes' Theorem, based on observations of one or more these pools or fluxes [@Dietze2017a,@Wikle2007].
+In essence, data assimilation "nudges" the model back towards the data, with the size of the nudge proportional to the relative uncertainties in the model (prior) and the data (Likelihood).
+The resulting posterior estimate is able to fuse numerous data constraints on the different pools and fluxes, is robust to large amounts of missing observations in space, time, and pool/flux, and provides a rich uncertainty estimate.
+Finally, unlike monitoring-only systems, data assimilation systems can seamlessly transition from reconstructing the past (a.k.a.
+reanalysis), to real-time monitoring (a.k.a.
+nowcast), to predictive forecasts [@Dietze2017a].
+A handful of different terrestrial carbon data assimilation systems are now in opperations (e.g., CLM-DART, CCDAS, PEcAn), with applications at site [@Raiho2020], regional [@Raczka2021], continental [@Dokoohaki2022], and global [@Fox2022, @Rayner2005] scales.
+
+The potential advantages of applying state data assimilation to address the disturbance problem in carbon MRV are straightforward.
+Ensembles of model runs could capture the stochasticity in natural disturbance while also providing a process-based representation of both disturbance and carbon cycle processes and numerous possible sources of fine-scale heterogeneity (current vegetation conditions, climate, soils, etc.).
+The assimilation would then update the predictions based on what actually happened, formally accounting for the uncertainty in the disturbance observations, as well as fusing in numerous other data constraints (e.g., inventory, sensor, remote sensing).
+These observations may variously either reinforce or cast doubt upon the disturbance attribution, while also informing our understanding of pre- and post-disturbance heterogeneity.
+Data assimilation even has the potential to capture disturbances that are not represented explicitly in the models but are apparent in the observations.
+
+The problem is that, when it comes to assimilating disturbance, conventional data assimilation doesn't actually work (Figure \@ref(fig:enkf)).
+For reasons we'll explain in the _Methods_, instead of nudging a disturbed system to a disturbed state in the model, it nudges it to a "no man's land" in between the disturbed and undisturbed state that doesn't represent either state well.
+It can then take multiple iterations of the assimilation for these nudges to get the model back on track, in effect predicting gradual ecosystem degredation rather than a discrete disturbance and recovery, during which time this gradual decline will be misrepresented in MRV accounting (missing removals, fire combustion, etc.) and misinterpreted by the model (e.g.
+missing recruitment and respiration responses).
+
+Here we present a potential solution -- a novel assimilation algorithm that accounts for both the discrete nature of disturbance events and the continuous variability in disturbance magnitude.
+This assimilation is based on the idea of representing the ensemble forecast as a multimodal mixture module, with different modes capturing the undisturbed state and multiple alternative disturbances, and adding a Multinomial probability to the assimilation to allow it to discretely jump between states.
+In the following sections we will more formally explain the disturbance assimilation problem and derive our proposed solution.
+We will then apply this statistical model to both simulated data with known disturbances and real-world observations, making use of the Very Simple Ecosystem Model (VSEM) for predictions and remotely-sensed Landtrendr biomass and disturbance data products as observations, using the ponderosa pine ( _Pinus ponderosa_ ) ecosystems of central Oregon as a testbed [@Kennedy2018, @BayesianTools].
+Overall, this manuscript aims to provide a proof-of-concept for the idea of disturbance assimilation, laying the foundation for future applications and algorith improvements.
 

### 2023-01-12 Paper.Rmd
knitr debugging

@@
 output:
-  pdf_document: 
+  bookdown::pdf_document2: 
     fig_caption: yes
@@
     keep_tex: true
-  html_document: 
+  bookdown::html_document2: 
     fig_caption: yes
@@
 They are also places where management and policy can have the greatest impact on carbon pools and fluxes, in terms of avoiding transitions, increasing resilience, and managing restoration and recovery after transitions occur.
-Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense, and there is mounting evidence that C stocks are likely to be more sensitive to climate effects on disturbance regimes than they are to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell, @Andrews2018].
+Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense, and there is mounting evidence that C stocks are likely to be more sensitive to climate effects on disturbance regimes than they are to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell2020, @Andrews2018].
 While there are a variety of approaches employed to monitor and understand the C cycle processes for intact ecosystems, when it comes to disturbance the status quo is based on detecting disturbances (e.g., via remote sensing) and then using a simple look up table to quantify C cycle impacts (CITE).
@@
 The alternative to look-up tables is to rely on process-based models to capture the spatiotemporal in the C cycle.
-However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), with major classes of disturbance being absent from most models (e.g., forest pests, [@Hicke2022, @DietzeMatthes2014]).
+However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), with major classes of disturbance being absent from most models (e.g., forest pests, [@Hicke2012, @DietzeMatthes2014]).
 Ultimately, MRV needs to be done with the disturbances that actually happened, not those that were predicted.

### 2023-01-12 discussion.Rmd
knitr debugging

@@
     * Efficient (semi-)analytic approximations or LUT
-    * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly [@McCabe]
+    * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly [@McCabe2019]
   * Multisite needs a way of estimating $\rho$ by site with cross site correlations, rather than unique $\rho$ for all permutations

### 2023-01-12 methods.Rmd
knitr debugging

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Methods
@@
 
-```{r concept, echo=FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
+```{r concept, echo=FALSE, message = FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
 Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities.
@@
 
-$$c_{A|k} = Multinomial\left( n_k, {\vec{\delta p_t}}\over{\sum \vec{\delta p_t}}  \right)$$
+$$c_{A|k} = Multinomial\left( n_k, \frac{\vec{\delta p_t}}{\sum \vec{\delta p_t}}  \right)$$
 
@@
 
+m = matrix(1:4,nrow=2,byrow = TRUE)
+layout(m)
+
 ## Cover
@@
 }
-if(!exists("dis04")) 
+if(!exists("dis04"))  dis04 = raster::raster("dis04.tif")
 dname = c("Intact","Other","Cut","Fire","Pest")
@@
 Finally, based on preliminary results we performed a series of additional assimilation experiments where we manually altered the observation error, R, or process error, Q, to assess the sensitivity of our results.
-Table (\@ref(table:QRexp)) lists the multipliers that were used to adjust R and Q.
+Table (\@ref(tab:QRexp)) lists the multipliers that were used to adjust R and Q.
 The skill of these assimilation experiments were assessed the same as the original.

### 2023-01-12 results.Rmd
Draft introduction
knitr debugging

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Results
@@
 ```{r sssd, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
+
+m = matrix(1:3,nrow=3,byrow = TRUE)
+layout(m)
+
 load("exp2.Rdata") # Generated by SingleSite_SingleDisturb_MultiPool.Rmd
@@
 
-```{r multiPseudo, echo=FALSE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.height=4,fig.width=3}
+```{r multiPseudo, echo=FALSE, message = FALSE, fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.height=4,fig.width=3}
 
@@
 Best fit parameters are given in Supplemental Table \@ref(tab:parameters).
-The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibrate)).
+The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibrate) ).
 Achieving this fit came at the cost of nontrivial systematic biases in both leaf carbon (0.064 $kgC/m^2$, 13%) and NEE (0.528 $kgC/m^2/year$, 69%).
@@
 
-```{r ORorig, echo=FALSE,fig.asp=0.75,fig.cap="Filter divergence and disturbance detection for central Oregon test case.
+```{r ORorig, echo=FALSE, fig.asp=0.75, fig.cap="Filter divergence and disturbance detection for central Oregon test case.
 Left column is the default configuration while the right column is a reduced observation error experiment (Exp 2).
@@
 Filter divergence is a known issue for Kalman Filters, not something unique to our multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data.
-Figure \@ref(fig:ORorig)a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed.
-Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:ORorig)b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`).
+Figure \@ref(fig:ORorig) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed.
+Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:ORorig) b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`).
 In other words, sites with low final biomass (i.e., large previous disturbance) showed a larger divergence and the divergence declined as the final biomass increased.
 Divergence also varied by disturbance type (p < `r format(anova(div.class[[1]])$"Pr(>F)"[2],digits=3,scientific=TRUE)`), with clear cuts having the lowest divergence for a given disturbance size and forest pests being lower than the fires and "other".
-Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:ORorig)c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`).
+Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:ORorig) c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`).
 However, detection was rarely successful for AGB disturbances that removed less than 7.5 $kgC/m^2$ and did not exceed 50% until disturbances were larger than `r bm=coef(dist.qb.abs[[1]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
@@
 The causes of filter divergence are known to be some combination of model overconfidence (e.g., insufficient process error, Q) and high observation error (R) in the data.
-As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(table:QRexp)) to verify that without filter divergence the multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so.
-Figure \@ref(fig:ORorig)d-f show the results for Experiment 2, where we experimented with reducing the observation error, R.
-In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:ORorig)d).
-We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:ORorig)e).
+As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(tab:QRexp)) to verify that without filter divergence the multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so.
+Figure \@ref(fig:ORorig) d-f show the results for Experiment 2, where we experimented with reducing the observation error, R.
+In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:ORorig) d).
+We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:ORorig) e).
 Divergence is no longer significantly affected by disturbance magnitude (p < `r format(summary(div.class[[2]])$coefficients[2,4],digits=3)`) and the only significant effect of disturbance type was a small reduction in divergence for clearcuts (p < `r format(summary(div.class[[2]])$coefficients[3,4],digits=3)`).
-Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:ORorig)f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
+Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:ORorig) f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
 
@@
 ```
+
+<!-- Future direction: rerun using disturbance data product as priors on disturbance probability -->

### 2023-01-26 Paper.Rmd
Fix multiple citation style

@@
 They are also places where management and policy can have the greatest impact on carbon pools and fluxes, in terms of avoiding transitions, increasing resilience, and managing restoration and recovery after transitions occur.
-Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense, and there is mounting evidence that C stocks are likely to be more sensitive to climate effects on disturbance regimes than they are to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell2020, @Andrews2018].
+Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense, and there is mounting evidence that C stocks are likely to be more sensitive to climate effects on disturbance regimes than they are to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell2020; @Andrews2018].
 While there are a variety of approaches employed to monitor and understand the C cycle processes for intact ecosystems, when it comes to disturbance the status quo is based on detecting disturbances (e.g., via remote sensing) and then using a simple look up table to quantify C cycle impacts (CITE).
@@
 The alternative to look-up tables is to rely on process-based models to capture the spatiotemporal in the C cycle.
-However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), with major classes of disturbance being absent from most models (e.g., forest pests, [@Hicke2012, @DietzeMatthes2014]).
+However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), with major classes of disturbance being absent from most models (e.g., forest pests, [@Hicke2012; @DietzeMatthes2014]).
 Ultimately, MRV needs to be done with the disturbances that actually happened, not those that were predicted.
@@
 The basic idea is that a process model is run numerous times to make a probabilistic prediction of the full carbon cycle (pools and fluxes) from one point in time to the next, sampling over the uncertainties about the current state of the system and possibly also parameter, driver, and process (residual) uncertainties.
-This multivariate prediction represents our (Bayesian) _prior_ understanding of the carbon cycle that is then updated, using Bayes' Theorem, based on observations of one or more these pools or fluxes [@Dietze2017a,@Wikle2007].
+This multivariate prediction represents our (Bayesian) _prior_ understanding of the carbon cycle that is then updated, using Bayes' Theorem, based on observations of one or more these pools or fluxes [@Dietze2017a; @Wikle2007].
 In essence, data assimilation "nudges" the model back towards the data, with the size of the nudge proportional to the relative uncertainties in the model (prior) and the data (Likelihood).
@@
 nowcast), to predictive forecasts [@Dietze2017a].
-A handful of different terrestrial carbon data assimilation systems are now in opperations (e.g., CLM-DART, CCDAS, PEcAn), with applications at site [@Raiho2020], regional [@Raczka2021], continental [@Dokoohaki2022], and global [@Fox2022, @Rayner2005] scales.
+A handful of different terrestrial carbon data assimilation systems are now in opperations (e.g., CLM-DART, CCDAS, PEcAn), with applications at site [@Raiho2020], regional [@Raczka2021], continental [@Dokoohaki2022], and global [@Fox2022; @Rayner2005] scales.
 
@@
 In the following sections we will more formally explain the disturbance assimilation problem and derive our proposed solution.
-We will then apply this statistical model to both simulated data with known disturbances and real-world observations, making use of the Very Simple Ecosystem Model (VSEM) for predictions and remotely-sensed Landtrendr biomass and disturbance data products as observations, using the ponderosa pine ( _Pinus ponderosa_ ) ecosystems of central Oregon as a testbed [@Kennedy2018, @BayesianTools].
+We will then apply this statistical model to both simulated data with known disturbances and real-world observations, making use of the Very Simple Ecosystem Model (VSEM) for predictions and remotely-sensed Landtrendr biomass and disturbance data products as observations, using the ponderosa pine ( _Pinus ponderosa_ ) ecosystems of central Oregon as a testbed [@Kennedy2018; @BayesianTools].
 Overall, this manuscript aims to provide a proof-of-concept for the idea of disturbance assimilation, laying the foundation for future applications and algorith improvements.

### 2023-01-31 discussion.Rmd
Discussion: Simulated data and OR test case text

@@
 
-* Interpretation
-  * TBD based on results
+## Algorithm development and simulated data experiments
 
-* Issues
-  * Need a more detail exploration of rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
-  * Would like to add look-back capabilities (ensemble backward smoothing) to update assignments based on more recent info
-    * related: with multi-PFT need to check for, and protect against, long term PFT leakage
-  * Computational demand it high for problems with many grid cells
-    * Efficient (semi-)analytic approximations or LUT
-    * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly [@McCabe2019]
-  * Multisite needs a way of estimating $\rho$ by site with cross site correlations, rather than unique $\rho$ for all permutations
-    * for $\rho$ handle the covariances of nearby sites -- disturbance in one usually increases the probability of disturbance in others
-    * similarly, need to be able to compose Multisite $\mu_f$ and $P_f$, from individual site values with reasonable approximations of cross-[site x pool] correlations
+At a high-level, the simulated data experiments demonstrate that the Multinomial filter is able to address the fundamental problem that traditional data assimilation encounters when faced with disturbance events.
+Rather than nudging the posterior state stranded into a "no man's land" between the disturbed and undisturbed states, the Multinomial successfully makes a discrete jump in state.
+Through the simulated data experiments we also demonstrated: (1) That the algorithm is robust to false positives -- once in the disturbed state it does not interpret further dips in LAI as new disturbances since those data are consistent with the observation error around the model's predicted recovery trajectory; (2) That the algorithm is successfully able to fuse observations of multiple pools simultaneously; (3) That it can distinguish among multiple types of disturbance; (4) That it can accommodate disturbances that result in a discrete change in plant functional type (e.g., forest to grassland or agriculture); and (5) That the sharing of information across space in multi-site assimilations, which normally increases precision and handles missing observations, is not thrown off by disturbance events (e.g., reducing biomass in undisturbed sites).
+From a data assimilation perspective, the algorithm works as intended and is superior to conventional alternatives at detecting disturbance.
 
-* Discussion of similar approaches
-[@Ye2020]
+From a disturbance detection perspective, three things set the Multinomial assimilation apart from other conventional time-series based disturbance detection algorithms (as opposed to static image classifications), such as Landtrendr [@KENNEDY2010], CCDC (Continuous Change Detection and Classification) [@ZHU2014; @Pasquarella2017; @ZHU2020] and BFAST (Break detection For Additive Season and Trend) [@VERBESSELT2012].
+First, most other algorithms flag disturbances as departures from the predicted value -- a rejection of the "status quo" hypothesis.
+With the multinomial, we are instead treating disturbance detection as a choice between competing hypotheses, assessing how much weight to place on the "status quo" versus different alternative disturbance options.
+Second, most of these algorithms rely on a moving-window approach as part of their "status quo" prediction, which requires accessing and processing a time-series of data from the past every time an prediction is updated.
+The iterative algorithms that form the basis for data assimilation (e.g., Kalman Filter) have on occasion been applied to disturbance detection (using statistical, rather than process-based model) and have demonstrated greater computational efficiency [@Ye2020].
+@Ye2020 avoid the "no man's land" nudge problem by the simple expedient of performing threshold-based disturbance detection prior to assimilation and not assimilating disturbed sites.
+Third, other algorithms are relying on simple statistical models for their predictions of the status quo, such as harmonic time-series models, while the Multinomial model is relying on the predictions of a process-based ecosystem model.
+This has three distinct advantages: (1) We are leveraging greater process understanding in the status quo prediction, potentially making the algorithm more robust in the face of extreme events that are not disturbances (e.g., unusually cold or dry) but which alter the magnitude or phenology of seasonal cycles; (2) While not included in the simple "toy" model used in these analyses, process-based models also have the potential to leveraging greater process understanding in predicting the likelihood of disturbances, for example by accounting for fuel loads, fuel moisture, and ignition rates in predicting the likelihood of fire; (3) Because process-based models are predicting multiple pools and fluxes that can be compared to multiple distinct types of data, the Multinomial algorithm provides a natural way to fuse multiple types of data that would either reinforce a disturbance detection (if all sources are pointing in the same direction) or call it into question.
+While conventional disturbance detection algorithms are overwhelmingly based on multispectral remote sensing from either a single sensor or a pre-harmonized fusion of multiple sensors [e.g., Harmonized Landsat Sentinel-2, @CHEN2021; @SHANG2022], data assimilation approaches can simultanously integrate a wide range of relevant data types, for example lidar, microwave, thermal, solar-induced fluorescence, and field data.
+Data assimilation can also integrate a range of optical sensors (both multispectral and hyperspectral) based on their spectral response functions, rather than requiring pre-harmonization [@Shiklomanov2021; @Zhang2023]).
 
-* Calibration was "quick and dirty" because the focus was on disturbance detection not default dynamics.
-For actual carbon monitoring and forecasting we would recommend a more extensive multi-site Bayesian calibration so that parameter uncertainty can be propagated [CITE istem 2018, dietze 2017], preferably employing hierarchical approaches to also capture parameter heterogeneity [CITE istem 2021].
-Similarly, would recommend a more robust estimation and propagation of process error as part of the data assimilation approach itself [CITE raiho].
+## Oregon case study
 
-* Implications  (borrow text from CMS proposals)
-  * Improving MRV
-  * Recovery forecasting
-  * Disturbance nowcasting
-    * requires reducing temporal resolution
+The most obvious issue with the initial application of the Multinomial filter to real-world data was the high incidence of filter divergence -- of the model becoming overconfident in itself and thus diverging from the observations.
+Filter divergence is not unique to this algorithm, but something that can occur in traditional filters as well. 
+Indeed, there's no reason to belive that filter divergence is any more common in the Multinomial filter than in more traditional filters, such as EnKF, because in the absence of disturbance the Multinomial filter behaves exactly as EnKF would.
+The underlying causes of filter divergence are a result of either an underaccounting of model uncertainty or an overestimation of observation error in the data, both of which are likely at play in this example.
+On the model side, forecast uncertainty is associated with five factors: initial condition uncertainty, driver uncertainty, parameter uncertainty, parameter heterogenity, and process error [@Dietze2017b].
+Since our goal here was a simple proof of concept, rather than a real world application, we did not formally account for uncertainty in model drivers (i.e., meteorology) or parameters. 
+Alternative meteorological driver data products exist that account for uncertainties, such as the ERA5 ensemble reanalysis product [@Hersbach2020], and methods exist for propagating the uncertainties associated with downscaling such regional data products [@Rollinson2017].
+Similarly, for actual carbon monitoring and forecasting we would recommend a more extensive multi-site Bayesian calibration so that parameter uncertainty can be propagated [@Fer2018; @Dietze2017a], preferably employing hierarchical approaches to also capture parameter heterogeneity (a.k.a.
+random effects), which can be substantial [@Fer2021].
+Likewise, we would recommend a more robust estimation and propagation of process error as part of the data assimilation approach itself [@Raiho2020].
+Ironically, the success of the existing model calibration at predicting AGB at our single calibration sites contributed to the underestimation of process error, since we failed to account for the uncertainties associated with extrapolating this single calibration to other sites across the landscape.
+Given the strong topographic and climatic gradients of the central Cascades, landscape heterogeneity in model performance is likely nontrivial.
 
-* Future directions
+In addition to underestimating model uncertainty, the AGB data product we are using, which relies on a single fixed residual error estimate, is likely overestimating the observation uncertainty. 
+First, the assumption of constant variance is unlikely to be true because AGB is non-negative, which means the current interval estimates for low biomass sites include a substantial probability of negative biomass.
+More often the uncertainties in AGB are heteroskedastic and increasing in absolute magnitude as AGB increases, a property shared with the allometeric models used to estimate AGB in the field calibration data [@Dietze2008].
+This heteroskedasticity particularly important for disturbance detection because it means we are likely significantly overestimating the observation error associated with low biomass (disturbed) conditions, contributing directly to the Multinomial model's inability to distinguish such biomass drops from observation error.
+Second, the AGB data product is does not distinguish random versus systematic errors [@Cameron2022], and in particular is not accounting for the spatial and temporal autocorrelation in observation errors.
+In reality, a site with higher than predicted AGB in one year is likely to have higher than predicted AGB the next year (positive temporal autocorrelation) and adjacent pixel are also likely to have higher than predicted AGB (positive spatial autocorrelation).
+These positive autocorrelations mean we are likely overestimating the uncertainty associated with change detection, since a substantial portion of this systematic observation error should cancel out [@Kennedy2023].
+Overall, the failures of the default calibration, which ignored driver uncertainty, parameter uncertainty, and parameter heterogeneity while understimating process error and overestimating observation error, provide an illustrative case study in the care that must be taken when applying data assimilation approaches in general, and the Multinomial filter in particular, to real-world data
+
+In contrast to our "off-the-shelf" application of the Multinomial filter, our computational experiments demonstrated that when we increase the model uncertainty and reduce the observation uncertainty that we significantly reduced the rate of filter divergence and were able to reliably detect all but the smallest disturbance events.
+Indeed, filter divergence declined, and disturbance detection increased, systematically as a function of the magnitude of the observation error. 
+Somewhat surprisingly the computational experiments suggested that the Multinomial filter tended to be more sensitive to observation error than to model error, and suggesting that applications should prioritize making sure that observation errors are correctly accounted.
+Indeed, doing a better job at propagating model uncertainties correctly only becomes important once the observation errors are constrained.
+
+## Future Directions: Current algorithm
+
+It is also worth noting that, unlike in the simulated data experiment, our real-world application only relied on a single data constraint.
+If we had leveraged additional data constraints, such as LAI, we would have gained additional confidence in our ability to detect real disturbances, not just because LAI data products often have lower observation errors but because the addition of additional data reduces the overall observation error and the positive correlation between LAI and AGB would allow the assimilation algorithm to borrow strength indirectly (i.e.
+observations of LAI would help constrain AGB and observations of AGB would help constrain LAI).
+In practice this is nontrivial because LAI data products based on different observations (e.g. MODIS) have a different spatial resolution, which would need to be addressed. 
+Alternatively, a data product based on LandSAT would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e.
+errors in data products based on the same underlying spectra would not be independent).
+
+* Lessons learned from computational experiments
+
+* OREGON: FUTURE DIR
+
+* multiple constraints
+    * also would benifit from more products (e.g. LAI)
+
+  * disturbance products are 'seeing' more than is in AGB product alone
+
   * Disturbance priors need to be spatially heterogenous and evolve temporally
@@
     * shared spatial information (e.g. dispersal kernels)
-  * Scaling up -> spatially implicit
-    * ED2
-    * Contagion
 
-* OREGON: FUTURE DIR
-  * disturbance products are 'seeing' more than is in AGB product alone
-    * also would benifit from more products (e.g. LAI)
+## Future Directions: Algorithm Development
+
   * would be good to add 'memory' to algorithm so further evidence of disturbance reinforces initial detection (similar to BLC); would allow reduction in time step and greater sensor fusion
@@
         * rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
+
+* Future directions
+
+  * Scaling up -> spatially implicit
+    * ED2
+    * Contagion
+  * Scaling up -> emulation
+    * Applied at scale full model would be both computationally prohibitive and highly redundant (similar sites, such as spatially adjacent, are going to have very similar inputs and output)
+    * run full Bayesian model for a subset of locations stratified by relevant variables (IC, observed data, possibly also met, soils, etc.)
+    * build ML emulator to predict outcome of full Bayesian model
+    * use ML emulator to predict full model; interpolate in env/obs space -> map to physical space
+
+* Issues
+  * Need a more detail exploration of rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
+  * Would like to add look-back capabilities (ensemble backward smoothing) to update assignments based on more recent info
+    * related: with multi-PFT need to check for, and protect against, long term PFT leakage
+  * Computational demand it high for problems with many grid cells
+    * Efficient (semi-)analytic approximations or LUT
+    * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly [@McCabe2019]
+  * Multisite needs a way of estimating $\rho$ by site with cross site correlations, rather than unique $\rho$ for all permutations
+    * for $\rho$ handle the covariances of nearby sites -- disturbance in one usually increases the probability of disturbance in others
+    * similarly, need to be able to compose Multisite $\mu_f$ and $P_f$, from individual site values with reasonable approximations of cross-[site x pool] correlations
+
+## Conclusion
+
+* Implications  (borrow text from CMS proposals)
+  * Improving MRV
+  * Recovery forecasting
+  * Disturbance nowcasting
+    * requires reducing temporal resolution

### 2023-01-31 discussion.Rmd
Future Directions: Current algorithm

@@
 
-It is also worth noting that, unlike in the simulated data experiment, our real-world application only relied on a single data constraint.
-If we had leveraged additional data constraints, such as LAI, we would have gained additional confidence in our ability to detect real disturbances, not just because LAI data products often have lower observation errors but because the addition of additional data reduces the overall observation error and the positive correlation between LAI and AGB would allow the assimilation algorithm to borrow strength indirectly (i.e.
+In the following subsections we discuss ways that the proof-of-concept presented here could be applied and built upon.
+We start by addressing future directions that make use of the current algorithm "as is", discussing ways that we could better leverage multiple data constraints, improved disturbance priors, and existing disturbance data products.
+In terms of multiple data constraints, it is worth noting that, unlike in the simulated data experiment, our real-world application only relied on a single data constraint.
+If we had leveraged additional data constraints, we would have gained additional confidence in our ability to detect real disturbances.
+One way to to do this is to bring in additional data on the same pool, in this case AGB, reducing the overall observation error.
+This could be done through some combination of field data [e.g., Forest Inventory and Analysis, CITE], lidar [e.g., GEDI, CITE], microwave vegetation optical depth [e.g., SMAP, CITE Koenings], or radar [e.g., BIOMASS, CITE].
+This was not done in this example because data products based on different observations have a different spatial resolutions, which would need to be addressed.
+Fusing in a data assimilation framework is well established but was judged beyond the scope of the current study because it requires running a fully spatial assimilation rather than a simple site-scale timeseries model.
+Basically, one either needs to run the assimilation at a fine spatial resolution and use an observation operator to map many model grid cells to one observation, or one could run the assimilation at a coarse spatial resolution and use an observation operator to map one model grid cell to many observations [@Dietze2017a].
+
+Another way would be to bring in additional data on different pools, such as LAI, which often has lower observation errors.
+In addition to the direct increase in data volumes, the positive correlation between LAI and AGB would allow the assimilation algorithm to borrow strength indirectly (i.e.
 observations of LAI would help constrain AGB and observations of AGB would help constrain LAI).
-In practice this is nontrivial because LAI data products based on different observations (e.g. MODIS) have a different spatial resolution, which would need to be addressed. 
-Alternatively, a data product based on LandSAT would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e.
+Existing LAI data products [e.g., MODIS, CITE] have the same spatial mismatch issue discussed above.
+Alternatively, a data product based on LandSAT would be at the same resolution but would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e.
 errors in data products based on the same underlying spectra would not be independent).
+In addition to bringing in additional data constratins on different pools, we could also constrain our disturbance estimates using observations on different ecosystem fluxes and processes, such as eddy covariance (carbon, water, and energy fluxes), solar induced fluorescence (a proxy for gross primary productivity), thermal imaging (a proxy for water stress and evapotranspiration).
 
-* Lessons learned from computational experiments
+In addition to constraining our estimates of disturbance through the ecosystem state variables, we can also improve them by improving our priors on the disturbance probability. 
+In our simple case study we relied on static and spatially-homogenous prior probabilities on different disturbance rates, an approach somewhat similar to the simplest state-and-transition models of land use and disturbance [@Daniel2016].
+More complex state-and-transition models are possible that allow these priors to be spatially heterogenous (e.g., relying on spatial covariates) or time evolving (e.g., relying on temporal covariates).
+Dynamic models are also possible that rely on the current state of a pixel (e.g., increasing fuel loads increase the probability of fire), the state of adjacent pixels (e.g., spatially contagious disturbances such as fire), or even those involving more distant spatial relations (e.g., dispersal kernels of biotic pests and pathogens) (CITATIONS).
+In addition to relying on external state-and-transition models, one of the appeals of working with process-based ecosystem models is that many of they include submodules that represent different types of disturbance internally (e.g., fire, pests).
+In these cases we can either extract the model's internally predicted probabilities or simply rely on the frequency of events predicted across the ensemble of model runs.
 
-* OREGON: FUTURE DIR
+The other way to improve upon the prior estimates of disturbance probability is to rely on external disturbance data that's not part of the assimilation itself. 
+For example, in our Oregon case study it is clear that the Landtrendr disturbance product is "seeing" more about disturbance than is being expressed in the AGB product alone, as the the disturbance product is looking for different information (e.g., weighting bands differently, training to distinguish among types of disturbance not just a loss of biomass).
+Therefore, one alternative to relying on models to predict disturbance priors is to use the disturbance product and its classification error statistics to specify priors itself to identify the error.
+This approach could be particularly valuable for distinguishing among alternative disturbance types (e.g., forest wildfire, forestry, and conversion to agriculture or development), especially for disturbance products that make use of bands outside of the normal optical range (e.g., thermal-based active fire products).
+Applying this approach was not done in our test case because we wanted to keep the disturbance product separate for purposes of validation, but would be relatively straightforward to implement.
+It's important to note that when using external disturbance data as priors we're still not prescribing disturbances deterministically, during the assimilation observations about ecosystem pools and fluxes would thus either reinforce these products or call them into question.
+Finally, it is worth nothing that relying on processed-based models and external disturbance data products are not mutually exclusive.
+The fusion of disturbance products and disturbance models could be part of the future development of the Multinomial filter itself (see next subsection), but may be simpler to handle as part of the prior construction.
+For example, one could start with the models as priors and then update these using one or more disturbance product observations [e.g., using the Bayesian Updating of Land Cover (BULC) algorithm, @CARDILLE2016; @Crowley2019] and then use these posteriors as the priors in the data assimilation.
 
-* multiple constraints
-    * also would benifit from more products (e.g. LAI)
+## Future Directions: Computational challenges
 
-  * disturbance products are 'seeing' more than is in AGB product alone
-
-  * Disturbance priors need to be spatially heterogenous and evolve temporally
-    * Connect to state-and-transition literature
-    * shared spatial information (e.g. dispersal kernels)
-
-## Future Directions: Algorithm Development
-
-  * would be good to add 'memory' to algorithm so further evidence of disturbance reinforces initial detection (similar to BLC); would allow reduction in time step and greater sensor fusion
-
-* Run just the Assimilation step for simulated disturbance that vary in magnitude, process error, observation error, etc to detect sensitivity
-* Approximation of not simulating disturbed ensemble members and relying SOLEY on ensemble adjustment
 * Analytical approximation of sequential beta-binomial then KF vs full joint model
 
-Leftover text from Methods:
-  + x SS Multi-Disturbance - distinct
-  + x SS Multi-Disturbance w/ distinct, PFT switching
-      * mention parallel PFT ensembles
-      * need to protect against long-term PFT leakage; related to more general need to "look back"
-  + x Simple Multi-site: spatial cov
-  + Future
-      * SSMD - initially overlapping but diverge
-        * how to handle updating of assignment based on new information?
-      * SSMD - transition probabilities change with state
-        * rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
-
-* Future directions
-
-  * Scaling up -> spatially implicit
-    * ED2
-    * Contagion
 * Scaling up -> emulation
@@
 
-* Issues
-  * Need a more detail exploration of rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
-  * Would like to add look-back capabilities (ensemble backward smoothing) to update assignments based on more recent info
-    * related: with multi-PFT need to check for, and protect against, long term PFT leakage
-  * Computational demand it high for problems with many grid cells
-    * Efficient (semi-)analytic approximations or LUT
-    * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly [@McCabe2019]
+## Future Directions: Algorithm Development
+
   * Multisite needs a way of estimating $\rho$ by site with cross site correlations, rather than unique $\rho$ for all permutations
@@
     * similarly, need to be able to compose Multisite $\mu_f$ and $P_f$, from individual site values with reasonable approximations of cross-[site x pool] correlations
+
+* would be good to add 'memory' to algorithm so further evidence of disturbance reinforces initial detection (similar to BLC); would allow reduction in time step and greater sensor fusion
+  * example case: initially overlapping but diverge
+  * Would like to add look-back capabilities (ensemble backward smoothing) to update assignments based on more recent info
+
+* PFT switching
+      * mention parallel PFT ensembles
+      * need to protect against long-term PFT leakage; related to more general need to "look back"
+
+* Scaling up -> spatially implicit
+  * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly 
+    * ED2
+    * Contagion [@McCabe2019]
 

### 2023-02-01 discussion.Rmd
Finished first pass at Discussion

@@
 
-* Analytical approximation of sequential beta-binomial then KF vs full joint model
+One of the challenges in scaling up the Multinomial filter is the computational cost involved.
+Unlike a traditional EnKF, the Analysis step of the data assimilation does not have a full analytically solution, and instead needs to be evaluated numerically, for example through standard Bayesian MCMC approaches as was done in this paper.
+However, MCMC is too computationally costly to apply using brute force on anything larger than an individual landscape.
+Here we discuss two possible approximations to the full algorithm, one analytical and the other based on model emulation.
+Analytically, instead of solving the full Multinomial filter jointly, it may be possible to approximate the full model by decompose the filter sequentially. 
+First, one would analytically apply EnKF to each alternative disturbance state, analytically generating independent posteriors for each $X^L_t$.
+Next, one would then apply a Dirlichet-Multinomial model to the problem of updating $\vec{\rho_t}$, treating $X^L_t$ as known rather than jointly estimated.
+The math for this approach has yet to be fully flushed out, and it remains to be seen how well this approximation works in practice and what information would be lost by solving the model conditionally rather than jointly.
 
-* Scaling up -> emulation
-    * Applied at scale full model would be both computationally prohibitive and highly redundant (similar sites, such as spatially adjacent, are going to have very similar inputs and output)
-    * run full Bayesian model for a subset of locations stratified by relevant variables (IC, observed data, possibly also met, soils, etc.)
-    * build ML emulator to predict outcome of full Bayesian model
-    * use ML emulator to predict full model; interpolate in env/obs space -> map to physical space
+An alternative to analytical approximation is to rely on machine learning approaches to build an emulator (a.k.a.
+surrogate model) [CITE].
+While emulators are often used to approximate process-based models, predicting what the model would have output given a set of inputs, here we want to construct an emulator to that Analysis step, predicting what the Multinomial filter would output (probability of disturbance, ecosystem pool means and variances) as a function of the Forecast model state and the observed data constraints.
+** Find citations for papers that have emulated other SDA**.
+Emulators addresses not only the computational cost of scaling up but also the observation that similar sites are going to make similar predictions (e.g., adjacent LANDSAT pixels), and thus fully gridded models are going to perform highly redundant calculations.
+Emulators still require running the model and analysis (in this case the Multinomial filter), but rather than evaluating every location the filter is run for a subset of locations stratified by relevant variables (initial conditions, observed data constraints, and posssibly also covariates like climate, soils, etc.).
+Flexible machine learning approaches are then used to build an emulator that predicts the outcome of full Bayesian model.
+For places where the full models was not run the emulator is then used to predict the full model, which in essence involves interpolating in environmental space and mapping this back to physical space.
 
@@
 
-  * Multisite needs a way of estimating $\rho$ by site with cross site correlations, rather than unique $\rho$ for all permutations
-    * for $\rho$ handle the covariances of nearby sites -- disturbance in one usually increases the probability of disturbance in others
-    * similarly, need to be able to compose Multisite $\mu_f$ and $P_f$, from individual site values with reasonable approximations of cross-[site x pool] correlations
+In this final section we discuss three future directions for improving upon the Multinomial filter: accounting for space when estimating the probability of disturbance, $\vec{\rho_t}$; adding memory to the algorithm to be able to better detect disturbance occurrence and type; and scaling up using spatially implicit approaches that track fractional area.
 
-* would be good to add 'memory' to algorithm so further evidence of disturbance reinforces initial detection (similar to BLC); would allow reduction in time step and greater sensor fusion
-  * example case: initially overlapping but diverge
-  * Would like to add look-back capabilities (ensemble backward smoothing) to update assignments based on more recent info
+In terms of spatial disturbance probability, the current filter does estimate a whole vector of probabilities, one for each site, and thus estimates will include a posterior covariance structure across sites, but the mdoel is not written in a way that explicitly changes the probability of disturbance based on spatial adjacency.
+In practice, almost all types of disturbance involve some degree of shared risk, if not true spatial contagion, whereby a disturbance at one site usually increases the probability of disturbance in near-by sites.
+Accounting for this could be done by adding a hierarchical layer to the estimation of $\rho$ that includes a spatial covariance structure.
+Because $\rho$ is bound between zero and one, this could be done by including a link function (e.g., logit) and modeling the covariance as a multivariate Normal with a standard distance-dependent covariance model (e.g., exponential, Gaussian, Matern) [@Banerjee2014].
+In practice, estimating larger spatial covariance matrices is computationally demanding and often more efficiently approximated using spatial basis functions and random effects [@Wikle2019].
 
-* PFT switching
-      * mention parallel PFT ensembles
-      * need to protect against long-term PFT leakage; related to more general need to "look back"
+As for the idea of incorporating memory into the Multinomial filter, one limitation of the existing approach is that (like most data assimilation filters) it is strictly Markovian - states are updated solely based on the current model state and current observations.
+Imagine the case where an anomalously low value is observed for AGB but at the next time point AGB returns to normal.
+At the time that the low observation is made it may trigger the Multinomial filter to predict a high probability of disturbance, but after the next observation becomes available it becomes much more likely that the low value was simply observation error.
+In this case what we want is the ability not just to restore the AGB to the undisturbed state, but also to jump back to the previous observation and assign it to be undisturbed as well.
+As another example, imagine the case where a disturbance does occur but the observations are compatible with multiple types of disturbance (e.g., clear cut versus conversion to agriculture).
+The algorithm may assign both a 50/50 probability at that time point, and then at the next time point it may update that assignment based on new information (e.g., the new year's vegetation is herbaceous).
+Rather than interpreting the 50% probability assigned to clear cut as a new transition at the second time point, what we'd really like to do is jump back in time and say, based on new information, that we're now much more confident that the original disturbance was agricultural conversion.
+Furthermore, because some PFT transitions are not easily reversible (e.g., a grassland cannot shift to a mature forest in one year), this capability to look back would also help protect against the risk that assigning a non-zero probability to a PFT transition could cause more and more of the ensemble to "leak" into that state.
+Implementing this "memory" feature is more challenging, as it requires writing a statistical model that accounts for multiple points in time and more complex multi-temporal disturbance state scenarios, as well as the informatics pipelines to save previous states and restart model ensembles from an earlier point in time.
 
-* Scaling up -> spatially implicit
-  * work at larger spatial scales and account for subgrid disturbance heterogeneity implicitly 
-    * ED2
-    * Contagion [@McCabe2019]
+The final potential algorithm improvement would be to adopt a spatially implicit approach to tracking disturbance.
+Spatially implicit approaches are a common feature of global vegetation models, which might represent a larger (e.g.
+1-by-1 degree) grid cell as containing multiple vegetation types, each with a fractional area, rather than keeping track of where in each grid cell those vegetation types are located.
+A subset of these models also keep track of both disturbance events and time since disturbance (a.k.a.
+patch age) in this way, updating fractional areas dynamically.
+Many of these models are derived from the size-and-age-structured (SAS) approximation introduced in the Ecosystem Demography (ED) model [@Moorcroft2001].
+The idea is that if you consider a patch of vegetation of a specific age and vegetation type, a disturbance might only affect part of the patch, which would cause it to split into two patches (disturbed and undisturbed) each with a new fractional area.
+If this pattern were repeated over time the number of patches would increase exponentially over time, so ED also implements a patch fusion scheme that merges older patches back together when they are judged to have converged to equivalent states.
+More recently, @McCabe2019 proposed a way to account not just for fractional area but also patch adjacency using a spatially implicit approach, allowing for the model to better account for spatial contagion and the distribution of patch sizes.
+Conceptually, the disturbance data assimilation approach we've presented here should be able to be updated to keep track of fractional areas of disturbance instead of a discrete disturbed/undisturbed state, for example by adopting a Dirlichet data model instead of a Multinomial.
 
@@
 
-* Implications  (borrow text from CMS proposals)
-  * Improving MRV
-  * Recovery forecasting
-  * Disturbance nowcasting
-    * requires reducing temporal resolution
+Over the course of this paper we discussed the potential advantages of model-data assimilation for improving MRV of the terrestrial carbon cycle, the importance of disturbance to carbon MRV, and the limitations of current data assimilation algorithms to account for disturbance.
+We then derived a new data assimilation algorithm, the Multinomial filter, based on the idea of accounting for the discrete nature of disturbance, and demonstated the ability of this algorithm to capture a range of simulated and real-world disturbances.
+We then discussed ways that this algorithm could be further improved.
+Overall, the Multinomial filter has the potential to improve not only MRV but also the near real-time disturbance detection, with important implications at scales from national (UN reporting) down to local (natural climate solutions).
+By allowing a more realistic representation of disturbance in process-based models, the algorithm also opens the door to the incorporation of ecosystem recovery forecasts, making possible more informed and efficient management of post-disturbance recovery.

### 2023-02-01 methods.Rmd
Discussion: Simulated data and OR test case text
formatting

@@
 
-```{r concept, echo=FALSE, message = FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
+```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
 Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities.
@@
 Ensemble members experienced disturbance stochastically with a 5% annual probability.
-For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 [UNITS?] for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).
+For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 $kgC/m^2$ for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).
 
@@
 To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_).
-The disturbance product **need text and citation from Robert** [CITE].
+The disturbance product **need text from Robert** [@KENNEDY2010]. 
 To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.
 
-```{r map, echo=FALSE, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances"}
+```{r map, echo=FALSE, fig.asp=1, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances"}
 ## maps of pre-disturbance biomass, cover, and 2004 disturbance
@@
 if(!exists("bio04")) bio04 = raster::raster("bio04.tif")
-raster::plot(bio04,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main="Landtrendr Biomass")
+raster::plot(bio04/10,col=c('#ffffff','#edf8fb','#ccece6','#99d8c9','#66c2a4','#2ca25f','#006d2c'),asp=1,main=expression("Landtrendr Biomass" ~ (kgC/m^2)))
 
@@
 The difference between the predicted and observed biomass at the end of the assimilation was taken as a metric of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations.
-**ONCE RESULTS ARE FINALIZED, COME BACK AND FILL IN WHICH SPECIFIC ANALYSES WERE USED**.
+To standardize this into a unitless metric, the difference in biomass was divided by half the width of the Analysis 95% CI.
+Across sites the extent of filter divergence was analyzed as a function of disturbance type and post-disturbance AGB using linear models.
+Similarly, the ability of the algorithm to detect disturbance, represented as the mean posterior estimate of the disturbance probability, was assessed as a function of disturbance type and post-disturbance AGB using generalized linear models with a logit link and quasibinomial family.
 
@@
 The skill of these assimilation experiments were assessed the same as the original.
-In addition, changes in skill across experiments were assessed as a function of R and Q.
+In addition, changes in skill and filter divergence across experiments were assessed as a function of R and Q focusing on two response variables: (1) the filter coverage, defined as the proportion of sites where our divergence metric falls between -1 and 1 (i.e., where the difference between the observation and Analysis is less than half the width of the Analysis 95% CI); and (2) the detection threshold, defined as the change in AGB where the probability of detecting a disturbance is 50% as inferred by the GLM of disturbance size versus probability of disturbance.
 

### 2023-02-01 results.Rmd
formatting

@@
 
-```{r sssd, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
+```{r sssd, echo=FALSE,fig.asp=1.25,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
 
@@
 #Figures showing all 3 pools cov, C accounting / attribution 
-pool.names=c("Cleaf","Csoil","Cwood")
+pool.names=c("Leaf","Soil","Wood")
+scaled = c(1,10,1)
 for(i in 1:3){
@@
   })
+  CI.Na = CI.Na/scaled[i]
 
@@
   if(rng[1]<1) rng[1] = 0
-  plot(No[,i],
+  plot(No[,i]/scaled[i],
        ylim = rng,type='n',
-       ylab=pool.names[i],xlab="time")
+       ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time")
   ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
   lines(CI.Na[2,],col="#b2df8a",lwd=2)
-  points(No[,i])
+  points(No[,i]/scaled[i])
   #lines(Ntrue[,i],lty=2,lwd=2)
@@
      xlim=range(Forecast[[10]]$Nf[,1]),
-     xlab=pool.names[1],ylab=pool.names[3],
+     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),ylab=bquote(.(pool.names[3])~(kgC/m^2)),
      cex.lab=1.25,asp = 1)
+.(labNames[1]) ~ x^2
 points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
@@
 
-```{r multiPseudo, echo=FALSE, message = FALSE, fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.height=4,fig.width=3}
+```{r multiPseudo, echo=FALSE, message = FALSE, fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.asp=1.25}
 
@@
      ylim = rng,type='n',
-     ylab=pool.names[i],xlab="time",
+     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
      main="Multi-Disturbance")
@@
      xlim=range(Forecast[[10]]$Nf[,1]),
-     xlab=pool.names[1],ylab=pool.names[3],
+     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),ylab=bquote(.(pool.names[3]) ~ (kgC/m^2)),
      cex.lab=1.25,asp = 1)
@@
      ylim = rng,type='n',
-     ylab=pool.names[i],xlab="time",
+     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
      main="Multi-Disturbance, PFT switching")
@@
      xlim=range(Forecast[[10]]$Nf[,1]),
-     xlab=pool.names[1],ylab=pool.names[3],
+     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),
+     ylab=bquote(.(pool.names[3]) ~ (kgC/m^2)),
      cex.lab=1.25)
@@
      ylim = rng,type='n',
-     ylab=pool.names[i],xlab="time",
+     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
      main="Multi-Site, Single-Disturbance")
@@
      xlim=range(Forecast10$Nf[,1,1]),
-     xlab=pool.names[1],ylab=pool.names[3],
+     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),
+     ylab=bquote(.(pool.names[3]) ~ (kgC/m^2)),
      cex.lab=1.25,asp = 1,type='n')
@@
 Best fit parameters are given in Supplemental Table \@ref(tab:parameters).
-The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibrate) ).
+The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibration) ).
 Achieving this fit came at the cost of nontrivial systematic biases in both leaf carbon (0.064 $kgC/m^2$, 13%) and NEE (0.528 $kgC/m^2/year$, 69%).
@@
 
-```{r ORorig, echo=FALSE, fig.asp=0.75, fig.cap="Filter divergence and disturbance detection for central Oregon test case.
+```{r oregon, echo=FALSE, fig.asp=1, fig.cap="Filter divergence and disturbance detection for central Oregon test case.
 Left column is the default configuration while the right column is a reduced observation error experiment (Exp 2).
@@
 exp.pretty.name = c("(a) Example site, Original","(c) Example site, Low Obs Error")
-pool.names=c("Cleaf","Csoil","AGB")
+pool.names=c("Leaf","Soil","AGB")
 i = 3 ## Stem
@@
        type='n',
-       ylab=pool.names[i],xlab="time",
+       ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
        main=exp.pretty.name[ex])
@@
 Filter divergence is a known issue for Kalman Filters, not something unique to our multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data.
-Figure \@ref(fig:ORorig) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed.
-Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:ORorig) b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`).
+Figure \@ref(fig:oregon) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed.
+Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:oregon) b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`).
 In other words, sites with low final biomass (i.e., large previous disturbance) showed a larger divergence and the divergence declined as the final biomass increased.
 Divergence also varied by disturbance type (p < `r format(anova(div.class[[1]])$"Pr(>F)"[2],digits=3,scientific=TRUE)`), with clear cuts having the lowest divergence for a given disturbance size and forest pests being lower than the fires and "other".
-Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:ORorig) c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`).
+Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:oregon) c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`).
 However, detection was rarely successful for AGB disturbances that removed less than 7.5 $kgC/m^2$ and did not exceed 50% until disturbances were larger than `r bm=coef(dist.qb.abs[[1]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
@@
 As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(tab:QRexp)) to verify that without filter divergence the multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so.
-Figure \@ref(fig:ORorig) d-f show the results for Experiment 2, where we experimented with reducing the observation error, R.
-In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:ORorig) d).
-We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:ORorig) e).
+Figure \@ref(fig:oregon) d-f show the results for Experiment 2, where we experimented with reducing the observation error, R.
+In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:oregon) d).
+We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:oregon) e).
 Divergence is no longer significantly affected by disturbance magnitude (p < `r format(summary(div.class[[2]])$coefficients[2,4],digits=3)`) and the only significant effect of disturbance type was a small reduction in divergence for clearcuts (p < `r format(summary(div.class[[2]])$coefficients[3,4],digits=3)`).
-Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:ORorig) f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
+Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:oregon) f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
 
@@
 
-```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=0.75 }
-m = matrix(1:3,nrow=3,byrow = TRUE)
+```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=1 , fig.width=4}
+m = matrix(1:2,nrow=2,byrow = TRUE)
 layout(m)
@@
 thresh = sapply(dist.qb.abs,function(x){bm=coef(x); -bm[1]/bm[2]})
-plot(Rmult[Rs],thresh,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab=expression("Detection Threshold" ~ (kgC/m^2)),main="(c) Disturbance Detection Threshold")
+plot(Rmult[Rs],thresh,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab=expression("Detection Threshold" ~ (kgC/m^2)),main="(b) Disturbance Detection Threshold")
 legend("topleft",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2)

### 2023-02-01 supplemental.Rmd
formatting

--- 2023-01-02 supplemental.Rmd
@@
 
-```{r calibration,fig.cap="Predicted and observed woody biomass, leaf biomass, and NEE at US-Me2",echo=FALSE}
+```{r calibration, echo=FALSE, fig.asp=1.25, fig.cap="Predicted and observed woody biomass, leaf biomass, and NEE at US-Me2"}
 load("fit.RData")
+
+m = matrix(1:3,nrow=3,byrow = TRUE)
+layout(m)
 
@@
 
-```{r parameters,echo=FALSE}
+```{r parameters, echo=FALSE}
 param.names = c("LUE","GAMMA","tauS","tauW","Av","sd_NEE","sd_leaf","sd_wood")
 param.desc = c("Light Use Efficiency","Fraction of GPP allocated to autotrophic respiration","Soil C residence time","Wood C residence time","Fraction of NPP allocated to leaves","NEE standard deviation","leaf C standard deviation","wood C standard deviation")
-param.table = cbind(param.names,param.desc,mle$par)
+param.table = cbind(param.names,param.desc,format(mle$par,digits=3,scientific=TRUE))
 colnames(param.table) <- c("Parameter","Description","Value")

### 2023-02-01 methods.Rmd
formatting

@@
 For the assimilation experiment, each site was assimilated separately.
-In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^2$ based on [@Kennedy2018] Figure 2.
+In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^4$ based on [@Kennedy2018] Figure 2.
 Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017).
-The difference between the predicted and observed biomass at the end of the assimilation was taken as a metric of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations.
-To standardize this into a unitless metric, the difference in biomass was divided by half the width of the Analysis 95% CI.
+One thing we wanted to assess with the Multinomial filter is the extent of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations.
+As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval.
+If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.
+
 Across sites the extent of filter divergence was analyzed as a function of disturbance type and post-disturbance AGB using linear models.
@@
 ```
-
-  * Score: filter divergence, disturbance detection & skill
-  * Regress vs: pre-disturbance biomass, size of disturbance (abs & rel), type, etc
-
-Divergence
-
-As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval.
-If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.

### 2023-02-01 results.Rmd
formatting

@@
 load("exp3.Rdata") ## Generated by SingleSite_MultiDisturb_MultiPool.Rmd
-pool.names=c("Cleaf","Csoil","Cwood")
+pool.names=c("Leaf","Soil","Wood")
 i = 3
@@
 load("exp4.Rdata") ## SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
-pool.names=c("Cleaf","Csoil","Cwood")
+pool.names=c("Leaf","Soil","Wood")
 i = 3
@@
 load("exp5trim.Rdata") ##Generated by TwoSite_SingleDisturb_MultiPool.Rmd
-pool.names=c("Cleaf","Csoil","Cwood")
+pool.names=c("Leaf","Soil","Wood")
 rng = range(CI.Na)
@@
 
-```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=1 , fig.width=4}
+```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=1.5 , fig.width=4.5}
 m = matrix(1:2,nrow=2,byrow = TRUE)

### 2023-03-31 Paper.Rmd
Google Drive track changes

@@
 ```
+\newpage
 
@@
 
-Current approaches to bottom-up disturbance monitoring rely heavily on the detection of land-use, land-use change, and forestry (LULUCF) through remote sensing, but often account for ecosystem impacts using simple look-up tables.
-By contrast, process models are frequently used to analyze and predict disturbance dynamics in greater detail.
-Once observations are available, however, we need to update predictions, especially for stochastic processes such as disturbance.
+1.
+Current approaches to bottom-up disturbance monitoring rely heavily on the detection of land-use, land-use change, and forestry through remote sensing, but often account for ecosystem impacts using simple look-up tables.
+By contrast, process models are frequently used to analyze and predict disturbance dynamics in greater mechanistic detail.
+Once observations are available, however, we need to update model predictions, especially for stochastic processes such as disturbance.
 State data assimilation (SDA) is designed specifically to update predictions, nudging modeled states back toward reality in proportion to the uncertainties in the model and the data, but current SDA algorithms are designed to update continuous states, not discrete disturbances.
+2.
 Here we develop a new Bayesian SDA algorithm that combines a discrete Multinomial state-and-transition framework with conventional ensemble filtering SDA approaches.
 To demonstrate the potential for assimilating disturbance, we applied the Multinomial SDA to the Very Simple Ecosystem Model (VSEM), performing both simulated data experiments with known disturbances and testing the algorithm against real-world disturbances detected in the LandTrender data product for central Oregon.
-
-With simulated disturbance we demonstrate the ability to not only detect discrete disturbance events but also avoid false positives.
+3.
+With simulated disturbance we not only demonstrate the ability to detect discrete disturbance events but also to avoid false positives.
 We also demonstrated the ability to fuse multiple data types to successfully distinguish different disturbance types, and to probabilistically capture vegetation type ‘switching’ events within the assimilation and ensemble forecast.
-To apply this to real-world data we calibrated VSEM against eddy-covariance and ancillary data from the Ameriflux US-Me2 tower.
+To apply this to real-world data we calibrated VSEM against eddy-covariance and ancillary data from the AmeriFlux US-Me2 tower.
 We then selected 356 conifer forest sites for testing, using the Landtrendr disturbance product to stratify by four disturbance types (cut, burn, pest, and other).
 We then assimilated the 30m Landtrendr annual aboveground biomass product from 1990-2017 and assessed the rate of disturbance detection.
-Assimilating just AGB, our assimilation was sensitive to disturbances that reduced biomass by 1.5 kg/m2 but underpredicted defoliation disturbances, which we expect would be improved by also assimilating LAI.
+With just AGB, our assimilation was sensitive to disturbances that reduced biomass by >1.5 kg/m2 but underpredicted defoliation disturbances, which could be improved by including LAI in the assimilation.
+4.
 Moving forward, the SDA framework provides an exciting opportunity to fuse multiple data sources to holistically improve real-time disturbance detection, impact assessment (e.g.
@@
 
-# Intro
+\newpage
 
-Researchers have devoted considerable time and resources to developing "bottom-up" terrestrial carbon (C) cycle data products at landscape- to global-scales (**CITE**, e.g.
-SOCCR2).
-Such efforts have drawn upon various combinations of remote sensing, inventory data, field-based sensors (e.g., eddy covariance), and process-based models with the aims of improving our understanding of C cycle processes, improving predictions of future changes in the C cycle, and improving monitoring, reporting, and verification (MRV) or C stocks and fluxes (**CITE**).
+# Introduction
+
+Researchers have devoted considerable time and resources to developing "bottom-up" terrestrial carbon (C) cycle data products at landscape- to global-scales [@SOCCR2].
+Such efforts have drawn upon various combinations of remote sensing, inventory data, field-based sensors (e.g., eddy covariance), and process-based models with the aims of improving both our understanding of C cycle processes and the monitoring, reporting, and verification (MRV) of C stocks and fluxes [@AR6WG3; @SOCCR2].
 MRV in particular is driven by a combination of local, national, and international reporting requirements (e.g.
-UN Framework Convention on Climate Change) and increasingly by the demand for natural climate solutions and carbon market certification (**CITE**).
+UN Framework Convention on Climate Change) and increasingly by the demand for natural climate solutions and carbon market certification [@NOVICK2022; @sands1992united].
 
 A particularly critical part of carbon MRV is the impact of land use, land use change, forestry (LULUCF) and natural disturbance (e.g.
-wildfire, pests, wind), which can have major impacts on the carbon cycle, quickly turning C sinks into large sources, and are responsible for a large fraction of interannual C cycle variability (CITE).
-They are also places where management and policy can have the greatest impact on carbon pools and fluxes, in terms of avoiding transitions, increasing resilience, and managing restoration and recovery after transitions occur.
-Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense, and there is mounting evidence that C stocks are likely to be more sensitive to climate effects on disturbance regimes than they are to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell2020; @Andrews2018].
-While there are a variety of approaches employed to monitor and understand the C cycle processes for intact ecosystems, when it comes to disturbance the status quo is based on detecting disturbances (e.g., via remote sensing) and then using a simple look up table to quantify C cycle impacts (CITE).
-While the closely related fields of disturbance detection and land use, land-use change, and forestry (LULUCF) monitoring have made considerable advances in recent years (CITE), the reliance on simple static look-up tables to quantify emissions is clearly missing considerable variability across space, time, and disturbance intensity.
+wildfire, pests, wind), which can have major impacts on carbons stocks and accumulation, quickly turning C sinks into large sources [@harris2016; @Law2018; @Hudiburg_2019] .
+**TODO: Bev says delete: Natural and anthropogenic disturbance events are also places where management and policy can have the greatest impact on carbon pools and fluxes, in terms of increasing resilience, managing restoration and recovery, and avoiding transitions to alternative system states.
+Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense [@seidl_forest_2017; Dale2001].
+Compounding this, C stocks are likely to be more sensitive to changes in disturbance regimes than to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell2020; @Andrews2018].** While there are a variety of approaches for monitoring and understanding C cycle processes for intact ecosystems, when it comes to disturbance the status quo is based on detecting disturbances (e.g., via remote sensing) and then using a simple look up table (i.e., emission factors) to quantify C cycle impacts [@eggleston2006; @Mitchell2017; @BULLOCK2021; @harris2021].
+Active sensors such as lidar and SAR more directly capture changes in aboveground biomass, but do not inform other pools and fluxes [@Mitchell2017].
+While the closely related fields of disturbance detection and classification, and LULUCF monitoring have made considerable advances [@KENNEDY2010; @ZHU2020; @VERBESSELT2012], the reliance on simple static look-up tables to quantify emissions is clearly missing considerable variability across space, time, and disturbance intensity.
 
-The alternative to look-up tables is to rely on process-based models to capture the spatiotemporal in the C cycle.
-However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), with major classes of disturbance being absent from most models (e.g., forest pests, [@Hicke2012; @DietzeMatthes2014]).
-Ultimately, MRV needs to be done with the disturbances that actually happened, not those that were predicted.
+An alternative to look-up tables is to rely on process-based models to capture the spatiotemporal changes in the C cycle post disturbance.
+However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), where major classes of disturbance are often absent (e.g., forest pests, [@Hicke2012; @DietzeMatthes2014]).
+Ultimately, MRV efforts should reflect the actual disturbance events, not those that were predicted.
 On the other hand, prescribing disturbance in models ignores the uncertainty in disturbance detection and brings us back to using look-up tables to prescribe how the disturbance is represented in the model.
 
-State data assimilation offers a promising path forward for addressing these challenges and is already being to see increasing applications in other aspects of carbon MRV [@Baatz2021].
-The basic idea is that a process model is run numerous times to make a probabilistic prediction of the full carbon cycle (pools and fluxes) from one point in time to the next, sampling over the uncertainties about the current state of the system and possibly also parameter, driver, and process (residual) uncertainties.
-This multivariate prediction represents our (Bayesian) _prior_ understanding of the carbon cycle that is then updated, using Bayes' Theorem, based on observations of one or more these pools or fluxes [@Dietze2017a; @Wikle2007].
-In essence, data assimilation "nudges" the model back towards the data, with the size of the nudge proportional to the relative uncertainties in the model (prior) and the data (Likelihood).
-The resulting posterior estimate is able to fuse numerous data constraints on the different pools and fluxes, is robust to large amounts of missing observations in space, time, and pool/flux, and provides a rich uncertainty estimate.
+**TODO: Bev suggests replacing the above paragraph with the IPCC AR6 text: Technological barriers and opportunities Monitoring, reporting, and verification.
+Development of satellite technologies to assess potential deforestation has grown in recent years with the release of 30 m data by Hansen et al.
+(2013), however, these data only captures tree cover loss, and increasing accuracy over time may limit its use for trend analysis (Ceccherini et al.
+2020; Palahí et al.
+2021).
+Datasets on forest losses are less well developed for reforestation and afforestation.
+As Mitchell et al.
+(2017) point out, there has been significant improvement in the ability to measure changes in tree and carbon density on sites using satellite data, but these techniques are still evolving and improving and they are not yet available for widespread use.**
+
+State data assimilation (SDA) offers a promising alternative to these two approaches, directly addressing their challenges while building on their respective strengths, and is already being used in other aspects of carbon MRV [@Baatz2021].
+The basic idea behind SDA is that a process model is run numerous times to make a probabilistic prediction of the full carbon cycle (pools and fluxes) from one point in time to the next.
+It does this by sampling over the uncertainties about the current state of the system, while providing the option also sampling the parameter, driver, and process (residual) uncertainties.
+This multivariate prediction represents our (Bayesian) _prior_ understanding of the carbon cycle that is then updated, using Bayes' Theorem, based on observations of one or more of these pools or fluxes [@Dietze2017a; @Wikle2007].
+In essence, data assimilation “nudges” the model back towards the data, with the size of the nudge proportional to the relative uncertainties in the model (prior) and the data (Likelihood).
+The resulting posterior estimates are able to fuse numerous data constraints simultaneously, are robust to large amounts of missing observations in space and time, and provide a comprehensive uncertainty estimate.
+SDA can even update pools and fluxes that are not directly observed, which it does by leveraging model ensemble covariance between observed and unobserved variables.
 Finally, unlike monitoring-only systems, data assimilation systems can seamlessly transition from reconstructing the past (a.k.a.
@@
 nowcast), to predictive forecasts [@Dietze2017a].
-A handful of different terrestrial carbon data assimilation systems are now in opperations (e.g., CLM-DART, CCDAS, PEcAn), with applications at site [@Raiho2020], regional [@Raczka2021], continental [@Dokoohaki2022], and global [@Fox2022; @Rayner2005] scales.
+A handful of different terrestrial carbon data assimilation systems are now in operation (e.g., CLM-DART, CCDAS, PEcAn), with applications at site [@Raiho2020], regional [@Raczka2021], continental [@Dokoohaki2022], and global [@Fox2022; @Rayner2005] scales.
 
-The potential advantages of applying state data assimilation to address the disturbance problem in carbon MRV are straightforward.
-Ensembles of model runs could capture the stochasticity in natural disturbance while also providing a process-based representation of both disturbance and carbon cycle processes and numerous possible sources of fine-scale heterogeneity (current vegetation conditions, climate, soils, etc.).
-The assimilation would then update the predictions based on what actually happened, formally accounting for the uncertainty in the disturbance observations, as well as fusing in numerous other data constraints (e.g., inventory, sensor, remote sensing).
-These observations may variously either reinforce or cast doubt upon the disturbance attribution, while also informing our understanding of pre- and post-disturbance heterogeneity.
-Data assimilation even has the potential to capture disturbances that are not represented explicitly in the models but are apparent in the observations.
+The potential advantages of applying SDA to address the disturbance problem in carbon MRV are straightforward.
+Ensembles of model runs could capture the stochasticity of natural disturbance regimes while also providing a process-based representation of both disturbance and carbon cycle processes across numerous possible sources of fine-scale heterogeneity (current vegetation conditions, climate, soils, etc.).
+The assimilation would then update the predictions based on what actually happened, formally accounting for the uncertainty in the disturbance observations, as well as allowing for the integration of other data constraints (e.g., inventory, sensor, remote sensing).
+These observations could then either reinforce or cast doubt upon the disturbance classification, while also informing our understanding of pre- and post-disturbance heterogeneity.
+Data assimilation even has the potential to capture disturbances that are not represented explicitly in a model but are apparent in the observations (e.g.
+insect defoliation).
 
-The problem is that, when it comes to assimilating disturbance, conventional data assimilation doesn't actually work (Figure \@ref(fig:enkf)).
+The problem is that, when it comes to assimilating disturbance, conventional data assimilation does not actually work (Figure \@ref(fig:enkf)).
 For reasons we'll explain in the _Methods_, instead of nudging a disturbed system to a disturbed state in the model, it nudges it to a "no man's land" in between the disturbed and undisturbed state that doesn't represent either state well.
-It can then take multiple iterations of the assimilation for these nudges to get the model back on track, in effect predicting gradual ecosystem degredation rather than a discrete disturbance and recovery, during which time this gradual decline will be misrepresented in MRV accounting (missing removals, fire combustion, etc.) and misinterpreted by the model (e.g.
+It can then take multiple iterations of the assimilation for these nudges to get the model back on track, in effect predicting gradual ecosystem degradation rather than a discrete disturbance and recovery, during which time this gradual decline will be misrepresented in MRV accounting (missing removals, fire combustion, etc.) and misinterpreted by the model (e.g.
 missing recruitment and respiration responses).
@@
 We will then apply this statistical model to both simulated data with known disturbances and real-world observations, making use of the Very Simple Ecosystem Model (VSEM) for predictions and remotely-sensed Landtrendr biomass and disturbance data products as observations, using the ponderosa pine ( _Pinus ponderosa_ ) ecosystems of central Oregon as a testbed [@Kennedy2018; @BayesianTools].
-Overall, this manuscript aims to provide a proof-of-concept for the idea of disturbance assimilation, laying the foundation for future applications and algorith improvements.
+Overall, this manuscript aims to provide a proof-of-concept for the idea of disturbance assimilation, laying the foundation for future applications and algorithm improvements.
 
@@
 
-NASA CMS, PEcAn, NEFI
+MCD was supported by funds from NASA (20-CMS20-0030) and the National Science Foundation (NSF 1458021, NSF 1638577, NSF 1702996, NSF 2213580).
+
+RK was supported by NASA grants NNX16AQ25G and NN12AP76G
+
+**COAUTHORS: ADD ACKNOWLEDGEMENTS **
+
+# Bibliography
+
+<div id="refs"></div>
+\newpage
+
+# (APPENDIX) Appendix {-}
 
@@
 ```
-
-# Bibliography

### 2023-03-31 discussion.Rmd
formatting
Google Drive track changes

@@
 
-At a high-level, the simulated data experiments demonstrate that the Multinomial filter is able to address the fundamental problem that traditional data assimilation encounters when faced with disturbance events.
+Taken as a whole, the simulated data experiments demonstrate that the Multinomial filter is able to address the fundamental problem that traditional data assimilation encounters when faced with disturbance events.
 Rather than nudging the posterior state stranded into a "no man's land" between the disturbed and undisturbed states, the Multinomial successfully makes a discrete jump in state.
@@
 From a disturbance detection perspective, three things set the Multinomial assimilation apart from other conventional time-series based disturbance detection algorithms (as opposed to static image classifications), such as Landtrendr [@KENNEDY2010], CCDC (Continuous Change Detection and Classification) [@ZHU2014; @Pasquarella2017; @ZHU2020] and BFAST (Break detection For Additive Season and Trend) [@VERBESSELT2012].
-First, most other algorithms flag disturbances as departures from the predicted value -- a rejection of the "status quo" hypothesis.
-With the multinomial, we are instead treating disturbance detection as a choice between competing hypotheses, assessing how much weight to place on the "status quo" versus different alternative disturbance options.
-Second, most of these algorithms rely on a moving-window approach as part of their "status quo" prediction, which requires accessing and processing a time-series of data from the past every time an prediction is updated.
+First, most other algorithms flag disturbances as departures from the predicted value -- a rejection of the undisturbed "status quo" hypothesis.
+With the Multinomial, we are instead treating disturbance detection as a choice between competing hypotheses, assessing how much weight to place on the undisturbed "status quo" versus different alternative disturbance options.
+Second, most of these algorithms rely on a moving-window approach as part of their "status quo" prediction, which requires accessing and processing a time-series of data from the past every time a prediction is updated.
 The iterative algorithms that form the basis for data assimilation (e.g., Kalman Filter) have on occasion been applied to disturbance detection (using statistical, rather than process-based model) and have demonstrated greater computational efficiency [@Ye2020].
-@Ye2020 avoid the "no man's land" nudge problem by the simple expedient of performing threshold-based disturbance detection prior to assimilation and not assimilating disturbed sites.
+@Ye2020 avoids the "no man's land" nudge problem by the simple expedient of performing threshold-based disturbance detection prior to assimilation and not assimilating disturbed sites.
 Third, other algorithms are relying on simple statistical models for their predictions of the status quo, such as harmonic time-series models, while the Multinomial model is relying on the predictions of a process-based ecosystem model.
-This has three distinct advantages: (1) We are leveraging greater process understanding in the status quo prediction, potentially making the algorithm more robust in the face of extreme events that are not disturbances (e.g., unusually cold or dry) but which alter the magnitude or phenology of seasonal cycles; (2) While not included in the simple "toy" model used in these analyses, process-based models also have the potential to leveraging greater process understanding in predicting the likelihood of disturbances, for example by accounting for fuel loads, fuel moisture, and ignition rates in predicting the likelihood of fire; (3) Because process-based models are predicting multiple pools and fluxes that can be compared to multiple distinct types of data, the Multinomial algorithm provides a natural way to fuse multiple types of data that would either reinforce a disturbance detection (if all sources are pointing in the same direction) or call it into question.
-While conventional disturbance detection algorithms are overwhelmingly based on multispectral remote sensing from either a single sensor or a pre-harmonized fusion of multiple sensors [e.g., Harmonized Landsat Sentinel-2, @CHEN2021; @SHANG2022], data assimilation approaches can simultanously integrate a wide range of relevant data types, for example lidar, microwave, thermal, solar-induced fluorescence, and field data.
+This has three distinct advantages: (1) We are leveraging greater process understanding in the status quo prediction, potentially making the Multinomial model more robust in the face of extreme events that are not disturbances (e.g., unusually cold or dry) but which alter the magnitude or phenology of seasonal cycles; (2) While not included in the simple "toy" model used in these analyses, process-based models also have the potential to leveraging greater process understanding in predicting the likelihood of disturbances, for example by accounting for fuel loads, fuel moisture, and ignition rates in predicting the likelihood of fire; (3) Because process-based models are predicting multiple pools and fluxes that can be compared to multiple distinct types of data, the Multinomial algorithm provides a natural way to fuse multiple types of data that would either reinforce a disturbance detection (if all sources are pointing in the same direction) or call it into question.
+While conventional disturbance detection algorithms are overwhelmingly based on multispectral remote sensing from either a single sensor or a pre-harmonized fusion of multiple sensors [e.g., Harmonized Landsat Sentinel-2, @CHEN2021; @SHANG2022], data assimilation approaches can simultaneously integrate a wide range of relevant data types, for example lidar, microwave, thermal, solar-induced fluorescence, and field data.
 Data assimilation can also integrate a range of optical sensors (both multispectral and hyperspectral) based on their spectral response functions, rather than requiring pre-harmonization [@Shiklomanov2021; @Zhang2023]).
@@
 
+** TODO from bev: Example of EnKF used with our ponderosa pine data: Spadavecchia, L., M.
+Williams, B.E.
+Law.
+2011.
+Uncertainty in predictions of forest carbon dynamics - separating driver error from model error.
+Ecological Applications 21:1506–1522.
+Doi:10.1890/09-1183.1.
+Cite Quaife, T., P.
+Lewis, M.
+De Kauwe, M.
+Williams, B.E.
+Law, M.
+Disney, P.
+Bowyer.
+2008.
+Assimilating canopy reflectance data into an ecosystem model with an ensemble Kalman filter.
+Remote Sensing of Environment 112:1347-1364.
+**
+
 The most obvious issue with the initial application of the Multinomial filter to real-world data was the high incidence of filter divergence -- of the model becoming overconfident in itself and thus diverging from the observations.
 Filter divergence is not unique to this algorithm, but something that can occur in traditional filters as well. 
-Indeed, there's no reason to belive that filter divergence is any more common in the Multinomial filter than in more traditional filters, such as EnKF, because in the absence of disturbance the Multinomial filter behaves exactly as EnKF would.
+Indeed, there's no reason to believe that filter divergence is any more common in the Multinomial filter than in more traditional filters, such as EnKF, because in the absence of disturbance the Multinomial filter behaves exactly as EnKF would.
 The underlying causes of filter divergence are a result of either an underaccounting of model uncertainty or an overestimation of observation error in the data, both of which are likely at play in this example.
-On the model side, forecast uncertainty is associated with five factors: initial condition uncertainty, driver uncertainty, parameter uncertainty, parameter heterogenity, and process error [@Dietze2017b].
+On the model side, forecast uncertainty is associated with five factors: initial condition uncertainty, driver uncertainty, parameter uncertainty, parameter heterogeneity, and process error [@Dietze2017b].
 Since our goal here was a simple proof of concept, rather than a real world application, we did not formally account for uncertainty in model drivers (i.e., meteorology) or parameters. 
@@
 First, the assumption of constant variance is unlikely to be true because AGB is non-negative, which means the current interval estimates for low biomass sites include a substantial probability of negative biomass.
-More often the uncertainties in AGB are heteroskedastic and increasing in absolute magnitude as AGB increases, a property shared with the allometeric models used to estimate AGB in the field calibration data [@Dietze2008].
-This heteroskedasticity particularly important for disturbance detection because it means we are likely significantly overestimating the observation error associated with low biomass (disturbed) conditions, contributing directly to the Multinomial model's inability to distinguish such biomass drops from observation error.
-Second, the AGB data product is does not distinguish random versus systematic errors [@Cameron2022], and in particular is not accounting for the spatial and temporal autocorrelation in observation errors.
+More often the uncertainties in AGB are heteroskedastic and increasing in absolute magnitude as AGB increases, a property shared with the allometric models used to estimate AGB in the field calibration data [@van2005; @berner2016].
+This heteroskedasticity is particularly important for disturbance detection because it means we are likely significantly overestimating the observation error associated with low biomass (disturbed) conditions, contributing directly to the Multinomial model's inability to distinguish such biomass drops from observation error.
+Second, the Landtrendr AGB data product does not distinguish random versus systematic errors [@Cameron2022], and in particular is not accounting for the spatial and temporal autocorrelation in observation errors.
 In reality, a site with higher than predicted AGB in one year is likely to have higher than predicted AGB the next year (positive temporal autocorrelation) and adjacent pixel are also likely to have higher than predicted AGB (positive spatial autocorrelation).
 These positive autocorrelations mean we are likely overestimating the uncertainty associated with change detection, since a substantial portion of this systematic observation error should cancel out [@Kennedy2023].
-Overall, the failures of the default calibration, which ignored driver uncertainty, parameter uncertainty, and parameter heterogeneity while understimating process error and overestimating observation error, provide an illustrative case study in the care that must be taken when applying data assimilation approaches in general, and the Multinomial filter in particular, to real-world data
+Overall, the failures of the default calibration, which ignored driver uncertainty, parameter uncertainty, and parameter heterogeneity while underestimating process error and overestimating observation error, provide an illustrative case study in the care that must be taken when applying data assimilation approaches in general, and the Multinomial filter in particular, to real-world data
 
@@
 Indeed, filter divergence declined, and disturbance detection increased, systematically as a function of the magnitude of the observation error. 
-Somewhat surprisingly the computational experiments suggested that the Multinomial filter tended to be more sensitive to observation error than to model error, and suggesting that applications should prioritize making sure that observation errors are correctly accounted.
+Somewhat surprisingly the computational experiments suggested that the Multinomial filter tended to be more sensitive to observation error than to model error, which implies that applications should prioritize making sure that observation errors are well constrained.
 Indeed, doing a better job at propagating model uncertainties correctly only becomes important once the observation errors are constrained.
@@
 If we had leveraged additional data constraints, we would have gained additional confidence in our ability to detect real disturbances.
-One way to to do this is to bring in additional data on the same pool, in this case AGB, reducing the overall observation error.
+One way to do this is to bring in additional data on the same pool, in this case AGB, reducing the overall observation error.
 This could be done through some combination of field data [e.g., Forest Inventory and Analysis, CITE], lidar [e.g., GEDI, CITE], microwave vegetation optical depth [e.g., SMAP, CITE Koenings], or radar [e.g., BIOMASS, CITE].
-This was not done in this example because data products based on different observations have a different spatial resolutions, which would need to be addressed.
-Fusing in a data assimilation framework is well established but was judged beyond the scope of the current study because it requires running a fully spatial assimilation rather than a simple site-scale timeseries model.
+This was not done in this example because data products based on different observations have different spatial resolutions, which would need to be addressed.
+Fusing in a data assimilation framework is well established but was judged beyond the scope of the current study because it requires running a fully spatial assimilation rather than a simple site-scale time series model.
 Basically, one either needs to run the assimilation at a fine spatial resolution and use an observation operator to map many model grid cells to one observation, or one could run the assimilation at a coarse spatial resolution and use an observation operator to map one model grid cell to many observations [@Dietze2017a].
@@
 errors in data products based on the same underlying spectra would not be independent).
-In addition to bringing in additional data constratins on different pools, we could also constrain our disturbance estimates using observations on different ecosystem fluxes and processes, such as eddy covariance (carbon, water, and energy fluxes), solar induced fluorescence (a proxy for gross primary productivity), thermal imaging (a proxy for water stress and evapotranspiration).
+In addition to bringing in additional data constraints on different pools, we could also constrain our disturbance estimates using observations on different ecosystem fluxes and processes, such as eddy covariance (carbon, water, and energy fluxes), solar induced fluorescence (a proxy for gross primary productivity), thermal imaging (a proxy for water stress and evapotranspiration).
 
 In addition to constraining our estimates of disturbance through the ecosystem state variables, we can also improve them by improving our priors on the disturbance probability. 
-In our simple case study we relied on static and spatially-homogenous prior probabilities on different disturbance rates, an approach somewhat similar to the simplest state-and-transition models of land use and disturbance [@Daniel2016].
-More complex state-and-transition models are possible that allow these priors to be spatially heterogenous (e.g., relying on spatial covariates) or time evolving (e.g., relying on temporal covariates).
-Dynamic models are also possible that rely on the current state of a pixel (e.g., increasing fuel loads increase the probability of fire), the state of adjacent pixels (e.g., spatially contagious disturbances such as fire), or even those involving more distant spatial relations (e.g., dispersal kernels of biotic pests and pathogens) (CITATIONS).
-In addition to relying on external state-and-transition models, one of the appeals of working with process-based ecosystem models is that many of they include submodules that represent different types of disturbance internally (e.g., fire, pests).
+In our simple case study we relied on static and spatially-homogeneous prior probabilities on different disturbance rates, an approach somewhat similar to the simplest state-and-transition models of land use and disturbance [@Daniel2016].
+It is possible to use more complex state-and-transition models that would allow these priors to be spatially heterogeneous (e.g., relying on spatial covariates) or time evolving (e.g., relying on temporal covariates).
+Similarly, it is also possible to build state-and-transition that leverage the current state of a pixel (e.g., increasing fuel loads increase the probability of fire), the state of adjacent pixels (e.g., spatially contagious disturbances such as fire), or even those involving more distant spatial relations (e.g., dispersal kernels of biotic pests and pathogens) [@johnstone2011; @hudgins2017].
+In addition to relying on external state-and-transition models, one of the appeals of working with process-based ecosystem models is that many of them include submodules that represent different types of disturbance internally (e.g., fire, pests).
 In these cases we can either extract the model's internally predicted probabilities or simply rely on the frequency of events predicted across the ensemble of model runs.
@@
 The other way to improve upon the prior estimates of disturbance probability is to rely on external disturbance data that's not part of the assimilation itself. 
-For example, in our Oregon case study it is clear that the Landtrendr disturbance product is "seeing" more about disturbance than is being expressed in the AGB product alone, as the the disturbance product is looking for different information (e.g., weighting bands differently, training to distinguish among types of disturbance not just a loss of biomass).
+For example, in our Oregon case study it is clear that the Landtrendr disturbance product is "seeing" more about disturbance than is being expressed in the AGB product alone, as the disturbance product is looking for different information (e.g., weighting bands differently, training to distinguish among types of disturbance not just a loss of biomass).
 Therefore, one alternative to relying on models to predict disturbance priors is to use the disturbance product and its classification error statistics to specify priors itself to identify the error.
@@
 It's important to note that when using external disturbance data as priors we're still not prescribing disturbances deterministically, during the assimilation observations about ecosystem pools and fluxes would thus either reinforce these products or call them into question.
-Finally, it is worth nothing that relying on processed-based models and external disturbance data products are not mutually exclusive.
+Finally, it is worth noting that relying on process-based models and external disturbance data products are not mutually exclusive.
 The fusion of disturbance products and disturbance models could be part of the future development of the Multinomial filter itself (see next subsection), but may be simpler to handle as part of the prior construction.
@@
 Here we discuss two possible approximations to the full algorithm, one analytical and the other based on model emulation.
-Analytically, instead of solving the full Multinomial filter jointly, it may be possible to approximate the full model by decompose the filter sequentially. 
+Analytically, instead of solving the full Multinomial filter jointly, it may be possible to approximate the full model by decomposing the filter sequentially. 
 First, one would analytically apply EnKF to each alternative disturbance state, analytically generating independent posteriors for each $X^L_t$.
-Next, one would then apply a Dirlichet-Multinomial model to the problem of updating $\vec{\rho_t}$, treating $X^L_t$ as known rather than jointly estimated.
-The math for this approach has yet to be fully flushed out, and it remains to be seen how well this approximation works in practice and what information would be lost by solving the model conditionally rather than jointly.
+Next, one would then apply a Dirichlet-Multinomial model to the problem of updating $\vec{\rho_t}$, treating $X^L_t$ as known rather than jointly estimated.
+The math for this approach has yet to be fully fleshed out, and it remains to be seen how well this approximation works in practice and what information would be lost by solving the model conditionally rather than jointly.
 
@@
 While emulators are often used to approximate process-based models, predicting what the model would have output given a set of inputs, here we want to construct an emulator to that Analysis step, predicting what the Multinomial filter would output (probability of disturbance, ecosystem pool means and variances) as a function of the Forecast model state and the observed data constraints.
-** Find citations for papers that have emulated other SDA**.
-Emulators addresses not only the computational cost of scaling up but also the observation that similar sites are going to make similar predictions (e.g., adjacent LANDSAT pixels), and thus fully gridded models are going to perform highly redundant calculations.
-Emulators still require running the model and analysis (in this case the Multinomial filter), but rather than evaluating every location the filter is run for a subset of locations stratified by relevant variables (initial conditions, observed data constraints, and posssibly also covariates like climate, soils, etc.).
-Flexible machine learning approaches are then used to build an emulator that predicts the outcome of full Bayesian model.
-For places where the full models was not run the emulator is then used to predict the full model, which in essence involves interpolating in environmental space and mapping this back to physical space.
+Emulators address not only the computational cost of scaling up but also the observation that similar sites are going to make similar predictions (e.g., adjacent LANDSAT pixels), and thus fully gridded models are going to perform highly redundant calculations.
+Emulators still require running the model and analysis (in this case the Multinomial filter), but rather than evaluating every location the filter is run for a subset of locations stratified by relevant variables (initial conditions, observed data constraints, and possibly also covariates like climate, soils, etc.).
+Flexible machine learning approaches are then used to build an emulator that predicts the outcome of a fully Bayesian model.
+For places where the full model was not run the emulator is then used to predict the full model, which in essence involves interpolating in environmental space and mapping this back to physical space.
 
@@
 
-In terms of spatial disturbance probability, the current filter does estimate a whole vector of probabilities, one for each site, and thus estimates will include a posterior covariance structure across sites, but the mdoel is not written in a way that explicitly changes the probability of disturbance based on spatial adjacency.
+In terms of spatial disturbance probability, the current filter does estimate a whole vector of probabilities, one for each site, and thus estimates will include a posterior covariance structure across sites, but the model is not written in a way that explicitly changes the probability of disturbance based on spatial adjacency.
 In practice, almost all types of disturbance involve some degree of shared risk, if not true spatial contagion, whereby a disturbance at one site usually increases the probability of disturbance in near-by sites.
@@
 The final potential algorithm improvement would be to adopt a spatially implicit approach to tracking disturbance.
-Spatially implicit approaches are a common feature of global vegetation models, which might represent a larger (e.g.
-1-by-1 degree) grid cell as containing multiple vegetation types, each with a fractional area, rather than keeping track of where in each grid cell those vegetation types are located.
+Spatially implicit approaches are a common feature of global vegetation models, which might represent a larger (e.g., 0.5-by-0.5 degree) grid cell as containing multiple vegetation types, each with a fractional area, rather than keeping track of where in each grid cell those vegetation types are located.
 A subset of these models also keep track of both disturbance events and time since disturbance (a.k.a.
@@
 More recently, @McCabe2019 proposed a way to account not just for fractional area but also patch adjacency using a spatially implicit approach, allowing for the model to better account for spatial contagion and the distribution of patch sizes.
-Conceptually, the disturbance data assimilation approach we've presented here should be able to be updated to keep track of fractional areas of disturbance instead of a discrete disturbed/undisturbed state, for example by adopting a Dirlichet data model instead of a Multinomial.
+Conceptually, the disturbance data assimilation approach we've presented here should be able to be updated to keep track of fractional areas of disturbance instead of a discrete disturbed/undisturbed state, for example by adopting a Dirichlet data model instead of a Multinomial.
 
@@
 
+** TODO from Alexis: In the previous sections it might be worth including more info on the pros/cons of additional development as it directly relates to MRV**
+
 Over the course of this paper we discussed the potential advantages of model-data assimilation for improving MRV of the terrestrial carbon cycle, the importance of disturbance to carbon MRV, and the limitations of current data assimilation algorithms to account for disturbance.
-We then derived a new data assimilation algorithm, the Multinomial filter, based on the idea of accounting for the discrete nature of disturbance, and demonstated the ability of this algorithm to capture a range of simulated and real-world disturbances.
+We then derived a new data assimilation algorithm, the Multinomial filter, based on the idea of accounting for the discrete nature of disturbance, and demonstrated the ability of this algorithm to capture a range of simulated and real-world disturbances.
 We then discussed ways that this algorithm could be further improved.

### 2023-03-31 methods.Rmd
unit change on obs error
Google Drive track changes

@@
 # Methods
+
+**TODO Dongchen: consider adding a conceptual diagram to explain the framework and data**
 
@@
 
-The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approached have with capturing discrete disturbances and as our jumping off point for developing more flexible methods.
-Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabalistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g.
+The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods.
+Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g.
 carbon pools) and sometimes model parameters.
@@
 
-Within the EnKF, the probabalistic Forecast step is performed using an ensemble of model runs that vary in at least initial conditions, but potentially also model parameters, drivers, and process error.
+Within the EnKF, the probabilistic Forecast step is performed using an ensemble of model runs that vary in at least initial conditions, but potentially also model parameters, drivers, and process error.
 The ensemble forecast approach is easy to apply to any model, is particularly good at capturing nonlinear predictions, and unlike many other approaches does not require solving for the adjoint of the model.
@@
 
-While this fusion of forecast and new data is an incredibly powerful concept, as alluded to in the _Introduction_ there are times that the Kalman Normal-Normal "nudge" produces results that don't always make sense.
+While this fusion of forecast and new data is an incredibly powerful concept, as alluded to in the _Introduction_ there are times that the Kalman Normal-Normal "nudge" produces results that do not make sense.
 Consider as a simple example a simulated data experiment where LAI growth follows a simple stochastic logistic model with growth rate $r$=0.1, equilibrium $K$=6, process error standard deviation 0.25, and observation error standard deviation 0.75.
@@
 Not only is the assimilated state wrong, but so is our estimate of the process -- a discrete disturbance and rapid recovery is behaving like gradual degradation, which will lead to incorrect inferences about related processes (litter, soil C, NPP, NEE, ET, albedo, etc.).
-Not also that including a disturbance probability in the process model does not solve this problem.
+It is also worth noting that including a disturbance probability in the process model does not, by itself, solve this problem.
 Assuming the disturbance probability is low (e.g.
-a background rates of <1%), at every point in time the forecast would include a small number of disturbed ensemble members, but the overwhelming majority would be undisturbed.
+a background rate of <1%), at every point in time the forecast would include a small number of disturbed ensemble members, but the overwhelming majority would be undisturbed.
 Therefore $\mu_{f,t}$ will be relatively unchanged and $P_{f,t}$ would see a small increase in variance.
-The fundamental problem is that even though the true forecast is now bimodal (a large peak at the undisturbed state and a much smaller peak at the disturbed state), the EnKF is still representing this as a unimodal Normal.
-Indeed, if the you increased the disturbance probability substantially (e.g.
-50%), the EnKF would treat this as a unimodal forecast half way between the disturbed and undisturbed states, a place where no ensemble members are present (Figure \@ref(fig:concept), top).
+The fundamental problem is that even though the true forecast is now bimodal (a large peak at the undisturbed state and a much smaller peak at the disturbed state), the EnKF is still representing this as a unimodal Normal centered around the undisturbed state.
+Indeed, if the disturbance probability were increased substantially (e.g.
+50%), the EnKF would treat this as a unimodal forecast halfway between the disturbed and undisturbed states, a place where no ensemble members are present (Figure \@ref(fig:concept), top).
 
@@
 
-```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
-Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities.
+** TODO from zoey for above figure: Labels could be more informative (i.e.
+EnKF = conventional and Multinomial = improved approach).
+Icon for "data" should match the icons rather than a line.
+Is the P(dist) bound at 0,1?
+I barely noticed the black line at first, it could maybe go in a second panel below with a shared x-axis and a distinct y-axis.**
+
+** TODO from zoey for below figure: I think this figure could be more streamlined to better emphasize the difference between the filters.
+Why change the assimilation color from orange to blue in this figure?
+The histogram, to me, seems more intuitive for displaying the data values rather than the forecasted values.
+The second panel doesn't have "forecast" in the color labels which also adds to confusion.
+The X axis could communicate the low-LAI = disturbed, high-LAI = undisturbed, without relying on the colors/legend.
+X axis for this and the Y axis for the above figure could just be "disturbance magnitude" **
+
+```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (gray histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
+Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabilities.
 The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
@@
 
-To address this fundamental flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodel, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges.
+To address this fundamental flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodal, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges.
 To begin, let's use $X$ to represent the latent (unobserved) "true" state of the system.
-Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent continuous state variables of interest, which we'll refer to as $X^L_{k,t}$.
-Similar to the EnKF we'll assume that the ensemble prediction for any one of these categories can be represented as a multivariate Normal based on the ensemble sample mean and covariance:
+Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent ($L$) continuous state variables of interest, which we'll refer to as $X^L_{k,t}$.
+Similar to the EnKF we'll assume that the ensemble forecast for any one of these categories can be represented as a multivariate Normal,
 
@@
 
-However, because we have $k$ categories we end up with a multimodel prediction.
+where $\vec{\mu}_{f,k,t}$ and $P_{f,k,t}$ are the ensemble forecast ($f$) sample mean and covariance for outcome $k$ at time $t$.
+However, because we have $k$ categories we end up with a multimodal prediction (i.e., a mixture of MVNs).
 Furthermore, these modes need not be distinct; it is perfectly fine in concept for these MVNs to be overlapping.
-In practice they more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabalistically.
-Indeed, to be able to represent our multimodel prediction as a single probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.
+In practice the more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabilistically.
+Indeed, to be able to represent our multimodal prediction as a single probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.
 
@@
 In practice, we will use Bayes' Theorem to integrate over the uncertainties in both our categorical assignment and continuous state variables, producing posterior Analysis probability distributions for both $X_t$ and $\vec{\rho_t}$.
-Unlike in the EnKF, where conjugacy allows us to calculate an analytical solution for the posterior, there is not simple analytical solution for the joint posterior distribution of $X_t$ and $\vec{\rho_t}$.
+Unlike in the EnKF, where conjugacy allows us to calculate an analytical solution for the posterior, there is no simple analytical solution for the joint posterior distribution of $X_t$ and $\vec{\rho_t}$.
 In this case we will use well-developed Bayesian Markov Chain Monte Carlo (MCMC) numerical algorithms to sample from the posterior distribution.
 Specifically, numerical analyses in this paper were conducted using JAGS v4.2.0 [@Plummer03] accessed from R v4.0.3 [@Rcore] using the rjags package v4.10 [@rjags].
+
+** TODO: Adjust fig 2 to make colors consistent between panels and with other figures**
 
@@
 $$c_{A|k} = Multinomial\left( n_k, \frac{\vec{\delta p_t}}{\sum \vec{\delta p_t}}  \right)$$
+
+** TODO: Check equation formatting **
 
@@
 
-Our forth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT.
+Our fourth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT.
 Specifically PFT 2 was identical to the VSEM default, but PFT 1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT.
@@
 
+** TODO: TRY AGAIN AT GETTING MAPS ON SAME PROJECTION **
+
 Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades.
@@
 
-Our study region was a Landsat tile over central Oregon, USA located between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude.
-For this tile we extracted data from three derived Landtrendr datasets (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html, tile h03v04):  biomass, cover, and disturbance.
+Our study region was a portion of central Oregon, USA that includes the Ameriflux tower used later to initialize the VSEM (see next section).
+We used biomass, cover, and disturbance data derived from time-series Landsat imagery processed with the LandTrendr algorithm under a separate project.
+Available on a tile-basis (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html), we downloaded these data for tile h03v04, which corresponds to the region between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude.
 All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover.
-The biomass product calibrated Landtrendr-processed Landsat data against field plot data using the gradient nearest neighbor (GNN) methodology to produce an annual biomass estimates [@Kennedy2018].
+The biomass product was constructed by linking Landtrendr-processed Landsat data with field plot data using the gradient nearest neighbor (GNN) methodology [@Ohmann2002] to produce annual biomass estimates [@Kennedy2018].
 The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S.
 National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018].
-To similify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_).
-The disturbance product **need text from Robert** [@KENNEDY2010]. 
+To simplify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_).
+The disturbance product linked LandTrendr-derived disturbance information [@KENNEDY2010] with photo-interpreted labels of disturbance to attribute drivers to each disturbance event [@KENNEDY2015].  
 To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.
@@
 Finally, because the process error within a dynamical data assimilation system is not identical to the calibration residual error, we estimated the leaf and wood process errors (Ql, Qw) based on the difference between the predicted and observed net annual changes in each of these pools.
-Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state.
+Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state (i.e.
+any change in modeled soil C was assumed to be an error).
 
@@
 Next, using the US-ME2 initial conditions we performed a >10k year spin-up, and then sampled carbon pool ensemble members from the second half of this time-series.
-Next, for each ensemble member we imposed at disturbance that reduced wood and leaf biomass to 1% of their equilibrum values, then simulated a 200-year recovery trajectory.
+Next, for each ensemble member we imposed at disturbance that reduced wood and leaf biomass to 1% of their equilibrium values, then simulated a 200-year recovery trajectory.
 Finally, for each ensemble member we selected the stand age that minimized the squared error between the modeled biomass and the initial biomass, and then selected the leaf and soil C pools at that stand age.
@@
 For the assimilation experiment, each site was assimilated separately.
-In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^4$ based on [@Kennedy2018] Figure 2.
+In the initial round of experiments the observation error was set to a standard deviation of $4.2\: kg/m^2$ based on [@Kennedy2018] Figure 2.
 Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017).
@@
 The skill of these assimilation experiments were assessed the same as the original.
-In addition, changes in skill and filter divergence across experiments were assessed as a function of R and Q focusing on two response variables: (1) the filter coverage, defined as the proportion of sites where our divergence metric falls between -1 and 1 (i.e., where the difference between the observation and Analysis is less than half the width of the Analysis 95% CI); and (2) the detection threshold, defined as the change in AGB where the probability of detecting a disturbance is 50% as inferred by the GLM of disturbance size versus probability of disturbance.
+In addition, changes in skill and filter divergence across experiments were assessed as a function of R and Q focusing on two response variables: (1) the filter coverage, defined as the proportion of sites where our divergence metric falls between -1 and 1 (i.e., where the difference between the observation and Analysis is less than half the width of the Analysis 95% CI); and (2) the detection threshold, defined as the change in AGB where the probability of detecting a disturbance is 50% as inferred by the GLM of disturbance size versus probability of disturbance detection.
 

### 2023-03-31 results.Rmd
Google Drive track changes

@@
 This posterior is still multimodal, but in this specific instance the weight assigned the undisturbed mode (0.023) is too low to be visually notable.
-Noteworthy is that the LAI in years 12 and 15 is actually lower than that in the disturbance year but the multinomial, yet the model still assigns this as being undisturbed.
+Noteworthy is that the LAI in years 12 and 15 is actually lower than that in the disturbance year, yet the model still assigns this as being undisturbed.
 This occurs because in these years the observed low LAI is compatible with the observation error around the undisturbed forecast, which has a higher prior probability than the odds of subsequent disturbance.
@@
 
+** TODO: in bivariate plots, data is hard to see **
+
 Experiment 3 (Figure \@ref(fig:multiPseudo) top row) illustrates the ability of the multinomial model to select between more than one disturbance.
@@
 grassland, shrubland, agriculture), demonstrating that the model can handle land use and land cover changes.
-Finally, experiment 5 (Figure \@ref(fig:multiPseudo) bottom row) illustrates the extension of the assimilation to multiple sites, one that experiences and disturbance and one that does not.
-In the years prior to the disturbance the multivariate nature of the assimilation allows us to borrow strength across sites, with observations at one site reducting uncertainty not just at that site but at the other site as well based on the cross-site forecast covariance.
+Finally, experiment 5 (Figure \@ref(fig:multiPseudo) bottom row) illustrates the extension of the assimilation to multiple sites, one that experiences a disturbance and one that does not.
+In the years prior to the disturbance the multivariate nature of the assimilation allows us to borrow strength across sites, with observations at one site reducing uncertainty not just at that site but at the other site as well based on the cross-site forecast covariance.
 That said,this example illustrates the ability of the multinomial model to avoid the impact of spurious correlations across sites during disturbance events.
@@
 
-```{r multiPseudo, echo=FALSE, message = FALSE, fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.asp=1.25}
+```{r multiPseudo, echo=FALSE, message = FALSE, fig.cap="Simulated experiments 3-5: Left hand column shows time series for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.asp=1.25}
 
-## figure with 3 rows, 2 col showing timeseries in rt and bivariate in left
+## figure with 3 rows, 2 col showing time series in rt and bivariate in left
 #par(mfrow=c(3,2))
@@
 
-Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from Ameriflux site US-ME2.
+Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from AmeriFlux site US-ME2.
 Best fit parameters are given in Supplemental Table \@ref(tab:parameters).
@@
 ## Remotely-sensed disturbance in Oregon
+
+** TODO: Dongchen wants examples for other disturbances and to know why the assimilation is so constrained **
 
@@
 ```
-In total we assimilated 365 example AGB timeseries from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively.
+In total we assimilated 365 example AGB time series from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively.
 Initial applications of the multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries.

### 2023-04-12 Paper.Rmd
additional citations

@@
 An alternative to look-up tables is to rely on process-based models to capture the spatiotemporal changes in the C cycle post disturbance.
-However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), where major classes of disturbance are often absent (e.g., forest pests, [@Hicke2012; @DietzeMatthes2014]).
+However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances and because the representation of disturbance is highly simplified in most models (@FisherKoven2020; @KEANE2015), where major classes of disturbance are often absent (e.g., forest pests, [@Hicke2012; @DietzeMatthes2014]).
 Ultimately, MRV efforts should reflect the actual disturbance events, not those that were predicted.

### 2023-04-12 discussion.Rmd
additional citations

@@
 One way to do this is to bring in additional data on the same pool, in this case AGB, reducing the overall observation error.
-This could be done through some combination of field data [e.g., Forest Inventory and Analysis, CITE], lidar [e.g., GEDI, CITE], microwave vegetation optical depth [e.g., SMAP, CITE Koenings], or radar [e.g., BIOMASS, CITE].
+This could be done through some combination of field data [e.g., Forest Inventory and Analysis, @burrill_2021], lidar [e.g., GEDI, @duncanson2022], microwave vegetation optical depth [e.g., SMAP, @konings2017], or radar [@sinha2015review].
 This was not done in this example because data products based on different observations have different spatial resolutions, which would need to be addressed.
@@
 observations of LAI would help constrain AGB and observations of AGB would help constrain LAI).
-Existing LAI data products [e.g., MODIS, CITE] have the same spatial mismatch issue discussed above.
+Existing LAI data products [e.g., MODIS, @yan2016] have the same spatial mismatch issue discussed above.
 Alternatively, a data product based on LandSAT would be at the same resolution but would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e.
@@
 An alternative to analytical approximation is to rely on machine learning approaches to build an emulator (a.k.a.
-surrogate model) [CITE].
+surrogate model) [@Fer2018].
 While emulators are often used to approximate process-based models, predicting what the model would have output given a set of inputs, here we want to construct an emulator to that Analysis step, predicting what the Multinomial filter would output (probability of disturbance, ecosystem pool means and variances) as a function of the Forecast model state and the observed data constraints.

### 2023-04-12 methods.Rmd
additional citations

@@
 To keep things simpler, this experiment reverted to a single disturbance type, same as experiment 2, except we dropped the disturbance probability to 2.5%.
-This scenario was designed to test a concern we had in previous research that a disturbance occurring at one site could cause the biomass to decline at other undisturbed sites because of the cross-site covariances [CITE Viskari, Hamze].
+This scenario was designed to test a concern we had in previous research that a disturbance occurring at one site could cause the biomass to decline at other undisturbed sites because of the cross-site covariances [@Dokoohaki2022].
 
@@
 
-For each location we next extracted PAR driver data from the Daymet product [CITE] using the daymetr R package [@daymetr].
+For each location we next extracted PAR driver data from the Daymet product [@thornton2021] using the daymetr R package [@daymetr].
 For each location the carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale.

### 2023-04-25 Paper.Rmd
Final edits for submission. Edits for resubmission to include double spacing and line numbers

--- 2023-03-31 Paper.Rmd
@@
 date: '`r Sys.Date()`'
+header-includes:
+    - \usepackage{setspace}\doublespacing
+    - \usepackage[left]{lineno}
+    - \linenumbers
 output:
@@
 We also demonstrated the ability to fuse multiple data types to successfully distinguish different disturbance types, and to probabilistically capture vegetation type ‘switching’ events within the assimilation and ensemble forecast.
-To apply this to real-world data we calibrated VSEM against eddy-covariance and ancillary data from the AmeriFlux US-Me2 tower.
-We then selected 356 conifer forest sites for testing, using the Landtrendr disturbance product to stratify by four disturbance types (cut, burn, pest, and other).
+To apply this to real-world data we selected 356 conifer forest sites for testing, using the Landtrendr disturbance product to stratify by four disturbance types (cut, burn, pest, and other).
 We then assimilated the 30m Landtrendr annual aboveground biomass product from 1990-2017 and assessed the rate of disturbance detection.
@@
 
-Researchers have devoted considerable time and resources to developing "bottom-up" terrestrial carbon (C) cycle data products at landscape- to global-scales [@SOCCR2].
-Such efforts have drawn upon various combinations of remote sensing, inventory data, field-based sensors (e.g., eddy covariance), and process-based models with the aims of improving both our understanding of C cycle processes and the monitoring, reporting, and verification (MRV) of C stocks and fluxes [@AR6WG3; @SOCCR2].
+Considerable time and resources have been devoted to developing "bottom-up" terrestrial carbon (C) cycle data products at landscape- to global-scales [@SOCCR2].
+Such efforts have drawn upon various combinations of remote sensing, inventory data, field-based sensors (e.g., eddy covariance), and process-based models with the dual aims of improving our understanding of C cycle processes and the monitoring, reporting, and verification (MRV) of C stocks and fluxes [@AR6WG3; @SOCCR2].
 MRV in particular is driven by a combination of local, national, and international reporting requirements (e.g.
@@
 
-A particularly critical part of carbon MRV is the impact of land use, land use change, forestry (LULUCF) and natural disturbance (e.g.
-wildfire, pests, wind), which can have major impacts on carbons stocks and accumulation, quickly turning C sinks into large sources [@harris2016; @Law2018; @Hudiburg_2019] .
-**TODO: Bev says delete: Natural and anthropogenic disturbance events are also places where management and policy can have the greatest impact on carbon pools and fluxes, in terms of increasing resilience, managing restoration and recovery, and avoiding transitions to alternative system states.
-Looking to the future, many types of disturbance are predicted to become more frequent and/or more intense [@seidl_forest_2017; Dale2001].
-Compounding this, C stocks are likely to be more sensitive to changes in disturbance regimes than to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell2020; @Andrews2018].** While there are a variety of approaches for monitoring and understanding C cycle processes for intact ecosystems, when it comes to disturbance the status quo is based on detecting disturbances (e.g., via remote sensing) and then using a simple look up table (i.e., emission factors) to quantify C cycle impacts [@eggleston2006; @Mitchell2017; @BULLOCK2021; @harris2021].
+Particularly critical parts of carbon MRV are the impacts of land use, land use change, forestry (LULUCF) and natural disturbance (e.g.
+wildfire, pests, wind), which can have major impacts on carbons stocks and accumulation, quickly turning C sinks into large sources [@harris2016; @Law2018; @Hudiburg_2019].
+Looking to the future, many types of disturbance are predicted to become more frequent and/or intense [@seidl_forest_2017; Dale2001].
+Compounding this, C stocks are likely to be more sensitive to changes in disturbance regimes than to the direct ecophysiological effects of climate and elevated $CO_2$ [@McDowell2020; @Andrews2018].
+Natural and anthropogenic disturbance events are also places where management and policy can have the greatest impact on carbon pools and fluxes, in terms of increasing resilience, managing restoration and recovery, and avoiding transitions to alternative system states.
+
+While there are a variety of approaches for monitoring and understanding C cycle processes for intact ecosystems, when it comes to disturbance the status quo is based on detecting disturbances (e.g., via remote sensing) and then using a simple look up table (i.e., emission factors) to quantify C cycle impacts [@eggleston2006; @Mitchell2017; @BULLOCK2021; @harris2021].
 Active sensors such as lidar and SAR more directly capture changes in aboveground biomass, but do not inform other pools and fluxes [@Mitchell2017].
@@
 An alternative to look-up tables is to rely on process-based models to capture the spatiotemporal changes in the C cycle post disturbance.
-However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances (CITE) and because the representation of disturbance is highly simplified in most models (CITE), where major classes of disturbance are often absent (e.g., forest pests, [@Hicke2012; @DietzeMatthes2014]).
+However, by themselves process-based models are going to misrepresent many disturbances, either under- or over-predicting individual events, both because there is an inherently stochastic component to disturbances and because the representation of disturbance is highly simplified in most models [@FisherKoven2020; @KEANE2015], where major classes of disturbance are often absent (e.g., forest pests, [@Hicke2012; @DietzeMatthes2014]).
 Ultimately, MRV efforts should reflect the actual disturbance events, not those that were predicted.
-On the other hand, prescribing disturbance in models ignores the uncertainty in disturbance detection and brings us back to using look-up tables to prescribe how the disturbance is represented in the model.
+On the other hand, prescribing disturbance in models ignores the uncertainty in disturbance detection and brings us back to using look-up tables to capture how the disturbance is represented in the model.
 
-**TODO: Bev suggests replacing the above paragraph with the IPCC AR6 text: Technological barriers and opportunities Monitoring, reporting, and verification.
-Development of satellite technologies to assess potential deforestation has grown in recent years with the release of 30 m data by Hansen et al.
-(2013), however, these data only captures tree cover loss, and increasing accuracy over time may limit its use for trend analysis (Ceccherini et al.
-2020; Palahí et al.
-2021).
-Datasets on forest losses are less well developed for reforestation and afforestation.
-As Mitchell et al.
-(2017) point out, there has been significant improvement in the ability to measure changes in tree and carbon density on sites using satellite data, but these techniques are still evolving and improving and they are not yet available for widespread use.**
+```{r conceptualFig, fig.width=7, fig.cap="Conceptual representation of a state data assimilation workflow.
+(1) Uncertainties in model inputs (e.g., initial conditions [IC], drivers, parameters) are sampled to generate ensembles of model inputs that (2) are fed into a model to produce probabilistic forecasts [blue].
+(3) When new observations [green] become available they are (4) analyzed statistically to produce an updated estimate of the state of the system [pink].
+Here we introduce a new Analysis algorith, the Multinomial Filter and test is using the Very Simple Ecosystem Model (VSEM) as the model and remotely sended aboveground biomass (AGB) as the data constraint.",echo=FALSE}
+knitr::include_graphics("WorkflowFigure.png",dpi=144)
+```
 
 State data assimilation (SDA) offers a promising alternative to these two approaches, directly addressing their challenges while building on their respective strengths, and is already being used in other aspects of carbon MRV [@Baatz2021].
-The basic idea behind SDA is that a process model is run numerous times to make a probabilistic prediction of the full carbon cycle (pools and fluxes) from one point in time to the next.
+The basic idea behind SDA is that a process model is run numerous times to make a probabilistic prediction of the full carbon cycle (pools and fluxes) from one point in time to the next (Figure \@ref(fig:conceptualFig)).
 It does this by sampling over the uncertainties about the current state of the system, while providing the option also sampling the parameter, driver, and process (residual) uncertainties.
@@
 In essence, data assimilation “nudges” the model back towards the data, with the size of the nudge proportional to the relative uncertainties in the model (prior) and the data (Likelihood).
-The resulting posterior estimates are able to fuse numerous data constraints simultaneously, are robust to large amounts of missing observations in space and time, and provide a comprehensive uncertainty estimate.
-SDA can even update pools and fluxes that are not directly observed, which it does by leveraging model ensemble covariance between observed and unobserved variables.
+Importantly, the resulting posterior estimates are able to fuse numerous data constraints simultaneously, are robust to large amounts of missing observations in space and time, and provide a comprehensive uncertainty estimate.
+SDA can even update pools and fluxes that are not directly observed, which it does by leveraging the model ensemble covariance between observed and unobserved variables.
 Finally, unlike monitoring-only systems, data assimilation systems can seamlessly transition from reconstructing the past (a.k.a.
@@
 
-MCD was supported by funds from NASA (20-CMS20-0030) and the National Science Foundation (NSF 1458021, NSF 1638577, NSF 1702996, NSF 2213580).
-
-RK was supported by NASA grants NNX16AQ25G and NN12AP76G
-
-**COAUTHORS: ADD ACKNOWLEDGEMENTS **
+MCD and SPS were supported by funds from NASA (20-CMS20-0030, 16-CMS16-0007), while MCD was additionally supported by the National Science Foundation (NSF 1458021, NSF 1638577, NSF 1702996, NSF 2213580).
+RK was supported by NASA grants NNX16AQ25G and NN12AP76G.
+We would like to thank members of MCD's Ecological Forecasting lab (Alexis Helgeson, Charlotte Malmborg, Zoey Werbin, Dongchen Zhang) for input they provided on the draft manuscript.
 

### 2023-04-25 StateTransition.Rmd
Final edits for submission. Edits for resubmission to include double spacing and line numbers

@@
 nlcd$ID = as.numeric(nlcd$ID)
-rat = dplyr::left_join(rat,nlcd,by="ID")
+#rat = dplyr::left_join(tibble::tibble(rat),nlcd,by="ID")
+rat = nlcd[as.integer(nlcd$ID) %in% rat$ID,]
 levels(cov04) <- rat 

### 2023-04-25 discussion.Rmd
Final edits for submission. Edits for resubmission to include double spacing and line numbers

@@
 Rather than nudging the posterior state stranded into a "no man's land" between the disturbed and undisturbed states, the Multinomial successfully makes a discrete jump in state.
-Through the simulated data experiments we also demonstrated: (1) That the algorithm is robust to false positives -- once in the disturbed state it does not interpret further dips in LAI as new disturbances since those data are consistent with the observation error around the model's predicted recovery trajectory; (2) That the algorithm is successfully able to fuse observations of multiple pools simultaneously; (3) That it can distinguish among multiple types of disturbance; (4) That it can accommodate disturbances that result in a discrete change in plant functional type (e.g., forest to grassland or agriculture); and (5) That the sharing of information across space in multi-site assimilations, which normally increases precision and handles missing observations, is not thrown off by disturbance events (e.g., reducing biomass in undisturbed sites).
-From a data assimilation perspective, the algorithm works as intended and is superior to conventional alternatives at detecting disturbance.
+Through the simulated data experiments we also demonstrated: (1) That the algorithm is robust to false positives -- once in the disturbed state it does not interpret further dips in biomass as new disturbances since those data are consistent with the observation error around the model's predicted recovery trajectory; (2) That the algorithm is successfully able to fuse observations of multiple pools simultaneously; (3) That it can distinguish among multiple types of disturbance; (4) That it can accommodate disturbances that result in a discrete change in plant functional type (e.g., forest to grassland or agriculture); and (5) That the sharing of information across space in multi-site assimilations, which normally increases precision and handles missing observations, is not thrown off by disturbance events (e.g., reducing biomass in undisturbed sites).
+From a data assimilation perspective, the algorithm works as intended and is superior to conventional data assimilation algorithms at detecting disturbance.
 
@@
 Second, most of these algorithms rely on a moving-window approach as part of their "status quo" prediction, which requires accessing and processing a time-series of data from the past every time a prediction is updated.
-The iterative algorithms that form the basis for data assimilation (e.g., Kalman Filter) have on occasion been applied to disturbance detection (using statistical, rather than process-based model) and have demonstrated greater computational efficiency [@Ye2020].
+By contrast, the iterative algorithms that form the basis for data assimilation (e.g., Kalman Filter) only need to look at one observation at a time.
+Such algorithms have on occasion been applied to disturbance detection (using statistical, rather than process-based model) and have demonstrated greater computational efficiency [@Ye2020].
 @Ye2020 avoids the "no man's land" nudge problem by the simple expedient of performing threshold-based disturbance detection prior to assimilation and not assimilating disturbed sites.
-Third, other algorithms are relying on simple statistical models for their predictions of the status quo, such as harmonic time-series models, while the Multinomial model is relying on the predictions of a process-based ecosystem model.
-This has three distinct advantages: (1) We are leveraging greater process understanding in the status quo prediction, potentially making the Multinomial model more robust in the face of extreme events that are not disturbances (e.g., unusually cold or dry) but which alter the magnitude or phenology of seasonal cycles; (2) While not included in the simple "toy" model used in these analyses, process-based models also have the potential to leveraging greater process understanding in predicting the likelihood of disturbances, for example by accounting for fuel loads, fuel moisture, and ignition rates in predicting the likelihood of fire; (3) Because process-based models are predicting multiple pools and fluxes that can be compared to multiple distinct types of data, the Multinomial algorithm provides a natural way to fuse multiple types of data that would either reinforce a disturbance detection (if all sources are pointing in the same direction) or call it into question.
+Third, other algorithms rely on simple statistical models for their predictions of the status quo, such as harmonic time-series models, while the Multinomial model can make use of predictions from process-based ecosystem models.
+This has three distinct advantages: (1) We are leveraging greater process understanding in the status quo prediction, potentially making the Multinomial model more robust in the face of extreme events that are not disturbances (e.g., unusually cold or dry) but which alter the magnitude or phenology of seasonal cycles; (2) While not included in the simple "toy" model used in these analyses, process-based models also have the potential to leverage greater process understanding in predicting the likelihood of disturbances, for example by accounting for fuel loads, fuel moisture, and ignition rates in predicting the likelihood of fire; (3) Because process-based models are predicting multiple pools and fluxes that can be compared to multiple distinct types of data, the Multinomial algorithm provides a natural way to fuse multiple types of data that would either reinforce a disturbance detection (if all sources are pointing in the same direction) or call it into question.
 While conventional disturbance detection algorithms are overwhelmingly based on multispectral remote sensing from either a single sensor or a pre-harmonized fusion of multiple sensors [e.g., Harmonized Landsat Sentinel-2, @CHEN2021; @SHANG2022], data assimilation approaches can simultaneously integrate a wide range of relevant data types, for example lidar, microwave, thermal, solar-induced fluorescence, and field data.
@@
 
-** TODO from bev: Example of EnKF used with our ponderosa pine data: Spadavecchia, L., M.
-Williams, B.E.
-Law.
-2011.
-Uncertainty in predictions of forest carbon dynamics - separating driver error from model error.
-Ecological Applications 21:1506–1522.
-Doi:10.1890/09-1183.1.
-Cite Quaife, T., P.
-Lewis, M.
-De Kauwe, M.
-Williams, B.E.
-Law, M.
-Disney, P.
-Bowyer.
-2008.
-Assimilating canopy reflectance data into an ecosystem model with an ensemble Kalman filter.
-Remote Sensing of Environment 112:1347-1364.
-**
-
 The most obvious issue with the initial application of the Multinomial filter to real-world data was the high incidence of filter divergence -- of the model becoming overconfident in itself and thus diverging from the observations.
@@
 The underlying causes of filter divergence are a result of either an underaccounting of model uncertainty or an overestimation of observation error in the data, both of which are likely at play in this example.
+
 On the model side, forecast uncertainty is associated with five factors: initial condition uncertainty, driver uncertainty, parameter uncertainty, parameter heterogeneity, and process error [@Dietze2017b].
@@
 First, the assumption of constant variance is unlikely to be true because AGB is non-negative, which means the current interval estimates for low biomass sites include a substantial probability of negative biomass.
-More often the uncertainties in AGB are heteroskedastic and increasing in absolute magnitude as AGB increases, a property shared with the allometric models used to estimate AGB in the field calibration data [@van2005; @berner2016].
+More often the uncertainties in AGB are heteroskedastic and increasing in absolute magnitude as AGB increases, a property shared with the allometric models used to estimate AGB in field calibration data [@van2005; @berner2016].
 This heteroskedasticity is particularly important for disturbance detection because it means we are likely significantly overestimating the observation error associated with low biomass (disturbed) conditions, contributing directly to the Multinomial model's inability to distinguish such biomass drops from observation error.
@@
 
-In contrast to our "off-the-shelf" application of the Multinomial filter, our computational experiments demonstrated that when we increase the model uncertainty and reduce the observation uncertainty that we significantly reduced the rate of filter divergence and were able to reliably detect all but the smallest disturbance events.
+In contrast to our "off-the-shelf" application of the Multinomial filter, our computational experiments demonstrated that when we increase the model uncertainty and reduce the observation uncertainty, we significantly reduced the rate of filter divergence and were able to reliably detect all but the smallest disturbance events.
 Indeed, filter divergence declined, and disturbance detection increased, systematically as a function of the magnitude of the observation error. 
@@
 
-## Future Directions: Current algorithm
+## Future Directions: Algorithm Development
 
@@
 We start by addressing future directions that make use of the current algorithm "as is", discussing ways that we could better leverage multiple data constraints, improved disturbance priors, and existing disturbance data products.
+We then discuss three possible longer-term improvements: accounting for space when estimating the probability of disturbance, $\vec{\rho_t}$; adding memory to the algorithm to be able to better detect disturbance occurrence and type; and using spatially-implicit approaches that track fractional area.
+
 In terms of multiple data constraints, it is worth noting that, unlike in the simulated data experiment, our real-world application only relied on a single data constraint.
@@
 This could be done through some combination of field data [e.g., Forest Inventory and Analysis, @burrill_2021], lidar [e.g., GEDI, @duncanson2022], microwave vegetation optical depth [e.g., SMAP, @konings2017], or radar [@sinha2015review].
-This was not done in this example because data products based on different observations have different spatial resolutions, which would need to be addressed.
-Fusing in a data assimilation framework is well established but was judged beyond the scope of the current study because it requires running a fully spatial assimilation rather than a simple site-scale time series model.
-Basically, one either needs to run the assimilation at a fine spatial resolution and use an observation operator to map many model grid cells to one observation, or one could run the assimilation at a coarse spatial resolution and use an observation operator to map one model grid cell to many observations [@Dietze2017a].
+This was not done in this example because data products based on different observations have different spatial resolutions.
+Data assimilation approaches for fusing data with different resolutions are well established but were judged beyond the scope of the current study because they requires running a fully-spatial assimilation, rather than a simple site-scale time series model.
+Basically, one either needs to run the assimilation at a fine spatial resolution and use an observation operator to map many model grid cells to one observation, or to run the assimilation at a coarse spatial resolution and use an observation operator to map one model grid cell to many observations [@Dietze2017a].
 
-Another way would be to bring in additional data on different pools, such as LAI, which often has lower observation errors.
-In addition to the direct increase in data volumes, the positive correlation between LAI and AGB would allow the assimilation algorithm to borrow strength indirectly (i.e.
+Another way to leverage additional data constraints would be to bring in observation of different pools, such as LAI, which often has lower observation errors.
+In addition to the direct increase in data volumes, the positive correlation between model predicted LAI and AGB would allow the assimilation algorithm to borrow strength indirectly (i.e.
 observations of LAI would help constrain AGB and observations of AGB would help constrain LAI).
@@
 errors in data products based on the same underlying spectra would not be independent).
-In addition to bringing in additional data constraints on different pools, we could also constrain our disturbance estimates using observations on different ecosystem fluxes and processes, such as eddy covariance (carbon, water, and energy fluxes), solar induced fluorescence (a proxy for gross primary productivity), thermal imaging (a proxy for water stress and evapotranspiration).
+In addition to bringing in additional data constraints on different pools, we could also constrain our disturbance estimates using observations on different ecosystem fluxes and processes, such as eddy covariance (carbon, water, and energy fluxes), solar induced fluorescence (a proxy for gross primary productivity), and thermal imaging (a proxy for water stress and evapotranspiration).
 
-In addition to constraining our estimates of disturbance through the ecosystem state variables, we can also improve them by improving our priors on the disturbance probability. 
-In our simple case study we relied on static and spatially-homogeneous prior probabilities on different disturbance rates, an approach somewhat similar to the simplest state-and-transition models of land use and disturbance [@Daniel2016].
-It is possible to use more complex state-and-transition models that would allow these priors to be spatially heterogeneous (e.g., relying on spatial covariates) or time evolving (e.g., relying on temporal covariates).
-Similarly, it is also possible to build state-and-transition that leverage the current state of a pixel (e.g., increasing fuel loads increase the probability of fire), the state of adjacent pixels (e.g., spatially contagious disturbances such as fire), or even those involving more distant spatial relations (e.g., dispersal kernels of biotic pests and pathogens) [@johnstone2011; @hudgins2017].
+In addition to constraining our estimates of disturbance through the ecosystem state variables, we could also improve our disturbance probability priors. 
+In our simple case study we relied on static and spatially-homogeneous prior probabilities on different disturbance rates, an approach analogous to the simplest state-and-transition models of land use and disturbance [@Daniel2016].
+It is possible to use more complex state-and-transition models that would allow priors to be spatially heterogeneous (e.g., relying on spatial covariates) or time evolving (e.g., relying on temporal covariates).
+Similarly, it is also possible to build state-and-transition models that leverage the current state of a pixel (e.g., increasing fuel loads increase the probability of fire), the state of adjacent pixels (e.g., spatially contagious disturbances such as fire), or even those involving more distant spatial relations (e.g., dispersal kernels of biotic pests and pathogens) [@johnstone2011; @hudgins2017].
 In addition to relying on external state-and-transition models, one of the appeals of working with process-based ecosystem models is that many of them include submodules that represent different types of disturbance internally (e.g., fire, pests).
@@
 For example, in our Oregon case study it is clear that the Landtrendr disturbance product is "seeing" more about disturbance than is being expressed in the AGB product alone, as the disturbance product is looking for different information (e.g., weighting bands differently, training to distinguish among types of disturbance not just a loss of biomass).
-Therefore, one alternative to relying on models to predict disturbance priors is to use the disturbance product and its classification error statistics to specify priors itself to identify the error.
+Therefore, one alternative to relying on models to predict disturbance priors is to use the disturbance product, and its classification error statistics, to specify priors.
 This approach could be particularly valuable for distinguishing among alternative disturbance types (e.g., forest wildfire, forestry, and conversion to agriculture or development), especially for disturbance products that make use of bands outside of the normal optical range (e.g., thermal-based active fire products).
 Applying this approach was not done in our test case because we wanted to keep the disturbance product separate for purposes of validation, but would be relatively straightforward to implement.
-It's important to note that when using external disturbance data as priors we're still not prescribing disturbances deterministically, during the assimilation observations about ecosystem pools and fluxes would thus either reinforce these products or call them into question.
+It's important to note that when using external disturbance data as priors we're still not prescribing disturbances deterministically.
+During the assimilation, additional observations about ecosystem pools and fluxes would thus either reinforce these products or call them into question.
 Finally, it is worth noting that relying on process-based models and external disturbance data products are not mutually exclusive.
@@
 For example, one could start with the models as priors and then update these using one or more disturbance product observations [e.g., using the Bayesian Updating of Land Cover (BULC) algorithm, @CARDILLE2016; @Crowley2019] and then use these posteriors as the priors in the data assimilation.
+
+In terms of spatial disturbance probability, the current filter estimates a whole vector of probabilities, one for each site, but does not explicitly account for spatial adjacency.
+In practice, almost all types of disturbance involve some degree of shared risk, if not true spatial contagion, whereby a disturbance at one site usually increases the probability of disturbance in near-by sites.
+Accounting for this could be done by adding a hierarchical layer to the estimation of $\rho$ that includes a spatial covariance structure, for example by using a link function (e.g., logit) and modeling the covariance as a multivariate Normal with a standard distance-dependent covariance model [@Banerjee2014] or spatial basis functions and random effects [@Wikle2019].
+
+As for the idea of incorporating memory, one limitation of the existing approach is that, like most data assimilation filters, it is strictly Markovian - states are updated solely based on the current model state and current observations.
+Imagine the case where an anomalously low AGB value is observed but at the next time point AGB returns to normal.
+At the time that the low observation is made it may trigger the Multinomial filter to predict a high probability of disturbance, but after the next observation becomes available it becomes much more likely that the low value was simply observation error.
+In this case what we want is the ability not just to restore the AGB to the undisturbed state, but also to jump back to the previous observation and assign it to be undisturbed as well.
+Similarly, if a disturbance occurs but observations are compatible with two types of disturbance (e.g., clear cut versus conversion to agriculture), the algorithm may assign both a 50/50 probability at that time point.
+At the next time point it may update that assignment based on new information (e.g., the new vegetation is a crop), in which case we also want to jump back in time and say that the original disturbance was agricultural conversion.
+Implementing this "memory" feature is more challenging, as it requires writing a statistical model that accounts for multiple points in time and more complex multi-temporal disturbance state scenarios, as well as the informatics pipelines to save previous states and restart model ensembles from an earlier point in time.
+
+A final potential algorithm improvement would be to adopt a spatially implicit approach to tracking disturbance.
+Spatially implicit approaches are a common feature of global vegetation models, which for computational reasons might represent a larger (e.g., 0.5-by-0.5 degree) grid cell as containing multiple vegetation types, each with a fractional area, rather than keeping track of where in each grid cell those vegetation types are located.
+A subset of these models also keep track of both disturbance events and time since disturbance (a.k.a.
+patch age) in this way, updating fractional areas dynamically [@Moorcroft2001].
+More recently, @McCabe2019 proposed a way to account not just for fractional area but also patch adjacency using a spatially implicit approach, allowing for the model to better account for spatial contagion and the distribution of patch sizes.
+Conceptually, the disturbance data assimilation approach we've presented here should be able to keep track of fractional areas of disturbance instead of a discrete disturbed/undisturbed state, for example by adopting a Dirichlet data model instead of a Multinomial.
 
@@
 Unlike a traditional EnKF, the Analysis step of the data assimilation does not have a full analytically solution, and instead needs to be evaluated numerically, for example through standard Bayesian MCMC approaches as was done in this paper.
-However, MCMC is too computationally costly to apply using brute force on anything larger than an individual landscape.
-Here we discuss two possible approximations to the full algorithm, one analytical and the other based on model emulation.
-Analytically, instead of solving the full Multinomial filter jointly, it may be possible to approximate the full model by decomposing the filter sequentially. 
-First, one would analytically apply EnKF to each alternative disturbance state, analytically generating independent posteriors for each $X^L_t$.
-Next, one would then apply a Dirichlet-Multinomial model to the problem of updating $\vec{\rho_t}$, treating $X^L_t$ as known rather than jointly estimated.
-The math for this approach has yet to be fully fleshed out, and it remains to be seen how well this approximation works in practice and what information would be lost by solving the model conditionally rather than jointly.
+Because MCMC is too computationally costly to apply using brute force on anything larger than an individual landscape, scaling up will require approximations to the full algorithm. 
+Analytically, instead of solving the full Multinomial filter jointly, it may be possible to approximate the full model by decomposing the filter sequentially, first applying EnKF to each alternative disturbance state, analytically generating independent posteriors for each $X^L_t$, and then applying a Dirichlet-Multinomial model to update $\vec{\rho_t}$, treating $X^L_t$ as known rather than jointly estimated.
+This approach has yet to be fully fleshed out and it remains to be seen how well this approximation works in practice.
 
@@
 surrogate model) [@Fer2018].
-While emulators are often used to approximate process-based models, predicting what the model would have output given a set of inputs, here we want to construct an emulator to that Analysis step, predicting what the Multinomial filter would output (probability of disturbance, ecosystem pool means and variances) as a function of the Forecast model state and the observed data constraints.
-Emulators address not only the computational cost of scaling up but also the observation that similar sites are going to make similar predictions (e.g., adjacent LANDSAT pixels), and thus fully gridded models are going to perform highly redundant calculations.
-Emulators still require running the model and analysis (in this case the Multinomial filter), but rather than evaluating every location the filter is run for a subset of locations stratified by relevant variables (initial conditions, observed data constraints, and possibly also covariates like climate, soils, etc.).
-Flexible machine learning approaches are then used to build an emulator that predicts the outcome of a fully Bayesian model.
-For places where the full model was not run the emulator is then used to predict the full model, which in essence involves interpolating in environmental space and mapping this back to physical space.
-
-## Future Directions: Algorithm Development
-
-In this final section we discuss three future directions for improving upon the Multinomial filter: accounting for space when estimating the probability of disturbance, $\vec{\rho_t}$; adding memory to the algorithm to be able to better detect disturbance occurrence and type; and scaling up using spatially implicit approaches that track fractional area.
-
-In terms of spatial disturbance probability, the current filter does estimate a whole vector of probabilities, one for each site, and thus estimates will include a posterior covariance structure across sites, but the model is not written in a way that explicitly changes the probability of disturbance based on spatial adjacency.
-In practice, almost all types of disturbance involve some degree of shared risk, if not true spatial contagion, whereby a disturbance at one site usually increases the probability of disturbance in near-by sites.
-Accounting for this could be done by adding a hierarchical layer to the estimation of $\rho$ that includes a spatial covariance structure.
-Because $\rho$ is bound between zero and one, this could be done by including a link function (e.g., logit) and modeling the covariance as a multivariate Normal with a standard distance-dependent covariance model (e.g., exponential, Gaussian, Matern) [@Banerjee2014].
-In practice, estimating larger spatial covariance matrices is computationally demanding and often more efficiently approximated using spatial basis functions and random effects [@Wikle2019].
-
-As for the idea of incorporating memory into the Multinomial filter, one limitation of the existing approach is that (like most data assimilation filters) it is strictly Markovian - states are updated solely based on the current model state and current observations.
-Imagine the case where an anomalously low value is observed for AGB but at the next time point AGB returns to normal.
-At the time that the low observation is made it may trigger the Multinomial filter to predict a high probability of disturbance, but after the next observation becomes available it becomes much more likely that the low value was simply observation error.
-In this case what we want is the ability not just to restore the AGB to the undisturbed state, but also to jump back to the previous observation and assign it to be undisturbed as well.
-As another example, imagine the case where a disturbance does occur but the observations are compatible with multiple types of disturbance (e.g., clear cut versus conversion to agriculture).
-The algorithm may assign both a 50/50 probability at that time point, and then at the next time point it may update that assignment based on new information (e.g., the new year's vegetation is herbaceous).
-Rather than interpreting the 50% probability assigned to clear cut as a new transition at the second time point, what we'd really like to do is jump back in time and say, based on new information, that we're now much more confident that the original disturbance was agricultural conversion.
-Furthermore, because some PFT transitions are not easily reversible (e.g., a grassland cannot shift to a mature forest in one year), this capability to look back would also help protect against the risk that assigning a non-zero probability to a PFT transition could cause more and more of the ensemble to "leak" into that state.
-Implementing this "memory" feature is more challenging, as it requires writing a statistical model that accounts for multiple points in time and more complex multi-temporal disturbance state scenarios, as well as the informatics pipelines to save previous states and restart model ensembles from an earlier point in time.
-
-The final potential algorithm improvement would be to adopt a spatially implicit approach to tracking disturbance.
-Spatially implicit approaches are a common feature of global vegetation models, which might represent a larger (e.g., 0.5-by-0.5 degree) grid cell as containing multiple vegetation types, each with a fractional area, rather than keeping track of where in each grid cell those vegetation types are located.
-A subset of these models also keep track of both disturbance events and time since disturbance (a.k.a.
-patch age) in this way, updating fractional areas dynamically.
-Many of these models are derived from the size-and-age-structured (SAS) approximation introduced in the Ecosystem Demography (ED) model [@Moorcroft2001].
-The idea is that if you consider a patch of vegetation of a specific age and vegetation type, a disturbance might only affect part of the patch, which would cause it to split into two patches (disturbed and undisturbed) each with a new fractional area.
-If this pattern were repeated over time the number of patches would increase exponentially over time, so ED also implements a patch fusion scheme that merges older patches back together when they are judged to have converged to equivalent states.
-More recently, @McCabe2019 proposed a way to account not just for fractional area but also patch adjacency using a spatially implicit approach, allowing for the model to better account for spatial contagion and the distribution of patch sizes.
-Conceptually, the disturbance data assimilation approach we've presented here should be able to be updated to keep track of fractional areas of disturbance instead of a discrete disturbed/undisturbed state, for example by adopting a Dirichlet data model instead of a Multinomial.
+While emulators are often used to approximate process-based models, predicting what the model would have output given a set of inputs, here we want to run the Multinomial filter for a subset of locations, stratified by relevant variables (initial conditions, observed data constraints, and possibly also covariates like climate, soils, etc.), and then construct an emulator to the Analysis step, predicting what the Multinomial filter would output (probability of disturbance, ecosystem pool means and variances) as a function of the Forecast model state and the observed data constraints.
+The emulator is then applied to places where the full model was not run, in essence interpolating in environmental space and mapping this back to physical space.
 
 ## Conclusion
-
-** TODO from Alexis: In the previous sections it might be worth including more info on the pros/cons of additional development as it directly relates to MRV**
 

### 2023-04-25 methods.Rmd
Final edits for submission. Edits for resubmission to include double spacing and line numbers

@@
 # Methods
-
-**TODO Dongchen: consider adding a conceptual diagram to explain the framework and data**
 
@@
 
-The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods.
+The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods.
 Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g.
-carbon pools) and sometimes model parameters.
+carbon pools) and sometimes model parameters (Figure \@ref(fig:conceptualFig)).
 These updated variables are then used as the initial conditions for the next Forecast round.
@@
 The ensemble forecast approach is easy to apply to any model, is particularly good at capturing nonlinear predictions, and unlike many other approaches does not require solving for the adjoint of the model.
-In EnKF the forecast, $f$, at any point in time, $t$, is then summarized in terms of a vector of means across ensemble members, $\mu_{f,t}$ and the sample covariance, $P_{f,t}$.
+In EnKF the multivariate forecast, $f$, at any point in time, $t$, is then summarized in terms of a vector of means across ensemble members, $\mu_{f,t}$ and the sample covariance, $P_{f,t}$.
 In our example we are interested in the three state variables in the VSEM, ${C_l,C_w,C_s}$, at a single site and assume an ensemble size of 200.
@@
 
-The Analysis step in the EnKF is the same as that in the standard Kalman Filter.
+The Analysis step in the EnKF is the same as the standard Kalman Filter.
 Because the forecast step from t-1 to t occurs prior to the observed data at time t, $Y_t$, the forecast of our three state variables is used as an informative _prior_ distribution and is assumed to follow a multivariate Normal (MVN) distribution with mean $\mu_{f,t}$ and covariance $P_{f,t}$.
@@
 Observation error is assumed to follow a MVN distribution with an observation error covariance matrix $R$.
-Furthermore, it is assumed that $R$ is known, which is not unreasonable as many derived data products and sensors report uncertainties, and uncertainties from field data can often be inferred from sampling theory or Monte Carlo resampling / bootstrapping.
+Furthermore, it is assumed that $R$ is known, which is not unreasonable as many derived data products and sensors report uncertainties, and uncertainties from field data can often be inferred from sampling theory or Monte Carlo resampling.
 In many cases new data do not exist for every state variable (e.g.
@@
 The subscript $a$ is used to refer to the Analysis posterior and distinguish this from the prior forecast $f$ mean and covariance for the same timestep.
-While the matrix notation in these equations can make these solutions look complicated, the underlying meaning is accessible.
+While the matrix notation can make these equations look complicated, the underlying meaning is accessible.
 First, the precision of the analysis (i.e.
@@
 This averaging can also be viewed as a 'nudge', pushing the forecast toward the new data, with the magnitude of that nudge determined by the uncertainties in both the forecast and data (i.e.
-highly precise data = big nudge, highly uncertain data = very small nudge).
+more precise data = larger nudge).
 
@@
   load("SSSD.RData")      ## simulation from SingleSite_SingleDisturbance.Rmd
+#  layout(matrix(c(1,1,1,2),nrow=4,ncol=1))
   plot(data$Y,ylim = range(rbind(CI.fN,data$Y)),xlim=xlim,type='n',
@@
   lines(CI.fN[3,],col="blue",lwd=2)         ## Multinomial median
-  lines(Dbar,lwd=2)                         ## Multinomial disturbance rate
-  points(data$Y,pch=18,col=2)               ## data
+  lines(Dbar,lwd=3)                         ## Multinomial disturbance rate
+  points(data$Y,pch=18,col="red")               ## data
   legend("topright",legend=c("EnKF","Multinomial","data","P(dist)"),
-       lwd=5,col=c("orange","blue","red","black")) ##"#b2df8a"
+       lwd=c(5,5,NA,5),pch=c(NA,NA,18,NA),col=c("orange","blue","red","black")) ##"#b2df8a"
+#  plot(Dbar,lwd=2,ylab="P(dist)")
 }
@@
 
-** TODO from zoey for above figure: Labels could be more informative (i.e.
-EnKF = conventional and Multinomial = improved approach).
-Icon for "data" should match the icons rather than a line.
-Is the P(dist) bound at 0,1?
-I barely noticed the black line at first, it could maybe go in a second panel below with a shared x-axis and a distinct y-axis.**
-
-** TODO from zoey for below figure: I think this figure could be more streamlined to better emphasize the difference between the filters.
-Why change the assimilation color from orange to blue in this figure?
-The histogram, to me, seems more intuitive for displaying the data values rather than the forecasted values.
-The second panel doesn't have "forecast" in the color labels which also adds to confusion.
-The X axis could communicate the low-LAI = disturbed, high-LAI = undisturbed, without relying on the colors/legend.
-X axis for this and the Y axis for the above figure could just be "disturbance magnitude" **
-
-```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (gray histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
-Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabilities.
-The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
+```{r concept, echo=FALSE, error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF  a multimodal forecast (gray histogram) is approximated by a unimodal Normal prior (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event.
+Bottom: In the Multinomial filter the same multimodal forecast is represented by a mixture of Normal priors (green) each weighted by their prior (forecast) probabilities.
+The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred.", message=FALSE, warning=FALSE,results='hide'}
 library(rjags)
@@
 lines(xseq,dnorm(xseq,mua,sqrt(Pa)),col='orange',lwd=3)
-legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","orange"))
+#legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","blue"))
 
@@
 lines(xseq,(1-p.dist)*dnorm(xseq,mu1,sqrt(P1)),col="#b2df8a",lwd=3)
-lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="brown",lwd=3)
+lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="#b2df8a",lwd=3)
 lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
 lines(density(as.matrix(jdat)),col="blue",lwd=3)
-legend("top",legend=c("undisturbed","disturbed","data","assimilation"),lwd=3,col=c("#b2df8a","brown","red","blue"))
+legend("top",legend=c("ensemble","prior","data","EnKF","Multinomial"),
+       lwd=3,col=c("gray","#b2df8a","red","orange","blue"))
 ```
@@
 
-To address this fundamental flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodal, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges.
-To begin, let's use $X$ to represent the latent (unobserved) "true" state of the system.
+To address this flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodal, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges.
+To begin, let's use $X_t$ to represent the latent (unobserved) "true" state of the system at time $t$.
 Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent ($L$) continuous state variables of interest, which we'll refer to as $X^L_{k,t}$.
@@
 However, because we have $k$ categories we end up with a multimodal prediction (i.e., a mixture of MVNs).
-Furthermore, these modes need not be distinct; it is perfectly fine in concept for these MVNs to be overlapping.
+Furthermore, these modes need not be distinct; in concept it is perfectly fine for these MVNs to be overlapping.
 In practice the more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabilistically.
-Indeed, to be able to represent our multimodal prediction as a single probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.
+Indeed, to be able to represent our prediction as a single multimodal probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.
 
@@
 
-This is a generalization of the Bernoulli distribution (Binomial with n=1) and for any individual draw from the Multinomial $\vec{\rho_t}$ will be a vector of length $k$ with a 1 in the category being assigned and a 0 for all other categories.
+This is a generalization of the Bernoulli distribution (Binomial with n=1).
+For any individual draw from the Multinomial, $\vec{\rho_t}$ will be a vector of length $k$ with a 1 in the category being assigned and a 0 for all other categories.
 
@@
 $$X_t = X^L_t \vec{\rho_t}$$
+
 which for any individual draw of $\vec{\rho_t}$ assigns $X_t$ that specific $X^L_{k,t}$.
@@
 
-** TODO: Adjust fig 2 to make colors consistent between panels and with other figures**
-
 ### Ensemble Adjustment
 
-When making ensemble forecasts using the Analysis from the traditional EnKF, one needs to sample new initial conditions from the Analysis joint posterior distribution.
-This work fine if the initial conditions are complete (100% of the model variables represented) and when there is no covariance with any of the other uncertainties being propagated (e.g.
-sampling ensemble drivers or posterior parameter distributions).
+When using the EnKF Analysis to initialize the next forecast, one needs to sample a new ensemble of initial conditions from the Analysis joint posterior distribution.
+This works fine if there is no covariance with any of the other uncertainties being propagated (e.g.
+sampling ensemble drivers or posterior parameter distributions) and the initial conditions are complete (100% of the model variables represented).
 In practice, this assumption is frequently violated, for example an above average ensemble member might be above average not just because of its initial condition, but also because it had favorable parameters or drivers, creating a complex covariance between states, parameters, and drivers that would be destroyed by randomly resampling initial conditions.
@@
 Mathematically, the approach is essentially the multivariate generalization of using the prior forecast mean and variance to calculate a Z-score for every ensemble member (i.e.
-express as an standard Normal anomaly with mean 0 and standard deviation 1), and then apply that Z-score to the new posterior analysis mean and variance.
+a standard Normal anomaly with mean 0 and standard deviation 1), and then apply that Z-score to the new posterior analysis mean and variance.
 
-In our Multinomial filter it is important to generalize this concept of ensemble adjustment to both discretely move ensemble members from one category to another, and then to preserve the covariances among state variables, and with drivers and parameters, when updating the discrete categorical assignments of individual ensemble members.
-To reassign forecasted ensemble members to new disturbance classes we first identify the "sink" categories that increased in probability between the prior, $p_{k,t}$, and the posterior, $\bar{\rho}_{k,t}$.
+In our Multinomial filter it is important to generalize this concept of ensemble adjustment to both discretely move ensemble members from one category to another, and then to preserve the covariances among state variables and with drivers and parameters, when updating the discrete categorical assignments of individual ensemble members.
+To reassign ensemble members to new disturbance classes we first identify the "sink" categories that increased in probability between the prior, $p_{k,t}$, and the posterior, $\bar{\rho}_{k,t}$.
 
@@
 
-$$c_{A|k} = Multinomial\left( n_k, \frac{\vec{\delta p_t}}{\sum \vec{\delta p_t}}  \right)$$
-
-** TODO: Check equation formatting **
+$$c_{A|k} = Multinomial\left( n_k,\, \frac{\vec{\delta p_t}}{\sum \vec{\delta p_t}}  \right)$$
 
 Conditional on this reassignment, the continuous states are updated according to [@Anderson2001] using the posterior statistics for that $X^L_{k,t}$.
-It is also worth noting that some cases of class reassignment may also require keeping track of categorical switches in Plant Functional Type (PFT), such as a transition from forest to agriculture, which implies a large jump in PFT parameters as well.
-One could apply the same [@Anderson2001] style adjustment to PFT parameters, but for simplicity here we just maintained alternative parameter ensembles for each PFT, with each ensemble member constructed with the identical anomalies around each PFT's mean..
+It is also worth noting that some cases of class reassignment may require keeping track of categorical switches in Plant Functional Type (PFT), such as a transition from forest to agriculture, which implies a large jump in PFT parameters as well.
+One could apply the same [@Anderson2001] style adjustment to PFT parameters, but for simplicity here we just maintained alternative parameter ensembles for each PFT, with each ensemble member constructed with the identical anomalies around each PFT's mean.
 
@@
 
-For the second experiment, we used the VSEM to extend the Multinomial assimilation model to a single site, single disturbance example with multiple pools.
+For the second experiment, we used the VSEM to extend the single site, single disturbance example to a model with multiple pools.
 In this and the following simulated data experiments we used VSEM's default parameters, drawing ensemble members from Normal distributions centered on the prior mean and assuming a standard deviation that was 1/10th the default uniform prior range.
@@
 Model initial conditions were generated by an initial 25 year ensemble spin-up prior to the 25 year experiment.
-Process error was added to the simulation annually assuming a zero-bound tobit distribution with a process error covariance matrix that was diagonal with variances of 0.1 for all pools.
+Process error was added to the simulation annually assuming a zero-censored Normal (a.k.a.
+tobit) distribution with a process error covariance matrix that was diagonal with variances of 0.1 for all pools.
 Ensemble members experienced disturbance stochastically with a 5% annual probability.
@@
 Our fourth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT.
-Specifically PFT 2 was identical to the VSEM default, but PFT 1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT.
-In addition, we established a transition matrix where disturbance 1 (stand clearing) transitions from PFT1 to PFT2, while disturbance 2 (defoliation) keeps the current PFT is.
-As with the previous scenario, the prescribed disturbance in the pseudodata was type 1 (stand clearing, transition from forest to shrub).
+Specifically PFT2 was identical to the VSEM default, but PFT1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT.
+In addition, we established a transition matrix where disturbance 1 (stand clearing) transitions from PFT1 (forest) to PFT2 (shrub), while disturbance 2 (defoliation) keeps the current PFT.
+The prescribed disturbance in the pseudodata was type 1 (stand clearing).
 
-Our fifth and final experiment extended the analysis to a two-site scenario with spatial covariance across sites and pools, but with a disturbance occurring in one site but not at the other.
+Our fifth experiment extended the analysis to a two-site scenario with spatial covariance across sites and pools, but with a disturbance occurring in one site but not at the other.
 To keep things simpler, this experiment reverted to a single disturbance type, same as experiment 2, except we dropped the disturbance probability to 2.5%.
@@
 
-** TODO: TRY AGAIN AT GETTING MAPS ON SAME PROJECTION **
+Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades, USA.
+In summary, we: (1) calibrated VSEM against field data; (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transitions; (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types; and then (4) assessed the ability of the algorithm to detect different disturbances types and magnitudes.
+Model calibration is discussed in the Supplementary material.
 
-Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades.
-In summary, we: (1) calibrated VSEM against field data, then (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transition, then (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types.
-We then assessed the ability of the algorithm to detect different disturbances types and magnitudes.
+### Remotely-sensed data sources
 
-### Study region and data sources
-
-Our study region was a portion of central Oregon, USA that includes the Ameriflux tower used later to initialize the VSEM (see next section).
-We used biomass, cover, and disturbance data derived from time-series Landsat imagery processed with the LandTrendr algorithm under a separate project.
+Data for our study region include biomass, cover, and disturbance data derived from time-series Landsat imagery processed with the LandTrendr algorithm under a separate project.
 Available on a tile-basis (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html), we downloaded these data for tile h03v04, which corresponds to the region between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude.
-All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover.
+All three products were 30m resolution and annual timestep; biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to the same projection as biomass and cover.
 The biomass product was constructed by linking Landtrendr-processed Landsat data with field plot data using the gradient nearest neighbor (GNN) methodology [@Ohmann2002] to produce annual biomass estimates [@Kennedy2018].
@@
 
-```{r map, echo=FALSE, fig.asp=1, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances"}
+```{r map, echo=FALSE, fig.asp=1, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances.
+Plots are in Albers Equal Area projection with axes in meters and cover a bounding box spanning roughly 43.5 to 45 North latitude by -121 to -123 West longitute and"}
 ## maps of pre-disturbance biomass, cover, and 2004 disturbance
@@
 if(!exists("cov04")) cov04 = raster::raster("cov04.tif")
+if(!exists("dis04"))  dis04 = raster::raster("dis04.tif")
 load("rat.RData")
-raster::plot(cov04,col=rat$color,asp=1,main="NLCD Land Cover",
+if(!exists("cov04b")){
+  #cov04b = raster::projectRaster(cov04,crs=crs(dis04))
+  #save(cov04b,file="cov04b.Rdata")
+  load("cov04b.Rdata")
+}
+raster::plot(cov04b,col=rat$color,asp=1,main="NLCD Land Cover",
             axis.args=list(at=seq_along(rat$description),labels=rat$description,cex.axis=0.4))
@@
 Because VSEM does not internally predict disturbances we instead generated these priors using the Landtrendr disturbance product.
-The disturbance probability prior was calculated simply as the frequency of each disturbance type over the whole time period and the subset of spatial domain that was classified as evergreen forest.
+The disturbance probability priors were calculated as the frequency of each disturbance type over the whole time period and the subset of spatial domain that was classified as evergreen forest.
 The post-disturbance pool size and transition matrix priors were similarly estimated using the remotely sensed biomass and cover estimates (respectively), this time focusing on a sample of 100 disturbance transitions per year for the years 1990-2003 because of the greater computational cost of this estimation process.
@@
 these were considered forest-to-forest transitions in the matrix, as were pixels that stayed forest the whole time).
-Similarly, in order to assess the change in biomass associated with disturbance we calculated the maximum biomass in the three years prior to disturbance and the minimum biomass in either the year of the disturbance or the two years after.
+Similarly, to assess the change in biomass associated with disturbance we calculated the maximum biomass in the three years prior to disturbance and the minimum biomass in either the year of the disturbance or the two years after.
 For all disturbance types we calculated the mean and standard deviation of the change in biomass both in absolute terms and as a proportional change.
-
-### Calibration of VSEM
-
-VSEM was calibrated primarily using data from 2002-2019 from the Ameriflux site US-ME2 (lat.lon), a mature Ponderosa pine stand.
-VSEM pools (leaf, wood, and soil C) were initialized using the earliest available site-specific data from the Ameriflux BADM (Biomass, Ancillary, Disturbance, Management) dataset.
-An additional ten years of leaf and wood biomass data were also available from the BADM, which we included as part of the calibration dataset.
-Also included was the cumulative annual tower NEE (2002-2019).
-Because we were working with annual data we assumed Normal likelihoods for all three response variables, and because our priority was the correct prediction of aboveground biomass we upweighted the wood likelihood by 4 and downweighted the NEE likelihood by 1/4.
-
-Prior to calibration we set the leaf area ratio (LAR) to 15.7 m2/kgC and leaf residence time (tauL) to 4.7 years based on data from the BADM and left the light extinction coefficient (KEXT) at its default (0.5).
-The remaining fit parameters were the light use efficiency (LUE), the fraction of GPP allocated to autotrophic respiration (GAMMA), the fraction of NPP allocated to leaves (Av), the wood and soil C residence times (tauW, tauS), and the residual standard deviations for each of our data constraints (sd_NEE, sd_leaf, sd_wood).
-The fit was done taking a Maximum Likelihood approach using base R's `optim` function.
-Finally, because the process error within a dynamical data assimilation system is not identical to the calibration residual error, we estimated the leaf and wood process errors (Ql, Qw) based on the difference between the predicted and observed net annual changes in each of these pools.
-Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state (i.e.
-any change in modeled soil C was assumed to be an error).
 
@@
 We specifically sampled disturbance locations from the year 2004, which is sufficiently deep in the past that there is an observable disturbance recovery trajectory, but with a sufficiently long time-series prior to disturbance for the data assimilation to come into steady-state with the biomass data product.
-For each selected location we extracted a time-series from the remotely-sensed biomass product for use as the data constraint within the assimilation algorithm.
-The cover and disturbance products were not assimilated, but were just used to construct priors (see above) and assess performance.
+For each selected location we extracted a time-series from the remotely-sensed biomass product as the assimilation data constraint.
+The cover and disturbance products were not assimilated, but were just used to construct priors and assess performance.
 
 For each location we next extracted PAR driver data from the Daymet product [@thornton2021] using the daymetr R package [@daymetr].
-For each location the carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale.
+For each location carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale.
 To start we used the first time point in each biomass time-series as the initial condition for the wood pool.

### 2023-04-25 results.Rmd
Final edits for submission. Edits for resubmission to include double spacing and line numbers

@@
 
-Overall, the results from the simulated data experiments (Table \@ref(tab:pseudoTable)) demonstrate that the proposed multinomial ensemble filter is able to successfully detect disturbances under a range of different conditions.
+Overall, the results from the simulated data experiments (Table \@ref(tab:pseudoTable)) demonstrate that the proposed Multinomial ensemble filter is able to successfully detect disturbances under a range of different conditions.
 
 Experiment 1, which aimed to capture a single disturbance at a single site for a single pool, was already depicted in Figure \@ref(fig:enkf) and Figure \@ref(fig:concept) when discussing the challenges facing traditional Kalman Filter variants.
-Specifically, Figure \@ref(fig:enkf) shows that the multinomial filter (blue) is able to capture the discrete nature of the disturbance at time 10, with the posterior probability of disturbance jumping from near zero before this event to 0.977 for the disturbance, and then dropping back down to near zero in subsequent years.
+Specifically, Figure \@ref(fig:enkf) shows that the Multinomial filter (blue) is able to capture the discrete nature of the disturbance at time 10, with the posterior probability of disturbance jumping from near zero before this event to 0.977 for the disturbance, and then dropping back down to near zero in subsequent years.
 Figure \@ref(fig:concept) bottom illustrates what is occurring in this disturbance year.
 As in the prior 9 years the forecast itself is multimodal, with most of the ensemble members predicting no disturbance and a small number predicting a disturbance event.
-The multinomial filter assigns a prior mean and variance to each mode based on the ensemble summary statistics, which is the same as a traditional EnKF except for the stratification by disturbance type, and then assigns each state a prior probability based on the ensemble frequency of being in each state (disturbed = dark red, undisturbed = green).
+The Multinomial filter assigns a prior mean and variance to each mode based on the ensemble summary statistics, which is the same as a traditional EnKF except for the stratification by disturbance type, and then assigns each state a prior probability based on the ensemble frequency of being in each state.
 When the observed data (red) suggests a large drop in LAI, the posterior (blue) not only updates the ensemble mean and variance of each mode, but also the probability of being in each mode.
-This posterior is still multimodal, but in this specific instance the weight assigned the undisturbed mode (0.023) is too low to be visually notable.
+This posterior is still multimodal, but in this specific instance the weight assigned the undisturbed mode (0.023) is too low to be visible.
 Noteworthy is that the LAI in years 12 and 15 is actually lower than that in the disturbance year, yet the model still assigns this as being undisturbed.
@@
 Experiment 2 expands the assimilation to a multivariate forecast, in this case including joint predictions for leaf, wood, and soil carbon, and with a disturbance to both the leaf and wood pools (Figure \@ref(fig:sssd)).
-Figure \@ref(fig:sssdBi) shows the multivariate nature of the detection of the disturbance event at time 10, with the bimodal forecast ensemble shown in purple (undisturbed in the top right, disturbed in the bottom left), an observation in black that is consistent with the disturbed state, and the posterior ensemble in green.
+Figure \@ref(fig:sssdBi) shows the multivariate nature of the detection of the disturbance event at time 10, with the bimodal forecast ensemble shown in green (undisturbed in the top right, disturbed in the bottom left), an observation in red that is consistent with the disturbed state, and the posterior ensemble in blue.
 Arrows depict the adjustment of a random subset of individual ensemble members from the forecast prior to the analysis posterior.
@@
        ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time")
-  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
-  lines(CI.Na[2,],col="#b2df8a",lwd=2)
-  points(No[,i]/scaled[i])
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('blue',0.5))
+  lines(CI.Na[2,],col="blue",lwd=2)
+  points(No[,i]/scaled[i],col="red",pch=18,cex=1.5)
   #lines(Ntrue[,i],lty=2,lwd=2)
@@
     legend("bottomright",legend=c("Assimilation","pseudodata"),
-       lwd=5,col=c("#b2df8a","black"))
+       lwd=c(5,NA),pch=c(NA,18),col=c("blue","red"))
   }
@@
 
-```{r sssdBi,echo=FALSE, fig.asp=1,fig.width=6, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
+```{r sssdBi,echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, results='hide', fig.asp=1,fig.width=6, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
 Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
@@
 
-plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("blue",alpha=1),
      ylim=range(Forecast[[10]]$Nf[,3]),
@@
 .(labNames[1]) ~ x^2
-points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
-points(No[10,1],No[10,3],cex=3,col=1)
-lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
-lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
+points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="#b2df8a",pch=18)
+points(No[10,1],No[10,3],cex=3,col="red")
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2),col="red",lwd=3)
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]),col="red",lwd=3)
 
@@
        legend=c("Forecast","Analysis","Data"),
-       pch=c(1,3,18),col=c("purple","#b2df8a",1),
+       pch=c(18,3,1),col=c("#b2df8a","blue","red"),
        cex = 1.5)
@@
 
-** TODO: in bivariate plots, data is hard to see **
-
-Experiment 3 (Figure \@ref(fig:multiPseudo) top row) illustrates the ability of the multinomial model to select between more than one disturbance.
+Experiment 3 (Figure \@ref(fig:multiPseudo) top row) illustrates the ability of the Multinomial model to select between more than one disturbance.
 In the bivariate plot (top right panel) we can see that the model can successfully distinguishing between two types of disturbance, one that involves the loss of both wood and leaf biomass (e.g.
@@
 In the years prior to the disturbance the multivariate nature of the assimilation allows us to borrow strength across sites, with observations at one site reducing uncertainty not just at that site but at the other site as well based on the cross-site forecast covariance.
-That said,this example illustrates the ability of the multinomial model to avoid the impact of spurious correlations across sites during disturbance events.
+That said, this example illustrates the ability of the Multinomial model to avoid the impact of spurious correlations across sites during disturbance events.
 The presence of the disturbance at site 1 did not cause a reduction in biomass in site 2 during the disturbance, nor did it increase the productivity in site 2 in the years post-disturbance despite the large increase observed at site 1.
@@
      main="Multi-Disturbance")
-ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
-lines(CI.Na[2,],col="#b2df8a",lwd=2)
-points(No[,i])
+ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('blue',0.5))
+lines(CI.Na[2,],col="blue",lwd=2)
+points(No[,i],col="red")
 legend("bottomleft",legend=c("Assimilation","pseudodata"),
-       lwd=5,col=c("#b2df8a","black"))
+       lwd=5,col=c("blue","red"))
 
-plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("blue",alpha=1),
      ylim=range(Forecast[[10]]$Nf[,3]),
-     xlim=range(Forecast[[10]]$Nf[,1]),
+     xlim=range(c(0,Forecast[[10]]$Nf[,1])),
      xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),ylab=bquote(.(pool.names[3]) ~ (kgC/m^2)),
      cex.lab=1.25,asp = 1)
-points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
-points(No[10,1],No[10,3],cex=3,col=1)
-lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
-lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
+points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="#b2df8a",pch=18)
+points(No[10,1],No[10,3],cex=3,col="red")
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2),col="red",lwd=3)
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]),col="red",lwd=3)
 
@@
        legend=c("Forecast","Analysis","Data"),
-       pch=c(1,3,18),col=c("purple","#b2df8a",1),
+       pch=c(18,3,1),col=c("#b2df8a","blue","red"),
        cex = 1.25)
@@
      main="Multi-Disturbance, PFT switching")
-ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
-lines(CI.Na[2,],col="#b2df8a",lwd=2)
-points(No[,i])
+ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('blue',0.5))
+lines(CI.Na[2,],col="blue",lwd=2)
+points(No[,i],col="red")
 legend("bottomleft",legend=c("Assimilation","pseudodata"),
-       lwd=5,col=c("#b2df8a","black"))
+       lwd=5,col=c("blue","red"))
 
-plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("blue",alpha=1),
      ylim=range(Forecast[[10]]$Nf[,3]),
-     xlim=range(Forecast[[10]]$Nf[,1]),
+     xlim=range(c(0,Forecast[[10]]$Nf[,1])),
      xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),
@@
      cex.lab=1.25)
-points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
-points(No[10,1],No[10,3],cex=3,col=1)
-lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
-lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
+points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="#b2df8a",pch=18)
+points(No[10,1],No[10,3],cex=3,col="red")
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2),col="red",lwd=3)
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]),col="red",lwd=3)
 
@@
      main="Multi-Site, Single-Disturbance")
-  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[3]]),CI.Na[[3]][1,],CI.Na[[3]][3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
-  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[6]]),CI.Na[[6]][1,],CI.Na[[6]][3,],col=ecoforecastR::col.alpha("green",0.7))
-lines(CI.Na[[3]][2,],col="#b2df8a",lwd=2)
-lines(CI.Na[[6]][2,],col="green",lwd=2)
-points(No[,3])
-points(No[,6],pch=2)
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[3]]),CI.Na[[3]][1,],CI.Na[[3]][3,],col=ecoforecastR::col.alpha('blue',0.7))
+  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[6]]),CI.Na[[6]][1,],CI.Na[[6]][3,],col=ecoforecastR::col.alpha("lightblue",0.7))
+lines(CI.Na[[3]][2,],col="blue",lwd=2)
+lines(CI.Na[[6]][2,],col="lightblue",lwd=2)
+points(No[,3],col="red")
+points(No[,6],pch=2,col="pink")
 legend("bottomleft",legend=c("Site 1","Site 2","pseudodata"),
-       lwd=5,col=c("#b2df8a","green","black"))
+       lwd=5,col=c("blue","lightblue","red"))
 
@@
 ss = floor(seq(1,nrow(Analysis10$Na),length=200)) ## subsample
-plot(Analysis10$Na[,1],Analysis10$Na[,3],pch=Analysis10$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
+plot(Analysis10$Na[,1],Analysis10$Na[,3],pch=Analysis10$Ca,col=ecoforecastR::col.alpha("blue",alpha=1),
      ylim=range(Forecast10$Nf[,3,1]),
-     xlim=range(Forecast10$Nf[,1,1]),
+     xlim=range(c(0,Forecast10$Nf[,1,1])),
      xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),
@@
      cex.lab=1.25,asp = 1,type='n')
-points(Forecast10$Nf[ss,1,1],Forecast10$Nf[ss,3,1],cex=0.5,col="purple",pch=18)
-points(Forecast10$Nf[ss,1,2],Forecast10$Nf[ss,3,2],cex=0.5,col="violet",pch=18)
-points(Analysis10$Na[ss,1],Analysis10$Na[ss,3],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("#b2df8a",alpha=0.8),cex=0.5)
-points(Analysis10$Na[ss,4],Analysis10$Na[ss,6],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("green",alpha=0.8),cex=0.5)
-points(No[10,1],No[10,3],cex=3,col=1)
-lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
-lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
-points(No[10,4],No[10,6],cex=3,col=1)
-lines(No[10,4]+c(-1,1)*sqrt(R[1,1]),rep(No[10,6],2))
-lines(rep(No[10,4],2),No[10,6]+c(-1,1)*sqrt(R[3,3]))
+points(Forecast10$Nf[ss,1,1],Forecast10$Nf[ss,3,1],cex=0.5,col="#b2df8a",pch=18)
+points(Forecast10$Nf[ss,1,2],Forecast10$Nf[ss,3,2],cex=0.5,col="green",pch=18)
+points(Analysis10$Na[ss,1],Analysis10$Na[ss,3],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("blue",alpha=0.8),cex=0.5)
+points(Analysis10$Na[ss,4],Analysis10$Na[ss,6],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("lightblue",alpha=0.8),cex=0.5)
+points(No[10,1],No[10,3],cex=3,col="red")
+lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2),lwd=3,col="red")
+lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]),lwd=3,col="red")
+points(No[10,4],No[10,6],cex=3,col="pink")
+lines(No[10,4]+c(-1,1)*sqrt(R[1,1]),rep(No[10,6],2),lwd=3,col="pink")
+lines(rep(No[10,4],2),No[10,6]+c(-1,1)*sqrt(R[3,3]),lwd=3,col="pink")
 
@@
        legend=c("Forecast","Analysis","Data"),
-       pch=c(1,3,18),col=c("purple","#b2df8a",1),
+       pch=c(18,3,1),col=c("#b2df8a","blue","red"),
        cex = 1)
@@
 
-## VSEM Calibration
-
-Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from AmeriFlux site US-ME2.
-Best fit parameters are given in Supplemental Table \@ref(tab:parameters).
-The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibration) ).
-Achieving this fit came at the cost of nontrivial systematic biases in both leaf carbon (0.064 $kgC/m^2$, 13%) and NEE (0.528 $kgC/m^2/year$, 69%).
-The RMSEs for leaf carbon (0.067 $kgC/m^2$, 13.6%) and NEE (0.612 $kgC/m^2/year$, 80%) were similar to the biases, suggesting that most of the model error was systematic, particularly for leaf carbon.
-
-The dynamic process errors for the leaf, wood, and soil pools, which were approximated as the RMSE of the _change_ in each pool each year, were 0.01, 0.03, and 0.15 $kgC/m^2/yr$.
-This represents a dynamic error of 2.1%, 0.6%, and 0.7% respectively relative to the size of the pool and a CV of 114% and 17% for leaf and wood relative to the observed annual biomass increment.
-
-Overall, while the inability of the calibration to capture multiple data constraints simultaneously does suggest room for improvement in the underlying model (@Cameron2022), the model performed well for our target variable (AGB) and appears adequate for the job of testing the multinomial model against real AGB data for a range of different disturbance events.
-
-```{r paramTable,echo=FALSE}
-# load("fit.RData")
-# varName = c(names(theta),"Ql","Qw","Qs")
-# varName[c(5,7)] = c("tauL","tauW")
-# paramTable = data.frame(varName,fit=format(c(theta,Ql,Qw,Qs),digits=3,scientific=TRUE))
-# knitr::kable(paramTable, caption = "VSEM parameter estimates")
-```
-
 ## Remotely-sensed disturbance in Oregon
-
-** TODO: Dongchen wants examples for other disturbances and to know why the assimilation is so constrained **
 
@@
 Bottom row shows how the probability of detecting the disturbance varies with disturbance magnitude"}
+
 ## summary figure showing didn't work great at default error rate
@@
        main=exp.pretty.name[ex])
-  ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
-  lines(time,CI.Na[3,],col="#b2df8a",lwd=2)
-  points(time,Yi[,i])
+  ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('blue',0.5))
+  lines(time,CI.Na[3,],col="blue",lwd=2)
+  points(time,Yi[,i],col="red",cex=1.5)
   if(ex == 1) {
     legend("bottomleft",legend=c("Assimilation","Data"),
-       lwd=5,col=c("#b2df8a","black"),cex=0.7)
+       lwd=5,col=c("blue","red"),cex=0.7)
   }
@@
 ```
+
 In total we assimilated 365 example AGB time series from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively.
-Initial applications of the multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries.
-Filter divergence is a known issue for Kalman Filters, not something unique to our multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data.
+Initial applications of the Multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries.
+Filter divergence is a known issue for Kalman Filters, not something unique to our Multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data.
 Figure \@ref(fig:oregon) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed.
@@
 The causes of filter divergence are known to be some combination of model overconfidence (e.g., insufficient process error, Q) and high observation error (R) in the data.
-As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(tab:QRexp)) to verify that without filter divergence the multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so.
-Figure \@ref(fig:oregon) d-f show the results for Experiment 2, where we experimented with reducing the observation error, R.
+As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(tab:QRexp)) to verify that without filter divergence the Multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so.
+Figure \@ref(fig:oregon) d-f show the results for Experiment 2, where we reduced the observation error, R.
 In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:oregon) d).
-We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:oregon) e).
+We also see a much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:oregon) e).
 Divergence is no longer significantly affected by disturbance magnitude (p < `r format(summary(div.class[[2]])$coefficients[2,4],digits=3)`) and the only significant effect of disturbance type was a small reduction in divergence for clearcuts (p < `r format(summary(div.class[[2]])$coefficients[3,4],digits=3)`).
-Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:oregon) f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
+Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:oregon) f), with the Multinomial filter successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.
 
-Looking across all of our experiments, the coverage of our filter divergence statistic improved as the observation error decreased (Figure \@ref(fig:divExp)a).
+Looking across all the experiments, the coverage of our filter divergence statistic improved as the observation error decreased (Figure \@ref(fig:divExp)a).
 Increasing process error had little impact at the original observation error, but improved coverage once the observation error was decreased.
-The threshold disturbance magnitude required to detect disturbances also improved (i.e., became smaller) as the observation error decreased (Figure \@ref(fig:divExp)b) but was insensitive to process error, at least for the values we explored in our computational experiments.
+The threshold disturbance magnitude required to detect disturbances also improved (i.e., became smaller) as the observation error decreased (Figure \@ref(fig:divExp)b) but was insensitive to process error, at least for the values explored in our computational experiments.
 

### 2023-04-25 supplemental.Rmd
Final edits for submission. Edits for resubmission to include double spacing and line numbers

@@
 # Supplemental material
+
+## VSEM Calibration: Methods
+
+VSEM was calibrated primarily using data from 2002-2019 from the Ameriflux site US-ME2 (lat.lon), a mature Ponderosa pine stand located within our study region.
+VSEM pools (leaf, wood, and soil C) were initialized using the earliest available site-specific data from the Ameriflux BADM (Biomass, Ancillary, Disturbance, Management) dataset.
+An additional ten years of leaf and wood biomass data were also available from the BADM, which we included as part of the calibration dataset.
+Also included was the cumulative annual tower NEE (2002-2019).
+Because we were working with annual data we assumed Normal likelihoods for all three response variables, and because our priority was the correct prediction of aboveground biomass we upweighted the wood likelihood by 4 and downweighted the NEE likelihood by 1/4.
+
+Prior to calibration we set the leaf area ratio (LAR) to 15.7 m2/kgC and leaf residence time (tauL) to 4.7 years based on data from the BADM and left the light extinction coefficient (KEXT) at its default (0.5).
+The remaining fit parameters were the light use efficiency (LUE), the fraction of GPP allocated to autotrophic respiration (GAMMA), the fraction of NPP allocated to leaves (Av), the wood and soil C residence times (tauW, tauS), and the residual standard deviations for each of our data constraints (sd_NEE, sd_leaf, sd_wood).
+Fitting was done via Maximum Likelihood using base R's `optim` function.
+Finally, because the process error within a dynamical data assimilation system is not identical to the calibration residual error, we estimated the leaf and wood process errors (Ql, Qw) based on the difference between the predicted and observed net annual changes in each of these pools.
+Because there was not an equivalent annually-observed soil carbon dataset, the soil carbon process error (Qs) was estimated based on the predicted mean annual rate of change under the simplifying assumption that the true soil carbon is in steady-state (i.e.
+any change in modeled soil C was assumed to be an error).
+
+## VSEM Calibration: Results
+
+Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from AmeriFlux site US-ME2.
+Best fit parameters are given in Supplemental Table \@ref(tab:parameters).
+The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibration) ).
+Achieving this fit came at the cost of nontrivial systematic biases in both leaf carbon (0.064 $kgC/m^2$, 13%) and NEE (0.528 $kgC/m^2/year$, 69%).
+The RMSEs for leaf carbon (0.067 $kgC/m^2$, 13.6%) and NEE (0.612 $kgC/m^2/year$, 80%) were similar to the biases, suggesting that most of the model error was systematic, particularly for leaf carbon.
+
+The dynamic process errors for the leaf, wood, and soil pools, which were approximated as the RMSE of the _change_ in each pool each year, were 0.01, 0.03, and 0.15 $kgC/m^2/yr$.
+This represents a dynamic error of 2.1%, 0.6%, and 0.7% respectively relative to the size of the pool and a CV of 114% and 17% for leaf and wood relative to the observed annual biomass increment.
+
+Overall, while the inability of the calibration to capture multiple data constraints simultaneously does suggest room for improvement in the underlying model (@Cameron2022), the model performed well for our target variable (AGB) and appears adequate for the job of testing the Multinomial model against real AGB data for a range of different disturbance events.
+
+```{r paramTable,echo=FALSE}
+# load("fit.RData")
+# varName = c(names(theta),"Ql","Qw","Qs")
+# varName[c(5,7)] = c("tauL","tauW")
+# paramTable = data.frame(varName,fit=format(c(theta,Ql,Qw,Qs),digits=3,scientific=TRUE))
+# knitr::kable(paramTable, caption = "VSEM parameter estimates")
+```
 

### 2024-05-17 Paper.Rmd
legacy changes from last submission Fall 2023

@@
 ---
-title: "Assimilating disturbance: Toward real-time monitoring and forecasting"
+title: |
+    | Ecological Applications
+    | Article
+    | Assimilating disturbance: Toward real-time monitoring and forecasting
 author:
-  - Michael C. Dietze, Earth and Environment, Boston University, Boston, MA
-  - Shawn P. Serbin, Environmental and Climate Sciences Department, Brookhaven National Laboratory, Upton, NY
-  - Robert Kennedy, Geography, Environmental Sciences, and Marine Resource Management, Oregon State University, Corvallis, OR
-  - Beverly E. Law, Forest Ecosystems & Society, Oregon State University, Corvallis, OR
-date: '`r Sys.Date()`'
+  - Michael C. Dietze 1
+  - Shawn P. Serbin 2
+  - Robert Kennedy 3
+  - Beverly E. Law 4
+  - 1 - Earth and Environment, Boston University, Boston, MA
+  - 2 - Environmental and Climate Sciences Department, Brookhaven National Laboratory, Upton, NY
+  - 3 - Geography, Environmental Sciences, and Marine Resource Mgmt, Oregon State University, Corvallis, OR
+  - 4 - Forest Ecosystems & Society, Oregon State University, Corvallis, OR
+  - Corresponding Author = Michael C. Dietze
+#date: '`r Sys.Date()`'
+keywords: Disturbance detection, Ensemble Data Assimilation, Natural Climate Solutions, Carbon Cycle
 header-includes:
@@
     number_sections: yes
-    toc: yes
+    toc: no
     keep_tex: true
@@
     toc: yes
+    keep_tex: true
+  bookdown::word_document2: 
+    fig_caption: yes
+    number_sections: yes
+    toc: no
     keep_tex: true
@@
 ```
+
+*Open Research Statement:*
+
+Data are already published and publicly available, with those publications properly cited in this submission. Data sets used for this research are as follows:
+
+* The Landtrendr aboveground biomass product [@Kennedy2018] and disturbance product [@KENNEDY2015] for tile h03v04 was downloaded from http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html.
+* The 2001 National Land Cover Dataset [@HOOPER2018] is available from https://www.mrlc.gov/.
+* DayMet meteorological data [@thornton2021] were downloaded using the daymetr R package [@daymetr].
+* Ameriflux eddy covariance data and supplemental BADM (Biological, Ancillary, Disturbance, and Management) metadata for site US-ME2 [@van2005, @berner2016] were downloaded from https://ameriflux.lbl.gov/.
+
+This submission uses novel code, which is provided, per our requirements, in an external repository to be evaluated during the peer review process.
+Code is available at https://github.com/mdietze/DisturbAssim
+
 \newpage
@@
 
-1.
 Current approaches to bottom-up disturbance monitoring rely heavily on the detection of land-use, land-use change, and forestry through remote sensing, but often account for ecosystem impacts using simple look-up tables.
@@
 State data assimilation (SDA) is designed specifically to update predictions, nudging modeled states back toward reality in proportion to the uncertainties in the model and the data, but current SDA algorithms are designed to update continuous states, not discrete disturbances.
-2.
 Here we develop a new Bayesian SDA algorithm that combines a discrete Multinomial state-and-transition framework with conventional ensemble filtering SDA approaches.
 To demonstrate the potential for assimilating disturbance, we applied the Multinomial SDA to the Very Simple Ecosystem Model (VSEM), performing both simulated data experiments with known disturbances and testing the algorithm against real-world disturbances detected in the LandTrender data product for central Oregon.
-3.
 With simulated disturbance we not only demonstrate the ability to detect discrete disturbance events but also to avoid false positives.
@@
 With just AGB, our assimilation was sensitive to disturbances that reduced biomass by >1.5 kg/m2 but underpredicted defoliation disturbances, which could be improved by including LAI in the assimilation.
-4.
 Moving forward, the SDA framework provides an exciting opportunity to fuse multiple data sources to holistically improve real-time disturbance detection, impact assessment (e.g.
@@
 
-# Bibliography
+# Author Contributions
+
+MCD: Conceptualization, funding acquisition, model and software development, analysis, original draft, reviewing and editing of writing.
+SPS: Conceptualization, funding acquisition, and review and editing of writing.
+RK and BEL: Data generation and curation, review and editing of writing.
+
+# Conflict of Interest Statement
+
+The authors certify that they have no affiliations with or involvement in any
+organization or entity with any financial or non-financial interest in
+the subject matter or materials discussed in this manuscript.
+
+\newpage
+# References
 

### 2024-09-04 discussion.Rmd
Summer 2024 revisions

@@
 The most obvious issue with the initial application of the Multinomial filter to real-world data was the high incidence of filter divergence -- of the model becoming overconfident in itself and thus diverging from the observations.
-Filter divergence is not unique to this algorithm, but something that can occur in traditional filters as well. 
+Filter divergence is not unique to this algorithm, but something that can occur in traditional filters as well [@kalnay2003; @lewis2006]. 
 Indeed, there's no reason to believe that filter divergence is any more common in the Multinomial filter than in more traditional filters, such as EnKF, because in the absence of disturbance the Multinomial filter behaves exactly as EnKF would.
@@
 Existing LAI data products [e.g., MODIS, @yan2016] have the same spatial mismatch issue discussed above.
-Alternatively, a data product based on LandSAT would be at the same resolution but would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e.
+Alternatively, a data product based on Landsat would be at the same resolution but would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e.
 errors in data products based on the same underlying spectra would not be independent).

### 2024-09-04 methods.Rmd
Summer 2024 revisions

@@
 The VSEM was selected for its speed and ease of understanding, but its fundamental structure is analogous to much more complex, large-scale terrestrial carbon models.
+Model calibration is discussed in the Supplementary material.
 
@@
 Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g.
-carbon pools) and sometimes model parameters (Figure \@ref(fig:conceptualFig)).
+carbon pools) (Figure \@ref(fig:conceptualFig)).
 These updated variables are then used as the initial conditions for the next Forecast round.
+Conceptually, EnKF can also be used to iteratively update both model states and parameters simultaneously, though for simplicity in the current application we focus on state updating.
 
@@
                   )
-knitr::kable(exps,caption="Simulated data experiments")
+knitr::kable(exps,caption="Simulated data experiments.
+Exp = Experiment number, pools = number of C pools, disturbances = number of disturbance types, PFTs = number of alternative plant functional types, sites = number of sites.")
 ```
@@
 In summary, we: (1) calibrated VSEM against field data; (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transitions; (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types; and then (4) assessed the ability of the algorithm to detect different disturbances types and magnitudes.
-Model calibration is discussed in the Supplementary material.
 
@@
 Finally, for each ensemble member we selected the stand age that minimized the squared error between the modeled biomass and the initial biomass, and then selected the leaf and soil C pools at that stand age.
-This results is an ensemble of leaf, wood, and soil carbon pools with a covariance structure that is consistent with the calibrated model's dynamics.
+This results in an ensemble of leaf, wood, and soil carbon pools with a covariance structure that is consistent with the calibrated model's dynamics.
 
@@
 QRexp = data.frame(experiment=1:6,Qmult=Qmult[Qs],Rmult=Rmult[Rs])
-knitr::kable(QRexp,caption="Data Assimilation experiments")
+knitr::kable(QRexp,caption="Data Assimilation experiments. Experiment = experiment number, Qmult = Process error multiplier, Rmult = Observation error multiplier.")
 ```

### 2024-09-04 results.Rmd
Summer 2024 revisions

@@
 ```{r sssdBi,echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, results='hide', fig.asp=1,fig.width=6, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10).
+Red lines indicate observation error (±1 S.D.).
 Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
@@
 Initial applications of the Multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries.
-Filter divergence is a known issue for Kalman Filters, not something unique to our Multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data.
+Filter divergence is a known issue for Kalman Filters, not something unique to our Multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data [@kalnay2003; @lewis2006].
 Figure \@ref(fig:oregon) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed.

### 2024-09-04 supplemental.Rmd
Summer 2024 revisions

@@
 ```{r parameters, echo=FALSE}
-param.names = c("LUE","GAMMA","tauS","tauW","Av","sd_NEE","sd_leaf","sd_wood")
-param.desc = c("Light Use Efficiency","Fraction of GPP allocated to autotrophic respiration","Soil C residence time","Wood C residence time","Fraction of NPP allocated to leaves","NEE standard deviation","leaf C standard deviation","wood C standard deviation")
-param.table = cbind(param.names,param.desc,format(mle$par,digits=3,scientific=TRUE))
-colnames(param.table) <- c("Parameter","Description","Value")
-knitr::kable(param.table,caption="Calibrated model parameters")
+parSel = c(3, 4, 6, 7, 8,9:11)
+refPars <- BayesianTools::VSEMgetDefaults()
+
+param.values = rep(NA,11)
+param.values[seq_along(refPars[,1])] = refPars[,1]
+param.values[parSel] = mle$par
+param.names = c(row.names(refPars)[-(9:11)],"sd_NEE","sd_leaf","sd_wood") #c("LUE","GAMMA","tauS","tauW","Av","sd_NEE","sd_leaf","sd_wood")
+param.desc = c("Light extinction coefficient (unitless)","Leaf area ratio (m2/kg)","Light Use Efficiency (kg C MJ-1 PAR)","Fraction of GPP allocated to autotrophic respiration (unitless)", "Leaf C residence time (days)","Soil C residence time (days)","Wood C residence time (days)","Fraction of NPP allocated to leaves (unitless)","NEE std deviation (kgC m-2 year-1)","leaf C std deviation (kgC m-1)","wood C std deviation (kgC m-1)")
+#c("Light Use Efficiency","Fraction of GPP allocated to autotrophic respiration","Soil C residence time","Wood C residence time","Fraction of NPP allocated to leaves","NEE standard deviation ","leaf C standard deviation","wood C standard deviation")
+calibrated = rep("Default",11)
+calibrated[parSel] = "Calibrated"
+param.table = cbind(param.names,param.desc,format(param.values,digits=3,scientific=TRUE),calibrated)
+colnames(param.table) <- c("Parameter","Description","Value","Type")
+knitr::kable(param.table,caption="Model parameters")
 ```

### 2025-11-21 OR_model_DisturbPrior.Rmd
in-progress versions of OR model that adds disturbance product priors

--- 2023-01-12 OR_original_model.Rmd
@@
 author: "Michael Dietze"
-date: "12/7/2020"
+date: "2024-05-27"
 output: html_document
@@
 source("multi.ens.adj.R")
+library(BayesianTools)
 ```
@@
 ```{r}
+## Load extracted site timeseries
+load("extract.h03v04.RData")
+
 ## Load disturbance rate, state&transition, disturbed biomass stats
 load("STprior.RData")
+dtime = 1985:2012
 colnames(dextract) <- paste0("t",dtime)
 dextract[is.na(dextract)] <- 0
-d.prior["UN"] = d.prior["UN"] + d.prior[6]                         ## change at runtime (removes <NA>)
+d.prior["0"] = d.prior["0"] + d.prior[6]  ## change at runtime (removes <NA>)
 d.prior = d.prior[-6]
@@
 Qsoil = Qs
-
-## Load extracted site timeseries
-load("extract.h03v04.RData")
 ```
@@
 ## functions
-```{r,echo=FALSE,results='hide',message=FALS}
+```{r,echo=FALSE,results='hide',message=FALSE}
 ##' Single Site, Single Disturbance
@@
 ##' @param THETA parameter matrix
-##' @param dist  disturbance list: p.dist, mu0[3], V0[3,3], alloc soil 
+##' @param dist  disturbance list: p.dist[1:NT], mu0[3], V0[3,3], alloc soil 
 SSSD.model <- function(Y,R,Q,PAR,Xic,THETA,dist){
@@
     } 
-    if(sum(!sel) <= 2 || class(pfD) == "try-error") {   ## backup values if the ensemble stats fail
+    if(sum(!sel) <= 2 || "try-error" %in% class(pfD)) {   ## backup values if the ensemble stats fail
       mufD = c(dist$mu0[1], mean(Nf[,2]),  dist$mu0[2])            
@@
     ## priors
-    priors <- list(p    = dist$p.dist,         ## disturbance rate
+    priors <- list(p    = dist$p.dist[t],         ## disturbance rate
                    mufN = colMeans(Nf[sel,]),    ## Not disturbed forecast means
@@
       run.adjustment = TRUE
-      if(FALSE){
-        d3 <-density(Xa[,3]) 
-        plot(d3)
-        lines(d3$x,dnorm(d3$x,update$Y[3],1/sqrt(R[3,3])),col=2)
-        lines(d3$x,dnorm(d3$x,priors$mufN[3],1/sqrt(priors$pfN[3,3])),col="blue")
-        p.kf = R[3,3]+priors$pfN[3,3]
-        mu.kf = priors$mufN[3]*priors$pfN[3,3]/p.kf+
-          update$Y[3]*R[3,3]/p.kf
-        lines(d3$x,dnorm(d3$x,mu.kf,1/sqrt(p.kf)),col=3,lty=2)
-
-        jtest <- "model{
-         Y ~ dmnorm(n[H],R[H,H])
-        }"
-        jtest <- "  ## SingleSite SingleDisturbance MultiPool
-    model{
-      ## prior
-      n[3] ~ dnorm(mufN[3],pfN[3,3])  
-      Y[H] ~ dnorm(n[H],R[H,H])
-      dummy <- mufD[1] + pfD[1,1] + p
-  }"
-      jtest <- "  ## SingleSite SingleDisturbance MultiPool
-    model{
-      ## prior
-      n[3] ~ dnorm(mufN[3],pfN[3,3])  
-      Y[H] ~ dnorm(n[3],R[H,H])
-  }"
-        update2 = list(mufN = update$mufN,
-                       pfN = update$pfN,
-                       Y=update$Y,R=R,H=3)
-        mtest <- rjags::jags.model(file=textConnection(jtest),
-                                   data = update,
-#                               data=list(n = c(NA,NA,mu.kf),
-#                                         R=matrix(p.kf,3,3),H=3),
-                               n.adapt=1000,n.chains=3,quiet=TRUE)
-        mdat <- rjags::coda.samples(mtest,variable.names=c("n"),n.iter=10000) 
-      summary(mdat)  
-      }
     } else {
@@
 Pmult = c(1)
-Ri = 4 #4
-Qi = 1 #2
+Ri = 2
+Qi = 1
 Pi = 1
-#i = 181  # 177:263 = #fire
+users_accuracy = c(1-d.prior[1],.77,.92,.77,.78,.78) #Kennedy et al 2015 table 6
+## 0 = none, 10 = other, 20 = cut, 30 = road, 40 = fire, 50 = biotic
 od = c(83:176,264:356)
-for(i in 83:356){
-  ofile = paste0("OR_SD_",
+for(i in 1:356){
+  ofile = paste0("OR_DP_",  ## DP = Disturbance Prior
                  formatC(i, width = 3, format = "d", flag = "0"),
@@
   print(ofile)
+  if(file.exists(ofile)) next
   NT = floor(ncol(rad)/365)
@@
   dtype = which(names(d.prior) == dextract[i,"t2004"])
+  if(dtype == 1) dtype = sample(2:5,1) ## if site is undisturbed, set alternative at random
   if(length(dtype)==0) break
+  p.dist = rep(d.prior[dtype]*Pmult[Pi],NT) ## set background rate
+  has.dist = which(dextract[i,] > 0)
+  p.dist[has.dist] = users_accuracy[dextract[i,has.dist]/10 + 1] 
   dist <- list(              ## disturbance parameters
-    p.dist = d.prior[dtype]*Pmult[Pi],     ## disturbance probability
+    #p.dist = d.prior[dtype]*Pmult[Pi],     ## disturbance probability
+    p.dist = p.dist,
     mu0 = c(0.2,PBBAR[4,dtype]),   ## % retention [leaf,agb], first parameter made up, should include soil
@@
                   dist = dist)
-  save(test,file=ofile)
+  save(test,dist,file=ofile)
+  test$Analysis[[20]]$mu
 }
@@
 ```{r}
-i = 181
+i = 188#181
+ofile = paste0("OR_DP_",  ## DP = Disturbance Prior
+                 formatC(i, width = 3, format = "d", flag = "0"),
+                 "_P",Pi,"Q",Qi,"R",Ri,".RData")
+load(ofile)
+NT = floor(ncol(rad)/365)
+Yi = matrix(NA,NT,3)  ## observed data
+Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
 pool.names=c("Cleaf","Csoil","Cstem")
@@
   plot(Yi[,i],
-       ylim = range(rbind(CI.Na,Yi[,i]),na.rm=TRUE),
+       ylim = range(c(range(rbind(CI.Na,Yi[,i]),na.rm=TRUE),0)),
        type='n',
@@
  TODO:
- * why is Analysis not contracting around the data?
-    * reducing obs error by /10 didn't solve
- * why are the disturbance forecast pools crazy?
- * assess disturbance posterior
-
-## Computer experiments
-* Run under defaults
-* Run under a grid of inc process error and decrease obs error and increase disturbance probability
+ * why is filter divergence such a problem even prior to disturbance and with prescribed disturbance probabilities?
 
@@
 runs <- dir(pattern="*.RData")
-runs <- runs[grepl("^OR",runs)]
+runs <- runs[grepl("^OR_DP",runs)]
 exp = sapply(strsplit(runs,"_",fixed=TRUE),function(x){
@@
 ## load experiments that haven't yet been summarized
-# elist <- list()   ## run once
+# eDPlist <- list()   ## run once
 t04 = 15            ## time of disturbance assessement (2004)
@@
 for(i in seq_along(exps)){
-  if(is.null(elist[[exps[i]]])){
+  if(is.null(eDPlist[[exps[i]]])){
     esel <- which(exp==exps[i])
@@
 
-    elist[[exps[i]]] = list(distrate=distrate,predist=predist,
+    eDPlist[[exps[i]]] = list(distrate=distrate,predist=predist,
                             distsize=distsize,divergence=divergence)
@@
 
-save(elist,obs.pre,obs.dist,obs.tlo,obs.class,file="elist.RData")
+save(eDPlist,obs.pre,obs.dist,obs.tlo,obs.class,file="eDPlist.RData")
 ```
@@
 ```{r}
-pdf("OR_exps_diagnostics.pdf")
+pdf("OR_DP_exps_diagnostics.pdf")
 trim <- function(x,lo=0.0001,hi=0.9999){
@@
   ## divergence
-  div = (elist[[i]]$divergence[,1] - obs.tlo)/
-    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
+  div = (eDPlist[[i]]$divergence[,1] - obs.tlo)/
+    (apply(eDPlist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
   div.reg    = lm(div~obs.tlo)
@@
   ## Disturbance detection
-  distrate = elist[[i]]$distrate
+  distrate = eDPlist[[i]]$distrate
   hist(distrate,main=exps[i])
@@
   ## vs proportional disturbance size
-  distprop = 1 - elist[[i]]$distsize[,1]/elist[[i]]$predist[,1] ## predicted
+  distprop = 1 - eDPlist[[i]]$distsize[,1]/eDPlist[[i]]$predist[,1] ## predicted
   distprop = 1 - obs.dist/obs.pre ## observed
@@
 dev.off()
-save(stats,file="stats.RData")
+save(stats,file="stats.DP.RData")
 ```
+# TODO
+* add figures looking at how AGB data updates disturbance probability relative to the priors
 
@@
 ```{r}
-Ps = as.numeric(substr(names(elist),2,2))
-Qs = as.numeric(substr(names(elist),4,4))
-Rs = as.numeric(substr(names(elist),6,6))
+Ps = as.numeric(substr(names(eDPlist),2,2))
+Qs = as.numeric(substr(names(eDPlist),4,4))
+Rs = as.numeric(substr(names(eDPlist),6,6))
 
@@
 ## Disturbance @ default
-summary(stats[[4]]$dist.combo)
+#summary(stats[[4]]$dist.combo)
 ## nothing matters
@@
 ## Disturbance @ P1Q2R2 (intermediate)
-summary(stats[[4]]$dist.combo)
+#summary(stats[[4]]$dist.combo)
 ## detection varies with divergence & disturbance type
 
-## Disturbance magnitude: P1Q1R4
-i = 3
+## Disturbance magnitude: P1Q1R2
+i = 2
 summary(stats[[i]]$dist.combo)  ## absolute magnitude of divergence is best predictor
@@
 
-## Disturbance type: P1Q1R4 (reduced divergence)
+## Disturbance type: P1Q1R2 (reduced divergence)
 stats[[i]]$dist.class
@@
 site = 188
-exp.name = "P1Q1R4" #"P1Q1R1"  ## manually run for each experiment **
+exp.name = "P1Q1R2" #"P1Q1R1"  ## manually run for each experiment **
 
@@
 ## model
-load(paste0("OR_SD_",site,"_",exp.name,".RData"),envir = viz)
+viz = rlang::env()
+load(paste0("OR_DP_",site,"_",exp.name,".RData"),envir = viz)
 pool.names=c("Cleaf","Csoil","Cstem")
@@
 }
-save(Yi,CI.Na,CI.D,file=paste0("OR_TS_",site,"_",exp.name,".RData"))
+save(Yi,CI.Na,CI.D,file=paste0("OR_DPTS_",site,"_",exp.name,".RData"))
 ```
@@
 ```{r}
-exp.name = "P1Q1R4" #"P1Q1R1"
+exp.name = "P1Q1R2" #"P1Q1R1"
 

### 2025-11-21 OR_model_DisturbPriorNIMBLE.Rmd
in-progress versions of OR model that adds disturbance product priors

--- 2025-11-21 OR_model_DisturbPrior.Rmd
@@
 ---
-title: "OR Multi Disturbance"
+title: "OR Multi Disturbance, NIMBLE version w/ Wishart"
 author: "Michael Dietze"
-date: "2024-05-27"
+date: "2024-05-28"
 output: html_document
@@
 library(BayesianTools)
+library(nimble)
 ```
@@
 ##' @param dist  disturbance list: p.dist[1:NT], mu0[3], V0[3,3], alloc soil 
-SSSD.model <- function(Y,R,Q,PAR,Xic,THETA,dist){
+SSSD.model <- function(Y,R,Q,PAR,Xic,THETA,dist,restart=FALSE){
+
+  uni = TRUE
 
@@
   Forecast = Analysis = list()
+  bq.stored <- rep(NA,NT)
+  bq.stored[1] <- ifelse(uni,1,4)
+  if(uni){
+    aq.stored <- rep(NA,NT)
+    aq.stored[1] <- Q[3,3]*bq.stored[1]
+  } else {
+    aq.stored <- array(NA, c(NT,3,3))
+    aq.stored[1,,] <- Q*bq.stored[1]
+  }
+  t0 = 1
 
-  AnalysisSSSDMP <- "  ## SingleSite SingleDisturbance MultiPool
-    model{
-      ## prior
-      muN ~ dmnorm(mufN,pfN)                ## forecast prior on Not disturbed
-      muD ~ dmnorm(mufD,pfD)                ## forecast prior on Disturbed
-      D ~ dbern(p)                          ## disturbance probability
-      ## Analysis 
-      n <- D*muD + (1-D)*muN                ## select between disturbed and undisturbed
-      Y[H] ~ dmnorm(n[H],R[H,H])                       ## Observation error
-  }"
+  ## restart
+  if(restart){
+    load("test.RData")
+    Na = Analysis[[t]]$Na
+    t0 = t
+  }
 
-  for(t in 1:NT){
+  AnalysisSSSDMP.uni <- nimble::nimbleCode({  ## univariate constraint
+    # Prior
+    muN[1:N] ~ dmnorm(mufN[1:N],prec=pfN[1:N,1:N]) ## forecast prior on Not disturbed
+    muD[1:N] ~ dmnorm(mufD[1:N],prec=pfD[1:N,1:N]) ## forecast prior on Disturbed
+    D ~ dbern(prob=p)                         ## disturbance probability
+    q ~ dgamma(bq,aq) ## wishart prior precision 
+    ## Likelihood
+    n[1:N] <- D*muD[1:N] + (1-D)*muN[1:N]     ## select between disturbed and undisturbed
+    X ~ dnorm(n[H], 1/sqrt(q))                ## add process error  
+    Y ~ dnorm(X,R)                            ## Observation error
+  })
+
+    AnalysisSSSDMP <- nimble::nimbleCode({
+    # Prior
+    muN[1:N] ~ dmnorm(mufN[1:N],prec=pfN[1:N,1:N]) ## forecast prior on Not disturbed
+    muD[1:N] ~ dmnorm(mufD[1:N],prec=pfD[1:N,1:N]) ## forecast prior on Disturbed
+    D ~ dbern(prob=p)                         ## disturbance probability
+    q[1:N, 1:N]  ~ dwish(R = aq[1:N, 1:N], df = bq) ## wishart prior precision
+    ## Likelihood
+    n[1:N] <- D*muD[1:N] + (1-D)*muN[1:N]     ## select between disturbed and undisturbed
+    #X[1:N]  ~ dmnorm(n[1:N], prec = q[1:N, 1:N])  ## add process error
+    X  ~ dnorm(n[H], 1/sqrt(q[H, H]))  ## add process error  ## HACK
+    #if(NY == 1){
+    #  Y ~ dnorm(X[H],R)                        ## Observation error
+    Y ~ dnorm(X,R)                        ## Observation error ## HACK
+    #} else {
+    #  for(i in 1:NY){                            
+    #    ns[i] <-X[H[i]]                        ## Observation operator
+    #  }
+    #  Y[1:NY] ~ dmnorm(ns[1:NY],R[1:NY,1:NY])    ## Observation error
+    #}
+  })
+
+  for(t in t0:NT){
 
@@
       tsel    = (t-1)*365 + 1:365             ## Select days from driver files
-      Nmu[i,] = VSEM(THETA[i,], PAR = PAR[tsel])[365,2:4] ## deterministic simulation
-      Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
+      Nf[i,] = VSEM(THETA[i,], PAR = PAR[tsel])[365,2:4] ## deterministic simulation
+      ##Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
+      ## Nimble Wishart version handles process error internally
     }
@@
       mufD  = colMeans(Nf[!sel,])
-      pfD = solve(var(Nf[!sel,]))#solve(diag(c(0.15,0.5,0.25)^2))
+      pfD = try(solve(var(Nf[!sel,])))#solve(diag(c(0.15,0.5,0.25)^2))
     } 
@@
     }
+    mufN  =colMeans(Nf[sel,])
+    varfN = var(Nf[sel,])
+    tooSmall = which(diag(varfN)/mufN < 0.001)
+    for(jj in seq_along(tooSmall)){
+      varfN[tooSmall[jj],tooSmall[jj]] = mufN[tooSmall[jj]]*0.001
+    }
+    pfN   = solve(varfN)
+    #if("try-error" %in% class(pfD)){
+    #  stop()
+    #}
     Nf <- tobit(Nf)
@@
     ## priors
+    aq = ifelse(uni,aq.stored[t],aq.stored[t,,])
     priors <- list(p    = dist$p.dist[t],       ## disturbance rate
-                   mufN = colMeans(Nf[sel,]),    ## Not disturbed forecast means
+                   mufN = mufN,   ## Not disturbed forecast means
                    mufD = mufD,                 ## Disturbed forecast means
-                   pfN  = solve(var(Nf[sel,])),  ## Not disturbed forecast precision
-                   pfD  = pfD  ## Disturbed forecast precision
+                   pfN  = pfN,                  ## Not disturbed forecast precision
+                   pfD  = pfD,                  ## Disturbed forecast precision
+                   aq   = aq,                   ## prior sum of squares
+                   bq   = bq.stored[t]          ## prior sample size
     )
@@
     ## Analysis
-    update = c(list(Y=Y[t,],R=solve(R)),priors,H = which(!is.na(Y[t,])))
-    if(length(update$H) > 0){ ## data observed, run filter
-      mod <- rjags::jags.model(file=textConnection(AnalysisSSSDMP),
-                               data=update,
-                               n.adapt=1000,n.chains=3,quiet=TRUE)
-      jdat <- rjags::coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000)
+    update = c(list(Y=Y[t,3],         ## Data                #### HACK
+                    R=solve(R)[3,3]), ## Observation error   #### HACK
+                    priors)
+    constants = list(N=3,                      ## number of state variables
+                     H = as.vector(which(!is.na(Y[t,]))))  ## observation vector
+    constants$NY = length(constants$H)
+    if(constants$NY > 0){ ## data observed, run filter
+      mod <- nimble::nimbleModel(code = AnalysisSSSDMP.uni, 
+                                 data = update, 
+                                 constants=constants)
+      conf <- nimble::configureMCMC(mod, print=FALSE)
+      conf$addMonitors(c("X","n","D","q"))   #,"muN","muD"
+      conf$printSamplers()
+      Rmcmc <- nimble::buildMCMC(conf)
+      Cmodel <- nimble::compileNimble(mod)
+      Cmcmc <- nimble::compileNimble(Rmcmc, project = mod, showCompilerOutput = FALSE)
+      dat  <- nimble::runMCMC(Cmcmc, niter = 10000, 
+                               nburnin = 3000, thin = 1, 
+                               nchains = 3,samplesAsCodaMCMC = TRUE)
 
       ## update parameters & save Analysis posterior
-      dat = as.matrix(jdat)
+      dat = as.matrix(dat)
       nsel <- grep(pattern="^n",x=colnames(dat))
+      xsel <- grep(pattern="^X",x=colnames(dat))
+      if(uni){
         Xa = dat[,nsel]
+        Xa[,3] = dat[,xsel]
+      } else {
+        Xa = dat[,xsel]
+      }
       ca = dat[,"D"]
+      ## update Wishart
+      iq   <- grep("q", colnames(dat)) # 9 =3*3 (mean of those columns to generate a vector of 9)
+      mq <- dat[, iq]  # Omega, Precision
+      dim.q <- sqrt(length(iq))
+      if(uni){
+        q.bar = mean(mq) ## currently all 1 very small #
+      } else {
+        q.bar <- matrix(apply(mq, 2, mean), dim.q, dim.q)  # Mean Omega, Precision
+      }
+      col <- matrix(1:dim.q ^ 2, dim.q, dim.q)
+      WV  <- matrix(0, dim.q, dim.q)
+      if(uni){
+        WV = q.bar/var(mq)
+      }else{
+        wish.df <- function(Om, X, i, j, col) { # degrees of freedom function  
+          (Om[i, j]^2 + Om[i, i] * Om[j, j]) / stats::var(X[, col])
+        }
+        for (ii in seq_len(dim.q)) {
+          for (jj in seq_len(dim.q)) {
+            WV[ii, jj] <- wish.df(q.bar, X = mq, i = ii, j = jj, col = col[ii, jj])
+          }
+        }
+      }
+      mean.WV <- mean(WV)
+      if (mean.WV < dim.q) {
+         mean.WV <- dim.q
+      }
+      V <- solve(q.bar) * mean.WV
+      if(uni){
+        aq.stored[t+1]       <- V
+      } else {
+        aq.stored[t + 1, , ] <- V
+      }
+      bq.stored[t + 1]       <- mean.WV
+
       run.adjustment = TRUE
@@
     }
+#hist(1/sqrt(mq))
 
@@
     )
+    save(Forecast,Analysis,aq.stored, bq.stored,
+         t,Y,R,Q,PAR,THETA,dist,file="test.RData")
+    # test = list(Forecast=Forecast,Analysis=Analysis,aq=aq.stored,bq=bq.stored)
   }
-  return(list(Forecast=Forecast,Analysis=Analysis))
+  return(list(Forecast=Forecast,Analysis=Analysis,aq=aq.stored,bq=bq.stored))
 }
@@
 Pmult = c(1)
-Ri = 2
+Ri = 1
 Qi = 1
@@
 od = c(83:176,264:356)
-for(i in 1:356){
-  ofile = paste0("OR_DP_",  ## DP = Disturbance Prior
+for(i in 181:280){
+  ofile = paste0("OR_NDP_",  ## N = Nimble, DP = Disturbance Prior
                  formatC(i, width = 3, format = "d", flag = "0"),
@@
   save(test,dist,file=ofile)
-  test$Analysis[[20]]$mu
+  print(test$Analysis[[20]]$mu)
 }
@@
 ```{r}
-i = 188#181
+i = 181#181
 ofile = paste0("OR_DP_",  ## DP = Disturbance Prior
@@
 
-    CI.Na = sapply(X = test$Analysis,function(x){
-      y = x$CI
-      sel.n = which(colnames(y) == paste0("n[",i,"]"))
-      y[,sel.n]
+    #CI.Na = sapply(X = test$Analysis,function(x){
+    #  y = x$CI
+    #  sel.n = which(colnames(y) == paste0("n[",i,"]"))
+    #  y[,sel.n]
+    #})
+    CI.Na = sapply(X = test$Analysis, function(x){
+      quantile(x$Na[,i],c(0.025,0.25,0.5,0.75,0.975))
     })
@@
   plot(Yi[,i],
-       ylim = range(c(range(rbind(CI.Na,Yi[,i]),na.rm=TRUE),0)),
+       ylim = range(c(range(c(range(CI.Na),Yi[,i]),na.rm=TRUE),0)),
        type='n',
@@
   ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
-  lines(CI.Na[3,],col="#b2df8a",lwd=2)
+  lines(CI.Na[3,],col="#b2df8a",lwd=2,type='b',pch=18)
   points(Yi[,i])
   lines(CI.D[3,])
+  if(ncol(CI.Na)==1){
+    points(c(1,1),c(CI.Na[1,],CI.Na[5,]),pch="-",col="#b2df8a")
+    points(c(1,1),c(CI.Nf[1,],CI.Nf[3,]),pch="-",col="blue")
+    points(1,CI.D[3,],pch=17)
+  }
 
@@
 
+## process error
+plot(test$bq, main="bq")
+plot(test$aq, main="aq")
+q = sapply(X = test$Analysis,function(x){
+      y = x$mu["q"]
+    })
+plot(q)
+plot(1/sqrt(q))
 ```
